---
title: "Human vs chimp CNCC: SNAI2 motif change vs H3K27ac/SNAI2 ChIP at ATAC regions"
output: html_document
date: "2026-02-13"
---

<!--
OVERVIEW
This notebook tests whether evolutionary changes in the SNAI2 DNA-binding MOTIF
(scored by FIMO) track with changes in H3K27ac and SNAI2 ChIP-seq signal between
human (H20961) and chimp (C3649) at ATAC-accessible regions (all coordinates hg38).

Pipeline:
  1. Load FIMO SNAI2 motif p-values for 4 samples (human hap1/hap2, chimp hap1/hap2).
  2. Outer-join the 4 tables on region + motif coordinates; fill motifs missing from a
     sample with p-value = 0.005, following Prescott et al.
  3. Keep only homozygous motifs (hap1 == hap2 p-value in each species); compute the
     per-motif human-vs-chimp p-value log-ratio; collapse to one value per region.
  4. Load per-region ChIP read counts, CPM-normalise by species-specific SNPsplit read depth,
     and compute input-normalised enrichment, then the human-vs-chimp log2FC ("enrich").
  5. Restrict to enhancer/distal regions overlapping SNAI2 (and/or H3K27ac) peaks.
  6. Correlate per-region motif change (x) with H3K27ac and SNAI2 ChIP change (y).

Only FIMO p-values are used (motif scores are not loaded/used anywhere).

Naming convention used below:
  fimo_*               : the four FIMO motif tables
  motifs_merged        : per-motif-instance table joined across the four samples
  region_motif_summary : per-region collapse of the motif metric
  region_data          : master per-ATAC-region table (motif summary + ChIP counts + log2FC)
  enh_*                : enhancer/distal subsets of region_data

All files needed
  -fimo calls at ATAC reference regions
  -ChIP counts at ATAC reference regions
  -mapping stats for CPM calculation
  -SNAI2 ChIP-seq peaks
  -Annotation file where each ATAC reference regions is labeled as promoter distal or not
-->

```{r}
# all libraries for the whole notebook
library(GenomicRanges)
library(ggplot2)
library(data.table)
library(ggpubr)
```


# H3K27ac and SNAI2 at all ATAC regions (human vs chimp)

To detect motif instances that have changed between species,
the full outer join of the fimo outputs was performed with missing matches assigned conservatively a p-value of 0.005 (as in Prescott et al).
The log ratio of the p-values for orthologous motifs was calculated and used in downstream analyses.



# Import SNAI2 FIMO p-values at ATAC regions

Load the SNAI2 FIMO motif p-values at all ATAC regions, one table per sample
(human/chimp x haplotype1/haplotype2). In each file V15 is the FIMO p-value;
V1-V4 are the ATAC region (chr/start/end/name) and V7-V9 are the matched motif
location (chr/start/end). Regions with no motif hit carry "." in the FIMO columns.
(The FIMO score in V14 is intentionally not loaded; only p-values are used.)

```{r}
fimo_human_hap1 <- fread(
  "/.../fimo_sorted_atATACregionsv3.bed")
setnames(fimo_human_hap1, c("V15"), c("pvalue.H_hap1_SNAI"))
fimo_human_hap1 <- fimo_human_hap1[, .(V1,V2,V3,V4,V7,V8,V9, pvalue.H_hap1_SNAI)]

fimo_human_hap2 <- fread(
  "/.../fimo_sorted_atATACregionsv3.bed")
setnames(fimo_human_hap2, c("V15"), c("pvalue.H_hap2_SNAI"))
fimo_human_hap2 <- fimo_human_hap2[, .(V1,V2,V3,V4,V7,V8,V9, pvalue.H_hap2_SNAI)]


fimo_chimp_hap1 <- fread(
  "/.../fimo_sorted_atATACregionsv3.bed")
setnames(fimo_chimp_hap1, c("V15"), c("pvalue.C_hap1_SNAI"))
fimo_chimp_hap1 <- fimo_chimp_hap1[, .(V1,V2,V3,V4,V7,V8,V9, pvalue.C_hap1_SNAI)]


fimo_chimp_hap2 <- fread(
  "/.../fimo_sorted_atATACregionsv3.bed")
setnames(fimo_chimp_hap2, c("V15"), c("pvalue.C_hap2_SNAI"))
fimo_chimp_hap2 <- fimo_chimp_hap2[, .(V1,V2,V3,V4,V7,V8,V9, pvalue.C_hap2_SNAI)]

# remove the ATAC regions that don't have a mapped motif (FIMO "." in the motif chr column)
# Using data.table syntax
fimo_human_hap1 <- fimo_human_hap1[V7 != "." ]
fimo_human_hap2 <- fimo_human_hap2[V7 != "." ]
fimo_chimp_hap1 <- fimo_chimp_hap1[V7 != "." ]
fimo_chimp_hap2 <- fimo_chimp_hap2[V7 != "." ]

# motif-location columns (originally V7,V8,V9) become the join key V5,V6,V7
colnames(fimo_human_hap1)[5:7]=c("V5","V6","V7")
colnames(fimo_human_hap2)[5:7]=c("V5","V6","V7")
colnames(fimo_chimp_hap1)[5:7]=c("V5","V6","V7")
colnames(fimo_chimp_hap2)[5:7]=c("V5","V6","V7")

```



Merge the four motif tables, filling in motifs that are missing from a sample
(full outer join, then impute p-value = 0.005, per Prescott et al.).

```{r}
# List the 4 motif tables (human/chimp x hap1/hap2)
fimo_tables <- list(
  fimo_human_hap1,
  fimo_human_hap2,
  fimo_chimp_hap1,
  fimo_chimp_hap2
)

# 1. Start with the first dataframe
# We join on V1, V2, V3 (ATAC region) AND V5, V6, V7 (Motif location)
motifs_merged <- fimo_tables[[1]]

# 2. Iteratively merge the rest
for(i in 2:length(fimo_tables)) {
  motifs_merged <- merge(motifs_merged, fimo_tables[[i]],
                      by = c("V1", "V2", "V3", "V4", "V5", "V6", "V7"),
                      all = TRUE)
}

# 3. Handle the Missing Values (NAs)
# Ensure p-value columns are numeric, then replace NA (motif missing from a sample) with 0.005
pval_cols <- grep("pvalue", names(motifs_merged), value = TRUE)

for (col in pval_cols) {
  # Convert to numeric (this handles cases where "." might still exist)
  set(motifs_merged, j = col, value = as.numeric(as.character(motifs_merged[[col]])))
  # Replace NA with 0.005
  motifs_merged[is.na(get(col)), (col) := 0.005]
}
```

Remove the motifs where the p-value differs between hap1 and hap2
(i.e. keep only motif instances where the two alleles of the same species are consistent).

```{r}
motifs_merged=motifs_merged[motifs_merged$pvalue.H_hap1_SNAI == motifs_merged$pvalue.H_hap2_SNAI & motifs_merged$pvalue.C_hap1_SNAI == motifs_merged$pvalue.C_hap2_SNAI,]
```

Calculate the human-vs-chimp p-value log-ratio (per motif, using hap1).

```{r}
# p-value log-ratio: positive => motif more significant (stronger) in human
motifs_merged$pval_logratio_h_c_SNAI = -log10( motifs_merged$pvalue.H_hap1_SNAI / motifs_merged$pvalue.C_hap1_SNAI)

```

Collapse the per-motif metric to one value per ATAC region (sum and mean).

```{r}
# 1. Define the column(s) you want to collapse
# It's cleaner to list them here in case you want to add more later
motif_metric_cols <- c("pval_logratio_h_c_SNAI")

# 2. Collapse by ATAC region (V1, V2, V3, V4)
# We use .SD (Subset of Data) to apply the sum/mean functions to our targets
region_motif_summary <- motifs_merged[, c(
    setNames(lapply(.SD, sum,  na.rm = TRUE), paste0("sum_",  motif_metric_cols)),
    setNames(lapply(.SD, mean, na.rm = TRUE), paste0("mean_", motif_metric_cols))
  ),
  by = .(V1, V2, V3, V4),
  .SDcols = motif_metric_cols]
```




```{r}
# keep a copy under a coordinate-keyed name (used in the merge with ChIP counts below)
region_motif_summary_coord <- region_motif_summary

```
Note: regions with no SNAI2 motif in ANY sample were dropped above (the V7 != "." filter),
so they are absent from this table rather than represented as NA. The only imputed values
are motifs present in some haplotype/sample but absent in another (filled to p-value = 0.005).
The correlation below is therefore conditioned on motif presence in >=1 sample.


# Import ChIP data for these regions

ATAC_region_counts_v3 is folder containing files with ChIP counts per region in this format:
chr/start/end/region_name/score/V6/COUNTS


```{r}
count_dir <- "/.../ATAC_region_counts_v3"



# --- read all files into one long table ---
read_counts <- function(species) {
  files <- list.files(count_dir,
                      pattern = paste0("ATAC_regions_v3_", species, "\\.bed$"),
                      full.names = TRUE)
  rbindlist(lapply(files, function(f) {
    dt <- fread(f)
    setnames(dt, 1:4, c("V1", "V2", "V3", "V4"))      # region key
    sample_name <- sub("_ATAC_regions_v3_.*", "", basename(f))
    sample_name <- sub("_counts$", "", sample_name)   # -> s26_ChIPlib8_HL130_inp
    count_col <- names(dt)[ncol(dt)]                  # <--  count is last col
    dt[, .(V1, V2, V3, V4,
           colname = paste0("count_", sample_name, "_", species),
           count   = as.numeric(get(count_col)))]
  }))
}

long_counts <- rbind(read_counts("human"), read_counts("chimp"))

# --- pivot to wide: one column per sample x species ---
wide_counts <- dcast(long_counts, V1 + V2 + V3 + V4 ~ colname, value.var = "count")

# --- join onto the collapsed ratio table (keep all collapsed regions) ---
region_data <- merge(region_motif_summary_coord, wide_counts,
               by = c("V1", "V2", "V3", "V4"), all.x = TRUE)



```





# Input-normalized enrichment (CPM) and human/chimp log2FC

Sample sheet: each ChIP paired to its matched input (same batch + rep + species).

```{r}
# VERIFY: each ChIP paired to its matched input (same batch + rep + species)
sample_sheet <- data.frame(
  id        = c("K27ac_rep1","K27ac_rep2","K27ac_rep3","K27ac_rep4",
                "SNAI2_rep1","SNAI2_rep2","SNAI2_rep3","SNAI2_rep4",
                "K27ac_lib8","SNAI2_lib8"),
  mark      = c(rep("K27ac",4), rep("SNAI2",4), "K27ac","SNAI2"),
  batch     = c(rep("CNCCP4",8), "ChIPlib8","ChIPlib8"),
  chip_base = c("s15_CNCCP4_HL130_rep1_aH3K27ac_deeper","s16_CNCCP4_HL130_rep2_aH3K27ac_deeper",
                "s17_CNCCP4_HL130_rep3_aH3K27ac_deeper","s18_CNCCP4_HL130_rep4_aH3K27ac_deeper",
                "s19_CNCCP4_HL130_rep1_aSNAI2_deeper","s20_CNCCP4_HL130_rep2_aSNAI2_deeper",
                "s21_CNCCP4_HL130_rep3_aSNAI2_deeper","s24_CNCCP4_HL130_rep4_aSNAI2_deeper",
                "s21_ChIPlib8_HL130_aH3K27ac","s20_ChIPlib8_HL130_aSNAI2"),
  inp_base  = c("inp7_CNCCP4_HL130_rep1","inp8_CNCCP4_HL130_rep2",
                "inp9_CNCCP4_HL130_rep3","inp10_CNCCP4_HL130_rep4",
                "inp7_CNCCP4_HL130_rep1","inp8_CNCCP4_HL130_rep2",
                "inp9_CNCCP4_HL130_rep3","inp10_CNCCP4_HL130_rep4",
                "s26_ChIPlib8_HL130_inp","s26_ChIPlib8_HL130_inp"),
  stringsAsFactors = FALSE
)
# (one input per CNCCP4 cell prep serves both marks of that rep).

col_for <- function(base, sp) paste0("count_", base, "_", sp)

# sanity: every column the sheet references actually exists in `region_data`
needed_cols <- c(sapply(c(sample_sheet$chip_base, sample_sheet$inp_base),
                   function(b) col_for(b, c("human","chimp"))))
stopifnot(all(unique(needed_cols) %in% names(region_data)))
```
CPM normalization using the species-specific SNPsplit read depth (read from the per-allele
mapping logs), then input-normalized enrichment and the human-vs-chimp log2FC.

```{r}
# ---- proper CPM normalization using species-specific SNPsplit read depth ----
# library size = deduplicated, uniquely-mapped read PAIRS assigned to the human (or chimp)
# allele by SNPsplit, parsed from each sample's species-specific mapping log. The human- and
# chimp-allele counts of a sample are each divided by their OWN allele's depth.
mapping_dirs <- list(
  human = "/.../mapping_human12",
  chimp = "/.../mapping_chimp12"
)

read_libsize <- function(base, species) {
  roots    <- mapping_dirs[[species]]
  log_file <- file.path(roots, base, paste0(base, "_log.txt"))
  log_file <- log_file[file.exists(log_file)]
  if (length(log_file) == 0) stop("mapping log not found for ", species, " sample: ", base)
  log_lines <- readLines(log_file[1])
  hit <- grep("mapped uniquely after PCR dedup", log_lines, value = TRUE)
  if (length(hit) == 0) stop("dedup read count not found in log for: ", base)
  as.numeric(sub("^[^0-9]*([0-9]+).*", "\\1", hit[1]))   # the count is the first number on that line
}

# one depth per (sample x species)
all_bases <- unique(c(sample_sheet$chip_base, sample_sheet$inp_base))
libsizes  <- list(
  human = setNames(vapply(all_bases, read_libsize, numeric(1), species = "human"), all_bases),
  chimp = setNames(vapply(all_bases, read_libsize, numeric(1), species = "chimp"), all_bases)
)
print(libsizes)   # sanity-check the depths

cpm_for <- function(base, sp) paste0("cpm_", base, "_", sp)   # companion to col_for()

# CPM = allele count / that allele's SNPsplit depth * 1e6
for (base in all_bases) {
  for (sp in c("human", "chimp")) {
    set(region_data, j = cpm_for(base, sp),
        value = region_data[[col_for(base, sp)]] / libsizes[[sp]][[base]] * 1e6)
  }
}

# per-sample, per-species input-normalized enrichment, then human-minus-chimp log2FC
for (i in seq_len(nrow(sample_sheet))) {
  id <- sample_sheet$id[i]

  for (sp in c("human", "chimp")) {
    chip_cpm <- region_data[[cpm_for(sample_sheet$chip_base[i], sp)]]
    inp_cpm  <- region_data[[cpm_for(sample_sheet$inp_base[i],  sp)]]
    set(region_data, j = paste0(id, "_enrich_", sp),
        value = log2((chip_cpm + 1) / (inp_cpm + 1)))
  }

  # input-normalized enrichment: difference of log2 enrichments (human minus chimp)
  set(region_data, j = paste0(id, "_logFC_h_c_enrich"),
      value = region_data[[paste0(id, "_enrich_human")]] -
              region_data[[paste0(id, "_enrich_chimp")]])
}


# mean across all reps of a mark
for (mk in c("K27ac", "SNAI2")) {
  reps <- sample_sheet$id[sample_sheet$mark == mk ]
  cols <- paste0(reps, "_logFC_h_c_enrich")
  set(region_data, j = paste0(mk, "_logFC_h_c_enrich_mean"),
      value = rowMeans(as.matrix(region_data[, ..cols]), na.rm = TRUE))
}
```


### Scatterplots

Select enhancers bound by SNAI2 with a change in motif from human to chimp.

### Enhancers
```{r}
anno=readRDS("/.../data.rds") ## data with annotations for each region to select only promoter-distal ones
anno_enh <- anno[anno$classification %in% c("other_distal", "enhancer"), ]


# region key is chr_start_end with the 0-based BED start (matches anno$region)
enh_regions <- region_data[paste(V1, V2, V3, sep = "_") %in% anno_enh$region]

```


### Peaks
```{r}
snai2_peaks <- fread("/.../SNAI2_peaks_min3reps_deeper.bed") ## SNAI2 ChIP peaks called in at least 3 replicates
snai2_peaks_gr <- GRanges(seqnames = snai2_peaks$V1, ranges = IRanges(start = snai2_peaks$V2 + 1, end = snai2_peaks$V3))

# GRanges from the enhancer regions (BED start V2 is 0-based -> +1 for 1-based IRanges)
enh_gr <- GRanges(seqnames = enh_regions$V1, ranges = IRanges(start = enh_regions$V2 + 1, end = enh_regions$V3))


# Overlap with both peak sets
hits_snai2 <- overlapsAny(enh_gr, snai2_peaks_gr)

# Keep regions overlapping SNAI2 peaks

enh_snai2 <- enh_regions[hits_snai2]

message("SNAI2 overlap:      ", sum(hits_snai2))

```

Using enrich: log2(ChIP/input) and then human minus chimp.

Using the regions that overlap SNAI2 peaks.
```{r}
plot_df=enh_snai2[enh_snai2$sum_pval_logratio_h_c_SNAI!=0,]

p1=ggplot(plot_df, aes(mean_pval_logratio_h_c_SNAI, K27ac_logFC_h_c_enrich_mean)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", color = "steelblue") +
  stat_cor(method = "pearson")  +
  #facet_wrap(~ class_label) +
  labs(x     = "Mean pval logratio SNAI2 motif (human/chimp)",
       y     = "H3K27ac log2FC (human/chimp)",
       title = "Corr H3K27ac change with SNAI2 motif change (human/chimp)") +
  theme_bw() +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold"))

p2=ggplot(plot_df, aes(mean_pval_logratio_h_c_SNAI, SNAI2_logFC_h_c_enrich_mean)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", color = "steelblue") +
  stat_cor(method = "pearson")  +
  #facet_wrap(~ class_label) +
  labs(x     = "Mean pval logratio SNAI2 motif (human/chimp)",
       y     = "SNAI2 log2FC (human/chimp)",
       title = "Corr SNAI2 change with SNAI2 motif change (human/chimp)") +
  theme_bw() +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold"))

ggarrange(p1,p2,ncol=1)
ggsave("/.../scatter_human_chimp_revised.pdf",
       height = 7, width = 5)

```
