---
title: "SNAI1/SNAI2 dTag RNA-seq: DESeq2 quantification and downstream analysis"
author: "Lucia Ichino"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)
```

# Overview

This document combines the two halves of the SNAI RNA-seq analysis into a single
reproducible workflow:

* **Part 1 — Quantification and differential expression.** Salmon transcript
  quantifications are imported with `tximport` and tested with `DESeq2` for three
  independent dTag time-course experiments. Results tables (raw and
  LFC-shrunken) are written to `data_dir`.
* **Part 2 — Downstream analysis and figures.** The results tables written in
  Part 1 are read back in, DEGs are counted and bar-plotted, genes are annotated
  by their distance to SNAI2 ChIP-seq peaks, and volcano / MA plots coloured by
  peak proximity are produced.

### Datasets

| Label in code | Library | Cell line / genotype | Timepoints (dTag) | Control | Replicates |
|---|---|---|---|---|---|
| `dbl`   | lib4 | SNAI1-FNV; SNAI2-KO (double perturbation) | 3h, 1 day, 3 days | DMSO | 2 clones × 2 reps |
| `snai1` | lib3 | SNAI1-FNV                                 | 3h, 1 day, 4 days | ctrl | 2 reps |
| `snai2` | lib2 | SNAI2-FNV                                 | 3h, 1 day, 4 days | ctrl | 3 reps |

Note on timepoints: the double-perturbation experiment ends at **3 days** while
the two single-degron experiments end at **4 days**. This is intentional (it
reflects how the experiments were performed). It is handled in the
bar plot by labelling the shared axis level `3-4days`.


Note on shrinkage: **apeglm** is used as the `lfcShrink()` estimator for all
three datasets, and the shrunken tables are the ones used for every downstream
figure (DEG counts, volcano, MA). 

### How to run

1. Edit the paths in the **Configuration** chunk below.
2. If you only have the processed DESeq2 tables (not the raw Salmon output),
   set `run_quantification <- FALSE` in that chunk. Part 1 will then be skipped
   and Part 2 will run off the tables already present in `data_dir`.
3. Knit, or run chunk by chunk.

### Dependencies

```{r dependencies, eval=FALSE}
# Bioconductor
BiocManager::install(c(
  "tximport", "DESeq2", "apeglm", "GenomicRanges",
  "AnnotationHub", "ensembldb", "ggrastr"
))

# CRAN
install.packages(c("tidyverse", "data.table", "ggplot2", "ggpubr", "reshape2"))
```

```{r libraries}
# --- Quantification / DE -----------------------------------------------------
library(tximport)
library(DESeq2)

# --- Data handling -----------------------------------------------------------
library(tidyverse)     # dplyr, readr, ggplot2, ...
library(data.table)

# --- Genomic annotation ------------------------------------------------------
library(GenomicRanges)
library(AnnotationHub)
library(ensembldb)

# --- Plotting ----------------------------------------------------------------
library(ggplot2)
library(ggpubr)        # ggarrange()
library(ggrastr)       # geom_point_rast(): keeps PDFs small for dense scatters
```

# Configuration


```{r config}
# ---- Master switch ----------------------------------------------------------
# TRUE  : run Part 1 (needs the raw Salmon quant.sf files and the tx2gene table)
# FALSE : skip Part 1 and use the DESeq2 tables already present in `data_dir`
run_quantification <- TRUE

# ---- Input: Salmon quantification directories -------------------------------
# Each directory is expected to contain one sub-directory per sample, each with
# a single quant.sf file.
salmon_dir_dbl   <- ".../"
salmon_dir_snai1 <- ".../"
salmon_dir_snai2 <- ".../"

# ---- Input: transcript-to-gene map used to build the Salmon index -----------
tx2gene_file <- file.path(
  "...",
  "Homo_sapiens.GRCh38.cdna.all.tx2gene.txt"
)

# ---- Input: SNAI2 ChIP-seq peaks (DMSO, relative to dTag, INPUT-subtracted) --
peak_file <- ".../peaks/s8_lib1_CNCCp4_SNAI2a3029_DMSO_r2_aSNAI2_reltodTag_peaks_minusINPUT_filtered.narrowPeak"

# ---- Output directories -----------------------------------------------------
data_dir <- ".../RNA_seq_processed_data"  # DESeq2 tables
fig_dir  <- ".../figures"                 # PDF figures
rda_dir  <- ".../SNAI_paper_Rproject"     # saved txi objects

for (d in c(data_dir, fig_dir, rda_dir)) {
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
}

# ---- Analysis parameters ----------------------------------------------------
padj_cut    <- 0.05   # adjusted p-value threshold used throughout
min_count   <- 5      # pre-filter: a gene must have >= min_count reads ...
min_samples <- 3      # ... in at least min_samples samples to be kept
```

# Part 1 — Salmon import and DESeq2

## Helper functions

The three experiments are analysed identically apart from their sample tables,
reference level and shrinkage estimator, so the shared logic is wrapped in
functions. This keeps the three blocks below short and guarantees they stay in
sync (the original script repeated the same ~40 lines three times, with objects
such as `ddsTxi`, `res_3h`, ... silently overwritten between blocks).

```{r part1-helpers}
#' List Salmon quant.sf files in a directory, named by sample
#'
#' Assumes the layout <salmon_dir>/<sample_name>/quant.sf. The sample name is
#' taken from the first path component, so it is robust to deeper nesting.
salmon_quant_files <- function(salmon_dir) {
  rel <- list.files(salmon_dir, pattern = "\\.sf$", recursive = TRUE)
  if (length(rel) == 0) stop("No *.sf files found under: ", salmon_dir)

  files        <- file.path(salmon_dir, rel)
  names(files) <- sub("/.*$", "", rel)

  stopifnot(all(file.exists(files)))
  files
}

#' Import Salmon quantifications, summarised to gene level
#'
#' countsFromAbundance = "lengthScaledTPM" gives length-corrected counts that
#' can be handed straight to DESeq2 (no separate offset needed).
import_salmon <- function(files, tx2gene) {
  tximport(
    files,
    type                = "salmon",
    tx2gene             = tx2gene,
    countsFromAbundance = "lengthScaledTPM"
  )
}

#' Fit a DESeq2 model to a dTag time course and extract all contrasts vs control
#'
#' @param txi          tximport object
#' @param sample_table data.frame with columns `condition` and `replicate`;
#'                     rownames MUST equal colnames(txi$counts), in that order
#' @param ref_level    level of `condition` used as the control (e.g. "DMSO")
#' @param contrasts    named character vector; names are the short tags used in
#'                     output filenames ("3h", "1d", "3d"/"4d"), values are the
#'                     corresponding `condition` levels ("3h", "1day", "3days")
#' @param shrink_type  LFC shrinkage estimator passed to lfcShrink()
#'
#' @return list(dds, res, res_shrunk); `res` and `res_shrunk` are named lists of
#'         DESeqResults ordered by padj, one per contrast.
run_deseq_timecourse <- function(txi,
                                 sample_table,
                                 ref_level,
                                 contrasts,
                                 shrink_type = c("apeglm", "normal"),
                                 design      = ~ condition + replicate,
                                 min_count   = 5,
                                 min_samples = 3) {

  shrink_type <- match.arg(shrink_type)

  # Guard against the classic silent bug: a sample table whose row order does
  # not match the column order of the count matrix.
  stopifnot(identical(rownames(sample_table), colnames(txi$counts)))

  dds <- DESeqDataSetFromTximport(txi, colData = sample_table, design = design)

  # Pre-filter low-count genes (>= min_count reads in >= min_samples samples).
  keep <- rowSums(counts(dds) >= min_count) >= min_samples
  message(sprintf("Pre-filter: keeping %d / %d genes", sum(keep), length(keep)))
  dds <- dds[keep, ]

  dds$condition <- relevel(dds$condition, ref = ref_level)
  dds <- DESeq(dds)

  # Unshrunken results: Wald test log2FC, p-value and padj.
  res <- lapply(contrasts, function(lvl) {
    r <- results(dds, contrast = c("condition", lvl, ref_level))
    r[order(r$padj), ]
  })

  # Shrunken results: lfcShrink() replaces log2FoldChange (and its SE) only —
  # pvalue/padj are carried over unchanged from the Wald test above.
  res_shrunk <- lapply(contrasts, function(lvl) {
    coef_name <- paste0("condition_", lvl, "_vs_", ref_level)
    stopifnot(coef_name %in% resultsNames(dds))
    r <- lfcShrink(dds, coef = coef_name, type = shrink_type)
    r[order(r$padj), ]
  })

  list(dds = dds, res = res, res_shrunk = res_shrunk)
}

#' Count significant up/down genes per contrast
#'
#' NOTE: the original script computed these counts by combining the padj of one
#' timepoint with the log2FoldChange of another (e.g. `res_3h$padj` with
#' `res_3d$log2FoldChange`). That was a copy-paste slip; here each contrast is
#' summarised against itself.
deg_summary <- function(res_list, padj_cut = 0.05) {
  data.frame(
    contrast = names(res_list),
    up       = vapply(res_list, function(r) sum(r$padj < padj_cut & r$log2FoldChange > 0, na.rm = TRUE), integer(1)),
    down     = vapply(res_list, function(r) sum(r$padj < padj_cut & r$log2FoldChange < 0, na.rm = TRUE), integer(1)),
    row.names = NULL
  )
}

#' Write raw and shrunken results tables with the naming convention used by Part 2
#'
#' Produces, for each contrast tag:
#'   DEseq2_<tag>_vs_ctrl_<suffix>.txt          (unshrunken)
#'   DEseq2_<tag>_vs_ctrl_<suffix>_shrunk.txt   (shrunken LFC)
write_deseq_tables <- function(fit, out_dir, suffix) {
  for (tag in names(fit$res)) {
    write.table(
      fit$res[[tag]],
      file  = file.path(out_dir, sprintf("DEseq2_%s_vs_ctrl_%s.txt", tag, suffix)),
      quote = FALSE, sep = "\t"
    )
    write.table(
      fit$res_shrunk[[tag]],
      file  = file.path(out_dir, sprintf("DEseq2_%s_vs_ctrl_%s_shrunk.txt", tag, suffix)),
      quote = FALSE, sep = "\t"
    )
  }
  invisible(NULL)
}
```

## Transcript-to-gene map

```{r tx2gene, eval=run_quantification}
tx2gene <- read_table(tx2gene_file, col_names = FALSE)
head(tx2gene)
```

## Dataset 1 — SNAI1-FNV; SNAI2-KO (lib4)

Design: 4 conditions (DMSO, 3h, 1day, 3days) × 2 clones (c5946b16, c5946b27) ×
2 replicates = 16 samples. Clone and replicate are combined into a single
4-level `replicate` factor, which is included in the design to absorb
clone/batch effects: `~ condition + replicate`.

Shrinkage uses **apeglm**, as for all three datasets in this document (see the
note on shrinkage estimators above).

```{r dbl-import, eval=run_quantification}
files_dbl <- salmon_quant_files(salmon_dir_dbl)
txi_dbl   <- import_salmon(files_dbl, tx2gene)

names(txi_dbl)
head(txi_dbl$counts)
```

```{r dbl-sampletable, eval=run_quantification}
# The condition/replicate vectors are positional: they assume the samples come
# back from list.files() in alphabetical order, i.e. 8 samples of clone
# c5946b16 followed by 8 of clone c5946b27, each block ordered
# 1day, 1day, 3days, 3days, 3h, 3h, DMSO, DMSO.
sampleTable_dbl <- data.frame(
  condition = factor(rep(rep(c("1day", "3days", "3h", "DMSO"), each = 2), 2)),
  replicate = factor(c(
    rep(c("c5946b16_rep1", "c5946b16_rep2"), 4),
    rep(c("c5946b27_rep1", "c5946b27_rep2"), 4)
  ))
)
rownames(sampleTable_dbl) <- colnames(txi_dbl$counts)

# ALWAYS eyeball this table: every sample name must sit next to the right
# condition. This is the single most common source of silent errors here.
sampleTable_dbl
```

```{r dbl-deseq, eval=run_quantification}
deseq_dbl <- run_deseq_timecourse(
  txi          = txi_dbl,
  sample_table = sampleTable_dbl,
  ref_level    = "DMSO",
  # names = output file tags; values = levels of `condition`
  contrasts    = c("3h" = "3h", "1d" = "1day", "3d" = "3days"),
  shrink_type  = "apeglm",
  min_count    = min_count,
  min_samples  = min_samples
)

resultsNames(deseq_dbl$dds)
```

```{r dbl-degs, eval=run_quantification}
# Summarised from the shrunken results so these numbers match the bar plot in
# Part 2 (padj is unaffected by shrinkage; only the sign of near-zero LFCs can
# differ from the unshrunken table).
deg_summary(deseq_dbl$res_shrunk, padj_cut)
```

```{r dbl-plotcounts, eval=run_quantification, fig.width=5, fig.height=4}
# Sanity check on a single gene (normalised counts per condition).
plotCounts(deseq_dbl$dds, gene = "UACA", intgroup = "condition")
```

```{r dbl-save, eval=run_quantification}
# The tximport object is saved so downstream analyses (e.g. TPM-level plots)
# don't have to re-import the Salmon files.
save(txi_dbl, file = file.path(rda_dir, "txi_SNAI1FNVSNAI2KO.rda"))

write_deseq_tables(deseq_dbl, data_dir, suffix = "SNAI1FNVSNAI2KO_ONLY_CNCC")
```

## Dataset 2 — SNAI1-FNV single degron (lib3)

Design: 4 conditions (ctrl, 3h, 1d, 4d) × 2 replicates = 8 samples.

```{r snai1-import, eval=run_quantification}
files_snai1 <- salmon_quant_files(salmon_dir_snai1)
txi_snai1   <- import_salmon(files_snai1, tx2gene)

head(txi_snai1$counts)
head(txi_snai1$abundance)
```

```{r snai1-sampletable, eval=run_quantification}
# Positional again: alphabetical order gives rep1 block (1d, 3h, 4d, ctrl)
# followed by rep2 block (1d, 3h, 4d, ctrl).
sampleTable_snai1 <- data.frame(
  condition = factor(rep(c("1d", "3h", "4d", "ctrl"), 2)),
  replicate = factor(rep(c("rep1", "rep2"), each = 4))
)
rownames(sampleTable_snai1) <- colnames(txi_snai1$counts)
sampleTable_snai1
```

```{r snai1-deseq, eval=run_quantification}
deseq_snai1 <- run_deseq_timecourse(
  txi          = txi_snai1,
  sample_table = sampleTable_snai1,
  ref_level    = "ctrl",
  contrasts    = c("3h" = "3h", "1d" = "1d", "4d" = "4d"),
  shrink_type  = "apeglm",
  min_count    = min_count,
  min_samples  = min_samples
)

resultsNames(deseq_snai1$dds)
deg_summary(deseq_snai1$res_shrunk, padj_cut)
```

```{r snai1-save, eval=run_quantification}
write_deseq_tables(deseq_snai1, data_dir, suffix = "SNAI1_CNCC")
```

## Dataset 3 — SNAI2-FNV single degron (lib2)

This Salmon directory holds **both** the SNAI2-FNV and the ZEB1-FNV libraries.
Only the first 12 samples (SNAI2-FNV: 4 conditions × 3 replicates) are used
here.

```{r snai2-import, eval=run_quantification}
files_snai2_all <- salmon_quant_files(salmon_dir_snai2)

# Inspect before subsetting — confirm that entries 1:12 really are the SNAI2
# samples and that none of the ZEB1 samples sneak in.
names(files_snai2_all)

files_snai2 <- files_snai2_all[1:12]
txi_snai2   <- import_salmon(files_snai2, tx2gene)

head(txi_snai2$counts)
```

```{r snai2-sampletable, eval=run_quantification}
# 3 replicates per condition, conditions in blocks of 3.
sampleTable_snai2 <- data.frame(
  condition = factor(rep(c("1d", "3h", "4d", "ctrl"), each = 3)),
  replicate = factor(rep(c("rep1", "rep2", "rep3"), 4))
)
rownames(sampleTable_snai2) <- colnames(txi_snai2$counts)
sampleTable_snai2
```

```{r snai2-deseq, eval=run_quantification}
deseq_snai2 <- run_deseq_timecourse(
  txi          = txi_snai2,
  sample_table = sampleTable_snai2,
  ref_level    = "ctrl",
  contrasts    = c("3h" = "3h", "1d" = "1d", "4d" = "4d"),
  shrink_type  = "apeglm",
  min_count    = min_count,
  min_samples  = min_samples
)

resultsNames(deseq_snai2$dds)
deg_summary(deseq_snai2$res_shrunk, padj_cut)
```

```{r snai2-save, eval=run_quantification}
write_deseq_tables(deseq_snai2, data_dir, suffix = "SNAI2_CNCC")
```

# Part 2 — Downstream analysis and figures

Part 2 reads the tables written above from `data_dir` rather than using the
in-memory objects, so it can be run on its own (`run_quantification <- FALSE`)
from the deposited processed data.

All three datasets are read from their `_shrunk` tables, so every figure below
is based on the same quantity: an apeglm-shrunken `log2FoldChange` paired with
the `pvalue`/`padj` from the unshrunken Wald test (`lfcShrink()` replaces the
LFC and its standard error but does not recompute p-values). The consequences
of that pairing are discussed under the volcano plots.

## Helper

```{r degs-helper}
#' Extract up- or down-regulated gene names from a DESeq2 results data.frame
get_degs <- function(df, direction = c("up", "down"), padj_cut = 0.05) {
  direction <- match.arg(direction)
  if (direction == "up") {
    rownames(df)[which(df$padj < padj_cut & df$log2FoldChange > 0)]
  } else {
    rownames(df)[which(df$padj < padj_cut & df$log2FoldChange < 0)]
  }
}
```

## Load log2FC tables — SNAI1-FNV; SNAI2-KO

```{r load-combined}
res_3h_dbl <- read.table(file.path(data_dir, "DEseq2_3h_vs_ctrl_SNAI1FNVSNAI2KO_ONLY_CNCC_shrunk.txt"))
res_1d_dbl <- read.table(file.path(data_dir, "DEseq2_1d_vs_ctrl_SNAI1FNVSNAI2KO_ONLY_CNCC_shrunk.txt"))
res_3d_dbl <- read.table(file.path(data_dir, "DEseq2_3d_vs_ctrl_SNAI1FNVSNAI2KO_ONLY_CNCC_shrunk.txt"))

DEGs_3h_up_comb <- get_degs(res_3h_dbl, "up", padj_cut)
DEGs_1d_up_comb <- get_degs(res_1d_dbl, "up", padj_cut)
DEGs_3d_up_comb <- get_degs(res_3d_dbl, "up", padj_cut)

DEGs_3h_dw_comb <- get_degs(res_3h_dbl, "down", padj_cut)
DEGs_1d_dw_comb <- get_degs(res_1d_dbl, "down", padj_cut)
DEGs_3d_dw_comb <- get_degs(res_3d_dbl, "down", padj_cut)
```

## Load log2FC tables — single degrons

These are the `_shrunk` tables, matching the double-perturbation dataset above.
The unshrunken tables (same names without the `_shrunk` suffix) are also written
by Part 1 if you need the raw Wald-test log2FC.

```{r load-individual}
res_3h_SNAI1 <- read.table(file.path(data_dir, "DEseq2_3h_vs_ctrl_SNAI1_CNCC_shrunk.txt"))
res_1d_SNAI1 <- read.table(file.path(data_dir, "DEseq2_1d_vs_ctrl_SNAI1_CNCC_shrunk.txt"))
res_4d_SNAI1 <- read.table(file.path(data_dir, "DEseq2_4d_vs_ctrl_SNAI1_CNCC_shrunk.txt"))

DEGs_3h_SNAI1_up_comb <- get_degs(res_3h_SNAI1, "up", padj_cut)
DEGs_1d_SNAI1_up_comb <- get_degs(res_1d_SNAI1, "up", padj_cut)
DEGs_4d_SNAI1_up_comb <- get_degs(res_4d_SNAI1, "up", padj_cut)

DEGs_3h_SNAI1_dw_comb <- get_degs(res_3h_SNAI1, "down", padj_cut)
DEGs_1d_SNAI1_dw_comb <- get_degs(res_1d_SNAI1, "down", padj_cut)
DEGs_4d_SNAI1_dw_comb <- get_degs(res_4d_SNAI1, "down", padj_cut)

res_3h_SNAI2 <- read.table(file.path(data_dir, "DEseq2_3h_vs_ctrl_SNAI2_CNCC_shrunk.txt"))
res_1d_SNAI2 <- read.table(file.path(data_dir, "DEseq2_1d_vs_ctrl_SNAI2_CNCC_shrunk.txt"))
res_4d_SNAI2 <- read.table(file.path(data_dir, "DEseq2_4d_vs_ctrl_SNAI2_CNCC_shrunk.txt"))

DEGs_3h_SNAI2_up_comb <- get_degs(res_3h_SNAI2, "up", padj_cut)
DEGs_1d_SNAI2_up_comb <- get_degs(res_1d_SNAI2, "up", padj_cut)
DEGs_4d_SNAI2_up_comb <- get_degs(res_4d_SNAI2, "up", padj_cut)

DEGs_3h_SNAI2_dw_comb <- get_degs(res_3h_SNAI2, "down", padj_cut)
DEGs_1d_SNAI2_dw_comb <- get_degs(res_1d_SNAI2, "down", padj_cut)
DEGs_4d_SNAI2_dw_comb <- get_degs(res_4d_SNAI2, "down", padj_cut)
```

## Bar plot of DEG numbers

A gene counts as a DEG if `padj < padj_cut`; the up/down split follows the sign
of the shrunken log2FC. Since shrinkage does not change `padj`, the totals per
bar are identical to the unshrunken ones — only the up/down assignment of a
small number of genes whose shrunken LFC crosses zero can differ.

The plotting `data.frame` is built from a **named** list, so each count stays
tied to its own label. (The original built four parallel vectors that had to
stay in sync by position — easy to silently misalign when edited.)

Names follow the pattern `<sample>_<UP|DW>_<timepoint>` and are split on `_` to
generate the three factor columns.

```{r barplot-data}
deg_lists <- list(
  SNAI1_UP_3h          = DEGs_3h_SNAI1_up_comb,
  SNAI1_UP_1day        = DEGs_1d_SNAI1_up_comb,
  `SNAI1_UP_3-4days`   = DEGs_4d_SNAI1_up_comb,
  SNAI2_UP_3h          = DEGs_3h_SNAI2_up_comb,
  SNAI2_UP_1day        = DEGs_1d_SNAI2_up_comb,
  `SNAI2_UP_3-4days`   = DEGs_4d_SNAI2_up_comb,
  `SNAI1/2_UP_3h`      = DEGs_3h_up_comb,
  `SNAI1/2_UP_1day`    = DEGs_1d_up_comb,
  `SNAI1/2_UP_3-4days` = DEGs_3d_up_comb,
  SNAI1_DW_3h          = DEGs_3h_SNAI1_dw_comb,
  SNAI1_DW_1day        = DEGs_1d_SNAI1_dw_comb,
  `SNAI1_DW_3-4days`   = DEGs_4d_SNAI1_dw_comb,
  SNAI2_DW_3h          = DEGs_3h_SNAI2_dw_comb,
  SNAI2_DW_1day        = DEGs_1d_SNAI2_dw_comb,
  `SNAI2_DW_3-4days`   = DEGs_4d_SNAI2_dw_comb,
  `SNAI1/2_DW_3h`      = DEGs_3h_dw_comb,
  `SNAI1/2_DW_1day`    = DEGs_1d_dw_comb,
  `SNAI1/2_DW_3-4days` = DEGs_3d_dw_comb
)

deg_counts <- do.call(rbind, lapply(names(deg_lists), function(nm) {
  parts <- strsplit(nm, "_")[[1]]
  data.frame(
    sample    = parts[1],
    DEG_type  = parts[2],
    dTag_time = parts[3],
    DEG_n     = length(deg_lists[[nm]])
  )
}))

deg_counts$dTag_time <- factor(deg_counts$dTag_time, levels = c("3h", "1day", "3-4days"))
deg_counts$sample    <- factor(deg_counts$sample,    levels = c("SNAI1", "SNAI2", "SNAI1/2"))

deg_counts
```

```{r barplot-plot, fig.width=4.5, fig.height=2.5}
p_bar <- ggplot(deg_counts, aes(x = sample, y = DEG_n, fill = dTag_time)) +
  geom_vline(xintercept = c(1.5, 2.5), linetype = "dashed", color = "grey75", linewidth = 0.6) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  facet_wrap(~DEG_type) +
  theme_classic() +
  scale_fill_manual(values = c("#D8BFD8", "#9932CC", "#4B0082")) +
  ylab("Differentially expressed genes n.") +
  xlab("") +
  labs(fill = "Timepoint") +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 11, color = "black"),
    axis.text.y     = element_text(color = "black"),
    strip.text      = element_text(size = 12, face = "bold"),
    legend.position = "right"
  )

p_bar
ggsave(file.path(fig_dir, "Barplot_DEGs_V3.pdf"), plot = p_bar, height = 2.5, width = 4.5)
```

## Peak-proximity annotation

Each gene is assigned to one of three mutually exclusive categories based on
the distance from its canonical TSS to the nearest SNAI2 ChIP-seq peak:

| Category | Definition | Colour |
|---|---|---|
| `peak <=1kb`     | peak within 1 kb of the TSS **and** no peak in the 1–50 kb ring | blue |
| `peak 1kb-50kb`  | at least one peak in the 1–50 kb ring (regardless of a promoter peak) | yellow |
| `no peak <=50kb` | no peak within 50 kb | grey/red |

### Canonical protein-coding TSSs

```{r peak-annotation-tss}
ah        <- AnnotationHub()
human_ens <- query(ah, c("EnsDb", "Homo sapiens"))

# Pinned to a specific EnsDb snapshot for reproducibility: AH116291 = Ensembl 111.
edb_latest <- ah[["AH116291"]]

tx_gr <- transcripts(
  edb_latest,
  filter = ~ gene_biotype == "protein_coding" & tx_is_canonical == TRUE
)

gene_map        <- genes(edb_latest, columns = c("gene_id", "gene_name"))
tx_gr$gene_name <- gene_map$gene_name[match(tx_gr$gene_id, gene_map$gene_id)]

# Collapse each transcript to its TSS (single bp, strand-aware).
tss_gr <- resize(tx_gr, width = 1, fix = "start")
seqlevelsStyle(tss_gr) <- "Ensembl"   # match the peak naming style used below

length(tss_gr)
```

### Overlap TSS windows with peaks

```{r peak-annotation-overlap}
proximal_bp <- 1000
distal_bp   <- 50000

peaks_np <- fread(peak_file, header = FALSE)
setnames(peaks_np, 1:3, c("chrom", "start", "end"))

# narrowPeak is BED-like: 0-based, half-open. +1 on the start converts to the
# 1-based, closed convention used by GRanges.
peaks_np_gr <- GRanges(
  seqnames = sub("^chr", "", peaks_np$chrom),
  ranges   = IRanges(peaks_np$start + 1, peaks_np$end)
)
seqlevelsStyle(peaks_np_gr) <- "Ensembl"

# Drop contigs that are absent from the annotation (scaffolds, chrM naming, ...)
peaks_np_gr <- keepSeqlevels(
  peaks_np_gr,
  intersect(seqlevels(peaks_np_gr), seqlevels(tss_gr)),
  pruning.mode = "coarse"
)

tss_distal_windows   <- GRanges(seqnames(tss_gr), IRanges(pmax(1L, start(tss_gr) - distal_bp),   start(tss_gr) + distal_bp))
tss_proximal_windows <- GRanges(seqnames(tss_gr), IRanges(pmax(1L, start(tss_gr) - proximal_bp), start(tss_gr) + proximal_bp))

ov_distal   <- findOverlaps(tss_distal_windows,   peaks_np_gr, ignore.strand = TRUE)
ov_proximal <- findOverlaps(tss_proximal_windows, peaks_np_gr, ignore.strand = TRUE)

pairs_distal_all <- data.frame(tss_idx = queryHits(ov_distal),   peak_idx = subjectHits(ov_distal))
pairs_proximal   <- data.frame(tss_idx = queryHits(ov_proximal), peak_idx = subjectHits(ov_proximal))

# Distal pairs: the peak overlaps the 50 kb window but NOT the 1 kb window.
pairs_distal <- dplyr::anti_join(pairs_distal_all, pairs_proximal, by = c("tss_idx", "peak_idx"))

# Genes with >= 1 peak in the 1kb-50kb ring (yellow), regardless of promoter peak
genes_has_distal    <- unique(tss_gr$gene_name[pairs_distal$tss_idx])
# Genes with a promoter peak (<= 1kb) but NO distal peak (blue)
genes_proximal_all  <- unique(tss_gr$gene_name[queryHits(ov_proximal)])
genes_only_promoter <- setdiff(genes_proximal_all, genes_has_distal)

message("Genes with peak <=1kb only (blue):    ", length(genes_only_promoter))
message("Genes with peak in 1kb-50kb (yellow): ", length(genes_has_distal))
```

### Flag the results tables

```{r peak-annotation-flag}
# Mutually exclusive proximity categories; matched on gene symbol (rownames).
flag_proximity <- function(df) {
  df$proximity <- "no peak <=50kb"
  df$proximity[rownames(df) %in% genes_has_distal]    <- "peak 1kb-50kb"  # yellow
  df$proximity[rownames(df) %in% genes_only_promoter] <- "peak <=1kb"     # blue
  df
}

res_3h_dbl <- flag_proximity(res_3h_dbl)
res_1d_dbl <- flag_proximity(res_1d_dbl)
res_3d_dbl <- flag_proximity(res_3d_dbl)

table(res_3d_dbl$proximity)
```





## MA plots



```{r ma-setup}
all_res <- list(res_3h_dbl, res_1d_dbl, res_3d_dbl)


xr_ma <- range(sapply(all_res, function(d) range(log10(d$baseMean + 1), na.rm = TRUE)))
yr_ma <- range(sapply(all_res, function(d) range(d$log2FoldChange,      na.rm = TRUE)))

make_snai_MA <- function(df, title_text) {
  df$plot_group <- "ns"
  df$plot_group[df$padj < padj_cut & df$proximity == "no peak <=50kb"] <- "sign_nopeak"
  df$plot_group[df$padj < padj_cut & df$proximity == "peak 1kb-50kb"]  <- "sign_50kb"
  df$plot_group[df$padj < padj_cut & df$proximity == "peak <=1kb"]     <- "sign_1kb"
  df <- df[order(df$plot_group == "ns", decreasing = TRUE), ]

  ggplot(df, aes(log10(baseMean + 1), log2FoldChange, color = plot_group, size = plot_group)) +
    geom_point_rast(alpha = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
    scale_color_manual(
      name   = "Peak proximity",
      values = c(ns = "grey70", sign_nopeak = "#0072B2", sign_50kb = "#E69F00", sign_1kb = "darkred"),
      labels = c(ns = "ns", sign_nopeak = "sig, no peak <=50kb", sign_50kb = "sig, peak <50kb", sign_1kb = "sig, peak <=1kb")
    ) +
    scale_size_manual(
      name   = "Peak proximity",
      values = c(ns = 0.5, sign_nopeak = 1.2, sign_50kb = 1.2, sign_1kb = 1.2),
      labels = c(ns = "ns", sign_nopeak = "sig, no peak <=50kb", sign_50kb = "sig, peak <50kb", sign_1kb = "sig, peak <=1kb")
    ) +
    scale_x_continuous(limits = xr_ma) +
    scale_y_continuous(limits = yr_ma) +
    coord_cartesian(ylim = c(-3.5, 3.5)) +
    guides(color = guide_legend(override.aes = list(alpha = 1, size = 3))) +
    theme_classic() +
    xlab("log10(baseMean + 1)") +
    ylab("Shrunk log2FoldChange (dTag/DMSO)") +
    ggtitle(title_text)
}
```

```{r ma-plot, fig.width=11, fig.height=3}
p1_ma <- make_snai_MA(res_3h_dbl, "SNAI1-FNV SNAI2-KO: 3h dTag vs DMSO")
p2_ma <- make_snai_MA(res_1d_dbl, "SNAI1-FNV SNAI2-KO: 1 day dTag vs DMSO")
p3_ma <- make_snai_MA(res_3d_dbl, "SNAI1-FNV SNAI2-KO: 3 days dTag vs DMSO")

p_ma <- ggarrange(p1_ma, p2_ma, p3_ma, nrow = 1, common.legend = TRUE, legend = "right")
p_ma
ggsave(file.path(fig_dir, "MA_alltimepoints_proximity_V1_shrunk.pdf"), plot = p_ma, height = 3, width = 11)
```

# Session info

```{r session-info}
sessionInfo()
```
