---
title: "SNAI1/SNAI2 dTag: differential H3K27ac and ATAC signal over consensus ATAC regions"
author: "Lucia Ichino"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)
```

# Overview

Differential analysis of two chromatin datasets from the SNAI1-FNV SNAI2-KO
dTag time course, both quantified over the same consensus ATAC peak set (2kb regins)
("ATAC regions v3"):

| Section | Assay | Counts directory | Samples |
|---|---|---|---|
| 1 | H3K27ac ChIP-seq (lib4) | `ChIP/lib4/ATAC_region_counts_v3` | 2 clones × 2 replicates × 4 timepoints = 16 |
| 2 | ATAC-seq (lib1)          | `ATAC/lib1ATAC/ATAC_region_counts_v3` | 2 clones × 2 replicates × 4 timepoints = 16 |

Both follow an identical workflow:

1. Read per-sample `bedtools coverage -counts` output and assemble a
   region × sample count matrix.
2. Fit `DESeq2` with `~ clone + replicate + condition` and test each dTag
   timepoint (3h, 1d, 3d) against DMSO.
3. Restrict the results to canonical chromosomes (chr1–22, chrX, chrY),
   dropping unplaced scaffolds, alt contigs and chrM.
4. Export results tables (CSV), MA plots (PDF), and the significant regions as
   UP/DOWN BED files for downstream genomic overlap.



### Dependencies

```{r dependencies, eval=FALSE}
BiocManager::install("DESeq2")
install.packages(c("tidyverse", "ggrastr", "ggpubr", "conflicted"))
```

```{r libraries}
library(tidyverse)   # dplyr, tidyr, purrr, readr, stringr, ggplot2
library(DESeq2)
library(ggrastr)     # geom_point_rast(): keeps dense MA plots small as PDF
library(ggpubr)      # ggarrange()

# Several packages here mask dplyr verbs (notably DESeq2/BiocGenerics).
# Fix the winners once, explicitly, rather than relying on attach order.
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("rename", "dplyr")
conflict_prefer("count",  "dplyr")
conflict_prefer("setdiff", "base")
```

# Configuration

All paths live here; nothing below this chunk should need editing.

```{r config}
# ---- Input: per-sample count files (bedtools coverage -counts output) -------
counts_dir_k27ac <- ".../ATAC_region_counts_v3"
counts_dir_atac  <- ".../ATAC_region_counts_v3"

# Both directories use the same filename convention.
counts_pattern <- "_counts_ATAC_regions_v3\\.bed$"

# ---- Output -----------------------------------------------------------------
out_dir <- ".../out"                    # CSV + BED
fig_dir <- ".../figures"                   # PDF figures

for (d in c(out_dir, fig_dir)) dir.create(d, showWarnings = FALSE, recursive = TRUE)

# ---- Analysis parameters ----------------------------------------------------
padj_cut    <- 0.05
min_count   <- 5   # a region needs >= min_count reads ...
min_samples <- 4   # ... in at least min_samples samples (= n per condition group)

# Canonical chromosomes. Everything else (unplaced/unlocalized scaffolds,
# alt haplotypes, chrEBV, ...) is dropped from the results before export.
# chrM is deliberately excluded: mitochondrial signal is an ATAC artefact and
# not meaningful for differential chromatin analysis. Add "chrM" here if you
# want to keep it.
canonical_chroms <- paste0("chr", c(1:22, "X", "Y"))

# Timepoints tested against DMSO. Names are the tags used in output filenames.
contrast_levels <- c("3h" = "3h", "1d" = "1d", "3d" = "3d")
```

# Helper functions

The two assays are processed identically, so the shared logic lives here. The
original script repeated the same ~120 lines per assay, with `dataPeaks`,
`dds`, `res_3h`, ... silently overwritten between sections — which made it easy
for one section's edits to drift away from another's.

## Reading counts

```{r helper-counts}
#' Read one `bedtools coverage -counts` BED file
#'
#' The count is always the LAST column; the first three are chrom/start/end.
#' The sample name is taken from the filename, up to "_counts".
read_coverage_bed <- function(path) {
  df <- readr::read_tsv(path, col_names = FALSE, show_col_types = FALSE)
  tibble(
    region = paste(df[[1]], df[[2]], df[[3]], sep = "_"),
    sample = sub("_counts.*", "", basename(path)),
    count  = df[[ncol(df)]]
  )
}

#' Build a region x sample count matrix from a directory of coverage files
#'
#' @return data.frame of integer counts, rownames = "chr_start_end"
load_region_counts <- function(counts_dir, pattern) {
  files <- list.files(counts_dir, pattern = pattern, full.names = TRUE)
  if (length(files) == 0) stop("No count files matching ", pattern, " in ", counts_dir)
  message("Found ", length(files), " count files in ", counts_dir)

  long_df <- purrr::map_dfr(files, read_coverage_bed)

  # Guard: duplicated regions within a sample would make pivot_wider produce
  # list-columns instead of numbers. Collapse exact duplicate rows first.
  long_df <- dplyr::distinct(long_df)
  stopifnot(!any(duplicated(long_df[c("region", "sample")])))

  wide <- tidyr::pivot_wider(long_df, names_from = sample, values_from = count)

  mat <- as.data.frame(wide[, setdiff(names(wide), "region"), drop = FALSE])
  rownames(mat) <- wide$region
  mat
}
```

## Sample metadata

```{r helper-metadata}
#' Derive condition / clone / replicate from sample names
#'
#' Sample names look like
#'   H3K27ac_c5946b16_DMSO_rep1
#'   ATAC_SNAI12_c5946b16_DMSO_rep1
#' so all three factors can be pulled out by pattern instead of being typed as
#' positional vectors.
#'
#' This matters: the original script reused one hard-coded sampleTable for both
#' assays, but the two assays' columns are ordered differently (H3K27ac is
#' clone-major, ATAC is replicate-major). In the ATAC section that swapped the
#' `clone` and `replicate` labels. Parsing the names removes that whole class
#' of error. (See the note in the ATAC section for why the swap did not change
#' the reported statistics.)
parse_sample_metadata <- function(sample_names) {
  meta <- data.frame(
    clone     = stringr::str_extract(sample_names, "c5946b[0-9]+"),
    condition = stringr::str_extract(sample_names, "(?<=_)(DMSO|3h|1d|3d)(?=_)"),
    replicate = stringr::str_extract(sample_names, "rep[0-9]+"),
    row.names = sample_names,
    stringsAsFactors = FALSE
  )

  if (anyNA(meta)) {
    stop("Could not parse metadata for: ",
         paste(rownames(meta)[!complete.cases(meta)], collapse = ", "))
  }

  meta$condition <- factor(meta$condition, levels = c("DMSO", "3h", "1d", "3d"))
  meta$clone     <- factor(meta$clone)
  meta$replicate <- factor(meta$replicate)
  meta
}
```

## Canonical chromosome filter

Region IDs are `chr_start_end`, so the chromosome is everything before the last
two underscores. Scaffolds, alt contigs and chrM are removed here — before any
region ID is split into BED coordinates — so that the exported tables, the MA
plots and the BED files all describe the same set of regions.

```{r helper-canonical}
#' Extract the chromosome from a "chr_start_end" region ID
#'
#' Strips the last two underscore-separated fields only, so contig names that
#' themselves contain underscores (chr1_KI270706v1_random) survive intact and
#' can be recognised as non-canonical.
region_chrom <- function(regions) {
  sub("^(.*)_[0-9]+_[0-9]+$", "\\1", regions)
}

#' Keep only regions on canonical chromosomes
#'
#' Reports what was dropped, so a naming mismatch (e.g. "1" vs "chr1") shows up
#' immediately rather than silently emptying the table.
keep_canonical_regions <- function(df, chroms = canonical_chroms, region_col = "region") {
  chrom <- region_chrom(df[[region_col]])
  keep  <- chrom %in% chroms

  if (!any(keep)) {
    stop("No regions on canonical chromosomes. Seqname style mismatch? ",
         "Found e.g.: ", paste(head(unique(chrom), 3), collapse = ", "),
         " but expected e.g.: ", paste(head(chroms, 3), collapse = ", "))
  }

  dropped <- sort(unique(chrom[!keep]))
  message(sprintf(
    "Canonical filter: kept %d / %d regions (%d dropped across %d non-canonical seqnames%s)",
    sum(keep), length(keep), sum(!keep), length(dropped),
    if (length(dropped)) paste0(": ", paste(head(dropped, 5), collapse = ", "),
                                if (length(dropped) > 5) ", ..." else "") else ""
  ))

  df[keep, , drop = FALSE]
}
```

## DESeq2

```{r helper-deseq}
#' Fit DESeq2 over regions and extract each timepoint vs DMSO
#'
#' @param counts_mat  region x sample integer counts
#' @param meta        colData; rownames must match colnames(counts_mat) in order
#' @param contrasts   named vector: names = output tags, values = condition levels
#'
#' `design = ~ clone + replicate + condition` puts the variable of interest last
#' and absorbs clone and replicate/batch effects as additive nuisance terms.
run_deseq_regions <- function(counts_mat,
                              meta,
                              contrasts,
                              ref_level   = "DMSO",
                              design      = ~ clone + replicate + condition,
                              min_count   = 5,
                              min_samples = 4) {

  stopifnot(identical(rownames(meta), colnames(counts_mat)))

  dds <- DESeqDataSetFromMatrix(
    countData = counts_mat,
    colData   = meta,
    design    = design
  )

  # Pre-filter weakly covered regions.
  keep <- rowSums(counts(dds) >= min_count) >= min_samples
  message(sprintf("Pre-filter: keeping %d / %d regions", sum(keep), length(keep)))
  dds <- dds[keep, ]

  dds$condition <- relevel(dds$condition, ref = ref_level)
  dds <- DESeq(dds)

  res <- lapply(contrasts, function(lvl) {
    r <- results(dds, contrast = c("condition", lvl, ref_level))
    r[order(r$padj), ]
  })

  list(dds = dds, res = res)
}

#' DESeqResults -> data.frame with an explicit `region` column
#'
#' The original wrote the tables with write.csv() and immediately read them back
#' so that the region IDs became a column ("X"). Doing the conversion in memory
#' is equivalent and avoids the round trip. NOTE: the CSVs written below now
#' have a named first column `region` instead of an empty header.
res_to_df <- function(res) {
  as.data.frame(res) %>%
    tibble::rownames_to_column("region") %>%
    mutate(sig = !is.na(padj) & padj < padj_cut)
}

#' Count significant regions per contrast
deg_summary <- function(res_dfs, padj_cut = 0.05) {
  data.frame(
    contrast = names(res_dfs),
    up       = vapply(res_dfs, function(d) sum(d$padj < padj_cut & d$log2FoldChange > 0, na.rm = TRUE), integer(1)),
    down     = vapply(res_dfs, function(d) sum(d$padj < padj_cut & d$log2FoldChange < 0, na.rm = TRUE), integer(1)),
    row.names = NULL
  )
}
```

## Plotting and export

```{r helper-plot-export}
#' MA plot: log2 mean signal vs log2 fold change, significant regions in red
ma_plot <- function(df, title_text, ylim = c(-3, 3)) {
  ggplot(df, aes(x = log2(baseMean), y = log2FoldChange, color = sig)) +
    geom_point_rast(size = 0.5, alpha = 0.2, raster.dpi = 300) +
    scale_color_manual(
      name   = paste0("padj < ", padj_cut),
      values = c("FALSE" = "grey20", "TRUE" = "red")
    ) +
    # coord_cartesian() clips the view without dropping points from the fit
    coord_cartesian(ylim = ylim) +
    theme_bw() +
    xlab("log2 BaseMean") +
    ylab("log2FoldChange (dTag/DMSO)") +
    ggtitle(title_text)
}

#' Write a vector of "chr_start_end" region IDs as a 3-column BED
#'
#' Expects regions that have already passed keep_canonical_regions().
#' Split on the LAST two underscores only: naive strsplit(x, "_") breaks on
#' contigs whose names themselves contain underscores
#' (e.g. chr1_KI270706v1_random).
write_regions_bed <- function(regions, file, chroms = canonical_chroms) {
  if (length(regions) == 0) {
    message("No regions to write for ", file)
    return(invisible(NULL))
  }

  m <- regmatches(regions, regexec("^(.*)_([0-9]+)_([0-9]+)$", regions))
  bad <- lengths(m) == 0
  if (any(bad)) stop("Unparseable region IDs, e.g.: ", regions[which(bad)[1]])

  bed <- data.frame(
    chr   = vapply(m, `[`, character(1), 2),
    start = as.integer(vapply(m, `[`, character(1), 3)),
    end   = as.integer(vapply(m, `[`, character(1), 4))
  )

  # Belt and braces: the canonical filter should already have removed these.
  if (!all(bed$chr %in% chroms)) {
    stop("Non-canonical seqnames reached write_regions_bed(): ",
         paste(unique(setdiff(bed$chr, chroms)), collapse = ", "))
  }

  # Sort in chromosome order (chr1, chr2, ... chr10, ...), not alphabetically.
  bed <- bed[order(factor(bed$chr, levels = chroms), bed$start), ]

  write.table(bed, file = file, sep = "\t",
              quote = FALSE, row.names = FALSE, col.names = FALSE)
  message("Wrote ", nrow(bed), " regions to ", file)
}

#' Write results CSVs plus UP/DOWN BEDs for every contrast
#'
#' @param label  goes into the filenames, e.g. "H3K27ac" or "ATAC"
#' @param csv_suffix  trailing part of the CSV name (kept for backwards
#'                    compatibility with the original filenames)
export_results <- function(res_dfs, out_dir, label, csv_suffix, padj_cut = 0.05) {
  for (tag in names(res_dfs)) {
    d <- res_dfs[[tag]]

    write.csv(
      d,
      file      = file.path(out_dir, sprintf("DEseq2_%s_vs_ctrl_%s.csv", tag, csv_suffix)),
      quote     = FALSE,
      row.names = FALSE
    )

    write_regions_bed(
      d$region[which(d$padj < padj_cut & d$log2FoldChange > 0)],
      file.path(out_dir, sprintf("DEseq2_%s_vs_ctrl_%s_UP.bed", tag, label))
    )
    write_regions_bed(
      d$region[which(d$padj < padj_cut & d$log2FoldChange < 0)],
      file.path(out_dir, sprintf("DEseq2_%s_vs_ctrl_%s_DOWN.bed", tag, label))
    )
  }
  invisible(NULL)
}
```

# 1. H3K27ac ChIP-seq (lib4)

## Load counts

```{r k27ac-load}
counts_k27ac_all <- load_region_counts(counts_dir_k27ac, counts_pattern)

# Explicit column order: clone-major, then replicate, then timepoint.
# Listing the samples by name (rather than trusting list.files() order) makes
# the matrix reproducible and fails loudly if a sample is missing.
samples_k27ac <- c(
  "H3K27ac_c5946b16_DMSO_rep1", "H3K27ac_c5946b16_3h_rep1", "H3K27ac_c5946b16_1d_rep1", "H3K27ac_c5946b16_3d_rep1",
  "H3K27ac_c5946b16_DMSO_rep2", "H3K27ac_c5946b16_3h_rep2", "H3K27ac_c5946b16_1d_rep2", "H3K27ac_c5946b16_3d_rep2",
  "H3K27ac_c5946b27_DMSO_rep1", "H3K27ac_c5946b27_3h_rep1", "H3K27ac_c5946b27_1d_rep1", "H3K27ac_c5946b27_3d_rep1",
  "H3K27ac_c5946b27_DMSO_rep2", "H3K27ac_c5946b27_3h_rep2", "H3K27ac_c5946b27_1d_rep2", "H3K27ac_c5946b27_3d_rep2"
)

setdiff(samples_k27ac, colnames(counts_k27ac_all))   # should be character(0)
counts_k27ac <- counts_k27ac_all[, samples_k27ac]

dim(counts_k27ac)
head(counts_k27ac)
```

## Sample table

```{r k27ac-meta}
meta_k27ac <- parse_sample_metadata(colnames(counts_k27ac))

# Always eyeball this: every sample must sit next to the right condition.
meta_k27ac
table(meta_k27ac$condition, meta_k27ac$clone)
```

## DESeq2

```{r k27ac-deseq}
fit_k27ac <- run_deseq_regions(
  counts_mat  = counts_k27ac,
  meta        = meta_k27ac,
  contrasts   = contrast_levels,
  ref_level   = "DMSO",
  min_count   = min_count,
  min_samples = min_samples
)

resultsNames(fit_k27ac$dds)

res_k27ac <- lapply(fit_k27ac$res, res_to_df)

# Drop scaffolds / alt contigs / chrM before anything downstream uses the
# region coordinates.
res_k27ac <- lapply(res_k27ac, keep_canonical_regions)

deg_summary(res_k27ac, padj_cut)
```

## MA plots

```{r k27ac-ma, fig.width=10, fig.height=3}
p_k27ac <- ggarrange(
  ma_plot(res_k27ac$`3h`, "H3K27ac: 3h dTag vs mock"),
  ma_plot(res_k27ac$`1d`, "H3K27ac: 1 day dTag vs mock"),
  ma_plot(res_k27ac$`3d`, "H3K27ac: 3 days dTag vs mock"),
  nrow = 1, common.legend = TRUE, legend = "right"
)

p_k27ac
ggsave(file.path(fig_dir, "MAplots_k27ac.pdf"), plot = p_k27ac, height = 3, width = 10)
```

## Export tables and BED files

```{r k27ac-export}
export_results(
  res_k27ac,
  out_dir    = out_dir,
  label      = "H3K27ac",                        # -> ..._H3K27ac_UP.bed / _DOWN.bed
  csv_suffix = "H3K27ac_ATACregionsv3",          # -> DEseq2_3h_vs_ctrl_H3K27ac_ATACregionsv3.csv
  padj_cut   = padj_cut
)
```

# 2. ATAC-seq (lib1)

Same consensus regions, same design, same contrasts as the H3K27ac section.

**Note on the sample table.** In the original script the ATAC columns were
ordered replicate-major (`b16 rep1`, `b27 rep1`, `b16 rep2`, `b27 rep2`) while
the hard-coded `sampleTable` assumed the clone-major order used for H3K27ac.
The effect was that the `clone` and `replicate` labels were exactly swapped.
It did not change any reported statistic — the two factors are additive
nuisance terms, and swapping them spans the same two columns of the model
matrix, so the `condition` estimates are identical — but the labels were wrong.
Parsing metadata from the sample names fixes this.

## Load counts

```{r atac-load}
counts_atac_all <- load_region_counts(counts_dir_atac, counts_pattern)

# Replicate-major order, as the files come.
samples_atac <- c(
  "ATAC_SNAI12_c5946b16_DMSO_rep1", "ATAC_SNAI12_c5946b16_3h_rep1", "ATAC_SNAI12_c5946b16_1d_rep1", "ATAC_SNAI12_c5946b16_3d_rep1",
  "ATAC_SNAI12_c5946b27_DMSO_rep1", "ATAC_SNAI12_c5946b27_3h_rep1", "ATAC_SNAI12_c5946b27_1d_rep1", "ATAC_SNAI12_c5946b27_3d_rep1",
  "ATAC_SNAI12_c5946b16_DMSO_rep2", "ATAC_SNAI12_c5946b16_3h_rep2", "ATAC_SNAI12_c5946b16_1d_rep2", "ATAC_SNAI12_c5946b16_3d_rep2",
  "ATAC_SNAI12_c5946b27_DMSO_rep2", "ATAC_SNAI12_c5946b27_3h_rep2", "ATAC_SNAI12_c5946b27_1d_rep2", "ATAC_SNAI12_c5946b27_3d_rep2"
)

setdiff(samples_atac, colnames(counts_atac_all))   # should be character(0)
counts_atac <- counts_atac_all[, samples_atac]

dim(counts_atac)
head(counts_atac)
```

## Sample table

```{r atac-meta}
meta_atac <- parse_sample_metadata(colnames(counts_atac))

meta_atac
table(meta_atac$condition, meta_atac$clone)
```

## DESeq2

```{r atac-deseq}
fit_atac <- run_deseq_regions(
  counts_mat  = counts_atac,
  meta        = meta_atac,
  contrasts   = contrast_levels,
  ref_level   = "DMSO",
  min_count   = min_count,
  min_samples = min_samples
)

resultsNames(fit_atac$dds)

res_atac <- lapply(fit_atac$res, res_to_df)

# Drop scaffolds / alt contigs / chrM before anything downstream uses the
# region coordinates.
res_atac <- lapply(res_atac, keep_canonical_regions)

deg_summary(res_atac, padj_cut)
```

## MA plots

```{r atac-ma, fig.width=10, fig.height=3}
p_atac <- ggarrange(
  ma_plot(res_atac$`3h`, "ATAC: 3h dTag vs mock"),
  ma_plot(res_atac$`1d`, "ATAC: 1 day dTag vs mock"),
  ma_plot(res_atac$`3d`, "ATAC: 3 days dTag vs mock"),
  nrow = 1, common.legend = TRUE, legend = "right"
)

p_atac
ggsave(file.path(fig_dir, "MAplots_ATACdata.pdf"), plot = p_atac, height = 3, width = 10)
```

## Export tables and BED files

```{r atac-export}
export_results(
  res_atac,
  out_dir    = out_dir,
  label      = "ATAC",                           # -> ..._ATAC_UP.bed / _DOWN.bed
  csv_suffix = "ATACdata_ATACregionsv3",         # -> DEseq2_3h_vs_ctrl_ATACdata_ATACregionsv3.csv
  padj_cut   = padj_cut
)
```

# Combined summary

```{r summary}
bind_rows(
  deg_summary(res_k27ac, padj_cut) %>% mutate(assay = "H3K27ac"),
  deg_summary(res_atac,  padj_cut) %>% mutate(assay = "ATAC")
) %>%
  select(assay, contrast, up, down)
```

# Session info

```{r session-info}
sessionInfo()
```
