---
title: "Define proximal/distal regulatory regions and annotate SNAI1/2 peaks"
output: html_document
date: "2025-09-04"
---

## Overview

This notebook annotates a reference set of ATAC-seq regulatory regions as
**promoter-proximal** or **distal** (enhancer / other) and then overlaps them
with SNAI1/2 ChIP-seq peaks.

Reference regions were defined as those overlapping across all four DMSO-treated
samples and normalised to a 2 kb window centred on the highest-scoring summit.

Outputs include panels used in **Figures 2 and 3**.

### Required input files (place in the directories configured below)

| File | Expected location |
|---|---|
| SNAI2 ChIP peaks (narrowPeak) | `peaks_dir` |
| Histone/ATAC signals at reference ATAC regions (2 kb window) | `regions_dir` |
| Histone/ATAC signals at reference ATAC regions (6 kb window) | `regions_dir` |
| ATAC DMSO signal at reference ATAC regions | `regions_dir` |
| DEseq2 BED files (H3K27ac and ATAC, UP/DOWN per timepoint) | `regions_dir` |
| DEseq2 CSV files (log2FC per timepoint/mark) | `regions_dir` |
| FIMO motif BED file (SNAI2 DBD, Jolma 2013) | `fimo_dir` |
| MEME motif files | `motif_dir` |

---

```{r setup, include=FALSE}
# ── Libraries ────────────────────────────────────────────────────────────────
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(purrr)

library(GenomicRanges)
library(rtracklayer)
library(ChIPseeker)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(clusterProfiler)
library(AnnotationHub)

library(ggpubr)
library(ggrastr)
library(universalmotif)
library(patchwork)

txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

library(lme4)       # lmer()
library(lmerTest)   # Satterthwaite df / p-values for fixed effects
library(emmeans)    # emtrends(): per-timepoint slopes + pairwise contrasts


```

```{r paths}
# ── User-configurable paths ───────────────────────────────────────────────────
# Set these to the root directories that match your local clone or data mount.
# All file reads and writes below use these variables; no paths are hard-coded.

peaks_dir   <- ".../peaks"       # directory containing narrowPeak files
regions_dir <- ".../ATAC_regions" # directory containing signal tables and DEseq2 results
fimo_dir    <- ".../fimo"         # directory containing FIMO output BED
motif_dir   <- ".../motifs"       # directory containing MEME motif files
figures_dir <- ".../figures"           # output directory for saved PDFs

dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)

# ── Input file names (edit here if filenames differ) ─────────────────────────
snai2_peak_file   <- "s8_lib1_CNCCp4_SNAI2a3029_DMSO_r2_aSNAI2_reltodTag_peaks_minusINPUT_filtered.narrowPeak"
histones_2kb_file <- "histones_at_ATACregions_v3.txt"
histones_6kb_file <- "histones_at_ATACregions_v3_6kb.txt"
atac_dmso_file    <- "ATACDMSO_at_ATACregions_v3.txt"
fimo_bed_file     <- "SNAI2_DBD_jolma2013_fimo/fimo_sorted.bed"
```

---

# 1. Import SNAI2 ChIP-seq peaks

```{r import-snai2-peaks}
# Import the MACS2 narrowPeak file as a GRanges object.
# The 'peak' metadata column holds the 0-based offset of the summit from the
# peak start, which is used later for summit-centred overlap resolution.
peaks_gr <- import(
  file.path(peaks_dir, snai2_peak_file),
  format = "narrowPeak"
)
message("Imported ", length(peaks_gr), " SNAI2 peaks")
```

---

# 2. Import histone mark and ATAC signals at reference ATAC regions

## 2a. Histone signals (2 kb and 6 kb windows)

```{r import-histone-signals}
# Helper: read a deeptools-style matrix text file that has a comment-line header
# starting with '#'. Returns a data frame with cleaned column names and a
# 'region' key column (chr_start_end).
read_deeptools_table <- function(path) {
  header_line <- readLines(path, n = 1)
  col_names   <- strsplit(sub("^#", "", header_line), "\t")[[1]]
  col_names   <- gsub("'", "", col_names)
  df          <- read_tsv(path, skip = 1, col_names = col_names,
                          show_col_types = FALSE)
  df$region   <- paste(df$chr, df$start, df$end, sep = "_")
  df
}

# 2 kb window table – used for histone signal values and as the region key
data1 <- read_deeptools_table(file.path(regions_dir, histones_2kb_file))

# 6 kb window table – broader context for H3K27me3; regions are in the same
# order as the 2 kb table, so we reuse the 2 kb region key
data2 <- read_deeptools_table(file.path(regions_dir, histones_6kb_file))
data2$region <- data1$region   # align region IDs (same peaks, larger window)

# Drop coordinate columns now that 'region' is the key; rename the duplicate
# K27me3 column in data1 to avoid clashing with the 6 kb version
data1 <- data1[, -(1:3)]
colnames(data1)[colnames(data1) == "H3K27me3_DMSO_merged.bw"] <- "old_k27me3"

# Keep only the additional 6 kb columns (drop the 2 kb redundant ones)
data2 <- data2[, -(1:8)]

# Merge the two windows into a single data frame keyed by 'region'
data <- merge(data1, data2, by = "region")
```

## 2b. Calculate histone mark log2 ratios

```{r calc-log2-ratios}
# H3K4me1 / H3K4me3 ratio: high values indicate enhancer-like chromatin;
# low values indicate promoter-like chromatin.
# A pseudocount of 10 is added to both numerator and denominator to stabilise
# ratios at low-signal regions and avoid log(0).
data$K4me1me3_log2ratio <- log2(
  (((data$H3K4me1_human1_rep1_SRR2096413.bw + data$H3K4me1_human3_rep1_SRR2096414.bw) / 2) + 10) /
  (((data$H3K4me3_human1_rep1_SRR2096417.bw + data$H3K4me3_human3_rep1_SRR2096419.bw) / 2) + 10)
)

# H3K27ac / H3K27me3 ratio: high values indicate active chromatin;
# low values indicate Polycomb-repressed regions.
data$K27acK27me3_log2ratio <- log2(
  (data$H3K27ac_DMSO_merged.bw + 10) /
  (data$H3K27me3_DMSO_merged.bw + 10)
)
```

## 2c. ATAC-seq signal (DMSO baseline)

```{r import-atac}
# Read the ATAC-seq signal table (same structure as the histone tables above).
# Only the 'region' key and the ATAC signal column are retained.
atac_df <- read_deeptools_table(file.path(regions_dir, atac_dmso_file))
atac_df  <- atac_df[, c("region",
                         setdiff(colnames(atac_df),
                                 c("chr", "start", "end", "region")))[1:2]]
data     <- merge(data, atac_df, by = "region")
```

---

# 3. Annotate regions as TSS-proximal or distal

Uses **Ensembl 111** canonical protein-coding TSSs (AH116291 from AnnotationHub).
A region is called *proximal* if its centre lies within **1 kb** of a canonical TSS.

```{r tss-annotation}
# Convert 'region' string keys back to GRanges for overlap operations
data_parsed <- data %>%
  separate(region, into = c("chr", "start", "end"),
           sep = "_", remove = FALSE, convert = TRUE)

data_gr <- GRanges(
  seqnames = data_parsed$chr,
  ranges   = IRanges(start = data_parsed$start,
                     end   = data_parsed$end)
)

# Retrieve Ensembl 111 EnsDb from AnnotationHub and extract canonical TSS positions
ah         <- AnnotationHub()
edb_latest <- ah[["AH116291"]]   # Ensembl 111, Homo sapiens

tx_gr  <- transcripts(edb_latest,
                       filter = ~ gene_biotype == "protein_coding" &
                                  tx_is_canonical == TRUE)
tss_gr <- resize(tx_gr, width = 1, fix = "start")

# Harmonise chromosome naming style (Ensembl → UCSC: e.g. "1" → "chr1")
seqlevelsStyle(tss_gr) <- "UCSC"

# Compute distance from each reference region to the nearest canonical TSS
nearest_hits            <- distanceToNearest(data_gr, tss_gr)
data_gr$dist_to_pc_TSS  <- mcols(nearest_hits)$distance
data_gr$within_1kb_TSS  <- data_gr$dist_to_pc_TSS <= 1000

message("TSS-proximal regions (≤1 kb): ",
        sum(data_gr$within_1kb_TSS), " / ", length(data_gr))

# Propagate the TSS-proximity flag back to the data frame
data <- data %>%
  mutate(within_1kb_TSS = ifelse(data_gr$within_1kb_TSS,
                                 "within_1kb_TSS",
                                 "not_within_1kb_TSS"))
```

---

# 4. Annotate SNAI1/2 peak overlap

```{r snai-overlap}
# Binary overlap: does any part of a SNAI2 peak intersect a reference region?
overlaps_peak <- overlapsAny(data_gr, peaks_gr)

data <- data %>%
  mutate(SNAI_peaks_all = ifelse(overlaps_peak,
                                 "SNAI_bound",
                                 "SNAI_NOT_bound"))

# Cross-tabulate TSS proximity vs SNAI binding as a sanity check
print(table(TSS_proximity = data$within_1kb_TSS,
            SNAI_binding  = data$SNAI_peaks_all))
```

---

# 5. Classify regions as promoter-proximal or distal

Classification logic (applied in this order of priority):

1. **Promoter** – within 1 kb of a canonical TSS.
2. **Promoter** (distal override) – high H3K4me3 signal (K4me1/me3 ratio ≤ −0.2), regardless of TSS distance.
3. **H3K27me3-enriched** – strong Polycomb repression (K27ac/K27me3 ratio ≤ −1).
4. **Enhancer** – distal, H3K4me1-enriched (ratio > −0.2) **and** active (K27ac/K27me3 ≥ 1).
5. **Other distal** – distal, H3K4me1-enriched but less strongly active (−1 < K27ac/K27me3 < 1).

```{r classify-regions}
# Step 1: TSS-proximal → promoter
data$classification <- ifelse(
  data$within_1kb_TSS == "within_1kb_TSS", "promoter", "other"
)

# Step 2: Distal but K4me3-high → reclassify as promoter
data$classification <- ifelse(
  data$within_1kb_TSS == "not_within_1kb_TSS" &
    data$K4me1me3_log2ratio <= -0.2 &
    data$K27acK27me3_log2ratio > -1,
  "promoter", data$classification
)

# Step 3: Polycomb-repressed
data$classification <- ifelse(
  data$within_1kb_TSS == "not_within_1kb_TSS" &
    data$K27acK27me3_log2ratio <= -1,
  "H3K27me3_enriched", data$classification
)

# Step 4: Active enhancer
data$classification <- ifelse(
  data$within_1kb_TSS == "not_within_1kb_TSS" &
    data$K4me1me3_log2ratio > -0.2 &
    data$K27acK27me3_log2ratio >= 1,
  "enhancer", data$classification
)

# Step 5: Other distal (H3K4me1-enriched but not strongly active)
data$classification <- ifelse(
  data$within_1kb_TSS == "not_within_1kb_TSS" &
    data$K4me1me3_log2ratio > -0.2 &
    data$K27acK27me3_log2ratio > -1 &
    data$K27acK27me3_log2ratio < 1,
  "other_distal", data$classification
)

print(table(data$classification))
```

## 5a. Scatter plot – all regions coloured by detailed classification


```{r}
cls_cols <- c(
  "promoter"          = "#0072B2",
  "enhancer"          = "#E69F00",
  "other_distal"      = "#F0E442",
  "H3K27me3_enriched" = "#009E73"
)

common_layers <- list(
  geom_point_rast(size = 0.2, alpha = 0.2),
  scale_color_manual(values = cls_cols),
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))),
  theme_bw(),
  labs(title = "", color = "Region type"),
  geom_vline(xintercept = -0.2, linetype = "dashed", color = "grey"),
  geom_hline(yintercept = -1,   linetype = "dashed", color = "grey"),
  coord_cartesian(xlim = range(data$K4me1me3_log2ratio, na.rm = TRUE),
                  ylim = range(data$K27acK27me3_log2ratio, na.rm = TRUE))
)

base_aes <- aes(x = K4me1me3_log2ratio,
                y = K27acK27me3_log2ratio,
                color = classification)

p1 <- ggplot(data, base_aes) + common_layers

p2 <- ggplot(data[data$SNAI_peaks_all %in% "SNAI_bound", ], base_aes) + common_layers

ggarrange(p1, p2, 
          ncol = 2, 
          common.legend = TRUE, 
          legend = "right",
  align = "h")

ggsave(file.path(figures_dir, "ATAC_regions_classification_plot_v4.pdf"),
       height = 3.5, width = 7.5)

```




---

# 6. Assign each SNAI peak to proximal or distal class and export BED files

Peaks overlapping more than one reference region are resolved by choosing the
region whose centre is closest to the MACS2 summit position.

These peaks were used for Heatmap in Figure 2 (deeptools computeMatrixand plotHeatmap)
```{r assign-snai-peaks}

# ── Reconstruct GRanges from the region string keys ──────────────────────────
# NOTE: BED coordinates are 0-based half-open intervals; GRanges are 1-based
# closed intervals, so we add 1 to the start.
region_parts <- do.call(rbind, strsplit(as.character(data$region), "_"))
data_gr <- GRanges(
  seqnames       = region_parts[, 1],
  ranges         = IRanges(start = as.numeric(region_parts[, 2]) + 1,
                           end   = as.numeric(region_parts[, 3])),
  classification = data$classification
)
data_centers <- start(data_gr) + (width(data_gr) / 2)

# ── Convert MACS2 summit offset to absolute genomic coordinates ───────────────
# The 'peak' column stores the 0-based offset from the peak start.
summit_positions <- start(peaks_gr) + peaks_gr$peak

# ── Find all peak–region overlaps (full peak body) ───────────────────────────
hits    <- findOverlaps(peaks_gr, data_gr)
hits_df <- data.frame(
  peak_idx   = queryHits(hits),
  region_idx = subjectHits(hits)
)

# Compute summit-to-region-centre distance for tie-breaking
hits_df$summit_pos    <- summit_positions[hits_df$peak_idx]
hits_df$region_center <- data_centers[hits_df$region_idx]
hits_df$distance      <- abs(hits_df$summit_pos - hits_df$region_center)

# Keep the closest region match for each peak
hits_df       <- hits_df[order(hits_df$distance), ]
hits_resolved <- hits_df[!duplicated(hits_df$peak_idx), ]

# ── Extract classified peaks ──────────────────────────────────────────────────
matched_peaks_gr               <- peaks_gr[hits_resolved$peak_idx]
matched_peaks_gr$matched_class <- data_gr$classification[hits_resolved$region_idx]

promoter_peaks <- matched_peaks_gr[matched_peaks_gr$matched_class == "promoter"]
distal_peaks   <- matched_peaks_gr[matched_peaks_gr$matched_class %in% c("enhancer", "other_distal")]

# Remove the temporary metadata column before export
promoter_peaks$matched_class <- NULL
distal_peaks$matched_class   <- NULL

# ── Export BED files ─────────────────────────────────────────────────────────
# rtracklayer::export.bed automatically converts back to 0-based BED coordinates.
export.bed(promoter_peaks, "/Volumes/storage/SNAI_paper/peaks/s8_lib1_CNCCp4_SNAI2a3029_DMSO_r2_aSNAI2_reltodTag_peaks_minusINPUT_filtered_PROXIMALcomb_v7.bed")
export.bed(distal_peaks,   "/Volumes/storage/SNAI_paper/peaks/s8_lib1_CNCCp4_SNAI2a3029_DMSO_r2_aSNAI2_reltodTag_peaks_minusINPUT_filtered_DISTALcomb_v7.bed")

message("Promoter-assigned peaks: ", length(promoter_peaks))
message("Distal-assigned peaks:   ", length(distal_peaks))

# ── Export BED files ─────────────────────────────────────────────────────────
# rtracklayer::export.bed automatically converts back to 0-based BED coordinates.
export.bed(promoter_peaks,
           file.path(peaks_dir,
                     "SNAI2_peaks_PROXIMALcomb_v7.bed"))
export.bed(distal_peaks,
           file.path(peaks_dir,
                     "SNAI2_peaks_DISTALcomb_v7.bed"))
```



---

# 7. Annotate regions with DEseq2 differential results (H3K27ac and ATAC)

Reads BED files produced by DEseq2 for significantly UP or DOWN regions at
3 h, 1 day, and 3 day dTag time points, and flags each reference region
accordingly (binary 1/0 columns).

```{r import-deseq-bed}
# Helper: read a DEseq2 BED file and return a vector of 'region' keys
read_deseq_bed <- function(fname) {
  df <- read_table(file.path(regions_dir, fname), col_names = FALSE,
                   show_col_types = FALSE)
  paste(df$X1, df$X2, df$X3, sep = "_")
}

# H3K27ac UP regions
H3K27ac_UP_3h <- read_deseq_bed("DEseq2_3h_vs_ctrl_H3K27ac_UP.bed")
H3K27ac_UP_1d <- read_deseq_bed("DEseq2_1d_vs_ctrl_H3K27ac_UP.bed")
H3K27ac_UP_3d <- read_deseq_bed("DEseq2_3d_vs_ctrl_H3K27ac_UP.bed")

# H3K27ac DOWN regions
H3K27ac_DW_3h <- read_deseq_bed("DEseq2_3h_vs_ctrl_H3K27ac_DOWN.bed")
H3K27ac_DW_1d <- read_deseq_bed("DEseq2_1d_vs_ctrl_H3K27ac_DOWN.bed")
H3K27ac_DW_3d <- read_deseq_bed("DEseq2_3d_vs_ctrl_H3K27ac_DOWN.bed")

# ATAC UP regions
ATAC_UP_3h <- read_deseq_bed("DEseq2_3h_vs_ctrl_ATAC_UP.bed")
ATAC_UP_1d <- read_deseq_bed("DEseq2_1d_vs_ctrl_ATAC_UP.bed")
ATAC_UP_3d <- read_deseq_bed("DEseq2_3d_vs_ctrl_ATAC_UP.bed")

# ATAC DOWN regions
ATAC_DW_3h <- read_deseq_bed("DEseq2_3h_vs_ctrl_ATAC_DOWN.bed")
ATAC_DW_1d <- read_deseq_bed("DEseq2_1d_vs_ctrl_ATAC_DOWN.bed")
ATAC_DW_3d <- read_deseq_bed("DEseq2_3d_vs_ctrl_ATAC_DOWN.bed")

# Flag each reference region (binary 1/0 columns)
flag_regions <- function(df, ref_regions, col_name) {
  df[[col_name]] <- as.integer(df$region %in% ref_regions)
  df
}

data <- data %>%
  flag_regions(H3K27ac_UP_3h, "H3K27ac_UP_3h") %>%
  flag_regions(H3K27ac_UP_1d, "H3K27ac_UP_1d") %>%
  flag_regions(H3K27ac_UP_3d, "H3K27ac_UP_3d") %>%
  flag_regions(H3K27ac_DW_3h, "H3K27ac_DW_3h") %>%
  flag_regions(H3K27ac_DW_1d, "H3K27ac_DW_1d") %>%
  flag_regions(H3K27ac_DW_3d, "H3K27ac_DW_3d") %>%
  flag_regions(ATAC_UP_3h,    "ATAC_UP_3h")    %>%
  flag_regions(ATAC_UP_1d,    "ATAC_UP_1d")    %>%
  flag_regions(ATAC_UP_3d,    "ATAC_UP_3d")    %>%
  flag_regions(ATAC_DW_3h,    "ATAC_DW_3h")    %>%
  flag_regions(ATAC_DW_1d,    "ATAC_DW_1d")    %>%
  flag_regions(ATAC_DW_3d,    "ATAC_DW_3d")
```

## 7a. Import DEseq2 log2 fold-changes

```{r import-deseq-log2fc}
# Each CSV contains one row per reference region with a log2FC column.
# File names encode the comparison (time point) and assay (mark):
#   DEseq2_<time>_vs_ctrl_<mark>_ATACregionsv3.csv
# The 'parts' vector is parsed to build a tidy column name: <mark>_<time>_log2FC.

deseq_files <- c(
  "DEseq2_3h_vs_ctrl_H3K27ac_ATACregionsv3",
  "DEseq2_1d_vs_ctrl_H3K27ac_ATACregionsv3",
  "DEseq2_3d_vs_ctrl_H3K27ac_ATACregionsv3",
  "DEseq2_3h_vs_ctrl_ATACdata_ATACregionsv3",
  "DEseq2_1d_vs_ctrl_ATACdata_ATACregionsv3",
  "DEseq2_3d_vs_ctrl_ATACdata_ATACregionsv3"
)

fc_list <- map(deseq_files, function(f) {
  parts <- strsplit(f, "_")[[1]]
  time  <- parts[2]   # "3h", "1d", or "3d"
  mark  <- parts[5]   # "H3K27ac" or "ATACdata"

  df <- read_csv(file.path(regions_dir, paste0(f, ".csv")),
                 show_col_types = FALSE)
  df <- df[, c(1, 3)]   # keep region ID and log2FC columns
  colnames(df) <- c("region", paste0(mark, "_", time, "_log2FC"))
  df
})

# Left-join all log2FC tables onto the main data frame
data <- purrr::reduce(
  fc_list,
  function(x, y) merge(x, y, by = "region", all.x = TRUE),
  .init = data
)
```

## 7b. Save annotated data

```{r save-data}
saveRDS(data, file = file.path(regions_dir, "data_v4.rds"))
# To reload: data <- readRDS(file.path(regions_dir, "data_v4.rds"))
```

---

# 8. Stacked-bar plots: composition of differential regions

## 8a. SNAI-bound UP regions only (detailed classification; Figure 2/3 panel)

```{r barplot-bound-up}

# Removed all the _DW_ columns from this vector
diff_cols <- c(
  "H3K27ac_UP_3h", "H3K27ac_UP_1d", "H3K27ac_UP_3d",
  "ATAC_UP_3h", "ATAC_UP_1d", "ATAC_UP_3d"
)

data <- data %>%
  mutate(across(all_of(diff_cols), as.numeric))

long <- data %>%
  dplyr::filter(SNAI_peaks_all == "SNAI_bound") %>% # take only bound
  dplyr::select(region, classification, all_of(diff_cols)) %>%
  pivot_longer(all_of(diff_cols),
               names_to = "diff_set",
               values_to = "is_diff") %>%
  dplyr::filter(is_diff == 1) %>%
  # Split the set name into mark / direction / time so we can facet
  separate(diff_set, into = c("mark", "direction", "time"),
           sep = "_", remove = FALSE) %>%
  mutate(
    time            = factor(time, levels = c("3h", "1d", "3d")),
    direction       = factor(direction, levels = c("UP")), # Removed "DW" here
    mark            = factor(mark, levels = c("H3K27ac", "ATAC")),
    classification = factor(classification,
                             levels = c("promoter",  "enhancer",
                                        "other_distal","H3K27me3_enriched"))
  )

totals <- long %>% dplyr::count(direction, mark, time, name = "n")

ggplot(long, aes(x = time, fill = classification)) +
  geom_bar(position = "fill") +
  geom_text(aes(label = after_stat(count)),
            stat = "count",
            position = position_fill(vjust = 0.5),
            size = 2.5) +
  #geom_text(data = totals,
  #          aes(x = time, y = 1.03, label = paste0("n=", n)),
  #          inherit.aes = FALSE, size = 2.8) +
  facet_wrap(~ mark + direction, ncol = 1) +
  scale_fill_manual(values = c(
    "promoter"      = "#0072B2",
    "enhancer"      = "#E69F00",
    "other_distal"  = "#F0E442",
    "H3K27me3_enriched" = "#009E73"
  )) +
  scale_y_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.08))) +
  labs(x = "Timepoint", y = "Fraction of differential regions",
       fill = "Region classification",
       title = "") + 
  theme_bw() +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold"))

ggsave(file.path(figures_dir, "ATAC_regions_diff_annotations_noK27me_bound_up.pdf"),
       height = 5, width = 4.5)
```

## 8b. All regions, UP and DOWN (full classification; supplementary figure)

```{r barplot-all-diff}
diff_cols_all <- c(
  "H3K27ac_UP_3h", "H3K27ac_UP_1d", "H3K27ac_UP_3d",
  "H3K27ac_DW_3h", "H3K27ac_DW_1d", "H3K27ac_DW_3d",
  "ATAC_UP_3h",    "ATAC_UP_1d",    "ATAC_UP_3d",
  "ATAC_DW_3h",    "ATAC_DW_1d",    "ATAC_DW_3d"
)

long_all <- data %>%
  dplyr::select(region, classification, all_of(diff_cols_all)) %>%
  pivot_longer(all_of(diff_cols_all), names_to = "diff_set", values_to = "is_diff") %>%
  dplyr::filter(is_diff == 1) %>%
  separate(diff_set, into = c("mark", "direction", "time"),
           sep = "_", remove = FALSE) %>%
  mutate(
    time            = factor(time,            levels = c("3h", "1d", "3d")),
    direction       = factor(direction,       levels = c("UP", "DW")),
    mark            = factor(mark,            levels = c("H3K27ac", "ATAC")),
    classification = factor(classification, levels = c("promoter", "enhancer", "other_distal","H3K27me3_enriched"))
  )

ggplot(long_all, aes(x = time, fill = classification)) +
  geom_bar(position = "fill") +
  geom_text(aes(label = after_stat(count)),
            stat = "count",
            position = position_fill(vjust = 0.5),
            size = 2.5) +
  facet_wrap(~ mark + direction, ncol = 1) +
  scale_fill_manual(values = c(
    "promoter"     = "#0072B2",
    "enhancer"     = "#E69F00",
    "other_distal" = "#F0E442"
  )) +
  scale_y_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.08))) +
  labs(x = "Timepoint", y = "Fraction of differential regions",
       fill = "Region classification") +
  theme_bw() +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold"))

ggsave(file.path(figures_dir, "ATAC_regions_diff_annotations_noK27me.pdf"),
       height = 7, width = 5)
```

## 8c. SNAI binding composition of differential regions

```{r barplot-snai-binding}
long_snai <- data %>%
  dplyr::select(region, SNAI_peaks_all, all_of(diff_cols_all)) %>%
  pivot_longer(all_of(diff_cols_all), names_to = "diff_set", values_to = "is_diff") %>%
  dplyr::filter(is_diff == 1) %>%
  separate(diff_set, into = c("mark", "direction", "time"),
           sep = "_", remove = FALSE) %>%
  mutate(
    time           = factor(time,           levels = c("3h", "1d", "3d")),
    direction      = factor(direction,      levels = c("UP", "DW")),
    mark           = factor(mark,           levels = c("H3K27ac", "ATAC")),
    SNAI_peaks_all = factor(SNAI_peaks_all, levels = c("SNAI_bound", "SNAI_NOT_bound"))
  )

ggplot(long_snai, aes(x = time, fill = SNAI_peaks_all)) +
  geom_bar(position = "fill") +
  geom_text(aes(label = after_stat(count)),
            stat = "count",
            position = position_fill(vjust = 0.5),
            size = 2.8) +
  facet_wrap(~ mark + direction, ncol = 1) +
  scale_fill_manual(values = c(
    "SNAI_bound"     = "#D55E00",
    "SNAI_NOT_bound" = "grey70"
  )) +
  scale_y_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.08))) +
  labs(x = "Timepoint", y = "Fraction of differential regions",
       fill = "SNAI binding") +
  theme_bw() +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold"))

ggsave(file.path(figures_dir, "ATAC_regions_diff_SNAI_binding_noK27me.pdf"),
       height = 7, width = 5)
```

---

# 9. Extract SNAI2 motif centres at distal peaks

Used as input for aggregate/heatmap plots centred on the SNAI motif.

```{r motif-centres}
SNAI_enhancer_peaks <- distal_peaks

# Read the FIMO output BED (0-based half-open coordinates)
fimo_df <- read.table(
  file.path(fimo_dir, fimo_bed_file),
  sep = "\t", header = FALSE, stringsAsFactors = FALSE, comment.char = "#"
)

motifs_gr <- GRanges(
  seqnames = fimo_df$V1,
  ranges   = IRanges(start = fimo_df$V2 + 1, end = fimo_df$V3),  # 0-based → 1-based
  strand   = if (ncol(fimo_df) >= 6) fimo_df$V6 else "*"
)

# Build ±100 bp windows around each SNAI2 peak summit
summit_pos     <- start(SNAI_enhancer_peaks) + SNAI_enhancer_peaks$peak
summit_windows <- GRanges(
  seqnames = seqnames(SNAI_enhancer_peaks),
  ranges   = IRanges(start = summit_pos - 100, end = summit_pos + 100)
)

# Retain only motif instances that fall within summit windows
motifs_in_peaks <- subsetByOverlaps(motifs_gr, summit_windows)
message("Motif instances within ±100 bp of SNAI2 summits: ", length(motifs_in_peaks))

# Collapse each motif to its 1 bp centre for downstream meta-peak analysis
motif_centers <- resize(motifs_in_peaks, width = 1, fix = "center")

export(motif_centers,
       file.path(peaks_dir, "SNAI2_peaks_DISTAL_v7_motifcenters.bed"),
       format = "BED")
```


---

# 10. Density plots: H3K27ac enrichment across region subsets

Compares baseline H3K27ac levels between all distal regions, SNAI-bound distal
regions, and those that become UP at each time point (only SNAI-bound).

```{r density-k27ac}
data_dist <- data[data$classification %in% c("other_distal", "enhancer"), ]

# Build the combined data frame with group labels
subsets <- list(
  all_distal = data_dist,
  bound      = data_dist[data_dist$SNAI_peaks_all == "SNAI_bound", ],
  UP_3h      = data_dist[data_dist$H3K27ac_UP_3h == 1 & data_dist$SNAI_peaks_all == "SNAI_bound", ],
  UP_1d      = data_dist[data_dist$H3K27ac_UP_1d == 1 & data_dist$SNAI_peaks_all == "SNAI_bound", ],
  UP_3d      = data_dist[data_dist$H3K27ac_UP_3d == 1 & data_dist$SNAI_peaks_all == "SNAI_bound", ]
)

for (nm in names(subsets)) subsets[[nm]]$group_K27ac <- nm

comb <- do.call(rbind, subsets)
comb$group_K27ac <- factor(comb$group_K27ac,
                            levels = c("all_distal", "bound", "UP_3h", "UP_1d", "UP_3d"))

# Build legend labels that include group sizes
counts     <- table(comb$group_K27ac)
new_labels <- setNames(paste0(names(counts), " (n=", counts, ")"), names(counts))

ggplot(comb, aes(x = log(H3K27ac_DMSO_merged.bw),
                 color = group_K27ac, fill = group_K27ac)) +
  geom_density(alpha = 0.2) +
  scale_color_discrete(labels = new_labels) +
  scale_fill_discrete(labels  = new_labels) +
  labs(x = "log(H3K27ac)", y = "Density",
       title = "H3K27ac baseline at UP regions") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        legend.title = element_blank())

ggsave(file.path(figures_dir, "SNAI_enh_strength_dist_k2only.pdf"),
       width = 5, height = 2.5)
```

---

# 11. Motif logos

```{r motif-logos-snai2}
# SNAI2 DBD motif (Jolma 2013) – used in main figures
motif_files_snai <- c(
  "SNAI2_JOLMA2013" = "SNAI2_DBD_jolma2013.meme"
)

motif_list_snai <- lapply(names(motif_files_snai), function(n) {
  m         <- read_meme(file.path(motif_dir, motif_files_snai[[n]]))
  m["name"] <- n
  m
})

view_motifs(motif_list_snai)
ggsave(file.path(figures_dir, "SNAI_motif_logos_jolma2013.pdf"),
       width = 5, height = 2)
```

```{r motif-logos-cobinding}
# Co-binding partner motifs: coordinator, NR2F1 (COUP-TFII), TFAP2A, SNAI
motif_files_co <- c(
  "coordinator" = "WAATNRRAAVCABMTGB.meme",
  "NR2F1"       = "COT2.H14CORE.1.P.B_meme_format.meme",
  "TFAP2A"      = "AP2A.H14CORE.0.PSM.A_meme_format.meme",
  "SNAI"        = "SNAI1.H14CORE.0.PSM.A_meme_format.meme"
)

motif_list_co <- lapply(names(motif_files_co), function(n) {
  m         <- read_meme(file.path(motif_dir, motif_files_co[[n]]))
  m["name"] <- n
  m
})

# Save each motif logo as a separate PDF
for (i in seq_along(motif_list_co)) {
  motif_name <- names(motif_files_co)[i]
  p          <- view_motifs(motif_list_co[[i]])
  ggsave(
    filename = file.path(figures_dir, paste0(motif_name, "_motif_logo.pdf")),
    plot     = p,
    width    = 5,
    height   = 2
  )
}
```




