---
title: "Fig4 — Linear model: TF binding changes predict H3K27ac change"
output: html_document
date: "2026-08-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# 0. Setup: paths, constants, libraries

All file paths and analysis constants live here so nothing downstream is hardcoded.

```{r libraries}
library(tidyverse)
library(GenomicRanges)
library(stringr)
```

```{r paths}
# ---- Input paths --------------------------------------------------------
DATA_RDS <- ".../data_v4.rds"

CHIP_DIR <- ".../DEseq"
FC_FILES <- c(
  TWIST1 = file.path(CHIP_DIR, "DEseq2_TWIST1_peaks.csv"),
  NR2F1  = file.path(CHIP_DIR, "DEseq2_NR2F1_peaks.csv"),
  AP2A   = file.path(CHIP_DIR, "DEseq2_AP2A_peaks.csv")
)

# ---- Output path --------------------------------------------------------
FIG_DIR <- paste0(
  ".../figures"
)
FIG_OUT <- file.path(FIG_DIR, "linear_model_K27ac_k10_v2.pdf")

# ---- Analysis constants -------------------------------------------------
OUTCOME         <- "H3K27ac_3h_log2FC"
KEEP_CLASSES    <- c("other_distal", "enhancer")
K_FOLDS         <- 10
SEED            <- 42

# Canonical chromosomes only: autosomes + sex chromosomes.
# chr1:23 is included so this also works on assemblies/organisms where chr23
# exists as a real autosome; in human, chr23 is simply absent and the filter
# is a no-op for it. Both "chr1" and "1" naming styles are accepted.
CANONICAL_CHR <- c(
  paste0("chr", c(as.character(1:23), "X", "Y")),
  c(as.character(1:23), "X", "Y")
)

stopifnot(file.exists(DATA_RDS), all(file.exists(FC_FILES)), dir.exists(FIG_DIR))
```

# 1. Helper functions

```{r helpers}
#' Parse "chr_start_end" region strings, anchored from the RIGHT.
#'
#' Splitting on "_" from the left breaks on scaffold names that themselves
#' contain underscores (e.g. "chr1_KI270706v1_random_100_200"). The regex below
#' takes the LAST two underscore-separated integer fields as start/end and
#' everything before them as the chromosome, so such names parse correctly
#' instead of silently turning into NA and being dropped.
parse_region <- function(x) {
  m <- stringr::str_match(x, "^(.+)_(\\d+)_(\\d+)$")
  tibble::tibble(
    chr   = m[, 2],
    start = as.numeric(m[, 3]),
    end   = as.numeric(m[, 4])
  )
}

#' Load a DESeq2 peak CSV (rownames in column `X` as "chr_start_end") as GRanges.
#'
#' Non-canonical chromosomes are dropped here on purpose, matching the filter
#' applied to the main data, so overlap counts are computed on a consistent
#' chromosome set.
load_fc_gr <- function(path, keep_chr = CANONICAL_CHR) {
  raw <- read.csv(path, stringsAsFactors = FALSE)
  stopifnot("X" %in% names(raw), "log2FoldChange" %in% names(raw))

  coords <- parse_region(raw$X)
  n_unparsed <- sum(is.na(coords$start))
  if (n_unparsed > 0) {
    warning(sprintf("%s: %d region strings could not be parsed and were dropped",
                    basename(path), n_unparsed))
  }

  out <- dplyr::bind_cols(coords, raw |> dplyr::select(-X)) |>
    dplyr::filter(!is.na(start), !is.na(end), chr %in% keep_chr)

  message(sprintf("%s: %d peaks loaded (%d after canonical-chr + parse filter)",
                  basename(path), nrow(raw), nrow(out)))

  makeGRangesFromDataFrame(out, keep.extra.columns = TRUE)
}

#' Mean log2FC of all peaks overlapping each query region.
#'
#' Returns a numeric vector of length(query_gr), positionally aligned with
#' query_gr (and therefore with the data.frame it was built from). Regions with
#' no overlapping peak get NA.
#'
#' NOTE: this is an UNWEIGHTED mean — a 50 bp sliver of overlap counts as much
#' as a fully contained peak. If that matters, switch to a width-weighted mean
#' or take the peak with the largest |log2FC|.
attach_fc_mean <- function(query_gr, fc_gr) {
  hits <- findOverlaps(query_gr, fc_gr)

  fc_values <- mcols(fc_gr)$log2FoldChange[subjectHits(hits)]
  # tapply names are the (sorted, unique) query indices as character strings
  mean_fcs  <- tapply(fc_values, queryHits(hits), mean, na.rm = TRUE)

  result <- rep(NA_real_, length(query_gr))
  result[as.integer(names(mean_fcs))] <- as.numeric(mean_fcs)
  # mean(all-NA, na.rm = TRUE) yields NaN; normalise those back to NA
  result[is.nan(result)] <- NA_real_
  result
}
```

# 2. Load data and restrict to canonical chromosomes

The chromosome filter is applied to `data` **before** building the GRanges, so
that `data_gr[i]` always corresponds to `data[i, ]`. Filtering after the fact
would break the positional indexing that `attach_fc_mean()` relies on.

```{r load-data}
data_raw <- readRDS(DATA_RDS)
stopifnot("region" %in% names(data_raw))

coords <- parse_region(data_raw$region)

data <- data_raw |>
  dplyr::bind_cols(coords) |>
  dplyr::filter(!is.na(start), !is.na(end), chr %in% CANONICAL_CHR)

cat(sprintf("Regions: %d total -> %d on canonical chromosomes\n",
            nrow(data_raw), nrow(data)))
cat("Chromosomes retained:\n"); print(sort(unique(data$chr)))

data_gr <- makeGRangesFromDataFrame(
  data,
  seqnames.field   = "chr",
  start.field      = "start",
  end.field        = "end",
  keep.extra.columns = FALSE
)

# Guard the positional-alignment assumption made by attach_fc_mean().
stopifnot(length(data_gr) == nrow(data))
```

# 3. Attach TF log2FC to each region

```{r attach-fc}
fc_grs <- purrr::map(FC_FILES, load_fc_gr)

PREDICTORS <- paste0(names(FC_FILES), "_log2FC")   # TWIST1_log2FC, ...

for (i in seq_along(fc_grs)) {
  data[[PREDICTORS[i]]] <- attach_fc_mean(data_gr, fc_grs[[i]])
}

# Coverage check: how many regions actually got a value for each TF?
data |>
  dplyr::select(all_of(PREDICTORS)) |>
  purrr::map_dfr(~ tibble(n_with_FC = sum(!is.na(.x)), n_NA = sum(is.na(.x))),
                 .id = "predictor") |>
  print()

summary(as.data.frame(data[, PREDICTORS]))
```

# 4. Model set

Eight models: the null (intercept-only) baseline, 3 singles, 3 pairs, 1 triple.
The null model is the reference line — any model whose cross-validated R²
overlaps it is not predicting anything.

```{r model-specs}
combos <- c(
  list(character(0)),                              # intercept-only
  unlist(lapply(1:3, function(k) combn(PREDICTORS, k, simplify = FALSE)),
         recursive = FALSE)
)

model_specs <- purrr::map(combos, function(vars) {
  rhs <- if (length(vars) == 0) "1" else paste(vars, collapse = " + ")
  list(
    label   = if (length(vars) == 0) "(intercept only)" else paste(vars, collapse = " + "),
    n_vars  = length(vars),
    formula = as.formula(paste(OUTCOME, "~", rhs))
  )
})
```

# 5. Modelling table

Complete cases only, so every model is fit and scored on exactly the same rows
and the comparison between them is fair.

**Caveat: requiring all three predictors
keeps only regions overlapping a peak in *all three* ChIP datasets. That is a
biased subset of your regions — the R² values describe those regions, not the
full ATAC set.

```{r model-table}
d <- data |>
  dplyr::filter(classification %in% KEEP_CLASSES) |>
  dplyr::select(chr, all_of(OUTCOME), all_of(PREDICTORS)) |>
  tidyr::drop_na()

cat(sprintf("Rows with outcome + all predictors: %d\n", nrow(d)))
cat(sprintf("Chromosomes represented: %d\n", dplyr::n_distinct(d$chr)))
```

# 6. Chromosome-blocked cross-validation

Random row-wise folds leak: neighbouring or overlapping ATAC regions share
signal, so a region in the training set can carry information about a nearly
identical region in the test set, inflating R². Holding out **whole
chromosomes** removes that leakage.

Chromosomes are packed greedily into `K_FOLDS` folds, largest first, so the
folds end up roughly balanced in row count despite very unequal chromosome
sizes.

```{r folds}
set.seed(SEED)

chr_sizes <- d |>
  dplyr::count(chr, name = "n") |>
  dplyr::arrange(dplyr::desc(n))

k_folds <- min(K_FOLDS, nrow(chr_sizes))   # can't have more folds than chromosomes

fold_load  <- rep(0, k_folds)
chr_to_fold <- integer(nrow(chr_sizes))
for (i in seq_len(nrow(chr_sizes))) {
  f <- which.min(fold_load)                # assign to the currently emptiest fold
  chr_to_fold[i] <- f
  fold_load[f] <- fold_load[f] + chr_sizes$n[i]
}

d$fold <- chr_to_fold[match(d$chr, chr_sizes$chr)]

# Fold composition — check no fold is pathologically small
d |>
  dplyr::group_by(fold) |>
  dplyr::summarise(n_regions = dplyr::n(),
                   chrs = paste(sort(unique(chr)), collapse = ",")) |>
  print(n = Inf)
```

```{r cross-validate}
#' Two complementary R² summaries per model:
#'   - per_fold : R² computed within each held-out chromosome block. Shows
#'                spread / stability across the genome. Uses the TEST-fold mean
#'                as the null, so it can go negative — that is meaningful, not a
#'                bug (it means the model does worse than the test fold's own mean).
#'   - pooled   : all out-of-fold predictions concatenated, then one R² against
#'                the global mean. This is the headline number: far less noisy
#'                than averaging fold-level R², and not distorted by small folds.
cv_out <- purrr::map(model_specs, function(spec) {
  y        <- d[[OUTCOME]]
  pred_oof <- rep(NA_real_, nrow(d))
  per_fold <- rep(NA_real_, k_folds)

  for (f in seq_len(k_folds)) {
    idx_te <- which(d$fold == f)
    idx_tr <- which(d$fold != f)

    m <- lm(spec$formula, data = d[idx_tr, ])
    p <- predict(m, newdata = d[idx_te, ])

    pred_oof[idx_te] <- p
    per_fold[f] <- 1 - sum((y[idx_te] - p)^2) /
                       sum((y[idx_te] - mean(y[idx_te]))^2)
  }

  pooled <- 1 - sum((y - pred_oof)^2) / sum((y - mean(y))^2)

  list(
    per_fold = tibble(model = spec$label, n_vars = spec$n_vars,
                      fold = seq_len(k_folds), r2 = per_fold),
    summary  = tibble(model = spec$label, n_vars = spec$n_vars,
                      r2_pooled = pooled, r2_median = median(per_fold))
  )
})

cv_r2      <- purrr::map_dfr(cv_out, "per_fold")
cv_summary <- purrr::map_dfr(cv_out, "summary") |> dplyr::arrange(dplyr::desc(r2_pooled))

print(cv_summary)
```

# 7. Figure

Boxes are the per-chromosome-block R² distribution; the number at the right of
each row is the pooled out-of-fold R² (the headline value). Fill encodes the
number of predictors.

```{r plot, fig.height=4.5, fig.width=8}
# Order rows by pooled R² so the labels read monotonically top-to-bottom
model_levels <- cv_summary |> dplyr::arrange(r2_pooled) |> dplyr::pull(model)

cv_r2      <- cv_r2      |> dplyr::mutate(model = factor(model, levels = model_levels))
cv_summary <- cv_summary |> dplyr::mutate(model = factor(model, levels = model_levels))

# Where to place the pooled-R² labels: just right of the widest box/point
x_max  <- max(cv_r2$r2, na.rm = TRUE)
x_min  <- min(cv_r2$r2, na.rm = TRUE)
x_pad  <- 0.12 * (x_max - x_min)

p <- ggplot(cv_r2, aes(x = model, y = r2)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_boxplot( alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 1.5) +
  geom_text(
    data = cv_summary,
    aes(x = model, y = x_max + x_pad * 0.35,
        label = sprintf("R² = %.3f", r2_pooled)),
    hjust = 0, size = 3.2, fontface = "bold", inherit.aes = FALSE
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(x_min - x_pad * 0.1, x_max + x_pad * 1.6)) +
  scale_fill_brewer(palette = "Blues", name = "# predictors") +
  theme_classic() +
  theme(plot.margin = margin(5.5, 12, 5.5, 5.5)) +
  labs(
    x = NULL,
    y = "Cross-validated R² (per held-out chromosome block)",
    title = paste("Predicting", OUTCOME),
    subtitle = sprintf(
      "%d-fold chromosome-blocked CV | %d regions | bold value = pooled out-of-fold R²",
      k_folds, nrow(d)
    )
  )

p
ggsave(FIG_OUT, plot = p, height = 4.5, width = 8)
cat("Saved:", FIG_OUT, "\n")
```

# 8. Session info

```{r session-info}
sessionInfo()
```