---
title: "ATAC vs H3K27ac log2FC across dTAG time points — linear mixed-effects model"
subtitle: "Supplementary Table S2 source code"
output: html_document
date: "2026-08-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 0. Purpose

Test whether the relationship between the SNAI1/2-depletion-induced change in
chromatin accessibility (ATAC log2FC, dTAG/DMSO) and the change in H3K27ac
(ChIP log2FC, dTAG/DMSO) differs across depletion time points (3h, 1d, 3d), at
distal SNAI1/2-bound regions that gain signal upon depletion.

Each genomic region contributes up to three observations (one per time point),
so observations are not independent. We therefore use a linear mixed-effects
model with a random intercept for region.


```{r libs}
library(lme4)        # lmer()
library(lmerTest)    # adds Satterthwaite df / p-values to the fixed-effect table
library(emmeans)     # emtrends(): estimated marginal slopes + contrasts
library(broom.mixed) # tidy() for lmer objects
library(ggplot2)
library(tidyr)
library(dplyr)       
library(conflicted)
conflicts_prefer(
  base::unname,
  lmerTest::lmer,
  dplyr::filter,
  dplyr::select,
  dplyr::mutate,
  dplyr::summarise,
  dplyr::rename,
  dplyr::count,
  .quiet = TRUE
)


# Several Bioconductor / plyr-style packages export functions named count(),
# mutate(), summarise(), rename() etc. If any of those are attached (e.g. from
# the session that produced data_v4.rds), they can mask the dplyr versions and
# produce cryptic errors such as
#   "Argument 'x' is not a vector: list".
# Every dplyr verb below is therefore namespace-qualified. Run this to see what
# is masking what in your session:
find("count"); find("mutate"); find("summarise")

# Output directory for the paper tables / figures
tab_dir <- "..."
fig_dir <- "..."

set.seed(1)  
```

# 1. Data

```{r load}
data <- readRDS(".../data_v4.rds")
```

## 1.1 Region selection

Inclusion criteria (all must hold):

* the region is called UP in **at least one** assay at **at least one** time
  point (H3K27ac or ATAC, 3h / 1d / 3d) — i.e. the region responds to SNAI1/2
  depletion in at least one readout;
* the region overlaps a SNAI1/2 peak (`SNAI_peaks_all == "SNAI_bound"`);
* the region is **distal** (`classification %in% c("enhancer","other_distal")`),
  i.e. promoters are excluded.

Note this is a *union* selection across assays and time points, so the same
region set is used at all three time points — which is what makes the
repeated-measures / random-intercept structure appropriate.

```{r select}
sel_alltp <- data %>%
  dplyr::filter(
    H3K27ac_UP_3d == 1 | ATAC_UP_3d == 1 |
    H3K27ac_UP_1d == 1 | ATAC_UP_1d == 1 |
    H3K27ac_UP_3h == 1 | ATAC_UP_3h == 1,
    SNAI_peaks_all == "SNAI_bound",
    classification %in% c("enhancer", "other_distal")
  )

n_regions_selected <- dplyr::n_distinct(sel_alltp$region)
n_regions_selected
```

## 1.2 Reshape to long format

One row per region x time point, with the matched ATAC and H3K27ac log2FC.

```{r reshape}
# H3K27ac log2FCs -> long
ac_long <- sel_alltp %>%
  dplyr::select(region, classification, SNAI_peaks_all,
                H3K27ac_3h_log2FC, H3K27ac_1d_log2FC, H3K27ac_3d_log2FC) %>%
  tidyr::pivot_longer(cols = tidyr::starts_with("H3K27ac_"),
                      names_to = "timepoint", values_to = "ac_l2fc") %>%
  dplyr::mutate(timepoint = gsub("H3K27ac_|_log2FC", "", timepoint))

# ATAC log2FCs -> long
atac_long <- sel_alltp %>%
  dplyr::select(region,
                ATACdata_3h_log2FC, ATACdata_1d_log2FC, ATACdata_3d_log2FC) %>%
  tidyr::pivot_longer(cols = tidyr::starts_with("ATACdata_"),
                      names_to = "timepoint", values_to = "atac_l2fc") %>%
  dplyr::mutate(timepoint = gsub("ATACdata_|_log2FC", "", timepoint))

# Match ATAC and H3K27ac on the SAME region and the SAME time point
df_long <- dplyr::inner_join(ac_long, atac_long, by = c("region", "timepoint"))

# 3h is the reference level (earliest time point)
df_long$timepoint <- factor(df_long$timepoint, levels = c("3h", "1d", "3d"))
df_long$region    <- factor(df_long$region)

# Drop incomplete rows explicitly, so that reported n matches what lmer used
# (lmer would silently drop them via na.omit otherwise).
df_long <- df_long %>% dplyr::filter(!is.na(ac_l2fc), !is.na(atac_l2fc))
df_long$region <- droplevels(df_long$region)
```

## 1.3 Sample sizes — needed for the table and the Methods

```{r n}
n_obs      <- nrow(df_long)
n_regions  <- dplyr::n_distinct(df_long$region)

# Regions contributing an observation at each time point
n_per_tp <- df_long %>%
  dplyr::group_by(timepoint) %>%
  dplyr::summarise(n_regions = dplyr::n(), .groups = "drop")

# How many regions have 1, 2 or 3 time points of data (should be mostly 3)
obs_per_reg <- df_long %>%
  dplyr::group_by(region) %>%
  dplyr::summarise(k = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(k) %>%
  dplyr::summarise(n_regions = dplyr::n(), .groups = "drop")

n_obs; n_regions; n_per_tp; obs_per_reg
```

# 2. Models

Two nested models, both fit with a random intercept for `region`:

* **reduced** — one common ATAC->H3K27ac slope shared by all time points; each
  time point gets its own intercept (main effect of `timepoint` only).
* **full** — the slope is allowed to differ by time point
  (`atac_l2fc * timepoint` = both main effects + their interaction).

A random *slope* per region (`(1 + atac_l2fc | region)`) is not estimable here:
with at most three observations per region there is no within-region
information to identify a region-specific slope. This is worth stating in the
Methods pre-emptively.

```{r fit}
full    <- lmerTest::lmer(ac_l2fc ~ atac_l2fc * timepoint + (1 | region), data = df_long)
reduced <- lmerTest::lmer(ac_l2fc ~ atac_l2fc + timepoint + (1 | region), data = df_long)

summary(full)
```

## 2.1 Likelihood-ratio test for the interaction

`anova()` on two `merMod` objects automatically refits both models by maximum
likelihood (REML likelihoods of models with different fixed effects are not
comparable) and returns the LRT. A significant result means the ATAC->H3K27ac
slope differs across time points.

```{r lrt}
lrt <- anova(reduced, full)
lrt
```

## 2.2 Estimated marginal slopes

`emtrends()` gives the slope of `ac_l2fc` on `atac_l2fc` within each time point,
i.e. exactly the three lines drawn in Figure 3E.

Degrees of freedom: with n in the tens of thousands, Kenward-Roger /
Satterthwaite df are computationally expensive and numerically identical to the
asymptotic limit, so we request asymptotic (normal-theory, z) inference
**explicitly** rather than letting `emmeans` fall back to it silently when
`lmerTest.limit` is exceeded. This is why the table reports z rather than t,
and it should be stated in the Methods.

```{r emtrends}
emt <- emtrends(full, ~ timepoint, var = "atac_l2fc", lmer.df = "asymptotic")

# (a) slope estimates with SE, 95% CI, and test against 0
slopes_vs0 <- summary(emt, infer = c(TRUE, TRUE), null = 0)

# (b) test against 1 -- H0: H3K27ac changes 1:1 with accessibility.
#     This is the test that actually backs the claim "H3K27ac increased more
#     slowly than accessibility", which is a statement about slope < 1, not
#     slope > 0. Unadjusted (3 pre-specified tests); note this in the legend.
slopes_vs1 <- summary(emt, infer = c(FALSE, TRUE), null = 1, adjust = "none")

# (c) pairwise comparisons between time points, Tukey-adjusted
pw <- summary(pairs(emt, adjust = "tukey"), infer = c(TRUE, TRUE))

slopes_vs0; slopes_vs1; pw
```

## 2.3 Descriptive correlation per time point



```{r cors}
cor_tab <- do.call(rbind, lapply(levels(df_long$timepoint), function(tp) {
  d <- df_long[df_long$timepoint == tp, ]
  p <- cor.test(d$atac_l2fc, d$ac_l2fc, method = "pearson")
  s <- suppressWarnings(
    cor.test(d$atac_l2fc, d$ac_l2fc, method = "spearman", exact = FALSE))
  data.frame(
    timepoint    = tp,
    n            = nrow(d),
    pearson_r    = unname(p$estimate),
    pearson_lcl  = p$conf.int[1],
    pearson_ucl  = p$conf.int[2],
    pearson_p    = p$p.value,
    spearman_rho = unname(s$estimate),
    spearman_p   = s$p.value,
    stringsAsFactors = FALSE
  )
}))
rownames(cor_tab) <- NULL

cor_tab
```

# 3. Build the supplementary table

## 3.1 Formatting helpers

```{r helpers}
# Report p-values honestly: below double.eps R cannot resolve them further, so
# report the machine limit rather than "0" or a spuriously precise number.
fmt_p <- function(p, digits = 3) {
  ifelse(is.na(p), NA_character_,
    ifelse(p < 2.2e-16, "< 2.2e-16", formatC(p, format = "e", digits = digits - 1)))
}
r3 <- function(x, d = 3) formatC(round(x, d), format = "f", digits = d)
```

## 3.2 Panels

```{r panels}
# ---- Panel A: model specification and data summary -------------------------
panel_A <- tibble::tibble(
  item = c(
    "Response variable",
    "Predictor",
    "Grouping factor (random effect)",
    "Full model formula",
    "Reduced model formula",
    "Estimation",
    "Degrees of freedom",
    "Multiple-testing correction (pairwise slope contrasts)",
    "Region set",
    "Regions (n)",
    "Observations (n region x time point)",
    "Observations per time point",
    "Software"
  ),
  value = c(
    "H3K27ac log2FC (dTAG/DMSO)",
    "ATAC log2FC (dTAG/DMSO)",
    "genomic region (random intercept)",
    "ac_l2fc ~ atac_l2fc * timepoint + (1|region)",
    "ac_l2fc ~ atac_l2fc + timepoint + (1|region)",
    "REML for parameter estimates; ML refit for the likelihood-ratio test",
    "asymptotic (normal-theory, z statistics)",
    "Tukey",
    "distal (enhancer / other_distal), SNAI1/2-bound, UP in H3K27ac or ATAC at >= 1 time point",
    format(n_regions, big.mark = ","),
    format(n_obs, big.mark = ","),
    paste(paste0(n_per_tp$timepoint, ": ", format(n_per_tp$n_regions, big.mark = ",")),
          collapse = "; "),
    paste0("R ", getRversion(),
           "; lme4 ", packageVersion("lme4"),
           "; lmerTest ", packageVersion("lmerTest"),
           "; emmeans ", packageVersion("emmeans"))
  )
)

# ---- Panel B: fixed effects of the full model ------------------------------
# Taken straight from summary() rather than via broom, so it does not depend on
# broom.mixed exposing the lmerTest df/p columns. Wald CIs (fast, and identical
# to profile CIs at this n).
fe <- as.data.frame(summary(full)$coefficients)
fe_ci <- confint(full, parm = rownames(fe), method = "Wald")

panel_B <- tibble::tibble(
  term     = rownames(fe),
  estimate = r3(fe[["Estimate"]]),
  SE       = r3(fe[["Std. Error"]]),
  `95% CI` = paste0("[", r3(fe_ci[, 1]), ", ", r3(fe_ci[, 2]), "]"),
  df       = if ("df" %in% names(fe)) formatC(fe[["df"]], format = "f", digits = 0) else "asymptotic",
  t        = r3(fe[["t value"]], 2),
  p        = if ("Pr(>|t|)" %in% names(fe)) fmt_p(fe[["Pr(>|t|)"]]) else NA_character_
)

# ---- Panel C: random effects / variance components -------------------------
vc <- as.data.frame(VarCorr(full))
v_region <- vc$vcov[vc$grp == "region"]
v_resid  <- vc$vcov[vc$grp == "Residual"]
panel_C <- tibble::tibble(
  component = c("region (intercept)", "Residual", "ICC (region)"),
  variance  = c(r3(v_region, 4), r3(v_resid, 4), ""),
  `SD`      = c(r3(sqrt(v_region), 4), r3(sqrt(v_resid), 4), ""),
  value     = c("", "", r3(v_region / (v_region + v_resid), 3))
)

# ---- Panel D: likelihood-ratio test, full vs reduced -----------------------
lrt_df <- as.data.frame(lrt)

# Column names of anova.merMod changed across lme4 versions
# (npar/Df for parameter count, Df/"Chi Df" for the test df), so fetch by name
# with fallbacks instead of positionally.
gc_ <- function(d, ...) {
  for (nm in c(...)) if (nm %in% names(d)) return(d[[nm]])
  rep(NA, nrow(d))
}

panel_D <- tibble::tibble(
  model    = c("reduced (common slope)", "full (slope x timepoint)"),
  npar     = as.character(gc_(lrt_df, "npar", "Df")),
  AIC      = r3(gc_(lrt_df, "AIC"), 1),
  BIC      = r3(gc_(lrt_df, "BIC"), 1),
  logLik   = r3(as.numeric(gc_(lrt_df, "logLik")), 1),
  deviance = r3(gc_(lrt_df, "deviance"), 1),
  chisq    = c("", r3(gc_(lrt_df, "Chisq")[2], 1)),
  df       = c("", as.character(gc_(lrt_df, "Chi Df", "Df")[2])),
  p        = c("", fmt_p(gc_(lrt_df, "Pr(>Chisq)")[2]))
)

# ---- Panel E: estimated marginal slopes, tested against 0 and against 1 ----
sl0 <- as.data.frame(slopes_vs0)
sl1 <- as.data.frame(slopes_vs1)
panel_E <- tibble::tibble(
  timepoint      = as.character(sl0$timepoint),
  n              = n_per_tp$n_regions[match(as.character(sl0$timepoint),
                                            as.character(n_per_tp$timepoint))],
  slope          = r3(sl0$atac_l2fc.trend),
  SE             = r3(sl0$SE),
  `95% CI`       = paste0("[", r3(sl0$asymp.LCL), ", ", r3(sl0$asymp.UCL), "]"),
  `z (H0: slope=0)` = r3(sl0$z.ratio, 2),
  `p (H0: slope=0)` = fmt_p(sl0$p.value),
  `z (H0: slope=1)` = r3(sl1$z.ratio, 2),
  `p (H0: slope=1)` = fmt_p(sl1$p.value)
)

# ---- Panel F: pairwise slope contrasts ------------------------------------
pw_df <- as.data.frame(pw)
panel_F <- tibble::tibble(
  contrast           = as.character(pw_df$contrast),
  `slope difference` = r3(pw_df$estimate),
  SE                 = r3(pw_df$SE),
  `95% CI`           = paste0("[", r3(pw_df$asymp.LCL), ", ", r3(pw_df$asymp.UCL), "]"),
  z                  = r3(pw_df$z.ratio, 2),
  `p (Tukey)`        = fmt_p(pw_df$p.value)
)

# ---- Panel G: descriptive correlations ------------------------------------
panel_G <- cor_tab %>%
  dplyr::transmute(
    timepoint = as.character(timepoint),
    n,
    `Pearson r` = r3(pearson_r),
    `Pearson 95% CI` = paste0("[", r3(pearson_lcl), ", ", r3(pearson_ucl), "]"),
    `Pearson p` = fmt_p(pearson_p),
    `Spearman rho` = r3(spearman_rho),
    `Spearman p` = fmt_p(spearman_p)
  )
```

## 3.3 Stack the panels into one CSV

```{r stack}
panels <- list(
  "A. Model specification and data summary"                                = panel_A,
  "B. Fixed effects, full model (Satterthwaite df)"                        = panel_B,
  "C. Variance components, full model"                                     = panel_C,
  "D. Likelihood-ratio test: full vs reduced (ML refit)"                   = panel_D,
  "E. Estimated marginal slopes of H3K27ac log2FC on ATAC log2FC by time point" = panel_E,
  "F. Pairwise comparisons of slopes between time points"                  = panel_F,
  "G. Descriptive correlation of ATAC and H3K27ac log2FC by time point"    = panel_G
)

ncol_total <- max(vapply(panels, ncol, integer(1)))

pad <- function(m, k) {
  if (ncol(m) < k) m <- cbind(m, matrix("", nrow(m), k - ncol(m)))
  m
}

blocks <- lapply(names(panels), function(nm) {
  d <- as.data.frame(lapply(panels[[nm]], as.character), stringsAsFactors = FALSE)
  colnames(d) <- colnames(panels[[nm]])
  rbind(
    pad(matrix(nm, 1, 1), ncol_total),                 # panel title
    pad(matrix(colnames(d), 1), ncol_total),           # column headers
    pad(as.matrix(d), ncol_total),                     # body
    matrix("", 1, ncol_total)                          # blank spacer
  )
})

header <- rbind(
  pad(matrix("Supplementary Table S2. Linear mixed-effects model of H3K27ac log2FC on ATAC log2FC across SNAI1/2 depletion time points at distal SNAI1/2-bound regions.", 1, 1), ncol_total),
  matrix("", 1, ncol_total)
)

out <- rbind(header, do.call(rbind, blocks))

footer <- rbind(
  pad(matrix("Notes: log2FC = log2 fold-change, dTAG / DMSO. Slopes are estimated marginal trends of H3K27ac log2FC on ATAC log2FC within each time point (emmeans::emtrends). Inference on slopes and contrasts is asymptotic (z); the fixed-effect table in panel B uses Satterthwaite degrees of freedom (lmerTest). Pairwise contrast p-values are Tukey-adjusted; tests of slope against 1 are unadjusted. Panel G correlations are descriptive and computed independently within each time point.", 1, 1), ncol_total)
)
out <- rbind(out, matrix("", 1, ncol_total), footer)

dir.create(tab_dir, recursive = TRUE, showWarnings = FALSE)
write.table(out,
            file = file.path(tab_dir, "SupplementaryTableS2_mixed_effects_ATAC_H3K27ac.csv"),
            sep = ",", row.names = FALSE, col.names = FALSE, na = "", qmethod = "double")
```

## 3.4 Optional: same panels as separate sheets in one .xlsx

Nicer for submission portals that dislike stacked CSVs. Skipped silently if
`openxlsx` is not installed.

```{r xlsx}
if (requireNamespace("openxlsx", quietly = TRUE)) {
  wb <- openxlsx::createWorkbook()
  sheet_names <- c("A_model_info", "B_fixed_effects", "C_variance",
                   "D_LRT", "E_slopes", "F_contrasts", "G_correlations")
  for (i in seq_along(panels)) {
    openxlsx::addWorksheet(wb, sheet_names[i])
    openxlsx::writeData(wb, sheet_names[i], names(panels)[i], startRow = 1)
    openxlsx::writeData(wb, sheet_names[i], panels[[i]], startRow = 3)
    openxlsx::setColWidths(wb, sheet_names[i], cols = 1:ncol(panels[[i]]), widths = "auto")
  }
  openxlsx::saveWorkbook(
    wb, file.path(tab_dir, "SupplementaryTableS2_mixed_effects_ATAC_H3K27ac.xlsx"),
    overwrite = TRUE)
} else {
  message("openxlsx not installed - CSV only.")
}
```

## 3.5 Also keep the two minimal tables used for figure annotation

```{r minimal}
slope_table <- panel_E %>% dplyr::select(timepoint, slope, `95% CI`)
contrast_table <- panel_F %>% dplyr::select(contrast, `slope difference`, z, `p (Tukey)`)

write.csv(slope_table,    file.path(tab_dir, "slope_table.csv"),    row.names = FALSE)
write.csv(contrast_table, file.path(tab_dir, "contrast_table.csv"), row.names = FALSE)

slope_table; contrast_table
```

# 4. Figure 3E

The red line per facet is the **population-level (fixed-effect) fit**, i.e.
`re.form = NA` — region random intercepts are excluded so the line shown is the
same slope reported in panel E.

```{r figure, fig.width=8, fig.height=3}
# Prediction grid spanning the observed ATAC range within each time point,
# so lines are not extrapolated beyond the data.
pred_grid <- do.call(rbind, lapply(levels(df_long$timepoint), function(tp) {
  x <- df_long$atac_l2fc[df_long$timepoint == tp]
  data.frame(timepoint = factor(tp, levels = levels(df_long$timepoint)),
             atac_l2fc = seq(min(x, na.rm = TRUE), max(x, na.rm = TRUE),
                             length.out = 100))
}))

pred_grid$ac_l2fc_pred <- predict(full, newdata = pred_grid, re.form = NA)

# Annotation: slope, CI and n per facet
annot <- panel_E
annot$timepoint <- factor(annot$timepoint, levels = levels(df_long$timepoint))
annot$label <- paste0("slope = ", annot$slope, "\n", annot$`95% CI`,
                      "\nn = ", format(annot$n, big.mark = ","))

p_facet <- ggplot(df_long, aes(x = atac_l2fc, y = ac_l2fc)) +
  geom_point(size = 1, alpha = 0.4, shape = 21, color = "grey40") +
  geom_line(data = pred_grid, aes(x = atac_l2fc, y = ac_l2fc_pred),
            color = "firebrick", linewidth = 1) +
  facet_wrap(~timepoint) +
  labs(x = "ATAC log2FC (dTag/DMSO)", y = "H3K27ac log2FC (dTag/DMSO)",
       title = "SNAI-bound distal regions - upregulated") +
  theme_bw() +
  theme(strip.background = element_blank())

p_out <- p_facet +
  geom_text(data = annot, aes(x = -0.3, y = 2, label = label),
            inherit.aes = FALSE, size = 3, hjust = 0)

p_out

dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
ggsave(file.path(fig_dir, "Scatter_ATAC_k27_v4_mixedeffectmodel_comb.pdf"),
       plot = p_out, height = 3, width = 8)
```

# 6. Session info

```{r session}
sessionInfo()
```