---
title: "Fiber-seq: nucleosome metaplot and merged-replicate clustering heatmap at SNAI2 motifs"
author: ""
date: "2026-05-18"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

<!--
================================================================================
 SCOPE OF THIS DOCUMENT
================================================================================
 Reduced to the two figures that go into the paper:

   FIGURE 1  metaplot_SNAI2alldistmotifs_merger_reps_v7.pdf
             Nucleosome-midpoint rate around all distal SNAI2 motifs,
             DMSO vs dTag, replicates exp2 + exp3 pooled.

   FIGURE 2  Heatmap_V5_mergedReps_plus_FC_SNAI2_K27acUPmotifs_
             proportional_v2_presel80bp.pdf
             Per-read nucleosome footprint heatmap for the K27ac-UP / log2FC-Q3
             motif subset: merged DMSO | merged dTag | per-cluster
             log2(dTag/DMSO) bars, faceted by k-means cluster.


 Input : fibertools "centered" tables. One row = one sequencing read (fiber)
         x one motif instance, with nucleosome / MSP / m6A calls already
         expressed in coordinates relative to the motif centre (position 0).
 Conditions: DMSO (control) vs dTag (3 h degron depletion of SNAI2).
 Replicates: exp2 (r1+r2 merged upstream) and exp3.
================================================================================
-->



# Setup

## Libraries

```{r setup-libraries, message=FALSE, warning=FALSE}
# Core data wrangling
library(data.table)   # fread(), fast in-place := operations
library(dplyr)
library(tidyr)        # separate_rows(), crossing(), pivot_*(), extract()
library(purrr)        # imap(), map(), imap_dfr()

# Plotting
library(ggplot2)
library(patchwork)    # wrap_plots() to compose the multi-panel figure
library(ggrastr)      # rasterise() — keeps PDFs small despite ~10^5 geom_rects
library(zoo)          # rollmean() for metaplot smoothing
library(scales)       # pretty_breaks()

# Matrix algebra for the occupancy matrix
library(Matrix)       # sparseMatrix() before densifying for kmeans()

# NOTE: the original script also loaded GenomicRanges, rtracklayer, tidyverse
# and stats. None of their functions are actually called below, so they are
# omitted here to keep the dependency list honest. Re-add if you extend the
# analysis (e.g. rtracklayer::import() for BED/bigWig overlays).
```

## Configuration — ALL FILE PATHS LIVE HERE

```{r config-paths}
# =============================================================================
# PATH INVENTORY  ††  THE ONLY PLACE IN THIS DOCUMENT WITH ABSOLUTE PATHS  ††
# -----------------------------------------------------------------------------
# Original literals, for your review before public release:
#
#  INPUT ROOTS
#    exp2 : .../fibertools
#    exp3 : .../fibertools
#
#  OUTPUT ROOT  
#    .../fiber-seq/
#
#  8 INPUT FILES = {exp2, exp3} x {DMSO, 3hdTag} x {v7 all-distal motifs,
#                  v6 K27acUP3h_K27Log2FCQ3 subset}
#
#  2 OUTPUT FILES (both written to FIG_DIR)
#    metaplot_SNAI2alldistmotifs_merger_reps_v7.pdf
#    Heatmap_V5_mergedReps_plus_FC_<figure_prefix>.pdf
#
# 
# =============================================================================

## ---- Roots ------------------------------------------------------------------
DATA_ROOT_EXP2 <- ".../fibertools"
DATA_ROOT_EXP3 <- ".../fibertools"

FIG_DIR <- file.path(
  "..."
)

## ---- Per-sample subdirectories ----------------------------------------------
# exp2 was sequenced as two runs (r1, r2) that were merged upstream of this
# script; exp3 is a single run. The directory name is repeated as the file-name
# stem in the fibertools output convention, hence the two variables per sample.
SAMPLE_DIRS <- list(
  DMSO_exp2 = list(root = DATA_ROOT_EXP2, stem = "fiber_exp2_DMSO_r1ANDr2"),
  dTag_exp2 = list(root = DATA_ROOT_EXP2, stem = "fiber_exp2_3hdTag_r1ANDr2"),
  DMSO_exp3 = list(root = DATA_ROOT_EXP3, stem = "fiber_exp3_DMSO"),
  dTag_exp3 = list(root = DATA_ROOT_EXP3, stem = "fiber_exp3_3hdTag")
)

# Builds <root>/<stem>/<stem>_<suffix>
fibertools_file <- function(key, suffix) {
  s <- SAMPLE_DIRS[[key]]
  file.path(s$root, s$stem, paste0(s$stem, "_", suffix))
}

## ---- Input file sets ---------------------------------------------------------
# Set A: v7 tables, ALL distal SNAI2 motifs. Used for FIGURE 1 (metaplot).
#        24 columns — includes the FIRE calls.
SUFFIX_V7_ALL_DISTAL <- "SNAI2motif_dist_v7_centered.txt"

# Set B: v6 tables, distal SNAI2 motifs restricted to those at loci with H3K27ac
#        UP at 3 h and in the top quartile (Q3) of K27ac log2FC.
#        Used for FIGURE 2 (clustering heatmap). 22 columns — no FIRE calls.
SUFFIX_V6_K27AC_Q3 <- "SNAI2motif_dist_v6_K27acUP3h_K27Log2FCQ3_centered.txt"

PATHS <- list(
  v7_all_distal = lapply(setNames(names(SAMPLE_DIRS), names(SAMPLE_DIRS)),
                         fibertools_file, suffix = SUFFIX_V7_ALL_DISTAL),
  v6_k27ac_q3   = lapply(setNames(names(SAMPLE_DIRS), names(SAMPLE_DIRS)),
                         fibertools_file, suffix = SUFFIX_V6_K27AC_Q3),
  fig_dir       = FIG_DIR
)

## ---- Sanity check ------------------------------------------------------------
# Fails loudly at the top of the document rather than halfway through a chunk.
check_paths <- function(paths, must_exist = TRUE) {
  flat    <- unlist(paths[c("v7_all_distal", "v6_k27ac_q3")], use.names = TRUE)
  missing <- flat[!file.exists(flat)]
  if (length(missing) && must_exist) {
    stop("Input file(s) not found:\n  ", paste(missing, collapse = "\n  "),
         "\n\nEdit the roots in the `config-paths` chunk.", call. = FALSE)
  }
  if (!dir.exists(paths$fig_dir)) {
    dir.create(paths$fig_dir, recursive = TRUE, showWarnings = FALSE)
  }
  invisible(TRUE)
}

# Set must_exist = FALSE if you want the document to knit on a machine that
# does not have the raw data mounted.
check_paths(PATHS, must_exist = TRUE)

str(PATHS, max.level = 2)
```

## Column schemas

```{r config-schemas}
# fibertools "centered" tables ship WITHOUT a header, so column names are
# assigned positionally. The two versions differ: v7 carries three extra FIRE
# columns where v6 has a single unlabelled column. Getting these out of order
# silently corrupts everything downstream, so they are defined once here.

# --- v7 (24 columns) ---------------------------------------------------------
FIBER_COLNAMES_V7 <- c(
  "chrom",                   # contig of the motif instance
  "centering_position",      # genomic coordinate of the motif centre (= 0 below)
  "strand",                  # motif strand; coordinates already strand-corrected
  "subset_sequence",
  "reference_start", "reference_end",   # read alignment span, genomic
  "query_name",              # read ID (NOT unique across motif instances)
  "RG", "HP",                # read group; haplotype tag
  "centered_query_start", "centered_query_end", "query_length",
  "centered_m6a_positions", "m6a_qual",       # comma-separated
  "centered_5mC_positions", "5mC_qual",       # comma-separated
  "centered_nuc_starts", "centered_nuc_ends", # nucleosome footprints, CSV
  "centered_msp_starts", "centered_msp_ends", # methyl-sensitive patches, CSV
  "centered_fire_starts", "centered_fire_ends", "fire_qual",
  "query_sequence"
)

# --- v6 (22 columns) ---------------------------------------------------------
# Identical up to centered_msp_ends; then one column of unknown provenance
# (kept as "unknown" — it is never read) and the sequence.
FIBER_COLNAMES_V6 <- c(
  "chrom", "centering_position", "strand", "subset_sequence",
  "reference_start", "reference_end", "query_name", "RG", "HP",
  "centered_query_start", "centered_query_end", "query_length",
  "centered_m6a_positions", "m6a_qual", "centered_5mC_positions", "5mC_qual",
  "centered_nuc_starts", "centered_nuc_ends",
  "centered_msp_starts", "centered_msp_ends",
  "unknown", "query_sequence"
)
```

---

# FIGURE 1 — Nucleosome midpoint metaplot

All distal SNAI2 motifs. Replicates are pooled *before* averaging, so each
fiber contributes equally and the replicate with more fibers contributes
proportionally more. The `rep` column is carried through so you can switch to a
mean-of-means (average within replicate first, then across replicates) if you
prefer that weighting.

```{r metaplot, fig.width=7.29, fig.height=4.51}
# =============================================================================
# Nucleosome midpoint metaplot around SNAI2 motifs
# Merges two biological replicates (exp2, exp3) per condition (DMSO, dTag)
# before computing per-fiber nucleosome midpoint density.
# =============================================================================

# ---- 0. Analysis parameters -------------------------------------------------
WINDOW    <- 1000   # +/- bp around the motif that is analysed
BIN_WIDTH <- 10     # bp per bin  -> 200 bins across the window
SMOOTH_K  <- 5      # rolling-mean width in BINS (= 50 bp) applied after averaging

# ---- 1. Load ----------------------------------------------------------------
# One row = one fiber (read) x one motif instance.
read_fiber_v7 <- function(path) {
  dt <- fread(path)
  stopifnot(ncol(dt) == length(FIBER_COLNAMES_V7))  # guard against schema drift
  setnames(dt, FIBER_COLNAMES_V7)
  dt
}

files <- list(
  list(sample = "DMSO", rep = "exp2", path = PATHS$v7_all_distal$DMSO_exp2),
  list(sample = "dTag", rep = "exp2", path = PATHS$v7_all_distal$dTag_exp2),
  list(sample = "DMSO", rep = "exp3", path = PATHS$v7_all_distal$DMSO_exp3),
  list(sample = "dTag", rep = "exp3", path = PATHS$v7_all_distal$dTag_exp3)
)

combined_df <- rbindlist(lapply(files, function(f) {
  read_fiber_v7(f$path)[, `:=`(sample = f$sample, rep = f$rep)]
}))

# Unique fiber identifier. Row number is already unique across the merged
# table, but making it an explicit column keeps downstream grouping
# unambiguous (query_name alone is NOT unique — a read can span several motifs).
combined_df[, unit_id := .I]

# ---- 2. Explode nucleosome calls -> one row per midpoint --------------------
# centered_nuc_starts / _ends are comma-separated strings of equal length per
# fiber. separate_rows() unnests them in parallel, so element i of starts pairs
# with element i of ends.
midpoints_df <- combined_df %>%
  filter(!is.na(centered_nuc_starts), !is.na(centered_nuc_ends)) %>%
  separate_rows(centered_nuc_starts, centered_nuc_ends, sep = ",") %>%
  filter(centered_nuc_starts != "", centered_nuc_ends != "") %>%   # trailing commas
  mutate(
    centered_nuc_starts = as.numeric(centered_nuc_starts),
    centered_nuc_ends   = as.numeric(centered_nuc_ends),
    nuc_midpoint        = (centered_nuc_starts + centered_nuc_ends) / 2
  )

# ---- 3. Per-fiber counts per bin, zero-filled -------------------------------
bins <- seq(-WINDOW, WINDOW - BIN_WIDTH, by = BIN_WIDTH)

per_unit_df <- midpoints_df %>%
  filter(nuc_midpoint >= -WINDOW, nuc_midpoint < WINDOW) %>%
  mutate(bin = floor(nuc_midpoint / BIN_WIDTH) * BIN_WIDTH) %>%   # left edge label
  dplyr::count(sample, rep, unit_id, bin, name = "count")

# CRITICAL for an unbiased denominator: every fiber must appear in every bin.
# Fibers with no nucleosome anywhere in the window are dropped by the count
# above, so they are re-introduced here as zeros. Without this the "mean per
# fiber" would silently be a mean over occupied fibers only.
all_units <- combined_df %>% distinct(sample, rep, unit_id)

per_unit_df <- all_units %>%
  tidyr::crossing(bin = bins) %>%
  left_join(per_unit_df, by = c("sample", "rep", "unit_id", "bin")) %>%
  mutate(count = tidyr::replace_na(count, 0))

# ---- 4. Aggregate: mean rate +/- SEM across fibers, then smooth -------------
# Note the order: average first, smooth second. Smoothing the SEM the same way
# as the mean keeps the ribbon width comparable to the smoothed line, though it
# is a display convenience rather than a rigorous CI for the smoothed estimate.
agg_df <- per_unit_df %>%
  group_by(sample, bin) %>%                     # replicates pooled here
  summarise(
    mean_rate = mean(count) / BIN_WIDTH,              # nucleosomes / fiber / bp
    sem       = (sd(count) / sqrt(n())) / BIN_WIDTH,  # SEM across fibers
    n_fibers  = n(),
    .groups   = "drop"
  ) %>%
  group_by(sample) %>%
  arrange(bin, .by_group = TRUE) %>%
  mutate(
    rollmean = zoo::rollmean(mean_rate, k = SMOOTH_K, fill = NA, align = "center"),
    rollsem  = zoo::rollmean(sem,       k = SMOOTH_K, fill = NA, align = "center"),
    lo       = rollmean - rollsem,
    hi       = rollmean + rollsem
  ) %>%
  ungroup() %>%
  filter(!is.na(rollmean))   # drops the (SMOOTH_K - 1)/2 bins at each edge

# ---- 5. Plot ----------------------------------------------------------------
# The dotted lines at +/-85 bp mark roughly one nucleosome half-width from the
# motif centre, i.e. where a nucleosome midpoint would sit if the nucleosome
# edge abutted the motif.
metaplot <- ggplot(agg_df, aes(x = bin, y = rollmean, color = sample, fill = sample)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = c(-85, 85), linetype = "dotted",
             color = "black", alpha = 0.8) +
  labs(
    title = "Nucleosome midpoint rate around SNAI2 motifs (distal, exp2+exp3)",
    x     = "Position relative to SNAI2 motif (bp)",
    y     = "Nucleosomes per fiber per bp (mean \u00b1 SEM)"
  ) +
  # Analysed over +/-1000 bp but displayed over +/-650 bp: the outer bins are
  # noisier because fewer reads span them.
  coord_cartesian(xlim = c(-650, 650)) +
  theme_bw()

metaplot

ggsave(
  file.path(PATHS$fig_dir, "metaplot_SNAI2alldistmotifs_merger_reps_v7.pdf"),
  plot = metaplot, width = 7.29, height = 4.51
)
```

```{r metaplot-qc}
# Quick reporting numbers for the figure legend: fibers per condition.
agg_df %>%
  group_by(sample) %>%
  summarise(n_fibers = max(n_fibers), .groups = "drop")
```

---

# FIGURE 2 — Merged-replicate clustering heatmap + fold change

Motif subset: distal SNAI2 motifs at loci with H3K27ac up at 3 h, restricted to
the top log2FC quartile (Q3).

## Load the K27ac-Q3 tables

```{r load-k27ac-q3}
read_fiber_v6 <- function(path) {
  dt <- fread(path)
  stopifnot(ncol(dt) == length(FIBER_COLNAMES_V6))
  setnames(dt, FIBER_COLNAMES_V6)
  dt
}

# Variable names kept as in the original script so the `samples` list below
# stays recognisable:
#   *_all_k27ac_K27Q3        -> exp2 (r1+r2 merged)
#   *_dist_K27ac_exp3_K27Q3  -> exp3
DMSO_SNAI2_fibertools_all_k27ac_K27Q3      <- read_fiber_v6(PATHS$v6_k27ac_q3$DMSO_exp2)
dTag_SNAI2_fibertools_all_k27ac_K27Q3      <- read_fiber_v6(PATHS$v6_k27ac_q3$dTag_exp2)
DMSO_SNAI2_fibertools_dist_K27ac_exp3_K27Q3 <- read_fiber_v6(PATHS$v6_k27ac_q3$DMSO_exp3)
dTag_SNAI2_fibertools_dist_K27ac_exp3_K27Q3 <- read_fiber_v6(PATHS$v6_k27ac_q3$dTag_exp3)

sapply(
  list(DMSO1 = DMSO_SNAI2_fibertools_all_k27ac_K27Q3,
       dTag1 = dTag_SNAI2_fibertools_all_k27ac_K27Q3,
       DMSO2 = DMSO_SNAI2_fibertools_dist_K27ac_exp3_K27Q3,
       dTag2 = dTag_SNAI2_fibertools_dist_K27ac_exp3_K27Q3),
  nrow
)
```

## Build occupancy matrices and cluster

All four samples are still loaded and clustered **jointly** — that is what makes
the per-cluster fold change meaningful. The replicate merge happens only at the
drawing step.

```{r clustering, fig.width=6, fig.height=4}
# ============================================================================
# Fiber-seq clustering: per-read nucleosome occupancy around SNAI2 motifs
#
# v2: nucleosome footprints are SIZE-SELECTED (width >= NUC_MIN_WIDTH) at the
#     parsing step, i.e. BEFORE the occupancy matrix is built and before
#     k-means. Sub-threshold nucleosome calls therefore contribute neither to
#     the clustering nor to the heatmaps. MSP calls are left untouched.
# ============================================================================

# ---- 0. Parameters ----------------------------------------------------------
plot_window    <- 800    # +/- bp drawn in the heatmaps
cluster_window <- 100    # +/- bp actually fed to k-means (the motif-proximal core)
bin_width      <- 10     # bp per column of the occupancy matrix -> 20 columns
k_val          <- 5      # number of k-means clusters
NUC_MIN_WIDTH  <- 80     # bp; size selection applied BEFORE clustering

snai_palette  <- c("#a50f15", "#fc8d59")     # the two replicate bars
fig_dir       <- PATHS$fig_dir               # <- from the config chunk

# NUC_MIN_WIDTH is baked into the filename, so changing the parameter cannot
# silently overwrite an earlier figure. With the default 80 this resolves to
# "SNAI2_K27acUPmotifs_proportional_v2_presel80bp".
figure_prefix <- sprintf("SNAI2_K27acUPmotifs_proportional_v2_presel%dbp",
                         NUC_MIN_WIDTH)

samples <- list(
  DMSO1 = DMSO_SNAI2_fibertools_all_k27ac_K27Q3,       # exp2 DMSO
  dTag1 = dTag_SNAI2_fibertools_all_k27ac_K27Q3,       # exp2 dTag 3 h
  DMSO2 = DMSO_SNAI2_fibertools_dist_K27ac_exp3_K27Q3, # exp3 DMSO
  dTag2 = dTag_SNAI2_fibertools_dist_K27ac_exp3_K27Q3  # exp3 dTag 3 h
)

# Only used to fix factor level order; the per-sample colour palette from the
# full script is not needed here because the merged panels use a single fill.
sample_levels <- c("DMSO1", "dTag1", "DMSO2", "dTag2")

# ---- 1. Per-sample matrix + footprint table ---------------------------------
# Returns: mat        = binary read x bin nucleosome occupancy (size-selected)
#          footprints = size-selected nucleosomes + all MSPs
#          size_stats = how many nucleosome calls survived selection
build_sample_matrix <- function(df, sample_name,
                                cluster_window = 100,
                                bin_width      = 10,
                                nuc_min_width  = 80) {

  dt <- as.data.table(df)

  # query_name repeats across motif instances, so the row index is appended to
  # make a genuinely unique read x motif key. The sample name is prefixed so
  # IDs stay unique after the four matrices are rbind()-ed together.
  dt[, unique_id := paste0(sample_name, "_", query_name, "_", .I)]

  # Turn two parallel comma-separated columns into a long start/end table.
  # strsplit + rep(..., lengths()) is much faster here than separate_rows().
  parse_intervals <- function(starts_col, ends_col) {
    s_list <- strsplit(starts_col, ",", fixed = TRUE)
    e_list <- strsplit(ends_col,   ",", fixed = TRUE)
    n_per  <- lengths(s_list)
    data.table(
      unique_id = rep(dt$unique_id, n_per),
      start     = suppressWarnings(as.numeric(unlist(s_list))),
      end       = suppressWarnings(as.numeric(unlist(e_list)))
    )[!is.na(start) & !is.na(end)]   # drops empty fields from trailing commas
  }

  nucs <- parse_intervals(dt$centered_nuc_starts, dt$centered_nuc_ends)
  msps <- parse_intervals(dt$centered_msp_starts, dt$centered_msp_ends)

  nucs[, `:=`(width = end - start + 1, type = "Nucleosome", Sample = sample_name)]
  msps[, `:=`(width = end - start + 1, type = "MSP",        Sample = sample_name)]

  # ---- SIZE SELECTION (before matrix construction / clustering) -------------
  # Removes short "nucleosome" calls that are more likely to be subnucleosomal
  # or mis-segmented. Doing it here rather than at plotting time means the
  # k-means partition itself is defined on the filtered signal.
  n_nuc_total <- nrow(nucs)
  nucs        <- nucs[width >= nuc_min_width]
  size_stats  <- data.table(
    Sample    = sample_name,
    n_total   = n_nuc_total,
    n_kept    = nrow(nucs),
    frac_kept = if (n_nuc_total > 0) nrow(nucs) / n_nuc_total else NA_real_
  )

  footprints <- rbind(nucs, msps)   # MSPs kept unfiltered, for optional overlays

  read_ids  <- dt$unique_id
  n_reads   <- length(read_ids)
  n_bins    <- (2 * cluster_window) %/% bin_width
  col_names <- as.character(seq(-cluster_window, cluster_window - bin_width,
                                by = bin_width))

  # Only nucleosomes overlapping the clustering window matter for the matrix.
  nucs_w <- nucs[end >= -cluster_window & start < cluster_window]

  # Edge case: a sample with nothing in the window still needs a conformable
  # all-zero matrix, otherwise the rbind() in step 2 fails.
  if (nrow(nucs_w) == 0) {
    mat <- matrix(0L, n_reads, n_bins, dimnames = list(read_ids, col_names))
    return(list(mat = mat, footprints = footprints, size_stats = size_stats))
  }

  # Clip footprints to the window, then convert bp -> 1-based column indices.
  nucs_w[, `:=`(s = pmax(start, -cluster_window), e = pmin(end, cluster_window - 1))]
  nucs_w[, `:=`(
    sb = (floor(s / bin_width) * bin_width + cluster_window) %/% bin_width + 1L,
    eb = (floor(e / bin_width) * bin_width + cluster_window) %/% bin_width + 1L
  )]
  nucs_w[, row_idx := match(unique_id, read_ids)]
  nucs_w[, nuc_id  := .I]

  # One row per (read, bin) covered by a nucleosome.
  expanded <- nucs_w[, .(col = seq.int(sb, eb)), by = .(row_idx, nuc_id)]

  smat <- Matrix::sparseMatrix(
    i = expanded$row_idx, j = expanded$col, x = 1,
    dims = c(n_reads, n_bins)
  )
  # sparseMatrix() SUMS duplicate (i, j) entries, so two nucleosomes touching
  # the same bin on the same read would give 2. Cap at 1 to keep it binary.
  smat@x <- pmin(smat@x, 1)

  # kmeans() needs a dense matrix; the clustering window is only 20 columns
  # wide, so densifying here is cheap.
  mat <- as.matrix(smat)
  storage.mode(mat) <- "integer"
  rownames(mat) <- read_ids
  colnames(mat) <- col_names

  list(mat = mat, footprints = footprints, size_stats = size_stats)
}

all_data <- imap(samples, ~ build_sample_matrix(
  .x, .y,
  cluster_window = cluster_window,
  bin_width      = bin_width,
  nuc_min_width  = NUC_MIN_WIDTH
))

# QC: fraction of nucleosome calls surviving size selection, per sample.
# These should be similar across DMSO and dTag; a large asymmetry would mean
# the filter itself is driving condition differences.
size_report <- rbindlist(map(all_data, "size_stats"))
print(size_report)

# ---- 2. Joint k-means clustering (on size-selected occupancy) ---------------
# All four samples are clustered TOGETHER so cluster identity is comparable
# across conditions; the condition effect is then read out as a shift in the
# proportion of reads per cluster (step 5), not as a change in cluster shape.
master_mat <- do.call(rbind, map(all_data, "mat"))

# Reads with no >=NUC_MIN_WIDTH nucleosome in the +/-100 bp window are all-zero
# rows; they are retained and will fall into the lowest-occupancy cluster.
cat(sprintf("All-zero reads in clustering window: %d / %d (%.1f%%)\n",
            sum(rowSums(master_mat) == 0), nrow(master_mat),
            100 * mean(rowSums(master_mat) == 0)))

set.seed(42)   # kmeans is stochastic — the seed makes cluster labels reproducible
km <- kmeans(master_mat, centers = k_val, nstart = 10, iter.max = 100)

# k-means labels are arbitrary. Relabel 1..k in ascending mean occupancy so
# cluster 1 is always the most open and cluster k the most nucleosome-dense,
# making figures comparable across re-runs and across parameter settings.
clust_signal <- tapply(rowMeans(master_mat), km$cluster, mean)
relabel      <- setNames(rank(clust_signal, ties.method = "first"),
                         names(clust_signal))
new_cluster  <- as.integer(relabel[as.character(km$cluster)])

# Recover Sample and read_id from the composite unique_id. Anchored alternation
# over the known sample names, so read IDs containing underscores are safe.
sample_regex <- paste0("^(", paste(names(samples), collapse = "|"), ")_(.+)$")

cluster_assignments <- data.frame(
  unique_id = names(km$cluster),
  cluster   = factor(new_cluster, levels = seq_len(k_val))
) %>%
  tidyr::extract(unique_id, into = c("Sample", "read_id"),
                 regex = sample_regex, remove = FALSE)

stopifnot(!anyNA(cluster_assignments$Sample))   # regex must match every row

# ---- 3. Footprints + cluster assignments ------------------------------------
# Nucleosomes in this table are already size-selected, so footprint_type just
# labels the two footprint classes.
all_footprints_df <- imap_dfr(all_data, ~ .x$footprints)

plot_df <- all_footprints_df %>%
  # inner_join drops footprints from reads that were not clustered (there
  # should be none — a mismatch here means the unique_id scheme broke).
  inner_join(cluster_assignments %>% select(unique_id, cluster), by = "unique_id") %>%
  mutate(footprint_type = ifelse(type == "Nucleosome",
                                 sprintf("Nucleosome (\u2265%dbp)", NUC_MIN_WIDTH),
                                 "MSP (open)")) %>%
  filter(end >= -plot_window, start < plot_window)

nuc_label <- sprintf("Nucleosome (\u2265%dbp)", NUC_MIN_WIDTH)

# Reads per cluster, pooled across samples. Reported as QC only: the
# proportional facet heights in the figure come from facet_grid(space = "free"),
# not from this table.
reads_per_cluster <- cluster_assignments %>%
  dplyr::count(cluster, name = "n_reads") %>%
  arrange(as.integer(as.character(cluster)))

print(reads_per_cluster)

# ---- 4. Nucleosome-only footprint table for the heatmap ---------------------
plot_xlim <- c(-plot_window, plot_window)
motif_pos <- 0     # coordinates are motif-centred by construction

# The per-sample local_index of the full script is not computed here: the
# merged panels re-derive their own y-positions over the pooled replicate set
# (see make_merged_panel), so only the factor levels matter at this stage.
plot_df_nuc <- plot_df %>%
  filter(footprint_type == nuc_label) %>%      # heatmaps show nucleosomes only
  mutate(
    Sample  = factor(Sample,  levels = sample_levels),
    cluster = factor(cluster, levels = as.character(seq_len(k_val)))
  )
```

## Fold-change bar panel

```{r bar-panel, fig.width=3, fig.height=8}
# The quantitative claim of the figure: does depletion redistribute reads
# between occupancy clusters, and does it do so in both replicates?
cluster_stats <- cluster_assignments %>%
  dplyr::count(Sample, cluster, name = "n_reads") %>%
  group_by(Sample) %>%
  mutate(percent = 100 * n_reads / sum(n_reads)) %>%   # within-sample composition
  ungroup()

log2_df <- cluster_stats %>%
  select(Sample, cluster, percent) %>%
  pivot_wider(names_from = Sample, values_from = percent) %>%
  mutate(
    `Rep 1 (dTag/DMSO)` = log2(dTag1 / DMSO1),
    `Rep 2 (dTag/DMSO)` = log2(dTag2 / DMSO2)
  ) %>%
  select(cluster, `Rep 1 (dTag/DMSO)`, `Rep 2 (dTag/DMSO)`) %>%
  pivot_longer(
    cols      = c(`Rep 1 (dTag/DMSO)`, `Rep 2 (dTag/DMSO)`),
    names_to  = "Replicate_Pair",
    values_to = "Log2Ratio"
  )

print(log2_df)

bar_panel <- ggplot(log2_df,
                    aes(x = Log2Ratio, y = Replicate_Pair, fill = Replicate_Pair)) +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 0.4) +
  geom_col(width = 0.7, color = "black", linewidth = 0.2) +
  # Same faceting as the heatmaps so bars line up row-wise with their cluster.
  facet_grid(cluster ~ ., scales = "free_y", space = "free", switch = "y") +
  scale_fill_manual(values = snai_palette) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 3)) +
  labs(x = "Log2(dTag / DMSO)", y = NULL, fill = NULL, title = "Fold change") +
  theme_classic() +
  theme(
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    panel.spacing     = unit(0, "lines"),
    strip.background  = element_blank(),
    strip.text.y.left = element_text(angle = 0, size = 7, face = "bold",
                                     margin = margin(r = 4, l = 2)),
    panel.border      = element_blank(),
    legend.position   = "bottom",
    plot.title        = element_text(size = 10)
  )
```

## Assemble: merged DMSO | merged dTag | log2FC bars

```{r heatmap-merged-reps, fig.width=8, fig.height=8}
# Replicates collapsed within condition, so the visual comparison is a single
# DMSO column against a single dTag column. The bar panel still shows the two
# replicates separately — that is where reproducibility is judged.
make_merged_panel <- function(samples_to_merge, merged_name, fill_col) {
  df_merged <- plot_df_nuc %>%
    dplyr::filter(Sample %in% samples_to_merge) %>%
    arrange(cluster, unique_id) %>%
    group_by(cluster) %>%
    # y-position within each cluster block, over the POOLED replicate set.
    # Reads are ordered by unique_id: arbitrary but stable. Change the
    # arrange() key above if you want reads sorted by footprint position.
    mutate(local_index = as.numeric(factor(unique_id, levels = unique(unique_id)))) %>%
    ungroup()

  ggplot(df_merged) +
    # rasterise() flattens the tens of thousands of rectangles into a bitmap
    # layer; without it the PDF is unopenable. Axes and text stay vector.
    rasterise(
      geom_rect(aes(xmin = pmax(start, plot_xlim[1]),
                    xmax = pmin(end,   plot_xlim[2]),
                    ymin = local_index - 0.45,
                    ymax = local_index + 0.45),
                fill = fill_col, color = NA),
      dpi = 150, dev = "ragg_png"
    ) +
    # space = "free" makes facet height proportional to reads per cluster.
    facet_grid(cluster ~ ., scales = "free_y", space = "free", switch = "y") +
    coord_cartesian(xlim = plot_xlim) +
    geom_vline(xintercept = motif_pos, linetype = "dashed",
               linewidth = 0.6, alpha = 0.7, color = "black") +
    theme_classic() +
    theme(
      axis.text.y       = element_blank(),   # individual reads are unlabelled
      axis.ticks.y      = element_blank(),
      axis.line         = element_blank(),
      panel.spacing     = unit(0, "lines"),  # clusters read as one stacked block
      strip.background  = element_blank(),
      strip.text.y.left = element_text(angle = 0, size = 7, face = "bold",
                                       margin = margin(r = 4, l = 2)),
      panel.border      = element_blank()
    ) +
    labs(title = merged_name, y = NULL, x = NULL)
}

p_dmso_merged <- make_merged_panel(c("DMSO1", "DMSO2"), "DMSO (merged)", "coral4")
p_dtag_merged <- make_merged_panel(c("dTag1", "dTag2"), "dTag (merged)", "coral4")

merged_combined <- wrap_plots(
  list(p_dmso_merged, p_dtag_merged, bar_panel),
  nrow = 1
) +
  plot_layout(widths = c(1, 1, 0.5)) +   # bar panel is half-width
  plot_annotation(
    title = paste("Merged replicates + per-cluster fold change:", figure_prefix)
  )

merged_combined

# cairo_pdf handles the UTF-8 "≥" in the labels; the default pdf() device will
# error or drop the glyph.
ggsave(file.path(fig_dir, paste0("Heatmap_V5_mergedReps_plus_FC_", figure_prefix, ".pdf")),
       plot = merged_combined, width = 8, height = 8, device = cairo_pdf)
```

---

# Session info

```{r session-info}
# Recorded for reproducibility — include the knitted HTML or this block when
# the code is deposited alongside the paper.
sessionInfo()
```
