---
title: "DAF-seq single-molecule chromatin analysis (fiberHMM v2.9.7)"
author: "Lucia Ichino"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = FALSE, dev = "ragg_png")
```

<!-- ==========================================================================
     WHAT THIS DOCUMENT DOES

     Single-molecule (DAF-seq / Fiber-seq style) analysis of nucleosome
     footprints, MSPs and TF footprints called by fiberHMM v2.9.7, at six
     enhancers, comparing DMSO vs dTag (TWIST1 degron, SNAI1/2 degron).

     Per enhancer it produces:
       1. a read-level "heatmap" of nucleosome footprints, k-means clustered
          on the local nucleosome pattern, with an ATAC track and DMSO->dTag
          cluster-flow ribbons;
       2. a log2(dTag/DMSO) bar panel of the % of reads in each cluster;
       3. an open/closed summary log2FC, where "open" clusters are defined
          automatically from MSP geometry around the motif;
       4. TF-footprint occupancy metaplots (all reads vs accessible reads);
       5. a cross-enhancer summary and a CMH-tested TF occupancy barplot.

     STRUCTURE
       Section 1  CONFIGURATION  <-- the ONLY place with absolute paths
       Section 2  Libraries
       Section 3  Enhancer registry (names, coordinates, motifs, k, clusters)
       Section 4  Sample / file-path helpers
       Section 5  Core functions
       Section 6  Per-enhancer analyses
       Section 7  Cross-enhancer summaries
     ========================================================================== -->


# 1. Configuration

<!-- ==========================================================================
     >>>>>>>>>>>>>>>>  THE ONLY PLACE WITH ABSOLUTE PATHS  <<<<<<<<<<<<<<<<

     Before pushing to GitHub, replace the four strings below with relative
     paths or environment lookups, e.g.

         DAF_ROOT <- Sys.getenv("DAF_ROOT", "data/fiberHMM")
         ATAC_DIR <- Sys.getenv("ATAC_DIR", "data/atac")
         FIG_DIR  <- "figures"
         ROBJ_DIR <- "cache"

     Nothing else in this document contains a machine-specific path.
     ========================================================================== -->

```{r config}
## ---- paths -----------------------------------------------------------------
DAF_ROOT <- "/Volumes/storage/DAFseq/DAFseq_pacbio_exp2_andplasmid_fiber/DAF_QC_marlowe"
ATAC_DIR <- file.path("/Users/luciaichino/Library/CloudStorage/Dropbox",
                      "lucia (1)/Lucia 09_05_16/postdoc/Wysocka_lab_personal_folder",
                      "seq_data/dataforIGV")
FIG_DIR  <- file.path("/Users/luciaichino/Library/CloudStorage/Dropbox",
                      "lucia (1)/Lucia 09_05_16/postdoc/Wysocka_lab_personal_folder",
                      "SNAI_paper/codes/single molecule/DAF/figures_min1")
ROBJ_DIR <- file.path(FIG_DIR, "min1Robj")

## ---- run options -----------------------------------------------------------
FIBERHMM_VER <- "fiberHMM_v2.9.7"

# TRUE  = recompute bundles + k-means and overwrite the cache
# FALSE = reuse cached .rds where present

REBUILD_CLUSTERS <- TRUE

## ---- k-means -----------------------------------------------------------
# kmeans(centers = <k>) seeds by SAMPLING ROWS, so the
# result depends on the order the bundles happen to be stacked in. Rows are now
# sorted into a canonical, locale-independent order before clustering, and the
# search is run much harder, so a run is reproducible from (reads, k, seed)
# alone.
KMEANS_SEED      <- 42L
KMEANS_NSTART    <- 200L   # was 25 -- see C7 for why 25 was not enough
KMEANS_ITER_MAX  <- 200L
# Full-data k-means at nstart = 200 is impractical here (~180k reads x ~400 bp
# per start). Instead the many random starts are run on a subsample to find
# good centres, then ONE deterministic pass over all reads refines them from
# those centres. Set to NULL to run every start on the full matrix.
KMEANS_SUBSAMPLE <- 20000L

## ---- analysis choices ------------------------------------------------------
TF_SCORE_THR <- 70L    # blockTq quality floor for TF footprints
TF_MAX_WIDTH <- 40L    # drop TF "footprints" wider than this (nucleosome-scale)
MIN_READ_COV <- 5L     # bp with fewer spanning reads than this are NA

LOW_COV_RULE <- "both"

MIN_READS_FC      <- 30L
MIN_READS_FC_ENH1 <- 110L

dir.create(FIG_DIR,  recursive = TRUE, showWarnings = FALSE)
dir.create(ROBJ_DIR, recursive = TRUE, showWarnings = FALSE)

fig  <- function(...) file.path(FIG_DIR,  ...)   # figure output path
robj <- function(...) file.path(ROBJ_DIR, ...)   # cached object path

# null-coalescing operator (base R only has it from 4.4.0)
`%||%` <- function(a, b) if (is.null(a)) b else a
```

# 2. Libraries

```{r libraries}
library(tidyverse)      # dplyr / tidyr / purrr / stringr / ggplot2 / readr
library(GenomicRanges)
library(rtracklayer)    # import.bw() for the ATAC tracks
library(patchwork)      # p1 | p2 composition
library(ggpubr)         # ggarrange();  FIX B7: was loaded mid-document
library(ggnewscale)     # second fill scale for the motif bands
library(ggrastr)        # rasterise the dense footprint rectangles
library(zoo)            # rollmean() in the midpoint metaplot
library(scales)         # comma(), pretty_breaks()

# fiberHMM writes very long BED lines (one block per footprint per read).
# Only relevant if a file is ever handed to readr as a *connection*.
Sys.setenv(VROOM_CONNECTION_SIZE = 500072)

# ---- conflicted -------------------------------------------------------------
# tidyverse + Bioconductor collide on a lot of common verb names (filter, lag,
# rename, select, slice, first, last, reduce, expand, the set operations...).
# Ambiguous calls in this document are namespaced explicitly, but if the
# `conflicted` package is attached these preferences settle the rest up front
# rather than letting them surface one error at a time mid-run.
if (requireNamespace("conflicted", quietly = TRUE) &&
    "conflicted" %in% loadedNamespaces()) {
  prefer <- function(pkg, fn) {
    try(conflicted::conflict_prefer(fn, pkg, quiet = TRUE), silent = TRUE)
  }
  for (f in c("filter", "lag", "select", "rename", "slice", "mutate", "count",
              "first", "last", "desc", "collapse", "summarise", "n")) prefer("dplyr", f)
  for (f in c("expand", "unnest", "complete", "extract"))              prefer("tidyr", f)
  for (f in c("map", "reduce", "discard", "compact", "simplify"))      prefer("purrr", f)
  for (f in c("setdiff", "intersect", "union", "setequal"))            prefer("base",  f)
  rm(prefer, f)
}
```

# 3. Enhancer registry

<!-- ==========================================================================
     One row per analysed window. `label` is the ONLY name that reaches a
     figure. Everything else in the document refers to enhancers by their
     internal key and calls enh_label() to render the name.

     `cluster_range` selects which reads enter the k-means and defines the
     matrix columns. `plot_range` is the (wider) window that gets drawn.
     `minset`        is the fiberHMM call set used -- see reviewer note Q1.
     `accessible`    are the cluster IDs treated as "open" in the TF
                     occupancy quantification (post-hoc, from the metaplots).
     ========================================================================== -->

```{r enhancer-registry}
ENH <- list(

  enh4 = list(
    label   = "Enh 1",
    chrom   = "chr20",
    minset  = "min1",                    # <-- Q1: the only enhancer on min1
    # clustering window spans BOTH the TWIST1 and the SNAI motif
    cluster = IRanges(17169646, 17170010),
    plot    = IRanges(17168966, 17170700),
    tf_plot = IRanges(17169600, 17170100),
    k       = 12L,
    accessible = "1",
    motifs  = tribble(
      ~TF,      ~xmin,     ~xmax,     ~col,
      "TWIST1",  17169866,  17169882, "turquoise3",
      "TFAP2A",  17169974,  17169985, "deepskyblue",
      "SNAI",    17169746,  17169754, "magenta3"
    ),
    # the TF metaplots used a different palette for the same three motifs
    motifs_tf = tribble(
      ~TF,      ~xmin,     ~xmax,     ~col,
      "TWIST1",  17169866,  17169882, "blue",
      "TFAP2A",  17169974,  17169985, "#6baed6",
      "SNAI",    17169746,  17169754, "purple"
    ),
    # alternative clustering windows tried during development:
    #   TWIST1 only : IRanges(17169794, 17170010)
    #   SNAI only   : IRanges(17169646, 17169854)
    peaks = list(
      snai   = list(motif = c(17169746, 17169754), tf = "SNAI"),
      twist1 = list(motif = c(17169866, 17169882), tf = "TWIST1")
    )
  ),

  enh10 = list(
    label   = "Enh 2",
    chrom   = "chr11",
    minset  = "min3",
    plot    = IRanges(12318593, 12320381),
    tf_plot = IRanges(12318950, 12320181),
    k       = 10L,
    # NO top-level `accessible` on purpose: this enhancer is clustered three
    # times, once per peak, so "cluster 1" is a different set of molecules in
    # each. The accessible sets live on the peak entries below, and
    # enh_accessible("enh10") now errors rather than guessing.
    motifs  = tribble(
      ~TF,       ~xmin,     ~xmax,     ~col,
      "TWIST1",   12319212,  12319228, "turquoise3",
      "SNAI1/2",  12319565,  12319573, "magenta3",
      "E-box",    12319892,  12319899, "turquoise4"
    ),
    # full motif set used for the TF metaplots (includes lower-confidence calls)
    motifs_tf = tribble(
      ~TF,       ~xmin,     ~xmax,     ~col,
      "TFAP2",    12319535,  12319546, "blue",
      "TWIST1",   12319212,  12319228, "blue",
      "SNAI",     12319565,  12319573, "purple",
      "?",        12319607,  12319626, "blue",
      "E-box",    12319892,  12319899, "turquoise4"
    ),
    # three separate clusterings, one per accessible peak
    peaks = list(
      peak1 = list(cluster = IRanges(12319112, 12319328),
                   motif = c(12319212, 12319228), tf = "TWIST1",
                   accessible = c("1", "2")),
      peak2 = list(cluster = IRanges(12319465, 12319665),
                   motif = c(12319565, 12319573), tf = "SNAI",
                   accessible = c("1")),
      # peak3$accessible is READ BY NOTHING today -- there is no peak-3 TF
      # metaplot chunk. Editing it has no effect until one is added.
      peak3 = list(cluster = IRanges(12319792, 12319999),
                   motif = c(12319892, 12319899), tf = "E-box",
                   accessible = c("1"))
    )
  ),

  enh8 = list(
    label   = "Enh 3",
    chrom   = "chr20",
    minset  = "min3",
    plot    = IRanges(57257593, 57259949),
    tf_plot = IRanges(57257900, 57259650),
    atac_plot = IRanges(57257593, 57259600),
    k       = 10L,
    accessible = "1",
    motifs  = tribble(
      ~TF,        ~xmin,     ~xmax,     ~col,
      "SNAI",      57258380,  57258388, "purple",
      "AP2A",      57258364,  57258374, "blue",
      "TWIST1",    57259398,  57259414, "blue",
      "TWIST1",    57259464,  57259480, "blue",
      "ZNF184?",   57259320,  57259332, "blue"
    ),
    peaks = list(
      peak1 = list(cluster = IRanges(57258280, 57258488),
                   motif = c(57258380, 57258388), tf = "SNAI",
                   motifs = tribble(~TF, ~xmin, ~xmax, ~col,
                                    "SNAI", 57258380, 57258388, "purple")),
      peak2 = list(cluster = IRanges(57259298, 57259514),
                   motif = c(57259398, 57259414), tf = "TWIST1",
                   motifs = tribble(~TF, ~xmin, ~xmax, ~col,
                                    "SNAI",   57258380, 57258388, "darkorchid",
                                    "TWIST1", 57259398, 57259414, "darkcyan"))
    )
  ),

  enh15 = list(
    label   = "Enh 4",
    chrom   = "chr6",
    minset  = "min3",
    # NB the section header in the old file quoted chr6:79,609,697-79,612,498
    # as the locus; the window actually drawn is the narrower one below.
    cluster = IRanges(79610335, 79610535),   # around the SNAI motif
    plot    = IRanges(79609721, 79611537),
    k       = 10L,
    accessible = "1",
    motifs  = tribble(
      ~TF,     ~xmin,     ~xmax,     ~col,
      "SNAI",   79610435,  79610443, "purple"
    ),
    # alternative clustering window tried: IRanges(79610635, 79610845) (TFAP2A)
    peaks = list(
      snai = list(motif = c(79610435, 79610443), tf = "SNAI")
    )
  ),

  enh30 = list(
    label   = "Enh 5",
    chrom   = "chr3",
    minset  = "min3",
    cluster = IRanges(69116902, 69117314),
    plot    = IRanges(69116276, 69117925),
    tf_plot = IRanges(69116800, 69118000),
    atac_plot = IRanges(69116600, 69117925),
    k       = 10L,
    accessible = "1",
    motifs  = tribble(
      ~TF,     ~xmin,     ~xmax,     ~col,
      "SNAI",   69117105,  69117112, "purple"
    ),
    # SOX9 (chr3:69117056-69117072) produced no footprint and is omitted
    motifs_tf = tribble(
      ~TF,        ~xmin,     ~xmax,     ~col,
      "SNAI",      69117105,  69117112, "purple",
      "STAT1/4",   69117023,  69117031, "blue",
      "AP2A",      69117131,  69117141, "blue",
      "NR2F1",     69117159,  69117165, "blue",
      "NR2F1",     69117169,  69117175, "blue",
      "ZNF596?",   69117861,  69117871, "blue"
    ),
    peaks = list(
      snai = list(motif = c(69117105, 69117112), tf = "SNAI")
    )
  ),

  enh33 = list(
    label   = "Enh 6",
    chrom   = "chr11",
    minset  = "min3",
    cluster = IRanges(8708990, 8709198),
    plot    = IRanges(8708345, 8709876),
    tf_plot = IRanges(8708750, 8709964),
    atac_plot = IRanges(8708500, 8709964),
    k       = 10L,
    accessible = "1",
    motifs  = tribble(
      ~TF,     ~xmin,    ~xmax,    ~col,
      "SNAI",   8709103,  8709110, "purple"
    ),
    motifs_tf = tribble(
      ~TF,       ~xmin,    ~xmax,    ~col,
      "SNAI",     8709103,  8709110, "purple",
      "TWIST1",   8709132,  8709147, "blue",
      "TFAP2A",   8709060,  8709071, "blue",
      "TFAP2A",   8709416,  8709427, "blue",
      "TFAP2A",   8709511,  8709523, "blue",
      "PRDM5",    8709453,  8709463, "blue",
      "PRDM5",    8709924,  8709934, "blue"
    ),
    peaks = list(
      snai = list(motif = c(8709103, 8709110), tf = "SNAI")
    )
  )
)

# ---- accessors -------------------------------------------------------------
# enh_label("enh10")           -> "Enh 2"
# enh_label("enh10", "peak 1") -> "Enh 2 peak 1"
enh_label <- function(key, suffix = NULL) {
  lab <- ENH[[key]]$label
  if (is.null(lab)) stop("No registry entry for enhancer key '", key, "'")
  if (is.null(suffix)) lab else paste(lab, suffix)
}

# "enh10" -> "Enh2"; used to build output filenames from the paper name so the
# PDFs on disk can no longer disagree with the titles inside them.
enh_slug <- function(key, suffix = NULL) {
  s <- gsub(" ", "", ENH[[key]]$label)
  if (is.null(suffix)) s else paste0(s, "_", gsub("[^A-Za-z0-9]+", "", suffix))
}

# fig_name("enh10", "heatmap", "peak1") -> <FIG_DIR>/Enh2_peak1_heatmap_v297.pdf
fig_name <- function(key, what, suffix = NULL, ext = "pdf") {
  fig(paste0(enh_slug(key, suffix), "_", what, "_v297.", ext))
}

# GRanges built from the registry, so chrom can never drift from coordinates
enh_gr <- function(key, which = c("cluster", "plot", "tf_plot", "atac_plot"),
                   peak = NULL) {
  which <- match.arg(which)
  cfg   <- ENH[[key]]
  ir    <- if (!is.null(peak) && !is.null(cfg$peaks[[peak]][[which]])) {
    cfg$peaks[[peak]][[which]]
  } else if (!is.null(cfg[[which]])) {
    cfg[[which]]
  } else {
    cfg$plot     # sensible fallback: tf_plot/atac_plot default to plot
  }
  GRanges(cfg$chrom, ir)
}

# Accessible ("open") cluster IDs, per enhancer or per peak.
#
# WHY THIS IS A FUNCTION AND NOT A BARE $accessible LOOKUP:
# compute_tf_metaplot(clusters = NULL) applies NO cluster filter and returns
# the all-reads curve. So a missing or mistyped `accessible` would not error --
# it would quietly draw every read under an "accessible reads" title. Erroring
# here makes that impossible.
enh_accessible <- function(key, peak = NULL) {
  cfg <- ENH[[key]]
  if (is.null(cfg)) stop("No registry entry for enhancer key '", key, "'")
  a <- if (!is.null(peak)) cfg$peaks[[peak]]$accessible else cfg$accessible
  if (is.null(a) || length(a) == 0) {
    stop("No `accessible` cluster set for '", key,
         if (!is.null(peak)) paste0("' peak '", peak, "'") else "'",
         "`. Enhancers clustered per peak (e.g. enh10) require the `peak` ",
         "argument, because cluster 1 means a different thing in each ",
         "clustering. Returning NULL here would silently pool EVERY cluster.")
  }
  as.character(a)
}

# motif band table for the plotting helpers
enh_motifs <- function(key, set = c("motifs", "motifs_tf"), peak = NULL) {
  set <- match.arg(set)
  cfg <- ENH[[key]]
  if (!is.null(peak) && !is.null(cfg$peaks[[peak]]$motifs)) return(cfg$peaks[[peak]]$motifs)
  if (!is.null(cfg[[set]])) return(cfg[[set]])
  cfg$motifs
}
```

# 4. Samples and file paths

```{r sample-helpers}
# ---- short sample name  <->  on-disk directory stem -------------------------
# "SNAI_rep1_DMSO5min"  ->  "SNAI1FVSNAI2KO_rep1_DMSO5min"
# "TWIST1_rep1_DMSO5min"->  "TWIST1FV_rep1_DMSO5min"
SAMPLE_PREFIX <- c(SNAI = "SNAI1FVSNAI2KO", TWIST1 = "TWIST1FV")

sample_stem <- function(samples) {
  tf   <- sub("_.*$", "", samples)
  rest <- sub("^[^_]+_", "", samples)
  unname(paste0(SAMPLE_PREFIX[tf], "_", rest))
}

# ---- canonical sample vectors ----------------------------------------------
SNAI_SAMPLES <- c(
  "SNAI_rep1_DMSO5min", "SNAI_rep1_dTag5min",
  "SNAI_rep2_DMSO5min", "SNAI_rep2_dTag5min",
  "SNAI_rep1_DMSO1min", "SNAI_rep1_dTag1min",
  "SNAI_rep2_DMSO1min", "SNAI_rep2_dTag1min"
)
TWIST1_SAMPLES <- sub("^SNAI", "TWIST1", SNAI_SAMPLES)
BOTH_SAMPLES   <- c(SNAI_SAMPLES, TWIST1_SAMPLES)

only <- function(samples, pattern) grep(pattern, samples, value = TRUE)

# ---- fiberHMM .bed path builder --------------------------------------------
# kind: "footprint" (nucleosomes), "tf" (TF footprints), "msp" (accessible patches)
daf_bed <- function(samples, minset, kind = c("footprint", "tf", "msp")) {
  kind <- match.arg(kind)
  stem <- sample_stem(samples)
  setNames(
    file.path(DAF_ROOT, minset, "results", stem, FIBERHMM_VER,
              paste0(stem, ".recalled_", kind, ".bed")),
    samples
  )
}

# convenience: the file set an enhancer needs, using ITS minset (see Q1)
enh_files <- function(key, samples, kind = "footprint") {
  daf_bed(samples, ENH[[key]]$minset, kind)
}

# ---- bulk ATAC bigWigs ------------------------------------------------------
ATAC_BW <- c(
  SNAI_DMSO   = file.path(ATAC_DIR, "ATAC_DMSO_merged.bw"),
  SNAI_dTag   = file.path(ATAC_DIR, "ATAC_3h_merged.bw"),
  TWIST1_DMSO = file.path(ATAC_DIR, "GSM7213525_TWIST1FV_dTAG0h_r1_ATAC.bw"),
  TWIST1_dTag = file.path(ATAC_DIR, "GSM7213526_TWIST1FV_dTAG3h_r1_ATAC.bw")
)

# map merged pseudo-sample names -> the right bigWig
atac_for <- function(merged_samples) {
  key <- paste0(ifelse(grepl("TWIST1", merged_samples), "TWIST1", "SNAI"), "_",
                ifelse(grepl("DMSO",   merged_samples), "DMSO",   "dTag"))
  setNames(unname(ATAC_BW[key]), merged_samples)
}

# ---- replicate merging ------------------------------------------------------
# "SNAI_rep1_DMSO5min" -> group "SNAI_DMSO_5min"
# Generated, never typed, so the merged names can never drift between sections.
merged_name <- function(samples) {
  tf   <- ifelse(grepl("^TWIST1", samples), "TWIST1", "SNAI")
  trt  <- ifelse(grepl("DMSO",    samples), "DMSO",   "dTag")
  time <- stringr::str_extract(samples, "\\d+min")
  paste(tf, trt, time, sep = "_")
}
merge_groups_for <- function(samples) split(samples, merged_name(samples))

# ---- palettes ---------------------------------------------------------------
twist_palette <- c("#08519c", "#6baed6")   # dark blue,  light blue
snai_palette  <- c("#a50f15", "#fc8d59")   # dark red,   light red

TF_FILL <- c(TWIST1 = "darkblue", SNAI = "darkred")   # read-rectangle colours

META_COLORS <- c(                                     # metaplot line colours
  TWIST1_DMSO = "#56B4E9", TWIST1_dTag = "#CC79A7",
  SNAI_DMSO   = "#99000D", SNAI_dTag   = "#FC9272"
)

# shared geometry for every linked heatmap, so the panels are comparable
LINKED_STYLE <- list(
  raster_dpi = 300, cluster_gap = 0.8, binding_site_linewidth = 0.4,
  lane_height = 180, atac_gap = 5, atac_height = 10,
  label_gap = 5, label_tier_gap = 8,
  sample_label_gap = 7, group_label_gap = 14
)

# ---- caching ----------------------------------------------------------------
# Replaces the old saveRDS(x, f); x <- readRDS(f) pattern (see C2).
cache_rds <- function(name, expr, rebuild = REBUILD_CLUSTERS) {
  f <- robj(paste0(name, ".rds"))
  if (!rebuild && file.exists(f)) {
    message("cache hit : ", basename(f))
    return(readRDS(f))
  }
  message("computing : ", basename(f))
  val <- force(expr)
  saveRDS(val, f)
  val
}
```

# 5. Core functions

## 5.1 Reading fiberHMM BED files

<!-- ==========================================================================
     fiberHMM writes one BED12-like line per READ, with one block per
     footprint. Two flavours appear in this project:

       13 columns : nucleosome footprint and MSP files (v2.9.7). Column 13 is
                    blockRgb.
       15 columns : TF footprint files. Columns 13-15 are blockTq / blockEl /
                    blockEr, where blockTq is the per-footprint quality byte
                    of fiberHMM's tf.QQQ annotation. blockEl / blockEr are
                    edge-sharpness bytes, NOT quality -- the score filter must
                    use blockTq.

     WHY NOT rtracklayer::import(): its BED/bigBed parser tries to coerce the
     list-valued extra columns to a single integer and fails. We parse as
     plain text instead.
     ========================================================================== -->

```{r fn-read-bed}
# 13-column layout (nucleosome / MSP files)
BED_COLS_13 <- c("chrom", "chromStart", "chromEnd", "name", "score", "strand",
                 "thickStart", "thickEnd", "itemRgb", "blockCount",
                 "blockSizes", "chromStarts", "blockRgb")

# 15-column layout (TF footprint files)
BED_COLS_15 <- c("chrom", "chromStart", "chromEnd", "name", "score", "strand",
                 "thickStart", "thickEnd", "itemRgb", "blockCount",
                 "blockSizes", "chromStarts", "blockTq", "blockEl", "blockEr")

#' Parse one fiberHMM .bed into reads + footprints
#'
#' @return list(reads, footprints)
#'   reads      : chrom, read_name, read_start, read_end, read_score, strand
#'   footprints : chrom, read_name, fp_start, fp_end, fp_width [, blockTq]
#'
#' COORDINATES: BED is 0-based half-open. Both tables are returned 1-based
#' inclusive, so a footprint of width w has fp_end - fp_start + 1 == w.
read_fiberhmm_bed <- function(path) {

  lines    <- read_lines(path)
  n_fields <- str_count(lines, "\t") + 1L

  # Auto-detect the layout from the modal field count.
  expected <- as.integer(names(which.max(table(n_fields))))
  if (!expected %in% c(13L, 15L)) {
    warning(sprintf("%s: unexpected field count %d", path, expected))
  }
  cols_to_use <- if (expected == 15L) BED_COLS_15 else BED_COLS_13

  # A rare fiberHMM writer bug fuses blockEl/blockEr with no delimiter. Drop
  # those rows rather than failing the whole import.
  # FIX B1: dropping rows here used to silently break compute_msp_motif_stats(),
  #         which recovered `chrom` by matching row POSITION against a separate
  #         readLines() of the same file. `chrom` is now carried through here,
  #         so a dropped row can no longer shift the chromosome assignment.
  bad <- which(n_fields != expected)
  if (length(bad) > 0) {
    warning(sprintf("%s: %d malformed row(s) dropped (lines: %s%s)",
                    path, length(bad), paste(head(bad, 10), collapse = ", "),
                    if (length(bad) > 10) ", ..." else ""))
    lines <- lines[-bad]
  }

  raw <- read_tsv(
    I(paste(lines, collapse = "\n")),
    col_names = cols_to_use,
    col_types = cols(.default    = col_character(),
                     chromStart  = "i", chromEnd   = "i",
                     score       = "i", thickStart = "i",
                     thickEnd    = "i", blockCount = "i")
  )

  reads_tbl <- raw %>%
    transmute(
      chrom,
      read_name  = name,
      read_start = chromStart + 1L,   # 0-based -> 1-based inclusive
      read_end   = chromEnd,
      read_score = score,
      strand
    )

  has_tq <- "blockTq" %in% names(raw)

  # FIX B3: blockTq used to be unnested unconditionally. For 13-column files it
  #         was created as vector("list", n) -- a length-0 entry per row --
  #         and unnesting that against length-k blockSizes is version-dependent
  #         at best. No-op for 15-column files; makes the 13-column path safe.
  fp_base <- raw %>%
    transmute(
      chrom,
      read_name   = name,                                    # BED col 4
      .read0      = chromStart,                              # 0-based read start
      blockSizes  = lapply(strsplit(blockSizes,  ","), as.integer),
      chromStarts = lapply(strsplit(chromStarts, ","), as.integer),
      blockTq     = if (has_tq) lapply(strsplit(blockTq, ","), as.integer) else NULL
    )

  unnest_cols <- if (has_tq) c("blockSizes", "chromStarts", "blockTq")
                 else        c("blockSizes", "chromStarts")

  fps_tbl <- fp_base %>%
    unnest(all_of(unnest_cols)) %>%
    mutate(
      fp_start = .read0 + chromStarts + 1L,                  # 1-based inclusive
      fp_end   = .read0 + chromStarts + blockSizes,
      fp_width = blockSizes
    ) %>%
    select(chrom, read_name, fp_start, fp_end, fp_width, any_of("blockTq"))

  list(reads = reads_tbl, footprints = fps_tbl)
}
```

## 5.2 Building per-sample bundles

```{r fn-bundles}
#' Reads overlapping cluster_range, plus ALL of those reads' footprints
#'
#' Footprints are deliberately NOT clipped here: the clustering matrix clips to
#' cluster_range, the plot clips to plot_range, and those are different windows.
build_sample_bundle <- function(path, sample_name, cluster_range) {

  obj          <- read_fiberhmm_bed(path)
  region_chrom <- as.character(seqnames(cluster_range))
  region_start <- start(cluster_range)
  region_end   <- end(cluster_range)

  reads <- obj$reads %>%
    dplyr::filter(chrom == region_chrom,
                  read_end >= region_start, read_start <= region_end) %>%
    mutate(unique_id = paste0(sample_name, "_", read_name))

  if (nrow(reads) == 0) return(NULL)

  footprints <- obj$footprints %>%
    dplyr::filter(read_name %in% reads$read_name) %>%
    mutate(unique_id = paste0(sample_name, "_", read_name))

  list(reads = reads, footprints = footprints,
       region_start = region_start, region_end = region_end)
}

#' Run build_sample_bundle() over a named vector of paths
build_all_bundles <- function(files, cluster_range) {

  missing <- files[!file.exists(files)]
  if (length(missing) > 0) {
    stop("File(s) not found:\n", paste(missing, collapse = "\n"))
  }

  bundles <- imap(files, ~ build_sample_bundle(.x, .y, cluster_range))

  empty <- names(Filter(is.null, bundles))
  if (length(empty) > 0) {
    message("Samples with 0 reads in cluster_range (dropped): ",
            paste(empty, collapse = ", "))
  }

  bundles <- compact(bundles)
  if (length(bundles) == 0) stop("No reads overlapped cluster_range in any sample.")
  bundles
}
```

## 5.3 Merging replicates for joint clustering

<!-- ==========================================================================
     Replicates are clustered TOGETHER so that a cluster means the same thing
     in rep1 and rep2, then split back apart for the per-replicate log2FC.

     Each read is re-stamped
         unique_id = "<merged_name>_<orig_sample>READSEP<orig_read_name>"
     cluster_reads() parses the merged name off the front and
     split_cluster_assignments() reverses the READSEP token.
     ========================================================================== -->

```{r fn-merge-split}
READSEP <- "READSEP"

merge_replicate_bundles <- function(bundles, merge_groups) {

  missing_samples <- base::setdiff(unlist(merge_groups), names(bundles))
  if (length(missing_samples) > 0) {
    stop("merge_replicate_bundles: sample(s) not in bundles: ",
         paste(missing_samples, collapse = ", "))
  }

  imap(merge_groups, function(orig_samples, merged_name) {

    parts <- bundles[orig_samples]

    restamp <- function(df, s_name) {
      df %>% mutate(
        orig_read_name = sub(paste0("^", s_name, "_"), "", unique_id),
        unique_id      = paste0(merged_name, "_", s_name, READSEP, orig_read_name)
      )
    }

    list(
      reads        = imap_dfr(parts, function(b, s) restamp(b$reads,      s)),
      footprints   = imap_dfr(parts, function(b, s) restamp(b$footprints, s)),
      region_start = parts[[1]]$region_start,
      region_end   = parts[[1]]$region_end
    )
  })
}

#' Expand merged cluster assignments back to one row per (replicate, read)
split_cluster_assignments <- function(cluster_assignments_merged) {

  merged_rows <- grepl(READSEP, cluster_assignments_merged$read_id, fixed = TRUE)

  if (!any(merged_rows)) {
    warning("split_cluster_assignments: no '", READSEP, "' token found. ",
            "Were these produced by merge_replicate_bundles()? Returning input.")
    return(cluster_assignments_merged)
  }

  cluster_assignments_merged %>%
    dplyr::filter(merged_rows) %>%
    mutate(
      orig_sample    = sub(paste0(READSEP, ".*$"), "", read_id),
      orig_read_name = sub(paste0("^.*", READSEP), "", read_id)
    ) %>%
    transmute(
      unique_id = paste0(orig_sample, "_", orig_read_name),
      cluster,
      Sample    = orig_sample,
      read_id   = orig_read_name
    )
}
```

## 5.4 Occupancy matrix and k-means clustering

```{r fn-cluster}
#' Binary nucleosome-occupancy matrix: rows = reads, columns = bp of cluster_range
bundle_to_matrix <- function(bundle) {

  region_start <- bundle$region_start
  region_end   <- bundle$region_end
  width_range  <- region_end - region_start + 1L

  reads <- bundle$reads
  mat   <- matrix(0, nrow = nrow(reads), ncol = width_range,
                  dimnames = list(reads$unique_id, NULL))

  ft <- bundle$footprints %>%
    dplyr::filter(fp_end >= region_start, fp_start <= region_end) %>%
    mutate(rel_start = pmax(1L,           fp_start - region_start + 1L),
           rel_end   = pmin(width_range,  fp_end   - region_start + 1L))

  row_idx <- match(ft$unique_id, rownames(mat))
  for (i in seq_len(nrow(ft))) {
    if (!is.na(row_idx[i])) mat[row_idx[i], ft$rel_start[i]:ft$rel_end[i]] <- 1
  }
  mat
}

#' k-means that does not depend on the order rows were stacked in
#'
#' PROBLEM : kmeans(centers = <a number>) chooses its initial
#' centres by SAMPLING ROWS of the matrix. Two runs with the same seed but a
#' different row order therefore start from different centres and can land in
#' different local optima. In this dataset that was worth ~11% of reads.
#'
#' FIX, in two parts:
#'   1. the caller sorts rows into a canonical order before calling this;
#'   2. the random restarts are run on a SUBSAMPLE to find good centres, then a
#'      single pass over the full matrix refines them. Passing `centers` as a
#'      matrix makes that final pass fully deterministic -- no sampling at all.
#'
#' Why subsample: nstart = 200 on ~180k x ~400 is hours of compute, and nstart
#' = 25 was demonstrably too few to converge on a stable partition. Running 200
#' starts on 20k rows explores far more of the space for a fraction of the cost,
#' and the full-data refinement recovers the exact centres.
#'
#' @return the kmeans object, plus $init_method describing how it was seeded
kmeans_stable <- function(mat, k,
                          seed      = KMEANS_SEED,
                          nstart    = KMEANS_NSTART,
                          iter_max  = KMEANS_ITER_MAX,
                          subsample = KMEANS_SUBSAMPLE) {

  set.seed(seed)

  if (is.null(subsample) || nrow(mat) <= subsample) {
    km <- kmeans(mat, centers = k, nstart = nstart, iter.max = iter_max)
    km$init_method <- sprintf("full data, nstart = %d", nstart)
  } else {
    idx <- sort(sample.int(nrow(mat), subsample))
    km0 <- kmeans(mat[idx, , drop = FALSE], centers = k,
                  nstart = nstart, iter.max = iter_max)
    # a degenerate subsample can yield an empty centre; refuse rather than
    # silently return fewer than k clusters
    if (nrow(km0$centers) < k) {
      stop("kmeans_stable: subsample produced only ", nrow(km0$centers),
           " of ", k, " centres. Increase KMEANS_SUBSAMPLE or lower k.")
    }
    km <- kmeans(mat, centers = km0$centers, iter.max = iter_max)
    km$init_method <- sprintf("centres from %d-row subsample, nstart = %d, then one full pass",
                              subsample, nstart)
  }

  if (!is.null(km$ifault) && km$ifault == 4L) {
    warning("kmeans did not converge in ", iter_max,
            " iterations -- raise KMEANS_ITER_MAX.")
  }
  km
}

#' Pool every sample's reads, k-means, then relabel clusters by "openness"
#'
#' After k-means the clusters are RENUMBERED by mean nucleosome occupancy,
#' ascending: cluster 1 = fewest nucleosomes = most open, cluster k = most
#' nucleosome-occupied. Set rank_clusters = FALSE to keep k-means' arbitrary
#' labels.
#'
#' @return tibble(unique_id, Sample, read_id, cluster) with attr "kmeans_info"
cluster_reads <- function(bundles, k, seed = KMEANS_SEED, rank_clusters = TRUE,
                          nstart = KMEANS_NSTART, subsample = KMEANS_SUBSAMPLE) {

  master_mat <- do.call(rbind, map(bundles, bundle_to_matrix))

  # FIX C7, part 1: canonical row order, so the clustering depends only on the
  # reads / k / seed and not on how the bundles happened to be stacked.
  # method = "radix" sorts in C collation, which is locale-independent --
  # plain order() on strings can differ between machines and locales.
  master_mat <- master_mat[order(rownames(master_mat), method = "radix"), ,
                           drop = FALSE]

  if (anyDuplicated(rownames(master_mat))) {
    stop("cluster_reads: duplicated read ids in the pooled matrix. ",
         "Two bundles are contributing the same unique_id.")
  }

  km         <- kmeans_stable(master_mat, k = k, seed = seed,
                              nstart = nstart, subsample = subsample)
  cluster_id <- km$cluster

  if (rank_clusters) {
    clust_signal <- tapply(rowMeans(master_mat), cluster_id, mean)
    relabel      <- setNames(rank(clust_signal, ties.method = "first"),
                             names(clust_signal))
    cluster_id   <- as.integer(relabel[as.character(cluster_id)])
  }

  sample_names <- names(bundles)
  ordered      <- sample_names[order(nchar(sample_names), decreasing = TRUE)]
  prefixes     <- paste0(ordered, "_")

  ids <- rownames(master_mat)
  hit <- rep(NA_integer_, length(ids))
  for (i in seq_along(prefixes)) {
    todo      <- is.na(hit) & startsWith(ids, prefixes[i])
    hit[todo] <- i
  }

  if (anyNA(hit)) {
    stop("cluster_reads: ", sum(is.na(hit)), " read id(s) matched no sample name. ",
         "First offender: ", ids[which(is.na(hit))[1]])
  }

  smp <- ordered[hit]

  out <- tibble(
    unique_id = ids,
    Sample    = smp,
    read_id   = substring(ids, nchar(smp) + 2L),   # drop "<Sample>_"
    cluster   = factor(cluster_id, levels = seq_len(k))
  )

  # Record how well this fit actually did, so two runs can be compared on
  # something better than eyeballing the panels. tot_withinss is the quantity
  # k-means minimises: lower is a better fit to the SAME data at the SAME k.
  attr(out, "kmeans_info") <- list(
    k             = k,
    seed          = seed,
    nstart        = nstart,
    subsample     = subsample,
    init_method   = km$init_method,
    n_reads       = nrow(master_mat),
    n_positions   = ncol(master_mat),
    tot_withinss  = km$tot.withinss,
    betweenss     = km$betweenss,
    prop_between  = km$betweenss / km$totss,
    iterations    = km$iter,
    cluster_sizes = as.integer(table(factor(cluster_id, levels = seq_len(k))))
  )
  out
}

#' Agreement between two clusterings of the same reads (Adjusted Rand Index)
#'
#' 1 = identical up to relabelling, ~0 = no better than chance. Use it to check
#' whether a partition is a property of the data or of the random seed.
adjusted_rand <- function(a, b) {
  tab <- table(a, b)
  nij <- sum(choose(tab, 2))
  ai  <- sum(choose(rowSums(tab), 2))
  bj  <- sum(choose(colSums(tab), 2))
  n   <- choose(sum(tab), 2)
  exp <- ai * bj / n
  (nij - exp) / ((ai + bj) / 2 - exp)
}

#' One row per footprint, tagged with its cluster, restricted to plot_range,
#' with a within-(Sample, cluster) stacking index for drawing.
prepare_plot_data <- function(bundles, cluster_assignments, plot_range) {

  imap_dfr(bundles, function(b, s_name) {
    b$footprints %>%
      transmute(unique_id, Sample = s_name,
                start = fp_start, end = fp_end, width = fp_width)
  }) %>%
    inner_join(cluster_assignments, by = c("unique_id", "Sample")) %>%
    dplyr::filter(start <= end(plot_range), end >= start(plot_range)) %>%
    arrange(Sample, cluster, unique_id) %>%
    group_by(Sample, cluster) %>%
    mutate(local_index = as.numeric(factor(unique_id, levels = unique(unique_id)))) %>%
    ungroup()
}
```

## 5.5 Shared coverage helpers

```{r fn-coverage}
#' Look up a per-sample fill colour: exact name, then regex, then a warned default
.resolve_fill <- function(s, fill_color, default = "grey30") {
  if (is.null(names(fill_color))) return(fill_color[[1]])
  if (s %in% names(fill_color))   return(unname(fill_color[[s]]))
  hit <- names(fill_color)[vapply(names(fill_color),
                                  function(p) grepl(p, s), logical(1))]
  if (length(hit) >= 1) return(unname(fill_color[[hit[1]]]))
  warning(sprintf("No fill colour matched sample '%s'; using %s", s, default))
  default
}

#' Per-bp integer coverage over [region_start, region_end] from a set of intervals
#'
#' FIX B2: intervals lying entirely OUTSIDE the region give s > e after
#'         clipping. `s:e` then counts downwards and adds coverage at the wrong
#'         positions instead of adding none. This matters because reads are
#'         selected on cluster_range but coverage is computed on plot_range,
#'         so a read outside plot_range is perfectly possible. Now dropped.
.coverage_vector <- function(starts, ends, region_start, region_end) {

  n_pos   <- region_end - region_start + 1L
  cov_vec <- integer(n_pos)
  if (length(starts) == 0) return(cov_vec)

  s <- pmax(starts, region_start) - region_start + 1L
  e <- pmin(ends,   region_end)   - region_start + 1L

  keep <- which(e >= s)                    # <-- the guard
  for (i in keep) cov_vec[s[i]:e[i]] <- cov_vec[s[i]:e[i]] + 1L
  cov_vec
}

#' Every read's extent + Sample, tagged with its cluster.
#' This is the per-bp read-coverage DENOMINATOR for every metaplot below.
reads_with_clusters <- function(bundles, cluster_assignments) {
  imap_dfr(bundles, function(b, s_name) {
    b$reads %>% transmute(Sample = s_name, unique_id, read_start, read_end)
  }) %>%
    inner_join(cluster_assignments %>% select(Sample, unique_id, cluster),
               by = c("Sample", "unique_id"))
}
```

## 5.6 Per-cluster log2 fold change

<!-- ==========================================================================
     `comparisons` is a named list; each element is a length-2 NAMED vector
         c(numerator = "<sample>", denominator = "<sample>")
     compute_log2fc_df() computes, per cluster, the % of that sample's reads
     falling in the cluster, then log2(numerator% / denominator%).
     ========================================================================== -->

```{r fn-log2fc}
#' @param min_reads     read-count floor for the "low coverage" flag
#' @param low_cov_rule  "both"   flag only if BOTH sides are below min_reads
#'                               (original behaviour -- see reviewer note Q2)
#'                      "either" flag if EITHER side is below min_reads
#'                               (the more conservative rule)
compute_log2fc_df <- function(cluster_assignments, comparisons,
                              min_reads    = MIN_READS_FC,
                              low_cov_rule = LOW_COV_RULE) {

  low_cov_rule <- match.arg(low_cov_rule, c("both", "either"))

  cluster_stats <- cluster_assignments %>%
    dplyr::count(Sample, cluster, name = "n_reads") %>%
    group_by(Sample) %>%
    mutate(percent = 100 * n_reads / sum(n_reads)) %>%
    ungroup()

  percent_wide <- cluster_stats %>%
    select(Sample, cluster, percent) %>%
    pivot_wider(names_from = Sample, values_from = percent, values_fill = 0)

  reads_wide <- cluster_stats %>%
    select(Sample, cluster, n_reads) %>%
    pivot_wider(names_from = Sample, values_from = n_reads, values_fill = 0)

  imap_dfr(comparisons, function(pair, label) {

    num <- pair[["numerator"]]
    den <- pair[["denominator"]]

    missing <- base::setdiff(c(num, den), names(percent_wide))
    if (length(missing) > 0) {
      stop("Comparison '", label, "' references sample(s) not in ",
           "cluster_assignments: ", paste(missing, collapse = ", "))
    }

    # a sample absent from reads_wide has 0 reads in every cluster
    num_reads <- if (num %in% names(reads_wide)) reads_wide[[num]] else rep(0L, nrow(reads_wide))
    den_reads <- if (den %in% names(reads_wide)) reads_wide[[den]] else rep(0L, nrow(reads_wide))

    low <- if (low_cov_rule == "both") {
      num_reads < min_reads & den_reads < min_reads
    } else {
      num_reads < min_reads | den_reads < min_reads
    }

    pct_num <- percent_wide[[num]]
    pct_den <- percent_wide[[den]]

    tibble(
      cluster        = percent_wide$cluster,
      Replicate_Pair = label,
      num_reads      = num_reads,
      den_reads      = den_reads,
      Log2Ratio      = log2(pct_num / pct_den),
      # A cluster emptied (or created) by the treatment gives +/-Inf. That is a
      # real result, not a defect -- but geom_col silently DROPS non-finite
      # values, so without this flag the strongest effect on the panel would be
      # the one bar you could not see. make_log2fc_panel() draws these clamped
      # to the panel edge with a marker.
      status = dplyr::case_when(
        num_reads == 0 & den_reads == 0 ~ "no reads",
        num_reads == 0                  ~ "complete loss",
        den_reads == 0                  ~ "complete gain",
        TRUE                            ~ "ok"
      )
    ) %>%
      left_join(tibble(cluster = reads_wide$cluster, low_coverage = low),
                by = "cluster")
  })
}

#' Horizontal log2FC bar panel, one facet row per cluster.
#' Designed to sit alongside a heatmap that uses the same
#' facet_grid(space = "free_y") cluster rows.
make_log2fc_panel <- function(log2fc_df, title = "Fold change",
                              palette = twist_palette,
                              xlim = NULL,
                              low_coverage_color = "grey70",
                              low_coverage_label = "low coverage",
                              show_boundary = TRUE) {

  if (!"low_coverage" %in% names(log2fc_df)) log2fc_df$low_coverage <- FALSE
  if (!"status" %in% names(log2fc_df))       log2fc_df$status <- "ok"

  # A cluster that the treatment empties gives log2 = -Inf (or +Inf if it
  # appears only after treatment). geom_col drops non-finite values without
  # comment, so the strongest effect on the panel would otherwise be invisible.
  # Clamp those bars to the panel edge and mark them, leaving every finite
  # value exactly as computed -- no pseudocount, nothing rescaled.
  finite_vals <- log2fc_df$Log2Ratio[is.finite(log2fc_df$Log2Ratio)]
  edge <- if (!is.null(xlim)) {
    max(abs(xlim))
  } else if (length(finite_vals) > 0) {
    max(abs(finite_vals)) * 1.15
  } else 1

  log2fc_df <- log2fc_df %>%
    mutate(
      is_boundary = !is.finite(Log2Ratio) & !is.na(Log2Ratio),
      Log2Plot    = ifelse(is_boundary, sign(Log2Ratio) * edge, Log2Ratio),
      # collapse the fill grouping so every flagged bar shares one grey category
      fill_group  = ifelse(low_coverage, low_coverage_label,
                           as.character(Replicate_Pair))
    )

  rp_levels   <- unique(as.character(log2fc_df$Replicate_Pair))
  fill_values <- setNames(palette[seq_along(rp_levels)], rp_levels)
  fill_values[low_coverage_label] <- low_coverage_color

  p <- ggplot(log2fc_df, aes(x = Log2Plot, y = Replicate_Pair, fill = fill_group)) +
    geom_vline(xintercept = 0, colour = "black", linetype = "dashed", linewidth = 0.4) +
    geom_col(width = 0.7, colour = "black", linewidth = 0.2) +
    facet_grid(cluster ~ ., scales = "free_y", space = "free_y", switch = "y") +
    scale_fill_manual(values = fill_values) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 3)) +
    labs(x = "Log2(ratio)", y = NULL, fill = NULL, title = title) +
    theme_classic() +
    theme(
      axis.text.y       = element_blank(),
      axis.ticks.y      = element_blank(),
      axis.line         = element_blank(),
      panel.spacing     = unit(0, "lines"),
      panel.border      = element_blank(),
      strip.background  = element_blank(),
      strip.text.y.left = element_text(angle = 0, size = 7, face = "bold",
                                       margin = margin(r = 4, l = 2)),
      legend.position   = "bottom",
      plot.title        = element_text(size = 10)
    ) +
    geom_vline(xintercept = -Inf, colour = "grey70", linewidth = 0.3)

  # arrowhead at the clamped end, so the bar reads as "runs off the scale"
  # rather than as a real value at the panel limit
  if (show_boundary && any(log2fc_df$is_boundary)) {
    bd <- log2fc_df %>% dplyr::filter(is_boundary)
    p <- p +
      geom_segment(data = bd,
                   aes(x = Log2Plot * 0.75, xend = Log2Plot,
                       y = Replicate_Pair, yend = Replicate_Pair),
                   inherit.aes = FALSE, colour = "black", linewidth = 0.3,
                   arrow = arrow(length = unit(0.05, "in"), type = "closed")) +
      geom_text(data = bd,
                aes(x = Log2Plot * 0.5, y = Replicate_Pair,
                    label = ifelse(Log2Ratio < 0, "0 reads after", "0 reads before")),
                inherit.aes = FALSE, colour = "black",
                size = 2.2, fontface = "italic", vjust = -1)
  }

  if (any(log2fc_df$low_coverage)) {
    lbl_df <- log2fc_df %>%
      dplyr::filter(low_coverage) %>%
      distinct(cluster, Replicate_Pair) %>%
      mutate(x = 0)
    p <- p + geom_text(data = lbl_df,
                       aes(x = x, y = Replicate_Pair, label = low_coverage_label),
                       inherit.aes = FALSE, colour = "black",
                       size = 2.5, fontface = "italic")
  }

  if (!is.null(xlim)) p <- p + coord_cartesian(xlim = xlim)
  p
}
```

## 5.7 Dense per-sample heatmap (independent panels)

```{r fn-dense}
#' One patchwork panel per sample, faceted by cluster.
#' Pass log2fc_df to append a fold-change bar panel on the right.
plot_fiberseq_dense <- function(plot_df, plot_range, sample_order,
                                vlines = NULL, vline_colors = NULL,
                                fill_color = TF_FILL,
                                binding_sites = NULL,
                                title = "Clustered Fiber-seq: dense per-cell view",
                                log2fc_df = NULL,
                                log2fc_title = "Fold change",
                                log2fc_palette = twist_palette,
                                log2fc_width = 0.5,
                                log2fc_xlim = NULL,
                                raster = TRUE, raster_dpi = 300,
                                xlab = NULL) {

  chr  <- as.character(seqnames(plot_range))
  xlab <- xlab %||% paste0("Genomic coordinate (", chr, ")")

  plot_df <- plot_df %>%
    dplyr::filter(Sample %in% sample_order) %>%
    mutate(Sample = factor(Sample, levels = sample_order))

  plot_xlim <- c(start(plot_range), end(plot_range))
  if (!is.null(vlines) && is.null(vline_colors)) {
    vline_colors <- rep("black", length(vlines))
  }

  make_sample_panel <- function(s) {

    df_s  <- plot_df %>% dplyr::filter(Sample == s)
    col_s <- .resolve_fill(s, fill_color)

    rect_layer <- geom_rect(
      data = df_s,
      aes(xmin = pmax(start, plot_xlim[1]),
          xmax = pmin(end,   plot_xlim[2]),
          ymin = local_index - 0.45,
          ymax = local_index + 0.45),
      fill = col_s, colour = NA
    )
    if (raster) {
      rect_layer <- ggrastr::rasterise(rect_layer, dpi = raster_dpi, dev = "ragg_png")
    }

    sample_label <- data.frame(
      x = mean(plot_xlim), y = Inf, label = s,
      cluster = factor(levels(df_s$cluster)[1], levels = levels(df_s$cluster))
    )

    p <- ggplot(df_s) +
      rect_layer +
      facet_grid(cluster ~ ., scales = "free_y", space = "free_y", switch = "y") +
      coord_cartesian(xlim = plot_xlim, clip = "off") +
      theme_classic() +
      theme(
        axis.text.y       = element_blank(),
        axis.ticks.y      = element_blank(),
        axis.line         = element_blank(),
        panel.spacing     = unit(0, "lines"),
        strip.background  = element_blank(),
        strip.text.y.left = element_text(angle = 0, size = 7,
                                         margin = margin(r = 4, l = 2)),
        panel.border      = element_blank(),
        plot.margin       = margin(t = 40, r = 5, b = 5, l = 5)
      ) +
      labs(title = NULL, y = NULL, x = xlab) +
      geom_text(data = sample_label, aes(x = x, y = y, label = label),
                vjust = -3, size = 3.5, inherit.aes = FALSE)

    if (!is.null(binding_sites)) {
      bs_labels <- binding_sites %>%
        mutate(cluster = factor(levels(df_s$cluster)[1],
                                levels = levels(df_s$cluster)))
      p <- p +
        geom_rect(data = binding_sites,
                  aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = col),
                  alpha = 1, inherit.aes = FALSE) +
        scale_fill_identity(guide = "none") +
        geom_text(data = bs_labels,
                  aes(x = (xmin + xmax) / 2, y = Inf, label = TF, colour = col),
                  vjust = -1, size = 3, fontface = "bold",
                  inherit.aes = FALSE, show.legend = FALSE) +
        scale_colour_identity(guide = "none")
    }
    p
  }

  plots <- lapply(sample_order, make_sample_panel)
  n_s   <- length(sample_order)

  if (!is.null(log2fc_df)) {
    bar_panel <- make_log2fc_panel(log2fc_df, title = log2fc_title,
                                   palette = log2fc_palette, xlim = log2fc_xlim)
    wrap_plots(c(plots, list(bar_panel)), nrow = 1) +
      plot_layout(widths = c(rep(1, n_s), log2fc_width)) +
      plot_annotation(title = title)
  } else {
    wrap_plots(plots, nrow = 1) + plot_annotation(title = title)
  }
}
```

## 5.8 Metaplots

<!-- ==========================================================================
     OCCUPANCY, both flavours:
         occupancy(bp) = footprints_covering(bp) / read_coverage(bp) * 100
     with bp below min_read_cov spanning reads set to NA.

     They differ ONLY in the read-coverage DENOMINATOR:
       NUCLEOSOME metaplots : normalised over WHOLE-SAMPLE depth, i.e. all of
                              that sample's reads at each bp across every
                              cluster, so per-cluster curves share one scale.
       TF metaplots         : normalised PER CLUSTER, i.e. only reads in the
                              cluster being drawn.
     ========================================================================== -->

```{r fn-nuc-metaplot}
#' Per-cluster nucleosome occupancy, normalised over whole-sample depth
compute_cluster_metaplot <- function(plot_df, cluster_assignments, plot_range,
                                     bundles, clusters = NULL,
                                     min_read_cov = MIN_READ_COV) {

  region_start <- start(plot_range)
  region_end   <- end(plot_range)
  pos_vec      <- region_start:region_end

  # denominator: per-bp read coverage from ALL reads of each sample
  reads_df        <- reads_with_clusters(bundles, cluster_assignments)
  samples_present <- unique(reads_df$Sample)

  read_cov_by_sample <- setNames(lapply(samples_present, function(s_name) {
    r <- reads_df[reads_df$Sample == s_name, ]
    .coverage_vector(r$read_start, r$read_end, region_start, region_end)
  }), samples_present)

  df <- plot_df
  if (!is.null(clusters)) {
    df <- df %>% dplyr::filter(as.character(cluster) %in% as.character(clusters))
  }

  groups <- df %>% distinct(Sample, cluster)

  purrr::map2_dfr(groups$Sample, groups$cluster, function(s_name, c_name) {
    # base-R subsetting to avoid NSE name collisions with the column names
    sub      <- df[df$Sample == s_name & df$cluster == c_name, ]
    fp_cov   <- .coverage_vector(sub$start, sub$end, region_start, region_end)
    read_cov <- read_cov_by_sample[[s_name]]

    tibble(Sample = s_name, cluster = c_name, pos = pos_vec,
           fp_coverage   = fp_cov,
           read_coverage = read_cov,
           normalized_occupancy = ifelse(read_cov >= min_read_cov,
                                         fp_cov / read_cov * 100, NA_real_))
  })
}

#' Draw a per-bp occupancy table: one panel per cluster, samples overlaid
plot_cluster_metaplot <- function(metaplot_df, plot_range,
                                  clusters = NULL, samples = NULL,
                                  free_y = FALSE,
                                  binding_sites = NULL,
                                  vlines = NULL, vline_colors = NULL,
                                  sample_colors = NULL,
                                  title = "Nucleosome footprint metaplot by cluster",
                                  subtitle = "Occupancy = footprints / read coverage x 100") {

  metaplot_df <- metaplot_df %>% mutate(cluster = as.character(cluster))

  if (!is.null(samples)) {
    metaplot_df <- metaplot_df %>%
      dplyr::filter(Sample %in% samples) %>%
      mutate(Sample = factor(Sample, levels = samples))
  }

  if (!is.null(clusters)) {
    clusters    <- as.character(clusters)
    metaplot_df <- metaplot_df %>%
      dplyr::filter(cluster %in% clusters) %>%
      mutate(cluster = factor(cluster, levels = clusters))
  } else {
    lvls        <- as.character(sort(as.integer(unique(metaplot_df$cluster))))
    metaplot_df <- metaplot_df %>% mutate(cluster = factor(cluster, levels = lvls))
  }

  if (!is.null(vlines) && is.null(vline_colors)) {
    vline_colors <- rep("black", length(vlines))
  }

  p <- ggplot(metaplot_df, aes(x = pos, y = normalized_occupancy))

  # motif bands first, on their own identity scales, then hand the fill/colour
  # scales back to the Sample aesthetics via ggnewscale
  if (!is.null(binding_sites)) {
    p <- p +
      geom_rect(data = binding_sites,
                aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = col),
                alpha = 0.15, inherit.aes = FALSE) +
      scale_fill_identity(guide = "none") +
      geom_text(data = binding_sites,
                aes(x = (xmin + xmax) / 2, y = Inf, label = TF, colour = col),
                vjust = 1.5, size = 3.5, fontface = "bold",
                inherit.aes = FALSE, show.legend = FALSE) +
      scale_colour_identity(guide = "none") +
      ggnewscale::new_scale_fill() +
      ggnewscale::new_scale_colour()
  }

  p <- p +
    geom_area(aes(fill = Sample), position = "identity", alpha = 0.1) +
    geom_line(aes(colour = Sample), linewidth = 0.8) +
    facet_wrap(~ cluster, scales = if (free_y) "free_y" else "fixed",
               ncol = 1, strip.position = "right") +
    coord_cartesian(xlim = c(start(plot_range), end(plot_range))) +
    theme_classic() +
    labs(title = title, subtitle = subtitle,
         x = "Genomic coordinate", y = "Occupancy %") +
    theme(legend.position  = "bottom",
          strip.background = element_blank(),
          strip.text       = element_text(face = "bold"),
          panel.spacing    = unit(0.2, "lines"))

  p <- p + if (!is.null(sample_colors)) {
    list(scale_colour_manual(values = sample_colors),
         scale_fill_manual(values   = sample_colors))
  } else {
    list(scale_colour_brewer(palette = "Set1"),
         scale_fill_brewer(palette   = "Set1"))
  }

  if (!is.null(vlines)) {
    for (i in seq_along(vlines)) {
      p <- p + geom_vline(xintercept = vlines[i], linetype = "dashed",
                          linewidth = 0.4, alpha = 0.5, colour = vline_colors[i])
    }
  }
  p
}
```

```{r fn-tf-metaplot}
#' Load TF footprint files and inherit each footprint's cluster from its
#' parent read's NUCLEOSOME cluster.
#'
#' A TF footprint is kept only if (a) its blockTq quality passes
#' score_threshold, (b) it is no wider than max_width, (c) it overlaps
#' plot_range, and (d) its read was assigned a cluster upstream.
build_tf_metaplot_data <- function(tf_files, cluster_assignments, plot_range,
                                   score_threshold = TF_SCORE_THR,
                                   max_width = TF_MAX_WIDTH) {

  missing <- tf_files[!file.exists(tf_files)]
  if (length(missing) > 0) {
    stop("TF file(s) not found:\n", paste(missing, collapse = "\n"))
  }

  region_chrom <- as.character(seqnames(plot_range))
  region_start <- start(plot_range)
  region_end   <- end(plot_range)

  clust_lookup <- cluster_assignments %>% select(Sample, unique_id, cluster)

  tf_df <- imap_dfr(tf_files, function(path, s_name) {

    raw_fps <- read_fiberhmm_bed(path)$footprints

    # blockTq is fiberHMM's tf.QQQ quality byte. blockEl / blockEr are
    # edge-sharpness bytes and must NOT be used as a quality score.
    fps <- raw_fps %>%
      transmute(
        Sample    = s_name,
        unique_id = paste0(s_name, "_", read_name),
        chrom,
        start     = fp_start,
        end       = fp_end,
        width     = fp_width,
        score     = if ("blockTq" %in% names(raw_fps)) raw_fps$blockTq
                    else rep(NA_integer_, nrow(raw_fps))
      )

    if (!is.null(score_threshold)) {
      if (all(is.na(fps$score))) {
        warning(sprintf("%s: score_threshold ignored -- no blockTq column", s_name))
      } else {
        fps <- fps %>% dplyr::filter(score >= score_threshold)
      }
    }
    if (!is.null(max_width)) fps <- fps %>% dplyr::filter(width <= max_width)

    fps %>% dplyr::filter(chrom == region_chrom,
                          start <= region_end, end >= region_start)
  })

  tf_df %>% inner_join(clust_lookup, by = c("Sample", "unique_id"))
}

#' Per-bp TF occupancy.
#' pool_clusters = FALSE : one curve per (sample, cluster), denominator = that
#'                         cluster's reads
#' pool_clusters = TRUE  : one curve per sample, denominator = all reads (or
#'                         all reads in `clusters`, if given)
compute_tf_metaplot <- function(tf_df, cluster_assignments, plot_range,
                                bundles, clusters = NULL, pool_clusters = FALSE,
                                min_read_cov = MIN_READ_COV) {

  region_start <- start(plot_range)
  region_end   <- end(plot_range)
  pos_vec      <- region_start:region_end

  reads_df <- reads_with_clusters(bundles, cluster_assignments)
  df       <- tf_df

  if (!is.null(clusters)) {
    df       <- df       %>% dplyr::filter(as.character(cluster) %in% as.character(clusters))
    reads_df <- reads_df %>% dplyr::filter(as.character(cluster) %in% as.character(clusters))
  }

  fp_cov <- function(sub) .coverage_vector(sub$start,      sub$end,      region_start, region_end)
  rd_cov <- function(sub) .coverage_vector(sub$read_start, sub$read_end, region_start, region_end)
  to_occ <- function(fp, rd) ifelse(rd >= min_read_cov, fp / rd * 100, NA_real_)

  if (pool_clusters) {
    purrr::map_dfr(unique(df$Sample), function(s_name) {
      fp <- fp_cov(df[df$Sample == s_name, ])
      rd <- rd_cov(reads_df[reads_df$Sample == s_name, ])
      tibble(Sample = s_name, pos = pos_vec,
             fp_coverage = fp, read_coverage = rd,
             normalized_occupancy = to_occ(fp, rd))
    })
  } else {
    groups <- df %>% distinct(Sample, cluster)
    purrr::map2_dfr(groups$Sample, groups$cluster, function(s_name, c_name) {
      fp <- fp_cov(df[df$Sample == s_name & df$cluster == c_name, ])
      rd <- rd_cov(reads_df[reads_df$Sample == s_name & reads_df$cluster == c_name, ])
      tibble(Sample = s_name, cluster = c_name, pos = pos_vec,
             fp_coverage = fp, read_coverage = rd,
             normalized_occupancy = to_occ(fp, rd))
    })
  }
}

#' Thin wrapper: per-cluster TF panel
plot_tf_metaplot <- function(tf_metaplot_df, plot_range, ...,
                             title = "TF footprint metaplot by cluster",
                             subtitle = "Occupancy = TF footprints / read coverage x 100 (per cluster)") {
  plot_cluster_metaplot(tf_metaplot_df, plot_range, ...,
                        title = title, subtitle = subtitle)
}

#' Single panel, clusters already pooled
plot_tf_metaplot_overall <- function(tf_metaplot_df, plot_range,
                                     samples = NULL,
                                     binding_sites = NULL,
                                     vlines = NULL, vline_colors = NULL,
                                     sample_colors = NULL,
                                     title = "TF footprint metaplot (all clusters)",
                                     subtitle = "Occupancy = TF footprints / read coverage x 100",
                                     xlab = "Genomic coordinate") {

  if (!is.null(samples)) {
    tf_metaplot_df <- tf_metaplot_df %>%
      dplyr::filter(Sample %in% samples) %>%
      mutate(Sample = factor(Sample, levels = samples))
  }
  if (!is.null(vlines) && is.null(vline_colors)) {
    vline_colors <- rep("black", length(vlines))
  }

  p <- ggplot(tf_metaplot_df, aes(x = pos, y = normalized_occupancy))

  if (!is.null(binding_sites)) {
    p <- p +
      geom_rect(data = binding_sites,
                aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = col),
                alpha = 0.15, inherit.aes = FALSE) +
      scale_fill_identity(guide = "none") +
      geom_text(data = binding_sites,
                aes(x = (xmin + xmax) / 2, y = Inf, label = TF, colour = col),
                vjust = 1.5, size = 3.5, fontface = "bold",
                inherit.aes = FALSE, show.legend = FALSE) +
      scale_colour_identity(guide = "none") +
      ggnewscale::new_scale_fill() +
      ggnewscale::new_scale_colour()
  }

  p <- p +
    geom_area(aes(fill = Sample), position = "identity", alpha = 0.1) +
    geom_line(aes(colour = Sample), linewidth = 0.8) +
    coord_cartesian(xlim = c(start(plot_range), end(plot_range))) +
    theme_classic() +
    labs(title = title, subtitle = subtitle, x = xlab, y = "Occupancy %") +
    theme(legend.position = "bottom")

  p <- p + if (!is.null(sample_colors)) {
    list(scale_colour_manual(values = sample_colors),
         scale_fill_manual(values   = sample_colors))
  } else {
    list(scale_colour_brewer(palette = "Set1"),
         scale_fill_brewer(palette   = "Set1"))
  }

  if (!is.null(vlines)) {
    for (i in seq_along(vlines)) {
      p <- p + geom_vline(xintercept = vlines[i], linetype = "dashed",
                          linewidth = 0.4, alpha = 0.5, colour = vline_colors[i])
    }
  }
  p
}

#' Average replicates of a per-bp metaplot table into one curve per condition.
#'
#' Replaces merge_reps() / merge_snai_reps(), which were defined five times in
#' the old document with three different output label conventions (C5).
average_replicates <- function(df, keep_timepoint = FALSE) {
  df %>%
    mutate(
      .tf   = ifelse(grepl("TWIST1", Sample), "TWIST1", "SNAI"),
      .trt  = ifelse(grepl("DMSO",   Sample), "DMSO",   "dTag"),
      .time = stringr::str_extract(Sample, "\\d+min"),
      condition = if (keep_timepoint) paste(.tf, .trt, .time, sep = "_")
                  else                paste(.tf, .trt, sep = "_")
    ) %>%
    group_by(condition, pos) %>%
    summarise(normalized_occupancy = mean(normalized_occupancy, na.rm = TRUE),
              .groups = "drop") %>%
    dplyr::rename(Sample = condition)
}
```

```{r fn-pooled-and-midpoint}
#' Pool a set of clusters into ONE occupancy curve per sample.
#' denominator = "cluster" : normalise over reads in those clusters only
#' denominator = "all"     : normalise over every read in the region
compute_pooled_cluster_metaplot <- function(plot_df, cluster_assignments, plot_range,
                                            bundles, clusters,
                                            min_read_cov = MIN_READ_COV,
                                            denominator = c("cluster", "all")) {

  denominator  <- match.arg(denominator)
  region_start <- start(plot_range)
  region_end   <- end(plot_range)
  pos_vec      <- region_start:region_end

  df <- plot_df %>%
    dplyr::filter(as.integer(as.character(cluster)) %in% as.integer(clusters))

  reads_df_all <- reads_with_clusters(bundles, cluster_assignments)
  reads_df     <- if (denominator == "cluster") {
    reads_df_all %>%
      dplyr::filter(as.integer(as.character(cluster)) %in% as.integer(clusters))
  } else reads_df_all

  purrr::map_dfr(unique(df$Sample), function(s_name) {
    fp <- .coverage_vector(df$start[df$Sample == s_name],
                           df$end[df$Sample == s_name],
                           region_start, region_end)
    rd <- .coverage_vector(reads_df$read_start[reads_df$Sample == s_name],
                           reads_df$read_end[reads_df$Sample == s_name],
                           region_start, region_end)
    tibble(Sample = s_name, pos = pos_vec,
           fp_coverage = fp, read_coverage = rd,
           normalized_occupancy = ifelse(rd >= min_read_cov, fp / rd * 100, NA_real_))
  })
}

#' Nucleosome MIDPOINT density per bin (positioning/phasing rather than
#' occupancy), smoothed with a rolling mean.
compute_midpoint_metaplot <- function(plot_df, cluster_assignments, plot_range,
                                      bundles, clusters = NULL,
                                      bin_width = 10L, smooth_k = 5L,
                                      min_read_cov = MIN_READ_COV) {

  region_start <- start(plot_range)
  region_end   <- end(plot_range)

  df <- plot_df
  if (!is.null(clusters)) {
    df <- df %>% dplyr::filter(as.integer(as.character(cluster)) %in% as.integer(clusters))
  }
  df <- df %>%
    mutate(midpoint = (start + end) / 2) %>%
    dplyr::filter(midpoint >= region_start, midpoint <= region_end)

  reads_df <- reads_with_clusters(bundles, cluster_assignments)
  if (!is.null(clusters)) {
    reads_df <- reads_df %>%
      dplyr::filter(as.integer(as.character(cluster)) %in% as.integer(clusters))
  }

  bins <- seq(region_start, region_end - bin_width, by = bin_width)

  purrr::map_dfr(unique(df$Sample), function(s_name) {

    r <- reads_df %>% dplyr::filter(Sample == s_name)

    bin_counts <- df %>%
      dplyr::filter(Sample == s_name) %>%
      mutate(bin = floor((midpoint - region_start) / bin_width) * bin_width + region_start) %>%
      dplyr::count(bin, name = "n_midpoints")

    # denominator: reads spanning each bin CENTRE
    bin_cov <- purrr::map_int(bins, function(b) {
      centre <- b + bin_width / 2
      sum(r$read_start <= centre & r$read_end >= centre)
    })

    tibble(bin = bins, read_coverage = bin_cov) %>%
      left_join(bin_counts, by = "bin") %>%
      mutate(
        n_midpoints          = tidyr::replace_na(n_midpoints, 0L),
        normalized_occupancy = ifelse(read_coverage >= min_read_cov,
                                      n_midpoints / read_coverage * 100, NA_real_)
      ) %>%
      arrange(bin) %>%
      mutate(normalized_occupancy = zoo::rollmean(normalized_occupancy,
                                                  k = smooth_k, fill = NA,
                                                  align = "center"),
             Sample = s_name)
  })
}
```

## 5.9 ATAC-seq signal tracks

```{r fn-atac}
#' Tidy per-bp bigWig signal restricted to plot_range, one row per (Sample, bp)
compute_atac_signal <- function(atac_files, plot_range) {

  missing <- atac_files[!file.exists(atac_files)]
  if (length(missing) > 0) {
    stop("bigWig file(s) not found:\n", paste(missing, collapse = "\n"))
  }

  region_start <- start(plot_range)
  region_end   <- end(plot_range)
  pos_vec      <- region_start:region_end

  imap_dfr(atac_files, function(path, s_name) {
    gr  <- rtracklayer::import.bw(path, which = plot_range)
    sig <- numeric(length(pos_vec))          # gaps stay 0, correct for bigWig
    if (length(gr) > 0) {
      s <- pmax(start(gr), region_start) - region_start + 1L
      e <- pmin(end(gr),   region_end)   - region_start + 1L
      for (i in which(e >= s)) sig[s[i]:e[i]] <- mcols(gr)$score[i]
    }
    tibble(Sample = s_name, pos = pos_vec, signal = sig)
  })
}

#' Standalone ATAC track, same visual conventions as the metaplots
plot_atac_signal <- function(atac_df, plot_range,
                             samples = NULL, facet = FALSE,
                             binding_sites = NULL,
                             vlines = NULL, vline_colors = NULL,
                             sample_colors = NULL,
                             title = "ATAC-seq signal", subtitle = NULL,
                             xlab = NULL) {

  region_start <- start(plot_range)
  region_end   <- end(plot_range)
  chr          <- as.character(seqnames(plot_range))
  xlab         <- xlab %||% paste0("Genomic coordinate (", chr, ")")

  df <- atac_df %>% dplyr::filter(pos >= region_start, pos <= region_end)
  if (!is.null(samples)) {
    df <- df %>%
      dplyr::filter(Sample %in% samples) %>%
      mutate(Sample = factor(Sample, levels = samples))
  }
  if (!is.null(vlines) && is.null(vline_colors)) {
    vline_colors <- rep("black", length(vlines))
  }

  p <- ggplot(df, aes(x = pos, y = signal))

  if (!is.null(binding_sites)) {
    p <- p +
      geom_rect(data = binding_sites,
                aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = col),
                alpha = 0.15, inherit.aes = FALSE) +
      scale_fill_identity(guide = "none") +
      geom_text(data = binding_sites,
                aes(x = (xmin + xmax) / 2, y = Inf, label = TF, colour = col),
                vjust = 1.5, size = 3.5, fontface = "bold",
                inherit.aes = FALSE, show.legend = FALSE) +
      scale_colour_identity(guide = "none") +
      ggnewscale::new_scale_fill() +
      ggnewscale::new_scale_colour()
  }

  p <- p +
    geom_area(aes(fill = Sample), position = "identity", alpha = 0.1) +
    geom_line(aes(colour = Sample), linewidth = 0.8) +
    coord_cartesian(xlim = c(region_start, region_end)) +
    theme_classic() +
    labs(title = title, subtitle = subtitle, x = xlab, y = "ATAC signal") +
    theme(legend.position  = "bottom",
          strip.background = element_blank(),
          strip.text       = element_text(face = "bold"))

  if (facet) p <- p + facet_wrap(~ Sample, ncol = 1, scales = "free_y",
                                 strip.position = "right")

  p <- p + if (!is.null(sample_colors)) {
    list(scale_colour_manual(values = sample_colors),
         scale_fill_manual(values   = sample_colors))
  } else {
    list(scale_colour_brewer(palette = "Set1"),
         scale_fill_brewer(palette   = "Set1"))
  }

  if (!is.null(vlines)) {
    for (i in seq_along(vlines)) {
      p <- p + geom_vline(xintercept = vlines[i], linetype = "dashed",
                          linewidth = 0.4, alpha = 0.5, colour = vline_colors[i])
    }
  }
  p
}
```

## 5.10 Linked dense heatmap (all samples in one coordinate space)

<!-- ==========================================================================
     Like plot_fiberseq_dense(), but every sample gets a horizontal "lane"
     inside ONE shared ggplot coordinate space rather than a separate patchwork
     panel. That is what makes the DMSO -> dTag cluster ribbons possible: a
     polygon cannot span two independent patchwork viewports.

     CLUSTER ROW HEIGHTS are proportional to that cluster's share of that
     sample's reads, so every lane fills the same total height regardless of
     depth -- exactly what facet_grid(scales = "free_y", space = "free_y")
     shows in plot_fiberseq_dense(). Because the shares differ between two
     linked samples, the connector is a TRAPEZOID, and its taper is the visual
     statement of how the cluster's share shifts between conditions.

     The genomic x axis is suppressed: each lane has its own local coordinate
     block, so one shared axis would be meaningless. A small "start-end" label
     is drawn under each lane instead. binding_sites and vlines are redrawn
     once per lane.
     ========================================================================== -->

```{r fn-linked}
plot_fiberseq_dense_linked <- function(plot_df, plot_range, sample_order,
                                       link_pairs = NULL,
                                       fill_color = TF_FILL,
                                       binding_sites = NULL,
                                       binding_site_alpha = 1,
                                       binding_site_linewidth = 0.6,
                                       n_label_tiers = 3,
                                       label_tier_gap = NULL,
                                       label_gap = NULL,
                                       vlines = NULL, vline_colors = NULL,
                                       link_fill = "grey50", link_alpha = 0.18,
                                       link_linetype = "22", link_color = "grey30",
                                       lane_gap_frac = 0.15,
                                       lane_height = 100,
                                       cluster_gap = 0,
                                       cluster_box = FALSE,
                                       cluster_box_color = "black",
                                       cluster_box_linewidth = 0.4,
                                       show_cluster_labels = TRUE,
                                       atac_df = NULL,          # compute_atac_signal() output
                                       atac_height = NULL,      # NULL = 10% of lane_height
                                       atac_gap = NULL,         # NULL =  5% of lane_height
                                       atac_ylim = NULL,        # NULL = shared max per scale group
                                       atac_scale_groups = NULL,# defaults to link_pairs
                                       atac_color = NULL,       # NULL = reuse fill_color
                                       atac_fill_alpha = 0.25,
                                       sample_label_gap = NULL,
                                       sample_display_labels = NULL,
                                       label_groups = NULL,
                                       group_label_gap = NULL,
                                       # decoration switches, for stacking several
                                       # panels that share one set of headers
                                       show_binding_labels = TRUE,
                                       show_lane_underline = TRUE,
                                       show_range_label = TRUE,
                                       margin_top = NULL,
                                       margin_bottom = 5,
                                       title = "Clustered Fiber-seq: linked cluster view",
                                       raster = TRUE, raster_dpi = 300,
                                       xlab = NULL) {

  chr          <- as.character(seqnames(plot_range))
  xlab         <- xlab %||% paste0("Genomic coordinate (", chr, ")")
  region_start <- start(plot_range)
  region_end   <- end(plot_range)
  region_width <- region_end - region_start + 1

  plot_df <- plot_df %>%
    dplyr::filter(Sample %in% sample_order) %>%
    mutate(Sample = factor(Sample, levels = sample_order))

  n_s       <- length(sample_order)
  gap_width <- region_width * lane_gap_frac
  lane_x0   <- setNames((seq_len(n_s) - 1) * (region_width + gap_width), sample_order)

  # ---- vertical layout, all relative to lane_height so the ATAC track, the
  #      motif label tiers and the sample name never collide -----------------
  has_atac         <- !is.null(atac_df)
  atac_gap         <- atac_gap         %||% (lane_height * 0.05)
  atac_height      <- atac_height      %||% (lane_height * 0.10)
  label_gap        <- label_gap        %||% (lane_height * 0.05)
  label_tier_gap   <- label_tier_gap   %||% (lane_height * 0.08)
  sample_label_gap <- sample_label_gap %||% (lane_height * 0.05)
  group_label_gap  <- group_label_gap  %||% (lane_height * 0.12)

  atac_top       <- if (has_atac) lane_height + atac_gap + atac_height else lane_height
  labels_base    <- atac_top + label_gap                                # tier-0 anchor
  # With show_binding_labels = FALSE there is no text above the lanes, so the
  # label tiers must not reserve their height -- otherwise a stacked panel
  # carries ~30 units of empty headroom it never uses.
  show_bs_labels <- !is.null(binding_sites) && show_binding_labels
  labels_top     <- if (show_bs_labels) {
    labels_base + (n_label_tiers - 1) * label_tier_gap                  # top tier
  } else atac_top
  sample_label_y <- labels_top + sample_label_gap
  group_label_y  <- sample_label_y + group_label_gap

  # ---- per (Sample, cluster) band geometry: share of that sample's reads ----
  band_stats <- plot_df %>%
    group_by(Sample, cluster) %>%
    summarise(n_reads = max(local_index), .groups = "drop") %>%
    group_by(Sample) %>%
    arrange(Sample, cluster) %>%
    mutate(
      idx          = row_number(),
      n_clust      = n(),
      avail_height = lane_height - cluster_gap * (n_clust - 1),
      height       = avail_height * n_reads / sum(n_reads),
      cum_before   = dplyr::lag(cumsum(height), default = 0) + (idx - 1) * cluster_gap,
      top          = lane_height - cum_before,
      bottom       = top - height
    ) %>%
    ungroup()

  # ---- one rectangle per footprint ------------------------------------------
  plot_df2 <- plot_df %>%
    left_join(band_stats %>% select(Sample, cluster, n_reads, top, bottom),
              by = c("Sample", "cluster")) %>%
    mutate(
      band_h   = top - bottom,
      y_center = bottom + (local_index - 0.5) / n_reads * band_h,
      y_half   = 0.45 / n_reads * band_h,
      x0       = lane_x0[as.character(Sample)],
      xmin     = x0 + pmax(start, region_start) - region_start,
      xmax     = x0 + pmin(end,   region_end)   - region_start,
      ymin     = y_center - y_half,
      ymax     = y_center + y_half,
      fill_hex = vapply(as.character(Sample), .resolve_fill,
                        character(1), fill_color = fill_color)
    )

  # ---- link geometry: shaded trapezoid + dashed top/bottom edges -------------
  link_polys <- NULL
  link_edges <- NULL
  if (!is.null(link_pairs) && length(link_pairs) > 0) {

    bad <- base::setdiff(unlist(link_pairs), sample_order)
    if (length(bad) > 0) {
      stop("link_pairs reference sample(s) not in sample_order: ",
           paste(bad, collapse = ", "))
    }

    link_geo <- purrr::map_dfr(seq_along(link_pairs), function(i) {
      sA <- link_pairs[[i]][1]; sB <- link_pairs[[i]][2]
      bA <- band_stats %>% dplyr::filter(Sample == sA)
      bB <- band_stats %>% dplyr::filter(Sample == sB)
      common <- base::intersect(as.character(bA$cluster), as.character(bB$cluster))
      if (length(common) == 0) {
        warning(sprintf("No shared clusters between '%s' and '%s'; link skipped.", sA, sB))
      }
      purrr::map_dfr(common, function(cl) {
        rA <- bA[as.character(bA$cluster) == cl, ]
        rB <- bB[as.character(bB$cluster) == cl, ]
        tibble(link_id = paste(sA, sB, cl, sep = "__"), cluster = cl,
               xA = lane_x0[sA] + region_width, xB = lane_x0[sB],
               topA = rA$top, botA = rA$bottom, topB = rB$top, botB = rB$bottom)
      })
    })

    link_polys <- purrr::pmap_dfr(link_geo, function(link_id, cluster, xA, xB,
                                                     topA, botA, topB, botB) {
      tibble(link_id = link_id, x = c(xA, xB, xB, xA), y = c(botA, botB, topB, topA))
    })
    link_edges <- bind_rows(
      link_geo %>% transmute(edge_id = paste0(link_id, "_top"),
                             x1 = xA, y1 = topA, x2 = xB, y2 = topB),
      link_geo %>% transmute(edge_id = paste0(link_id, "_bottom"),
                             x1 = xA, y1 = botA, x2 = xB, y2 = botB)
    )
  }

  p <- ggplot()

  # ---- ATAC track, drawn first so everything layers on top -------------------
  if (has_atac) {

    atac_plot_df <- atac_df %>% dplyr::filter(Sample %in% sample_order)

    # samples in the same scale group share a y maximum, so a DMSO/dTag pair is
    # visually comparable. Defaults to the link_pairs grouping.
    scale_groups <- atac_scale_groups %||% link_pairs
    group_lookup <- character(0)
    if (!is.null(scale_groups)) {
      for (i in seq_along(scale_groups)) group_lookup[scale_groups[[i]]] <- paste0("grp", i)
    }
    ungrouped <- base::setdiff(sample_order, names(group_lookup))
    group_lookup[ungrouped] <- ungrouped
    atac_plot_df$scale_group <- group_lookup[as.character(atac_plot_df$Sample)]

    if (is.null(atac_ylim)) {
      group_range <- atac_plot_df %>%
        group_by(scale_group) %>%
        summarise(y_lo = 0, y_hi = max(signal, na.rm = TRUE), .groups = "drop")
      atac_plot_df <- atac_plot_df %>% left_join(group_range, by = "scale_group")
    } else {
      atac_plot_df <- atac_plot_df %>% mutate(y_lo = atac_ylim[1], y_hi = atac_ylim[2])
    }

    atac_fill_lookup <- atac_color %||% fill_color
    baseline_y       <- lane_height + atac_gap

    atac_plot_df <- atac_plot_df %>%
      mutate(
        x0       = lane_x0[as.character(Sample)],
        x        = x0 + pos - region_start,
        sig_clip = pmin(pmax(signal, y_lo), y_hi),
        y        = baseline_y + (sig_clip - y_lo) / (y_hi - y_lo) * atac_height,
        col_hex  = vapply(as.character(Sample), .resolve_fill,
                          character(1), fill_color = atac_fill_lookup)
      )

    baseline_df   <- tibble(x    = lane_x0[sample_order],
                            xend = lane_x0[sample_order] + region_width,
                            y    = baseline_y)
    atac_label_df <- tibble(x = lane_x0[sample_order] + region_width * 0.02,
                            y = atac_top, label = "ATAC")

    atac_layer <- geom_ribbon(data = atac_plot_df,
                              aes(x = x, ymin = baseline_y, ymax = y,
                                  group = Sample, fill = col_hex),
                              alpha = atac_fill_alpha, inherit.aes = FALSE)
    if (raster) atac_layer <- ggrastr::rasterise(atac_layer, dpi = raster_dpi,
                                                 dev = "ragg_png")

    p <- p +
      geom_segment(data = baseline_df, aes(x = x, xend = xend, y = y, yend = y),
                   colour = "grey80", linewidth = 0.3, inherit.aes = FALSE) +
      atac_layer +
      scale_fill_identity(guide = "none") +
      geom_line(data = atac_plot_df,
                aes(x = x, y = y, group = Sample, colour = col_hex),
                linewidth = 0.4, inherit.aes = FALSE) +
      scale_colour_identity(guide = "none") +
      geom_text(data = atac_label_df, aes(x = x, y = y, label = label),
                hjust = 0, vjust = 0, size = 2.3, colour = "grey40",
                fontface = "italic", inherit.aes = FALSE)
  }

  # ---- motif guides, with labels staggered across n_label_tiers ---------------
  # show_binding_labels = FALSE keeps the dashed guides but stops them at the
  # top of the lanes and draws no text, so a stacked panel can carry the
  # vertical alignment cues without repeating the motif names.
  if (!is.null(binding_sites)) {
    bs_all <- purrr::map_dfr(sample_order, function(s) {
      binding_sites %>%
        arrange(xmin) %>%
        mutate(tier     = (row_number() - 1) %% n_label_tiers,
               x_center = (xmin + xmax) / 2,
               x        = lane_x0[s] + x_center - region_start,
               label_y  = if (show_bs_labels) labels_base + tier * label_tier_gap
                          else lane_height)
    })
    p <- p +
      geom_segment(data = bs_all,
                   aes(x = x, xend = x, y = 0, yend = label_y, colour = col),
                   linetype = "dashed", linewidth = binding_site_linewidth,
                   alpha = binding_site_alpha, inherit.aes = FALSE)
    if (show_bs_labels) {
      p <- p + geom_text(data = bs_all,
                         aes(x = x, y = label_y, label = TF, colour = col),
                         vjust = -0.1, size = 2.5, inherit.aes = FALSE,
                         show.legend = FALSE)
    }
    p <- p + scale_colour_identity(guide = "none")
  }

  if (!is.null(link_polys)) {
    p <- p +
      geom_polygon(data = link_polys, aes(x = x, y = y, group = link_id),
                   fill = link_fill, alpha = link_alpha, inherit.aes = FALSE) +
      geom_segment(data = link_edges, aes(x = x1, y = y1, xend = x2, yend = y2),
                   linetype = link_linetype, colour = link_color,
                   linewidth = 0.4, inherit.aes = FALSE)
  }

  rect_layer <- geom_rect(data = plot_df2,
                          aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
                          fill = plot_df2$fill_hex, colour = NA)
  if (raster) rect_layer <- ggrastr::rasterise(rect_layer, dpi = raster_dpi,
                                               dev = "ragg_png")
  p <- p + rect_layer

  if (cluster_box) {
    box_df <- purrr::map_dfr(sample_order, function(s) {
      band_stats %>% dplyr::filter(Sample == s) %>%
        mutate(xmin = lane_x0[s], xmax = lane_x0[s] + region_width)
    })
    p <- p + geom_rect(data = box_df,
                       aes(xmin = xmin, xmax = xmax, ymin = bottom, ymax = top),
                       fill = NA, colour = cluster_box_color,
                       linewidth = cluster_box_linewidth, inherit.aes = FALSE)
  }

  if (!is.null(vlines)) {
    if (is.null(vline_colors)) vline_colors <- rep("black", length(vlines))
    vl_df <- purrr::map_dfr(seq_along(vlines), function(i) {
      purrr::map_dfr(sample_order, function(s) {
        tibble(x = lane_x0[s] + vlines[i] - region_start, colour = vline_colors[i])
      })
    })
    p <- p +
      geom_segment(data = vl_df, aes(x = x, xend = x, y = 0, yend = lane_height,
                                     colour = colour),
                   linetype = "dashed", linewidth = 0.4, alpha = 0.5,
                   inherit.aes = FALSE) +
      scale_colour_identity(guide = "none")
  }

  # ---- per-lane label + underline (e.g. "DMSO" / "dTag") ---------------------
  lane_label_text <- if (!is.null(sample_display_labels)) {
    sample_display_labels[sample_order]
  } else sample_order

  # Drawing an EMPTY label still expands the y range to sample_label_y, so a
  # panel with blanked names kept its headroom. Skip the layers entirely when
  # there is nothing to show.
  if (show_lane_underline) {
    lane_underlines <- tibble(x    = lane_x0[sample_order],
                              xend = lane_x0[sample_order] + region_width,
                              y    = sample_label_y)
    p <- p + geom_segment(data = lane_underlines,
                          aes(x = x, xend = xend, y = y, yend = y),
                          colour = "black", linewidth = 0.4, inherit.aes = FALSE)
  }
  if (any(nzchar(lane_label_text))) {
    lane_labels <- tibble(x = lane_x0[sample_order] + region_width / 2,
                          y = sample_label_y, label = lane_label_text)
    p <- p + geom_text(data = lane_labels, aes(x = x, y = y, label = label),
                       vjust = -0.3, size = 3.5, inherit.aes = FALSE)
  }

  # ---- group label spanning several lanes ------------------------------------
  if (!is.null(label_groups)) {
    group_df <- imap_dfr(label_groups, function(members, glabel) {
      bad <- base::setdiff(members, sample_order)
      if (length(bad) > 0) {
        warning("label_groups entry '", glabel, "' references sample(s) not in ",
                "sample_order: ", paste(bad, collapse = ", "), " -- skipping.")
        return(NULL)
      }
      tibble(label = glabel,
             x0 = min(lane_x0[members]),
             x1 = max(lane_x0[members]) + region_width)
    })
    p <- p +
      geom_segment(data = group_df,
                   aes(x = x0, xend = x1, y = group_label_y, yend = group_label_y),
                   colour = "black", linewidth = 0.4, inherit.aes = FALSE) +
      geom_text(data = group_df,
                aes(x = (x0 + x1) / 2, y = group_label_y, label = label),
                vjust = -0.3, size = 3.8, inherit.aes = FALSE)
  }

  # Repeated under every lane, so worth switching off on all but the bottom
  # panel of a stack.
  if (show_range_label) {
    range_labels <- tibble(x = lane_x0[sample_order] + region_width / 2, y = 0,
                           label = paste0(comma(region_start), "-", comma(region_end)))
    p <- p + geom_text(data = range_labels, aes(x = x, y = y, label = label),
                       vjust = 2.2, size = 2.3, colour = "grey40", inherit.aes = FALSE)
  }

  if (show_cluster_labels) {
    lbl <- band_stats %>%
      mutate(x = lane_x0[as.character(Sample)], y = (top + bottom) / 2)
    p <- p + geom_text(data = lbl, aes(x = x, y = y, label = cluster),
                       hjust = 1.3, size = 2.8, inherit.aes = FALSE)
  }

  p +
    coord_cartesian(clip = "off") +
    theme_classic() +
    theme(axis.text = element_blank(), axis.ticks = element_blank(),
          axis.line = element_blank(), panel.grid = element_blank(),
          plot.margin = margin(
            t = margin_top %||% (if (!is.null(label_groups)) 50
                                 else if (has_atac) 40 else 20),
            r = 5, b = margin_bottom, l = 20)) +
    labs(title = title, x = xlab, y = NULL)
}
```

## 5.11 MSP geometry and the open / closed summary

<!-- ==========================================================================
     For one enhancer, collapse the k-means clusters into "open" and "closed"
     groups and compute log2FC(dTag / DMSO) of the % of reads in each group,
     per replicate pair.

     TWO WAYS TO DEFINE THE GROUPS

       (A) MANUAL   pass open_clusters and closed_clusters as integer vectors.

       (B) AUTO     pass msp_files + motif_range. A cluster is "open" if, among
                    its MSPs overlapping the motif:
                        mean_msp_width >= min_msp_width      AND
                        width_before   >= min_width_before   AND
                        width_after    >= min_width_after    AND
                        n_msp          >= min_n_msp
                    Every other cluster becomes "closed" (the complement, so no
                    cluster is left unclassified).

     ON THE THRESHOLDS : for the same pool of overlapping
     MSPs these three widths satisfy
         mean_msp_width = width_before + width_after + motif_width
     so with a small motif and 300 / 150 / 150 the mean_msp_width test is
     already implied by the other two and adds nothing. It is kept because it
     is harmless and self-documenting, and because it would matter for a very
     asymmetric overhang. In the methods, describe the criterion as "at least
     150 bp of accessible MSP on each side of the motif".

     clusters with fewer than min_n_msp motif-overlapping
     MSPs are called "closed" by default rather than trusted on sparse data.
     The function messages which clusters this affected -- read those messages.

     The output always carries a `clusters` column recording which cluster IDs
     went into that row's group, so the result is self-documenting even though
     the classification is data-driven.
     ========================================================================== -->

```{r fn-msp}
#' Mean MSP geometry around a motif, per cluster
#'
#' @param motif_range  a GRanges, or c(start, end) plus an explicit `chrom`.
#'   MSP .bed files pool footprints from several PCR amplicons, so filtering on
#'   numeric coordinates alone can match the wrong locus -- chrom is mandatory.
compute_msp_motif_stats <- function(msp_files, cluster_assignments_perrep,
                                    motif_range, chrom = NULL) {

  if (methods::is(motif_range, "GRanges")) {
    chrom       <- as.character(seqnames(motif_range))
    motif_start <- start(motif_range)
    motif_end   <- end(motif_range)
  } else {
    if (is.null(chrom)) {
      stop("compute_msp_motif_stats: provide `chrom` (or pass motif_range as a ",
           "GRanges). MSP bed files pool multiple amplicons, so filtering on ",
           "coordinates alone risks matching the wrong locus.")
    }
    motif_start <- motif_range[1]
    motif_end   <- motif_range[2]
  }

  # FIX B1: this used to re-read the file with readLines() and attach chrom by
  #         ROW POSITION. read_fiberhmm_bed() drops malformed rows, so one bad
  #         line shifted every chromosome assignment after it -- silently
  #         corrupting the amplicon filter and therefore the open/closed call.
  #         read_fiberhmm_bed() now carries `chrom` through directly.
  msps <- imap_dfr(msp_files, function(path, s_name) {
    read_fiberhmm_bed(path)$footprints %>%
      transmute(unique_id = paste0(s_name, "_", read_name),
                chrom, msp_start = fp_start, msp_end = fp_end)
  })

  target_chrom <- chrom
  stats <- msps %>%
    dplyr::filter(chrom == target_chrom,
                  msp_end >= motif_start, msp_start <= motif_end) %>%
    inner_join(cluster_assignments_perrep %>% dplyr::select(unique_id, cluster),
               by = "unique_id") %>%
    mutate(msp_width    = msp_end - msp_start + 1L,
           width_before = motif_start - msp_start,   # accessible bp 5' of motif
           width_after  = msp_end - motif_end) %>%   # accessible bp 3' of motif
    group_by(cluster) %>%
    summarise(mean_msp_width = mean(msp_width),
              width_before   = mean(width_before),
              width_after    = mean(width_after),
              n_msp          = dplyr::n(),
              .groups = "drop") %>%
    mutate(cluster = as.integer(as.character(cluster)))

  # keep clusters with zero overlapping MSPs as explicit NA rows
  all_clusters <- sort(unique(as.integer(as.character(cluster_assignments_perrep$cluster))))
  tibble(cluster = all_clusters) %>% left_join(stats, by = "cluster")
}
```

```{r fn-openclosed}
compute_openclosed_log2fc <- function(cluster_assignments_perrep,
                                      comparisons,
                                      enhancer_name,
                                      open_clusters    = NULL,
                                      closed_clusters  = NULL,
                                      msp_files        = NULL,
                                      motif_range      = NULL,
                                      chrom            = NULL,
                                      min_msp_width    = 300,
                                      min_width_before = 150,
                                      min_width_after  = 150,
                                      min_n_msp        = 20) {

  manual_mode <- !is.null(open_clusters) || !is.null(closed_clusters)
  auto_mode   <- !is.null(msp_files)     || !is.null(motif_range)

  if (manual_mode && auto_mode) {
    stop("Provide EITHER open_clusters/closed_clusters (manual) OR ",
         "msp_files/motif_range (auto), not both.")
  }
  if (!manual_mode && !auto_mode) {
    stop("Provide either open_clusters & closed_clusters, or msp_files & motif_range.")
  }
  if (manual_mode && (is.null(open_clusters) || is.null(closed_clusters))) {
    stop("Manual mode requires BOTH open_clusters and closed_clusters.")
  }
  if (auto_mode && (is.null(msp_files) || is.null(motif_range))) {
    stop("Auto mode requires BOTH msp_files and motif_range.")
  }

  if (auto_mode) {
    msp_stats    <- compute_msp_motif_stats(msp_files, cluster_assignments_perrep,
                                            motif_range, chrom = chrom)
    all_clusters <- sort(unique(as.integer(as.character(cluster_assignments_perrep$cluster))))

    is_open <- !is.na(msp_stats$mean_msp_width) & msp_stats$mean_msp_width >= min_msp_width &
               !is.na(msp_stats$width_before)   & msp_stats$width_before   >= min_width_before &
               !is.na(msp_stats$width_after)    & msp_stats$width_after    >= min_width_after &
               !is.na(msp_stats$n_msp)          & msp_stats$n_msp          >= min_n_msp

    open_clusters   <- as.integer(as.character(msp_stats$cluster[is_open]))
    closed_clusters <- base::setdiff(all_clusters, open_clusters)

    message(sprintf("[%s] auto-classified from MSP geometry: open = {%s}, closed = {%s}",
                    enhancer_name,
                    paste(sort(open_clusters),   collapse = ","),
                    paste(sort(closed_clusters), collapse = ",")))

    low_n <- msp_stats$cluster[!is.na(msp_stats$n_msp) & msp_stats$n_msp < min_n_msp]
    if (length(low_n) > 0) {
      message("  note: cluster(s) ", paste(sort(low_n), collapse = ","),
              " had < ", min_n_msp, " motif-overlapping MSPs -- called 'closed' ",
              "by default rather than trusted on sparse data (see Q6).")
    }
  }

  ca <- cluster_assignments_perrep %>%
    mutate(group = case_when(
      as.integer(as.character(cluster)) %in% open_clusters   ~ "open",
      as.integer(as.character(cluster)) %in% closed_clusters ~ "closed",
      TRUE ~ NA_character_
    )) %>%
    dplyr::filter(!is.na(group))

  # denominator = ALL reads in that sample. In auto mode open + closed is the
  # complete set, so this is the same thing. In manual mode with unassigned
  # clusters it is NOT -- the percentages will not sum to 100.
  total_per_sample <- cluster_assignments_perrep %>%
    dplyr::count(Sample, name = "total_reads")

  pct_wide <- ca %>%
    dplyr::count(Sample, group, name = "n_reads") %>%
    left_join(total_per_sample, by = "Sample") %>%
    mutate(percent = 100 * n_reads / total_reads) %>%
    select(Sample, group, percent) %>%
    pivot_wider(names_from = Sample, values_from = percent, values_fill = 0)

  result <- imap_dfr(comparisons, function(pair, label) {

    num <- pair[["numerator"]]
    den <- pair[["denominator"]]

    missing <- base::setdiff(c(num, den), names(pct_wide))
    if (length(missing) > 0) {
      stop("compute_openclosed_log2fc: sample(s) not found: ",
           paste(missing, collapse = ", "))
    }

    # Q4: the degraded TF is inferred from the numerator's sample name.
    # Warn rather than silently defaulting everything unmatched to SNAI.
    tf_lab <- if (grepl("TWIST1", num)) "TWIST1" else if (grepl("SNAI", num)) "SNAI" else {
      warning("Numerator '", num, "' matches neither TWIST1 nor SNAI; labelling SNAI.")
      "SNAI"
    }

    tibble(
      enhancer       = enhancer_name,
      Replicate_Pair = label,
      TF             = tf_lab,
      group          = pct_wide$group,
      Log2Ratio      = log2(pct_wide[[num]] / pct_wide[[den]])
    )
  })

  cluster_str      <- c(open   = paste(sort(open_clusters),   collapse = ","),
                        closed = paste(sort(closed_clusters), collapse = ","))
  result$clusters  <- cluster_str[result$group]
  result
}
```

## 5.12 Convenience wrappers used by the per-enhancer sections

```{r fn-wrappers}
#' Build bundles, merge replicates, cluster, and split back out -- cached.
#'
#' Returns list(bundles, bundles_merged, ca_merged, ca_perrep).
#' `cache_name` becomes <ROBJ_DIR>/<cache_name>.rds
run_clustering <- function(key, samples, cluster_range, k, cache_name,
                           rebuild = REBUILD_CLUSTERS) {

  out <- cache_rds(cache_name, {
    files   <- enh_files(key, samples, "footprint")
    bundles <- build_all_bundles(files, cluster_range)

    merged     <- merge_replicate_bundles(bundles, merge_groups_for(samples))
    ca_merged  <- cluster_reads(merged, k = k, seed = KMEANS_SEED)
    ca_perrep  <- split_cluster_assignments(ca_merged)

    list(bundles        = bundles,
         bundles_merged = merged,
         ca_merged      = ca_merged,
         ca_perrep      = ca_perrep,
         kmeans_info    = attr(ca_merged, "kmeans_info"),
         meta           = list(key = key, k = k, seed = KMEANS_SEED,
                               minset = ENH[[key]]$minset,
                               cluster_range = as.character(cluster_range),
                               samples = samples,
                               R_version = R.version.string,
                               run_date  = Sys.Date()))
  }, rebuild = rebuild)

  # Print the fit quality every time, cached or not. Two things to watch:
  # a cluster with very few reads is not a real class, and a low prop_between
  # means the partition is not carrying much of the variance.
  ki <- out$kmeans_info
  if (!is.null(ki)) {
    message(sprintf(
      "  %s (%s): %s reads x %d bp, k = %d\n    init: %s\n    tot.withinss = %s | between/total = %.3f | %d iterations\n    cluster sizes: %s",
      enh_label(key), cache_name, format(ki$n_reads, big.mark = ","),
      ki$n_positions, ki$k, ki$init_method,
      format(round(ki$tot_withinss), big.mark = ","),
      ki$prop_between, ki$iterations,
      paste(ki$cluster_sizes, collapse = ", ")))
    tiny <- which(ki$cluster_sizes < 0.01 * ki$n_reads)
    if (length(tiny) > 0) {
      warning(enh_label(key), ": cluster(s) ", paste(tiny, collapse = ", "),
              " hold under 1% of reads. Consider a smaller k.")
    }
  }
  out
}

#' The linked DMSO/dTag heatmap, with LINKED_STYLE applied.
#' `pairs` is a named list: "group label" = c(DMSO sample, dTag sample).
linked_heatmap <- function(plot_df_merged, plot_range, pairs, atac_df,
                           binding_sites, title, fill_color = TF_FILL) {

  sample_order <- unlist(pairs, use.names = FALSE)

  do.call(plot_fiberseq_dense_linked, c(
    list(
      plot_df               = plot_df_merged,
      plot_range            = plot_range,
      sample_order          = sample_order,
      atac_df               = atac_df,
      link_pairs            = unname(pairs),
      sample_display_labels = setNames(
        ifelse(grepl("DMSO", sample_order), "DMSO", "dTag"), sample_order),
      label_groups          = pairs,
      fill_color            = fill_color,
      binding_sites         = binding_sites,
      title                 = title
    ),
    LINKED_STYLE
  ))
}

#' TF metaplot for one condition pair, replicate-averaged
tf_panel <- function(meta_avg, plot_range, tf, binding_sites, title) {
  samples <- paste0(tf, c("_DMSO", "_dTag"))
  plot_tf_metaplot_overall(
    meta_avg, plot_range,
    samples       = samples,
    sample_colors = META_COLORS[samples],
    binding_sites = binding_sites,
    subtitle      = "",
    title         = title
  )
}
```

## 5.13 Cluster stability check (opt-in)

<!-- ==========================================================================
     WHY THIS EXISTS

     Two runs of the previous pipeline differing ONLY in the order the bundles
     were stacked (same reads, same k, same seed) disagreed on 11.4% of reads
     at Enh 1: one cluster dissolved entirely and another was assembled from
     thirds of three others. nstart = 25 was not finding a stable optimum.

     The canonical row order in cluster_reads() makes a run reproducible, but
     reproducible is not the same as stable -- it only guarantees you get the
     SAME answer every time. This chunk asks the harder question: does the
     partition survive changing the seed?

     HOW TO READ IT
       ARI >= 0.9   the partition is a property of the data; k is supportable
       ARI 0.7-0.9  broadly stable, but expect some clusters to shuffle
       ARI < 0.7    the partition is substantially an artefact of the seed.
                    Do not rest a claim on individual cluster identities at
                    this k -- either lower k, or report only the pooled
                    open/closed split, which is far more robust than the
                    individual clusters because it collapses exactly the
                    boundaries that are moving.

     Clusters a fixed subsample so every seed sees identical rows. This is a
     diagnostic, not the analysis. Set eval = TRUE to run it; it is slow.
     ========================================================================== -->

```{r cluster-stability, eval = FALSE}
#' Recluster the same reads under several seeds and report pairwise agreement
cluster_stability <- function(bundles, k, seeds = c(1, 42, 123, 2024, 7),
                              subsample = 20000L) {

  mat <- do.call(rbind, map(bundles, bundle_to_matrix))
  mat <- mat[order(rownames(mat), method = "radix"), , drop = FALSE]

  if (!is.null(subsample) && nrow(mat) > subsample) {
    set.seed(1L)                                   # same rows for every seed
    mat <- mat[sort(sample.int(nrow(mat), subsample)), , drop = FALSE]
  }

  fits <- purrr::map(seeds, function(s) {
    set.seed(s)
    kmeans(mat, centers = k, nstart = KMEANS_NSTART, iter.max = KMEANS_ITER_MAX)
  })
  names(fits) <- paste0("seed", seeds)

  pairs <- utils::combn(seq_along(seeds), 2)
  ari   <- purrr::map_dfr(seq_len(ncol(pairs)), function(i) {
    a <- pairs[1, i]; b <- pairs[2, i]
    tibble(seed_a = seeds[a], seed_b = seeds[b],
           ARI    = adjusted_rand(fits[[a]]$cluster, fits[[b]]$cluster))
  })

  wss <- purrr::map_dbl(fits, "tot.withinss")
  list(
    ari      = ari,
    withinss = wss,
    summary  = tibble(k = k, n = nrow(mat),
                      mean_ARI = mean(ari$ARI),
                      min_ARI  = min(ari$ARI),
                      withinss_cv = sd(wss) / mean(wss))
  )
}

# Enh 1 at the k actually used
stab <- cluster_stability(enh1$bundles_merged, k = ENH$enh4$k)
print(stab$summary)
print(stab$ari)

# Sweep k, to see whether a coarser partition is better determined
sweep_k <- purrr::map_dfr(c(4, 6, 8, 10, 12), function(kk) {
  cluster_stability(enh1$bundles_merged, k = kk, seeds = c(1, 42, 123))$summary
})
print(sweep_k)
```

# 6. Per-enhancer analyses

<!-- ==========================================================================
     Each section follows the same shape:
       (a) cluster        run_clustering() -> bundles / merged / assignments
       (b) heatmap        linked_heatmap() + make_log2fc_panel()
       (c) open/closed    compute_openclosed_log2fc(), assigned to a NAMED
                          object (never appended to a global -- see C4)
       (d) TF metaplots   build_tf_metaplot_data() -> compute_tf_metaplot()

     Every title and filename comes from enh_label() / fig_name(), so the paper
     names in section 3 are the single source of truth.
     ========================================================================== -->

## 6.1 Enh 1 (internal: enh4, chr20)

<!-- Clustered jointly across the TWIST1 and SNAI motifs.
     Q1: this is the ONE enhancer analysed on the `min1` fiberHMM calls. -->

```{r enh1-cluster}
key <- "enh4"

comparisons_5min <- list(
  "TWIST1 Rep 1"  = c(numerator = "TWIST1_rep1_dTag5min", denominator = "TWIST1_rep1_DMSO5min"),
  "TWIST1 Rep 2"  = c(numerator = "TWIST1_rep2_dTag5min", denominator = "TWIST1_rep2_DMSO5min"),
  "SNAI1/2 Rep 1" = c(numerator = "SNAI_rep1_dTag5min",   denominator = "SNAI_rep1_DMSO5min"),
  "SNAI1/2 Rep 2" = c(numerator = "SNAI_rep2_dTag5min",   denominator = "SNAI_rep2_DMSO5min")
)

enh1 <- run_clustering(
  key           = key,
  samples       = BOTH_SAMPLES,
  cluster_range = enh_gr(key, "cluster"),
  k             = ENH[[key]]$k,                 # 12
  cache_name    = "enh4_k12"
)
```

```{r enh1-heatmap, fig.width = 6, fig.height = 6}
plot_range <- enh_gr(key, "plot")

pairs_5min <- list(
  "SNAI1/2 depletion" = c("SNAI_DMSO_5min",   "SNAI_dTag_5min"),
  "TWIST1 depletion"  = c("TWIST1_DMSO_5min", "TWIST1_dTag_5min")
)
sample_order <- unlist(pairs_5min, use.names = FALSE)

atac_df        <- compute_atac_signal(atac_for(sample_order), plot_range)
plot_df_merged <- prepare_plot_data(enh1$bundles_merged, enh1$ca_merged, plot_range)

# Q3: Enh 1 alone used 110 rather than the default 30, with no stated reason.
# Both live in the config chunk now so the asymmetry is visible.
log2fc_df <- compute_log2fc_df(enh1$ca_perrep, comparisons_5min,
                               min_reads = MIN_READS_FC_ENH1)

p_heat <- linked_heatmap(plot_df_merged, plot_range,
                         pairs         = pairs_5min,
                         atac_df       = atac_df,
                         binding_sites = enh_motifs(key, "motifs"),
                         title         = enh_label(key))
p_heat
ggsave(fig_name(key, "heatmap_linked"), p_heat,
       width = 6, height = 6, device = cairo_pdf)
```

```{r enh1-log2fc, fig.width = 5, fig.height = 6}
p_snai  <- make_log2fc_panel(log2fc_df %>% dplyr::filter(grepl("SNAI",   Replicate_Pair)),
                             title = "SNAI",   palette = snai_palette)
p_twist <- make_log2fc_panel(log2fc_df %>% dplyr::filter(grepl("TWIST1", Replicate_Pair)),
                             title = "TWIST1", palette = twist_palette)

(p_snai | p_twist)
ggsave(fig_name(key, "clusterFC"), (p_snai | p_twist), width = 5, height = 6)
```

```{r enh1-openclosed}
# MSP files: DMSO 5 min only -- accessibility is defined in the untreated state
msp_files <- enh_files(key, only(BOTH_SAMPLES, "DMSO5min"), "msp")

enh1_summary_snai <- compute_openclosed_log2fc(
  cluster_assignments_perrep = enh1$ca_perrep,
  comparisons = comparisons_5min[grepl("SNAI", names(comparisons_5min))],
  enhancer_name = enh_label(key),
  msp_files     = msp_files,
  chrom         = ENH[[key]]$chrom,
  motif_range   = ENH[[key]]$peaks$snai$motif
)

enh1_summary_twist <- compute_openclosed_log2fc(
  cluster_assignments_perrep = enh1$ca_perrep,
  comparisons = comparisons_5min[grepl("TWIST1", names(comparisons_5min))],
  enhancer_name = enh_label(key),
  msp_files     = msp_files,
  chrom         = ENH[[key]]$chrom,
  motif_range   = ENH[[key]]$peaks$twist1$motif
)
```

```{r enh1-tf, fig.width = 6, fig.height = 10}
tf_range <- enh_gr(key, "tf_plot")
tf_bs    <- enh_motifs(key, "motifs_tf")

tf_df <- cache_rds(
  sprintf("tf_df_%s_thr%dw%d", enh_slug(key), TF_SCORE_THR, TF_MAX_WIDTH),
  build_tf_metaplot_data(enh_files(key, only(BOTH_SAMPLES, "1min"), "tf"),
                         enh1$ca_perrep, tf_range,
                         score_threshold = TF_SCORE_THR, max_width = TF_MAX_WIDTH)
)

# all reads (clusters pooled), replicates averaged
tf_all <- compute_tf_metaplot(tf_df, enh1$ca_perrep, tf_range,
                              bundles = enh1$bundles, pool_clusters = TRUE) %>%
  average_replicates()

# accessible reads only: cluster 1
tf_acc <- compute_tf_metaplot(tf_df, enh1$ca_perrep, tf_range,
                              bundles = enh1$bundles,
                              clusters = enh_accessible(key),
                              pool_clusters = TRUE) %>%
  average_replicates()

p1 <- tf_panel(tf_all, tf_range, "TWIST1", tf_bs, enh_label(key, "- all reads"))
p2 <- tf_panel(tf_acc, tf_range, "TWIST1", tf_bs, enh_label(key, "- cluster 1"))
p3 <- tf_panel(tf_all, tf_range, "SNAI",   tf_bs, enh_label(key, "- all reads"))
p4 <- tf_panel(tf_acc, tf_range, "SNAI",   tf_bs, enh_label(key, "- cluster 1"))

p_atac <- plot_atac_signal(compute_atac_signal(ATAC_BW["SNAI_DMSO"], tf_range),
                           tf_range, binding_sites = tf_bs, title = enh_label(key))

ggarrange(p_atac, p1, p2, p3, p4, nrow = 5, align = "hv", legend = "right",
          heights = c(0.7, 1, 1, 1, 1))
ggsave(fig_name(key, "TFmetaplot_all_and_accessible"), width = 6, height = 10)
```

```{r enh1-tf-qc, eval = FALSE}
# QC only (not a paper figure): per-replicate TWIST1 curves with the individual
# footprint peaks marked, before replicate averaging.
tf_overall_reps <- compute_tf_metaplot(tf_df, enh1$ca_perrep, tf_range,
                                       bundles = enh1$bundles, pool_clusters = TRUE)
plot_tf_metaplot_overall(
  tf_overall_reps, tf_range,
  samples      = only(TWIST1_SAMPLES, "1min"),
  vlines       = c(17169874, 17169902, 17169943, 17169979, 17169750),
  vline_colors = c("black", "black", "black", "black", "purple"),
  title        = enh_label(key)
)
```

## 6.2 Enh 2 (internal: enh10, chr11)

<!-- Three accessible peaks are clustered SEPARATELY, each on its own local
     window, because "is this molecule open" is a local question:
       peak 1  TWIST1 motif
       peak 2  SNAI motif   <- this clustering is the one reused in section 7
       peak 3  E-box
     All three are drawn over the same wide plot_range so the panels align. -->

```{r enh2-common}
key <- "enh10"

comparisons_5min <- list(
  "TWIST1 Rep 1"  = c(numerator = "TWIST1_rep1_dTag5min", denominator = "TWIST1_rep1_DMSO5min"),
  "TWIST1 Rep 2"  = c(numerator = "TWIST1_rep2_dTag5min", denominator = "TWIST1_rep2_DMSO5min"),
  "SNAI1/2 Rep 1" = c(numerator = "SNAI_rep1_dTag5min",   denominator = "SNAI_rep1_DMSO5min"),
  "SNAI1/2 Rep 2" = c(numerator = "SNAI_rep2_dTag5min",   denominator = "SNAI_rep2_DMSO5min")
)

# named list form used by the open/closed summaries
comparisons_oc <- list(
  "SNAI R1 5m (dTag/DMSO)"   = c(numerator = "SNAI_rep1_dTag5min",   denominator = "SNAI_rep1_DMSO5min"),
  "SNAI R2 5m (dTag/DMSO)"   = c(numerator = "SNAI_rep2_dTag5min",   denominator = "SNAI_rep2_DMSO5min"),
  "TWIST1 R1 5m (dTag/DMSO)" = c(numerator = "TWIST1_rep1_dTag5min", denominator = "TWIST1_rep1_DMSO5min"),
  "TWIST1 R2 5m (dTag/DMSO)" = c(numerator = "TWIST1_rep2_dTag5min", denominator = "TWIST1_rep2_DMSO5min")
)

plot_range   <- enh_gr(key, "plot")
pairs_5min   <- list(
  "SNAI1/2 depletion" = c("SNAI_DMSO_5min",   "SNAI_dTag_5min"),
  "TWIST1 depletion"  = c("TWIST1_DMSO_5min", "TWIST1_dTag_5min")
)
sample_order <- unlist(pairs_5min, use.names = FALSE)
atac_df      <- compute_atac_signal(atac_for(sample_order), plot_range)

msp_files    <- enh_files(key, only(BOTH_SAMPLES, "DMSO5min"), "msp")
```

### 6.2.1 Peak 1 (TWIST1 motif)

```{r enh2-peak1-cluster}
enh2p1 <- run_clustering(
  key           = key,
  samples       = BOTH_SAMPLES,
  cluster_range = enh_gr(key, "cluster", peak = "peak1"),
  k             = ENH[[key]]$k,                  # 10
  cache_name    = "enh10peak1_k10"
)
```

```{r enh2-peak1-heatmap, fig.width = 10, fig.height = 8}
plot_df_merged <- prepare_plot_data(enh2p1$bundles_merged, enh2p1$ca_merged, plot_range)
log2fc_df      <- compute_log2fc_df(enh2p1$ca_perrep, comparisons_5min)

p_heat <- linked_heatmap(plot_df_merged, plot_range,
                         pairs         = pairs_5min,
                         atac_df       = atac_df,
                         binding_sites = enh_motifs(key, "motifs"),
                         title         = enh_label(key, "peak 1"))
p_heat
ggsave(fig_name(key, "heatmap_linked", "peak1"), p_heat,
       width = 10, height = 8, device = cairo_pdf)
```

```{r enh2-peak1-log2fc, fig.width = 5, fig.height = 6}
p_snai  <- make_log2fc_panel(log2fc_df %>% dplyr::filter(grepl("SNAI",   Replicate_Pair)),
                             title = "SNAI",   palette = snai_palette,  xlim = c(-0.8, 0.8))
p_twist <- make_log2fc_panel(log2fc_df %>% dplyr::filter(grepl("TWIST1", Replicate_Pair)),
                             title = "TWIST1", palette = twist_palette, xlim = c(-2, 2))
(p_snai | p_twist)
ggsave(fig_name(key, "clusterFC", "peak1"), (p_snai | p_twist), width = 5, height = 6)
```

```{r enh2-peak1-1min-vs-5min, fig.width = 8, fig.height = 6}
# TWIST1 only: does the 1 min vs 5 min SsDddA labelling change the picture?
pairs_time <- list(
  "5 minutes SsDddA" = c("TWIST1_DMSO_5min", "TWIST1_dTag_5min"),
  "1 minute SsDddA"  = c("TWIST1_DMSO_1min", "TWIST1_dTag_1min")
)
order_time <- unlist(pairs_time, use.names = FALSE)

p_time <- do.call(plot_fiberseq_dense_linked, c(
  list(
    plot_df      = plot_df_merged,
    plot_range   = plot_range,
    sample_order = order_time,
    atac_df      = compute_atac_signal(atac_for(order_time), plot_range),
    # the extra middle link visualises the 5min-dTag -> 1min-DMSO step
    link_pairs   = list(c("TWIST1_DMSO_5min", "TWIST1_dTag_5min"),
                        c("TWIST1_DMSO_1min", "TWIST1_dTag_1min"),
                        c("TWIST1_dTag_5min", "TWIST1_DMSO_1min")),
    sample_display_labels = c(TWIST1_DMSO_5min = "DMSO 5 min",
                              TWIST1_dTag_5min = "dTag 5 min",
                              TWIST1_DMSO_1min = "DMSO 1 min",
                              TWIST1_dTag_1min = "dTag 1 min"),
    label_groups  = pairs_time,
    fill_color    = c(TWIST1 = "darkblue"),
    binding_sites = enh_motifs(key, "motifs"),
    title         = enh_label(key, "peak 1 - TWIST1, 1 min vs 5 min")
  ),
  LINKED_STYLE
))
p_time
ggsave(fig_name(key, "heatmap_TWIST1_1min_vs_5min", "peak1"), p_time,
       width = 8, height = 6, device = cairo_pdf)
```

```{r enh2-peak1-openclosed}
enh2_peak1_summary <- compute_openclosed_log2fc(
  cluster_assignments_perrep = enh2p1$ca_perrep,
  comparisons   = comparisons_oc,
  enhancer_name = "Enh 2 peak 1",
  msp_files     = msp_files,
  chrom         = ENH[[key]]$chrom,
  motif_range   = ENH[[key]]$peaks$peak1$motif        # TWIST1
)
```

```{r enh2-peak1-tf, fig.width = 12, fig.height = 6}
tf_range <- enh_gr(key, "tf_plot")
tf_bs    <- enh_motifs(key, "motifs_tf")

tf_df <- cache_rds(
  sprintf("tf_df_%s_peak1_thr%dw%d", enh_slug(key), TF_SCORE_THR, TF_MAX_WIDTH),
  build_tf_metaplot_data(enh_files(key, only(BOTH_SAMPLES, "1min"), "tf"),
                         enh2p1$ca_perrep, tf_range,
                         score_threshold = TF_SCORE_THR, max_width = TF_MAX_WIDTH)
)

tf_all <- compute_tf_metaplot(tf_df, enh2p1$ca_perrep, tf_range,
                              bundles = enh2p1$bundles, pool_clusters = TRUE) %>%
  average_replicates()

tf_acc <- compute_tf_metaplot(tf_df, enh2p1$ca_perrep, tf_range,
                              bundles = enh2p1$bundles,
                              clusters = enh_accessible(key, "peak1"),
                              pool_clusters = TRUE) %>%
  average_replicates()

p1 <- tf_panel(tf_all, tf_range, "TWIST1", tf_bs, enh_label(key, "- all reads"))
p2 <- tf_panel(tf_acc, tf_range, "TWIST1", tf_bs, enh_label(key, "peak 1 - clusters 1,2"))
p_atac <- plot_atac_signal(compute_atac_signal(ATAC_BW["SNAI_DMSO"], tf_range),
                           tf_range, binding_sites = tf_bs,
                           title = enh_label(key, "ATAC signal"))

ggarrange(p_atac, p1, p2, nrow = 3, align = "hv", legend = "right",
          heights = c(0.7, 1, 1))
ggsave(fig_name(key, "TFmetaplot_TWIST1_all_and_accessible", "peak1"),
       width = 12, height = 6, device = cairo_pdf)
```

```{r enh2-atac-wide, fig.width = 10, fig.height = 2}
atac_wide <- compute_atac_signal(ATAC_BW["SNAI_DMSO"], enh_gr(key, "tf_plot"))
p_atac_wide <- plot_atac_signal(atac_wide, enh_gr(key, "tf_plot"),
                                binding_sites = enh_motifs(key, "motifs_tf"),
                                title = enh_label(key, "ATAC signal"))
p_atac_wide
ggsave(fig_name(key, "ATAC_wide"), p_atac_wide, width = 10, height = 2)
```

### 6.2.2 Peak 2 (SNAI motif)

```{r enh2-peak2-cluster}
enh2p2 <- run_clustering(
  key           = key,
  samples       = BOTH_SAMPLES,
  cluster_range = enh_gr(key, "cluster", peak = "peak2"),
  k             = ENH[[key]]$k,
  cache_name    = "enh10peak2_k10"
)
```

```{r enh2-peak2-heatmap, fig.width = 10, fig.height = 8}
plot_df_merged <- prepare_plot_data(enh2p2$bundles_merged, enh2p2$ca_merged, plot_range)
log2fc_df      <- compute_log2fc_df(enh2p2$ca_perrep, comparisons_5min)

p_heat <- linked_heatmap(plot_df_merged, plot_range,
                         pairs         = pairs_5min,
                         atac_df       = atac_df,
                         binding_sites = enh_motifs(key, "motifs"),
                         title         = enh_label(key, "peak 2"))
p_heat
ggsave(fig_name(key, "heatmap_linked", "peak2"), p_heat,
       width = 10, height = 8, device = cairo_pdf)

p_snai  <- make_log2fc_panel(log2fc_df %>% dplyr::filter(grepl("SNAI",   Replicate_Pair)),
                             title = "SNAI",   palette = snai_palette,  xlim = c(-0.8, 0.8))
p_twist <- make_log2fc_panel(log2fc_df %>% dplyr::filter(grepl("TWIST1", Replicate_Pair)),
                             title = "TWIST1", palette = twist_palette, xlim = c(-2, 2))
(p_snai | p_twist)
ggsave(fig_name(key, "clusterFC", "peak2"), (p_snai | p_twist), width = 5, height = 6)
```

```{r enh2-peak2-openclosed}
enh2_peak2_summary <- compute_openclosed_log2fc(
  cluster_assignments_perrep = enh2p2$ca_perrep,
  comparisons   = comparisons_oc,
  enhancer_name = "Enh 2 peak 2",
  msp_files     = msp_files,
  chrom         = ENH[[key]]$chrom,
  motif_range   = ENH[[key]]$peaks$peak2$motif        # SNAI
)
```

```{r enh2-peak2-tf, fig.width = 12, fig.height = 4.5}
tf_range <- enh_gr(key, "tf_plot")
tf_bs    <- enh_motifs(key, "motifs_tf")

tf_df <- cache_rds(
  sprintf("tf_df_%s_peak2_thr%dw%d", enh_slug(key), TF_SCORE_THR, TF_MAX_WIDTH),
  build_tf_metaplot_data(enh_files(key, only(BOTH_SAMPLES, "1min"), "tf"),
                         enh2p2$ca_perrep, tf_range,
                         score_threshold = TF_SCORE_THR, max_width = TF_MAX_WIDTH)
)

tf_all <- compute_tf_metaplot(tf_df, enh2p2$ca_perrep, tf_range,
                              bundles = enh2p2$bundles, pool_clusters = TRUE) %>%
  average_replicates()
tf_acc <- compute_tf_metaplot(tf_df, enh2p2$ca_perrep, tf_range,
                              bundles = enh2p2$bundles,
                              clusters = enh_accessible(key, "peak2"),
                              pool_clusters = TRUE) %>%
  average_replicates()

p1 <- tf_panel(tf_all, tf_range, "SNAI", tf_bs, enh_label(key, "- all reads"))
p2 <- tf_panel(tf_acc, tf_range, "SNAI", tf_bs, enh_label(key, "peak 2 - open clusters"))

ggarrange(p1, p2, nrow = 2, align = "hv", common.legend = TRUE, legend = "right")
ggsave(fig_name(key, "TFmetaplot_SNAI_all_and_accessible", "peak2"),
       width = 12, height = 4.5, device = cairo_pdf)
```

### 6.2.3 Peak 3 (E-box)

```{r enh2-peak3-cluster}
enh2p3 <- run_clustering(
  key           = key,
  samples       = BOTH_SAMPLES,
  cluster_range = enh_gr(key, "cluster", peak = "peak3"),
  k             = ENH[[key]]$k,
  cache_name    = "enh10peak3_k10"
)
```

```{r enh2-peak3-heatmap, fig.width = 10, fig.height = 8}
plot_df_merged <- prepare_plot_data(enh2p3$bundles_merged, enh2p3$ca_merged, plot_range)
log2fc_df      <- compute_log2fc_df(enh2p3$ca_perrep, comparisons_5min)

p_heat <- linked_heatmap(plot_df_merged, plot_range,
                         pairs         = pairs_5min,
                         atac_df       = atac_df,
                         binding_sites = enh_motifs(key, "motifs"),
                         title         = enh_label(key, "peak 3"))
p_heat
ggsave(fig_name(key, "heatmap_linked", "peak3"), p_heat,
       width = 10, height = 8, device = cairo_pdf)

p_snai  <- make_log2fc_panel(log2fc_df %>% dplyr::filter(grepl("SNAI",   Replicate_Pair)),
                             title = "SNAI",   palette = snai_palette,  xlim = c(-0.8, 0.8))
p_twist <- make_log2fc_panel(log2fc_df %>% dplyr::filter(grepl("TWIST1", Replicate_Pair)),
                             title = "TWIST1", palette = twist_palette, xlim = c(-2, 2))
(p_snai | p_twist)
ggsave(fig_name(key, "clusterFC", "peak3"), (p_snai | p_twist),
       width = 5, height = 6, device = cairo_pdf)
```

```{r enh2-peak3-openclosed}
enh2_peak3_summary <- compute_openclosed_log2fc(
  cluster_assignments_perrep = enh2p3$ca_perrep,
  comparisons   = comparisons_oc,
  enhancer_name = "Enh 2 peak 3",
  msp_files     = msp_files,
  chrom         = ENH[[key]]$chrom,
  motif_range   = ENH[[key]]$peaks$peak3$motif        # E-box
)
```

### 6.2.4 Enh 2: opening across the three peaks

```{r enh2-peak-summary, fig.width = 4, fig.height = 3}
peak_levels <- c("Enh 2 peak 1", "Enh 2 peak 2", "Enh 2 peak 3")

df_plot <- bind_rows(enh2_peak1_summary, enh2_peak2_summary, enh2_peak3_summary) %>%
  dplyr::filter(group == "open") %>%
  mutate(
    enhancer    = factor(enhancer, levels = peak_levels),
    peak_label  = sub("^Enh 2 ", "", as.character(enhancer)),
    TF          = factor(TF, levels = c("TWIST1", "SNAI")),
    facet_label = factor(ifelse(TF == "TWIST1", "TWIST1 depletion", "SNAI1/2 depletion"),
                         levels = c("SNAI1/2 depletion", "TWIST1 depletion"))
  )

df_mean <- df_plot %>%
  group_by(enhancer, peak_label, TF, facet_label) %>%
  summarise(mean_log2fc = mean(Log2Ratio, na.rm = TRUE), .groups = "drop")

p_enh2_peaks <- ggplot(df_mean, aes(x = peak_label, y = mean_log2fc, fill = TF)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6,
           colour = "black", linewidth = 0.2) +
  # FIX B6: geom_point has no `linewidth`; the outline width is `stroke`
  geom_point(data = df_plot, aes(x = peak_label, y = Log2Ratio, fill = TF),
             position = position_dodge(width = 0.7),
             shape = 21, size = 1.5, colour = "black", stroke = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, colour = "black") +
  scale_fill_manual(values = TF_FILL) +
  facet_wrap(~ facet_label, ncol = 1, scales = "free_y") +
  labs(x = NULL, fill = NULL,
       y = "log2(dTag/DMSO) of % reads in accessible clusters (+/- 150 bp of motif)") +
  theme_classic() +
  theme(legend.position  = "none",
        axis.text.x      = element_text(angle = 45, hjust = 1),
        strip.background = element_blank(),
        strip.text       = element_text(face = "bold"))

p_enh2_peaks
ggsave(fig_name(key, "openclosed_by_peak"), p_enh2_peaks,
       width = 4, height = 3, device = cairo_pdf)
```

### 6.2.5 Enh 2: all three peak clusterings in one stacked figure

<!-- ==========================================================================
     The same reads and the same window drawn three times, clustered on a
     different peak each time: peak 1 (TWIST1) top, peak 2 (SNAI) middle,
     peak 3 (E-box) bottom.

     ATAC track, motif names, lane names and the DMSO/dTag group bars appear on
     the TOP panel only. The lower two keep the dashed motif guides so the eye
     can carry a vertical line down all three blocks, but with the label text
     blanked.

     Alignment: all three panels use the same plot_range and the same
     sample_order, so lane_x0 is identical and the columns line up exactly.
     Nothing here needs a new function -- it is the existing linked heatmap
     called three times with different decoration.

     The decoration is controlled by three switches on
     plot_fiberseq_dense_linked(): show_binding_labels, show_lane_underline and
     show_range_label, plus margin_top / margin_bottom. All default to the old
     behaviour, so no other figure changes. Turning the labels off also
     RECLAIMS their space -- an empty label still expanded the y range, so the
     lower blocks used to carry ~30 units of unused headroom each.
     ========================================================================== -->

```{r enh2-stacked-peaks, fig.width = 10, fig.height = 18}
key         <- "enh10"
stack_range <- enh_gr(key, "plot")

stack_pairs <- list(
  "SNAI1/2 depletion" = c("SNAI_DMSO_5min",   "SNAI_dTag_5min"),
  "TWIST1 depletion"  = c("TWIST1_DMSO_5min", "TWIST1_dTag_5min")
)
stack_order <- unlist(stack_pairs, use.names = FALSE)
stack_atac  <- compute_atac_signal(atac_for(stack_order), stack_range)

stack_motifs <- enh_motifs(key, "motifs")

#' One block of the stack.
#' @param clust  a run_clustering() result (enh2p1 / enh2p2 / enh2p3)
#' @param decorate TRUE only for the top block: ATAC, motif names, lane names,
#'   DMSO/dTag group bars. The lower blocks keep the dashed motif guides for
#'   vertical alignment but draw no header at all.
#' @param show_xlab TRUE only for the bottom block
stack_block <- function(clust, title, decorate = FALSE, show_xlab = FALSE) {

  pdf_merged <- prepare_plot_data(clust$bundles_merged, clust$ca_merged, stack_range)

  do.call(plot_fiberseq_dense_linked, c(
    list(
      plot_df       = pdf_merged,
      plot_range    = stack_range,
      sample_order  = stack_order,
      link_pairs    = unname(stack_pairs),
      fill_color    = TF_FILL,
      title         = title,
      binding_sites = stack_motifs,
      atac_df       = if (decorate) stack_atac  else NULL,
      label_groups  = if (decorate) stack_pairs else NULL,
      sample_display_labels = if (decorate) {
        setNames(ifelse(grepl("DMSO", stack_order), "DMSO", "dTag"), stack_order)
      } else NULL,

      # header drawn once, on the top block only
      show_binding_labels = decorate,   # guides stay, motif names do not
      show_lane_underline = decorate,   # the bare black rule above the lanes
      # the coordinate caption is identical in all 12 lanes -- bottom row only
      show_range_label    = show_xlab,

      # blocks 2 and 3 need no headroom, so pull them up under the one above
      margin_top    = if (decorate) NULL else 2,
      margin_bottom = if (show_xlab) 5 else 0,

      # "" is not NULL, so %||% leaves it alone and the axis title is blank
      xlab = if (show_xlab) NULL else ""
    ),
    LINKED_STYLE
  ))
}

p_peak1 <- stack_block(enh2p1, "Clustered on peak 1 (TWIST1 motif)", decorate = TRUE)
p_peak2 <- stack_block(enh2p2, "Clustered on peak 2 (SNAI1/2 motif)")
p_peak3 <- stack_block(enh2p3, "Clustered on peak 3 (E-box)", show_xlab = TRUE)

# Top block is taller because it alone carries the ATAC track, the motif names
# and the group bars. Blocks 2 and 3 now occupy almost exactly their lane
# height, so 1.35 / 1 / 1 keeps the three read blocks the same physical size.
p_enh2_stack <- (p_peak1 / p_peak2 / p_peak3) +
  plot_layout(heights = c(1.35, 1, 1)) +
  plot_annotation(title = enh_label(key))

p_enh2_stack
ggsave(fig_name(key, "heatmap_linked_all3peaks_stacked"), p_enh2_stack,
       width = 10, height = 15, limitsize = FALSE, device = cairo_pdf)
```

```{r enh2-stacked-peaks-with-fc, eval = FALSE, fig.width = 14, fig.height = 18}
# Variant: same stack with the per-cluster log2FC bars alongside each block.
# Each row is (heatmap | SNAI bars | TWIST1 bars); patchwork nests fine.
fc_pair <- function(clust) {
  d <- compute_log2fc_df(clust$ca_perrep, comparisons_5min)
  (make_log2fc_panel(d %>% dplyr::filter(grepl("SNAI",   Replicate_Pair)),
                     title = "SNAI",   palette = snai_palette,  xlim = c(-0.8, 0.8)) |
   make_log2fc_panel(d %>% dplyr::filter(grepl("TWIST1", Replicate_Pair)),
                     title = "TWIST1", palette = twist_palette, xlim = c(-2, 2)))
}

row1 <- (p_peak1 | fc_pair(enh2p1)) + plot_layout(widths = c(5, 2))
row2 <- (p_peak2 | fc_pair(enh2p2)) + plot_layout(widths = c(5, 2))
row3 <- (p_peak3 | fc_pair(enh2p3)) + plot_layout(widths = c(5, 2))

p_enh2_stack_fc <- (row1 / row2 / row3) + plot_layout(heights = c(1.35, 1, 1))
ggsave(fig_name(key, "heatmap_linked_all3peaks_stacked_withFC"), p_enh2_stack_fc,
       width = 14, height = 18, limitsize = FALSE, device = cairo_pdf)
```

## 6.3 SNAI1/2-only enhancers (Enh 3, Enh 4, Enh 5, Enh 6)

<!-- These four are analysed with the SNAI1/2 degron samples only, so they all
     share one comparison set and one two-lane heatmap layout. -->

```{r snai-only-common}
comparisons_snai <- list(
  "Rep1" = c(numerator = "SNAI_rep1_dTag5min", denominator = "SNAI_rep1_DMSO5min"),
  "Rep2" = c(numerator = "SNAI_rep2_dTag5min", denominator = "SNAI_rep2_DMSO5min")
)

# open/closed summaries use the same two pairs but labelled for the figure
comparisons_snai_oc <- list(
  "SNAI R1 5m (dTag/DMSO)" = c(numerator = "SNAI_rep1_dTag5min", denominator = "SNAI_rep1_DMSO5min"),
  "SNAI R2 5m (dTag/DMSO)" = c(numerator = "SNAI_rep2_dTag5min", denominator = "SNAI_rep2_DMSO5min")
)

pairs_snai   <- list("SNAI1/2 depletion" = c("SNAI_DMSO_5min", "SNAI_dTag_5min"))
order_snai   <- unlist(pairs_snai, use.names = FALSE)

#' Heatmap + log2FC bar for a two-lane SNAI-only enhancer
snai_heatmap_panel <- function(clust, key, peak = NULL, plot_range,
                               title, fc_xlim = NULL) {

  atac_df        <- compute_atac_signal(atac_for(order_snai), plot_range)
  plot_df_merged <- prepare_plot_data(clust$bundles_merged, clust$ca_merged, plot_range)
  log2fc_df      <- compute_log2fc_df(clust$ca_perrep, comparisons_snai)

  p_heat <- linked_heatmap(plot_df_merged, plot_range,
                           pairs         = pairs_snai,
                           atac_df       = atac_df,
                           binding_sites = enh_motifs(key, "motifs", peak = peak),
                           title         = title,
                           fill_color    = c(SNAI = "darkred"))

  p_bar <- make_log2fc_panel(log2fc_df, title = "log2FC",
                             palette = snai_palette, xlim = fc_xlim) +
    theme(legend.position = "right")

  (p_heat | p_bar) + plot_layout(widths = c(5, 1))
}

#' All-reads vs accessible-reads TF metaplots + ATAC, for a SNAI-only enhancer
snai_tf_figure <- function(clust, key, tf_range, cache_tag,
                           accessible = enh_accessible(key)) {

  tf_bs <- enh_motifs(key, "motifs_tf")

  tf_df <- cache_rds(
    sprintf("tf_df_%s_thr%dw%d", cache_tag, TF_SCORE_THR, TF_MAX_WIDTH),
    build_tf_metaplot_data(enh_files(key, only(SNAI_SAMPLES, "1min"), "tf"),
                           clust$ca_perrep, tf_range,
                           score_threshold = TF_SCORE_THR, max_width = TF_MAX_WIDTH)
  )

  tf_all <- compute_tf_metaplot(tf_df, clust$ca_perrep, tf_range,
                                bundles = clust$bundles, pool_clusters = TRUE) %>%
    average_replicates()
  tf_acc <- compute_tf_metaplot(tf_df, clust$ca_perrep, tf_range,
                                bundles = clust$bundles, clusters = accessible,
                                pool_clusters = TRUE) %>%
    average_replicates()

  p1 <- tf_panel(tf_all, tf_range, "SNAI", tf_bs, enh_label(key, "- all reads"))
  p2 <- tf_panel(tf_acc, tf_range, "SNAI", tf_bs,
                 enh_label(key, paste0("- cluster ", paste(accessible, collapse = ","))))
  p_atac <- plot_atac_signal(compute_atac_signal(ATAC_BW["SNAI_DMSO"], tf_range),
                             tf_range, binding_sites = tf_bs, title = enh_label(key))

  ggarrange(p_atac, p1, p2, nrow = 3, align = "hv", legend = "right",
            heights = c(0.7, 1, 1))
}
```

### 6.3.1 Enh 3 (internal: enh8, chr20) -- peak 1, SNAI motif

```{r enh3-peak1, fig.width = 10, fig.height = 6}
key <- "enh8"

enh3p1 <- run_clustering(
  key           = key,
  samples       = SNAI_SAMPLES,
  cluster_range = enh_gr(key, "cluster", peak = "peak1"),
  k             = ENH[[key]]$k,
  cache_name    = "ENH8peak1_k10"
)

p <- snai_heatmap_panel(enh3p1, key, peak = "peak1",
                        plot_range = enh_gr(key, "plot"),
                        title      = enh_label(key))
p
ggsave(fig_name(key, "heatmap_linked", "peak1"), p,
       width = 10, height = 6, device = cairo_pdf)
```

```{r enh3-peak1-openclosed}
msp_files <- enh_files(key, only(SNAI_SAMPLES, "DMSO5min"), "msp")

enh3_peak1_summary <- compute_openclosed_log2fc(
  cluster_assignments_perrep = enh3p1$ca_perrep,
  comparisons   = comparisons_snai_oc,
  enhancer_name = "Enh 3 peak 1",
  msp_files     = msp_files,
  chrom         = ENH[[key]]$chrom,
  motif_range   = ENH[[key]]$peaks$peak1$motif        # SNAI
)
```

```{r enh3-tf, fig.width = 12, fig.height = 6}
tf_range <- enh_gr(key, "tf_plot")
snai_tf_figure(enh3p1, key, tf_range, cache_tag = "Enh3_peak1")
ggsave(fig_name(key, "TFmetaplot_all_and_accessible", "peak1"),
       width = 12, height = 6)
```

```{r enh3-atac-wide, fig.width = 10, fig.height = 2}
r <- enh_gr(key, "atac_plot")
p_atac_wide <- plot_atac_signal(compute_atac_signal(ATAC_BW["SNAI_DMSO"], r), r,
                                binding_sites = enh_motifs(key, "motifs_tf"),
                                title = enh_label(key, "ATAC signal"))
p_atac_wide
ggsave(fig_name(key, "ATAC_wide"), p_atac_wide, width = 10, height = 2)
```

### 6.3.2 Enh 3 -- peak 2 (rightmost peak, TWIST1 motif)

```{r enh3-peak2, fig.width = 10, fig.height = 6}
enh3p2 <- run_clustering(
  key           = key,
  samples       = SNAI_SAMPLES,
  cluster_range = enh_gr(key, "cluster", peak = "peak2"),
  k             = ENH[[key]]$k,
  cache_name    = "ENH8peak2_k10"
)

p <- snai_heatmap_panel(enh3p2, key, peak = "peak2",
                        plot_range = enh_gr(key, "plot"),
                        title      = enh_label(key, "- peak 2"))
p
ggsave(fig_name(key, "heatmap_linked", "peak2"), p,
       width = 10, height = 6, device = cairo_pdf)
```

```{r enh3-peak2-openclosed}
enh3_peak2_summary <- compute_openclosed_log2fc(
  cluster_assignments_perrep = enh3p2$ca_perrep,
  comparisons   = comparisons_snai_oc,
  enhancer_name = "Enh 3 peak 2",
  msp_files     = msp_files,
  chrom         = ENH[[key]]$chrom,
  motif_range   = ENH[[key]]$peaks$peak2$motif        # TWIST1
)
```

```{r enh3-peak-summary, fig.width = 4, fig.height = 2.5}
df_plot <- bind_rows(enh3_peak1_summary, enh3_peak2_summary) %>%
  dplyr::filter(group == "open") %>%
  mutate(
    enhancer    = factor(enhancer, levels = c("Enh 3 peak 1", "Enh 3 peak 2")),
    TF          = factor(TF, levels = c("TWIST1", "SNAI")),
    facet_label = factor(ifelse(TF == "TWIST1", "TWIST1 depletion", "SNAI1/2 depletion"),
                         levels = c("SNAI1/2 depletion", "TWIST1 depletion"))
  )

df_mean <- df_plot %>%
  group_by(enhancer, TF, facet_label) %>%
  summarise(mean_log2fc = mean(Log2Ratio, na.rm = TRUE), .groups = "drop")

p_enh3_peaks <- ggplot(df_mean, aes(x = enhancer, y = mean_log2fc, fill = TF)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6,
           colour = "black", linewidth = 0.2) +
  geom_point(data = df_plot, aes(x = enhancer, y = Log2Ratio, fill = TF),
             position = position_dodge(width = 0.7),
             shape = 21, size = 1.5, colour = "black", stroke = 0.2) +   # FIX B6
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, colour = "black") +
  scale_fill_manual(values = TF_FILL) +
  facet_wrap(~ facet_label, ncol = 1, scales = "free_y") +
  labs(x = NULL, fill = NULL,
       y = "log2(dTag/DMSO) of % reads in accessible clusters (+/- 150 bp of SNAI1/2 motif)") +
  theme_classic() +
  theme(legend.position  = "none",
        axis.text.x      = element_text(angle = 45, hjust = 1),
        strip.background = element_blank(),
        strip.text       = element_text(face = "bold"))

p_enh3_peaks
ggsave(fig_name(key, "openclosed_by_peak"), p_enh3_peaks,
       width = 4, height = 2.5, device = cairo_pdf)
```

### 6.3.3 Enh 4 (internal: enh15, chr6)

<!-- NB this is the locus the previous version also titled "ENH4", the same
     string it used for internal enh4 (= Enh 1). Under the paper naming this
     locus IS Enh 4, and internal enh4 is Enh 1. -->

```{r enh4-analysis, fig.width = 10, fig.height = 6}
key <- "enh15"

enh4 <- run_clustering(
  key           = key,
  samples       = SNAI_SAMPLES,
  cluster_range = enh_gr(key, "cluster"),
  k             = ENH[[key]]$k,
  cache_name    = "ENH15_k10"
)

p <- snai_heatmap_panel(enh4, key,
                        plot_range = enh_gr(key, "plot"),
                        title      = enh_label(key),
                        fc_xlim    = c(-0.5, 0.5))
p
ggsave(fig_name(key, "heatmap_linked"), p, width = 10, height = 6, device = cairo_pdf)
```

```{r enh4-openclosed}
msp_files <- enh_files(key, only(SNAI_SAMPLES, "DMSO5min"), "msp")

enh4_summary <- compute_openclosed_log2fc(
  cluster_assignments_perrep = enh4$ca_perrep,
  comparisons   = comparisons_snai_oc,
  enhancer_name = enh_label(key),
  msp_files     = msp_files,
  chrom         = ENH[[key]]$chrom,
  motif_range   = ENH[[key]]$peaks$snai$motif
)
```

### 6.3.4 Enh 5 (internal: enh30, chr3)

```{r enh5-analysis, fig.width = 10, fig.height = 6}
key <- "enh30"

enh5 <- run_clustering(
  key           = key,
  samples       = SNAI_SAMPLES,
  cluster_range = enh_gr(key, "cluster"),
  k             = ENH[[key]]$k,
  cache_name    = "ENH30_k10"
)

p <- snai_heatmap_panel(enh5, key,
                        plot_range = enh_gr(key, "plot"),
                        title      = enh_label(key))
p
ggsave(fig_name(key, "heatmap_linked"), p, width = 10, height = 6, device = cairo_pdf)
```

```{r enh5-openclosed}
msp_files <- enh_files(key, only(SNAI_SAMPLES, "DMSO5min"), "msp")

enh5_summary <- compute_openclosed_log2fc(
  cluster_assignments_perrep = enh5$ca_perrep,
  comparisons   = comparisons_snai_oc,
  enhancer_name = enh_label(key),
  msp_files     = msp_files,
  chrom         = ENH[[key]]$chrom,
  motif_range   = ENH[[key]]$peaks$snai$motif
)
```

```{r enh5-tf, fig.width = 12, fig.height = 6}
tf_range <- enh_gr(key, "tf_plot")
snai_tf_figure(enh5, key, tf_range, cache_tag = "Enh5")
ggsave(fig_name(key, "TFmetaplot_all_and_accessible"), width = 12, height = 6)
```

```{r enh5-atac-wide, fig.width = 10, fig.height = 2}
r <- enh_gr(key, "atac_plot")
p_atac_wide <- plot_atac_signal(compute_atac_signal(ATAC_BW["SNAI_DMSO"], r), r,
                                binding_sites = enh_motifs(key, "motifs_tf"),
                                title = enh_label(key, "ATAC signal"))
p_atac_wide
ggsave(fig_name(key, "ATAC_wide"), p_atac_wide, width = 10, height = 2)
```

### 6.3.5 Enh 6 (internal: enh33, chr11)

```{r enh6-analysis, fig.width = 10, fig.height = 6}
key <- "enh33"

enh6 <- run_clustering(
  key           = key,
  samples       = SNAI_SAMPLES,
  cluster_range = enh_gr(key, "cluster"),
  k             = ENH[[key]]$k,
  cache_name    = "ENH33_k10"
)

p <- snai_heatmap_panel(enh6, key,
                        plot_range = enh_gr(key, "plot"),
                        title      = enh_label(key))
p
ggsave(fig_name(key, "heatmap_linked"), p, width = 10, height = 6, device = cairo_pdf)
```

```{r enh6-openclosed}
msp_files <- enh_files(key, only(SNAI_SAMPLES, "DMSO5min"), "msp")

enh6_summary <- compute_openclosed_log2fc(
  cluster_assignments_perrep = enh6$ca_perrep,
  comparisons   = comparisons_snai_oc,
  enhancer_name = enh_label(key),
  msp_files     = msp_files,
  chrom         = ENH[[key]]$chrom,
  motif_range   = ENH[[key]]$peaks$snai$motif
)
```

```{r enh6-tf, fig.width = 12, fig.height = 6}
tf_range <- enh_gr(key, "tf_plot")
snai_tf_figure(enh6, key, tf_range, cache_tag = "Enh6")
ggsave(fig_name(key, "TFmetaplot_all_and_accessible"), width = 12, height = 6)
```

```{r enh6-atac-wide, fig.width = 10, fig.height = 2}
r <- enh_gr(key, "atac_plot")
p_atac_wide <- plot_atac_signal(compute_atac_signal(ATAC_BW["SNAI_DMSO"], r), r,
                                binding_sites = enh_motifs(key, "motifs_tf"),
                                title = enh_label(key, "ATAC signal"))
p_atac_wide
ggsave(fig_name(key, "ATAC_wide"), p_atac_wide, width = 10, height = 2)
```

# 7. Cross-enhancer summaries

## 7.1 Assemble the open/closed table

<!-- ==========================================================================
     C4: the previous version grew one global table with
         all_enhancer_summary_v2 <- bind_rows(all_enhancer_summary_v2, x)
     spread over eight chunks. Re-running any chunk duplicated its rows and
     the duplicates then fed straight into the summary bar plot. Everything is
     bound exactly once, here, from named objects.
     ========================================================================== -->

```{r summary-assemble}
open_closed_all <- bind_rows(
  enh1_summary_snai,    # Enh 1, SNAI motif
  enh1_summary_twist,   # Enh 1, TWIST1 motif
  enh2_peak1_summary,   # Enh 2 peak 1 (TWIST1 motif)
  enh2_peak2_summary,   # Enh 2 peak 2 (SNAI motif)
  enh2_peak3_summary,   # Enh 2 peak 3 (E-box)
  enh3_peak1_summary,   # Enh 3 peak 1 (SNAI motif)
  enh3_peak2_summary,   # Enh 3 peak 2 (TWIST1 motif)
  enh4_summary,         # Enh 4
  enh5_summary,         # Enh 5
  enh6_summary          # Enh 6
)

stopifnot(!any(duplicated(open_closed_all)))    # guard against the old C4 bug

# which clusters ended up "open" at each locus -- worth eyeballing every run
open_closed_all %>%
  dplyr::filter(group == "open") %>%
  distinct(enhancer, clusters) %>%
  arrange(enhancer) %>%
  print(n = Inf)
```

## 7.2 Chromatin opening on TF degradation, one bar per enhancer

<!-- Each enhancer contributes the peak where the degraded TF actually binds:
       Enh 1  both motifs (SNAI + TWIST1) were quantified
       Enh 2  TWIST1 read out at peak 1, SNAI1/2 read out at peak 2
       Enh 3  the SNAI1/2 peak (peak 1)
       Enh 4-6  their single SNAI1/2 peak -->

```{r summary-plot, fig.width = 6, fig.height = 4}
plot_open_summary <- function(summary_df, enhancer_order, ylim = NULL,
                              tf_colors = TF_FILL) {

  df <- summary_df %>%
    dplyr::filter(group == "open") %>%
    mutate(enhancer = factor(enhancer, levels = enhancer_order))

  df_mean <- df %>%
    group_by(enhancer, TF) %>%
    summarise(mean_log2fc = mean(Log2Ratio, na.rm = TRUE), .groups = "drop")

  p <- ggplot(df_mean, aes(x = enhancer, y = mean_log2fc, fill = TF)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6,
             colour = "black", linewidth = 0.2) +
    geom_point(data = df, aes(x = enhancer, y = Log2Ratio, fill = TF),
               position = position_dodge(width = 0.7),
               shape = 21, size = 1.5, colour = "black", stroke = 0.2) +  # FIX B6
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, colour = "black") +
    scale_fill_manual(values = tf_colors) +
    labs(x = "Enhancer", fill = NULL,
         y = "log2(dTag/DMSO) of % reads in accessible clusters (+/- 150 bp of motif)") +
    theme_classic() +
    theme(legend.position = "right",
          axis.text.x     = element_text(angle = 45, hjust = 1))

  if (!is.null(ylim)) p <- p + coord_cartesian(ylim = ylim)
  p
}

summary_subset <- open_closed_all %>%
  dplyr::filter(
    (enhancer == "Enh 2 peak 1" & TF == "TWIST1") |     # TWIST1 at its own peak
    (enhancer == "Enh 2 peak 2" & TF == "SNAI")   |     # SNAI1/2 at its own peak
    enhancer %in% c("Enh 1", "Enh 3 peak 1", "Enh 4", "Enh 5", "Enh 6")
  ) %>%
  mutate(enhancer = recode(enhancer,
                           "Enh 2 peak 1" = "Enh 2",
                           "Enh 2 peak 2" = "Enh 2",
                           "Enh 3 peak 1" = "Enh 3"))

p_open <- plot_open_summary(
  summary_subset,
  enhancer_order = c("Enh 1", "Enh 2", "Enh 3", "Enh 4", "Enh 5", "Enh 6")
)
p_open
ggsave(fig("summary_open_closed_all_enhancers_v297.pdf"), p_open,
       width = 6, height = 4)
```

## 7.3 TF footprint occupancy: all reads vs accessible reads (CMH)

<!-- ==========================================================================
     SNAI1/2 DMSO vs dTag, 1 min DAF. Plotted: TFAP2A at Enh 2, TFAP2A at
     Enh 5, NR2F1 at Enh 5. The Enh 1 and Enh 6 sites are kept commented out in
     `sites` below; enh_config covers all four enhancers.

     LAYOUT  facet_grid(read_set ~ site), row 1 = all reads, row 2 = accessible
             reads, scales = "free_y" so each ROW gets its own y axis.

     OCCUPANCY (the bars)  same definition as plot_tf_metaplot_overall():
             per bp   -> TF footprints covering bp / read coverage at bp * 100
             per site -> mean of that over the motif's bp (bp below
                         min_read_cov are NA and excluded)

     READ SETS
       All reads        compute_tf_metaplot(pool_clusters = TRUE): denominator
                        is every read of that sample.
       Accessible reads compute_tf_metaplot(clusters = <open set>,
                        pool_clusters = TRUE): numerator AND denominator are
                        restricted to that enhancer's open cluster(s), the same
                        sets used in the per-enhancer TF metaplots above.

     STATISTICS
       Plotted: Cochran-Mantel-Haenszel on per-read footprint calls, stratified
         by replicate. For every read FULLY spanning the motif, "is there a TF
         footprint overlapping the motif?" -> 2 x 2 x 2 table
         (condition x footprinted x replicate). Note this tests the PERCENTAGE
         OF READS footprinted -- a per-molecule quantity, close to but not
         identical to the mean per-bp occupancy drawn as the bars. Both appear
         in the QC table.
       Printed only: paired t-test on the per-replicate occupancy values shown
         as points. n = 2 pairs, df = 1 

     Self-contained: reloads the cached clustering objects and the SNAI 1-min
     tf .bed files itself, so it only needs sections 1-5 to have been run.
     ========================================================================== -->

```{r cmh-config}
# Each enhancer must be paired with the tf .bed set matching the fiberHMM call
# set it was CLUSTERED on -- Enh 1 is min1, the rest are min3 (reviewer note Q1).
enh_config <- list(
  enh4  = list(cache = "enh4_k12",         accessible = enh_accessible("enh4")),
  # TFAP2A sits inside the peak-2 window, so the SNAI-centred peak-2 clustering
  # is the correct local accessibility call for it.
  enh10 = list(cache = "enh10peak2_k10",   accessible = enh_accessible("enh10", "peak2")),
  enh30 = list(cache = "ENH30_k10",        accessible = enh_accessible("enh30")),
  enh33 = list(cache = "ENH33_k10",        accessible = enh_accessible("enh33"))
)

# TF sites to quantify.
#   NR2F1 at Enh 5 = the two adjacent motifs treated as one window.
#   TFAP2A at Enh 6 = the site closest to the SNAI motif (8,709,103-8,709,110).
sites <- tribble(
  ~enh,     ~TF,       ~motif_start, ~motif_end,
  "enh10",  "TFAP2A",      12319535,   12319546,
  "enh30",  "TFAP2A",      69117131,   69117141,
  "enh30",  "NR2F1",       69117159,   69117175
# "enh4",   "TWIST1",      17169866,   17169882,
# "enh33",  "TFAP2A",       8709060,    8709071,
# "enh33",  "TWIST1",       8709132,    8709147
)

flank           <- 300L    # bp added around the motif(s) when building tf_df;
                           # only the motif bp are averaged
read_set_levels <- c("All reads", "Accessible reads")
```

```{r cmh-per-read-calls}
#' Per-read footprint calls: the numerator/denominator for the CMH test
#'
#' denominator = reads FULLY spanning the motif (a read stopping inside the
#'               motif cannot be scored), optionally restricted to `clusters`
#' numerator   = those reads carrying a TF footprint overlapping the motif
per_read_calls <- function(tf_df_e, ca_e, bundles_e, tf_files,
                           motif_start, motif_end, clusters = NULL) {

  reads <- reads_with_clusters(bundles_e, ca_e) %>%
    dplyr::filter(Sample %in% names(tf_files),
                  read_start <= motif_start, read_end >= motif_end)

  if (!is.null(clusters)) {
    reads <- reads %>%
      dplyr::filter(as.character(cluster) %in% as.character(clusters))
  }

  fp_ids <- tf_df_e %>%
    dplyr::filter(start <= motif_end, end >= motif_start) %>%
    dplyr::pull(unique_id) %>% unique()

  reads %>%
    transmute(Sample, unique_id,
              footprinted = factor(ifelse(unique_id %in% fp_ids, "yes", "no"),
                                   levels = c("yes", "no")))
}
```

```{r cmh-quantify}
res <- purrr::map(unique(sites$enh), function(enh_key) {

  cfg      <- enh_config[[enh_key]]
  enh_site <- sites %>% dplyr::filter(enh == enh_key)
  lab      <- enh_label(enh_key)

  message("=== ", enh_key, " -> ", lab, " ===")

  clust     <- readRDS(robj(paste0(cfg$cache, ".rds")))
  bundles_e <- clust$bundles
  ca_e      <- clust$ca_perrep
  tf_files  <- enh_files(enh_key, only(SNAI_SAMPLES, "1min"), "tf")

  # one window covering every site of this enhancer, plus flank
  win <- GRanges(ENH[[enh_key]]$chrom,
                 IRanges(min(enh_site$motif_start) - flank,
                         max(enh_site$motif_end)   + flank))

  tf_df_e <- cache_rds(
    sprintf("tf_df_occbar_%s_thr%dw%d", enh_key, TF_SCORE_THR, TF_MAX_WIDTH),
    build_tf_metaplot_data(tf_files, ca_e, win,
                           score_threshold = TF_SCORE_THR,
                           max_width       = TF_MAX_WIDTH)
  )

  # per-bp occupancy for the two read sets (replicates kept separate)
  meta <- bind_rows(
    compute_tf_metaplot(tf_df_e, ca_e, win, bundles = bundles_e,
                        pool_clusters = TRUE, min_read_cov = MIN_READ_COV) %>%
      mutate(read_set = "All reads"),
    compute_tf_metaplot(tf_df_e, ca_e, win, bundles = bundles_e,
                        clusters      = cfg$accessible,
                        pool_clusters = TRUE, min_read_cov = MIN_READ_COV) %>%
      mutate(read_set = "Accessible reads")
  )

  # (a) bars: mean per-bp occupancy over each motif window, per sample
  occ <- purrr::pmap_dfr(enh_site, function(enh, TF, motif_start, motif_end) {
    meta %>%
      dplyr::filter(pos >= motif_start, pos <= motif_end) %>%
      group_by(Sample, read_set) %>%
      summarise(occupancy  = mean(normalized_occupancy, na.rm = TRUE),
                mean_depth = mean(read_coverage),
                n_bp_used  = sum(!is.na(normalized_occupancy)),
                .groups    = "drop") %>%
      mutate(enh = enh, TF = TF, fig_label = lab,
             motif = sprintf("%s:%d-%d", ENH[[enh_key]]$chrom, motif_start, motif_end),
             acc_clusters = paste(cfg$accessible, collapse = ","))
  })

  # (b) per-read calls for the CMH test
  calls <- purrr::pmap_dfr(enh_site, function(enh, TF, motif_start, motif_end) {
    bind_rows(
      per_read_calls(tf_df_e, ca_e, bundles_e, tf_files,
                     motif_start, motif_end, clusters = NULL) %>%
        mutate(read_set = "All reads"),
      per_read_calls(tf_df_e, ca_e, bundles_e, tf_files,
                     motif_start, motif_end, clusters = cfg$accessible) %>%
        mutate(read_set = "Accessible reads")
    ) %>%
      mutate(enh = enh, TF = TF, fig_label = lab)
  })

  list(occ = occ, calls = calls)
})

add_keys <- function(df) {
  df %>%
    mutate(
      condition  = factor(ifelse(grepl("DMSO", Sample), "DMSO", "dTag"),
                          levels = c("DMSO", "dTag")),
      replicate  = ifelse(grepl("rep1", Sample), "Rep 1", "Rep 2"),
      site_label = factor(
        paste0(TF, "\n", fig_label),
        levels = unique(paste0(sites$TF, "\n",
                               purrr::map_chr(sites$enh, enh_label)))),
      read_set   = factor(read_set, levels = read_set_levels)
    )
}

occ_df   <- purrr::map_dfr(res, "occ")   %>% add_keys()
calls_df <- purrr::map_dfr(res, "calls") %>% add_keys()
```

```{r cmh-stats}
# ---- plotted: Cochran-Mantel-Haenszel, strata = replicates ------------------
cmh_one <- function(df) {
  tab <- df %>%
    dplyr::count(replicate, condition, footprinted, name = "n") %>%
    tidyr::complete(replicate, condition, footprinted, fill = list(n = 0L))
  arr  <- stats::xtabs(n ~ condition + footprinted + replicate, data = tab)
  # drop strata with an empty margin -- mantelhaen.test cannot use them
  keep <- apply(arr, 3, function(m) all(rowSums(m) > 0) && all(colSums(m) > 0))
  arr  <- arr[, , keep, drop = FALSE]
  if (dim(arr)[3] < 1) return(NA_real_)
  tryCatch(stats::mantelhaen.test(arr, exact = FALSE)$p.value,
           error = function(e) NA_real_)
}

p_cmh <- calls_df %>%
  group_by(site_label, read_set) %>%
  group_modify(~ tibble(n_reads = nrow(.x), p_value = cmh_one(.x))) %>%
  ungroup() %>%
  mutate(test = "CMH on per-read footprint calls (strata = replicates)")

# ---- printed only: paired t-test on the per-replicate occupancy values ------
# n = 2 pairs, df = 1. See reviewer note Q8: do not promote these to a figure.
p_paired_t <- occ_df %>%
  select(site_label, read_set, replicate, condition, occupancy) %>%
  pivot_wider(names_from = condition, values_from = occupancy) %>%
  group_by(site_label, read_set) %>%
  summarise(
    n_pairs = sum(!is.na(DMSO) & !is.na(dTag)),
    p_value = if (sum(!is.na(DMSO) & !is.na(dTag)) >= 2) {
      tryCatch(stats::t.test(DMSO, dTag, paired = TRUE)$p.value,
               error = function(e) NA_real_)
    } else NA_real_,
    .groups = "drop"
  ) %>%
  mutate(test = "paired t-test (n = replicate pairs)")

# ---- QC: what is behind every bar and every test ---------------------------
qc <- occ_df %>%
  select(site_label, motif, read_set, Sample, occupancy,
         mean_depth, n_bp_used, acc_clusters) %>%
  left_join(
    calls_df %>%
      group_by(site_label, read_set, Sample) %>%
      summarise(n_reads_spanning      = dplyr::n(),
                pct_reads_footprinted = 100 * mean(footprinted == "yes"),
                .groups = "drop"),
    by = c("site_label", "read_set", "Sample")
  ) %>%
  arrange(site_label, read_set, Sample)
print(qc, n = Inf)

bind_rows(p_cmh      %>% select(site_label, read_set, p_value, test),
          p_paired_t %>% select(site_label, read_set, p_value, test)) %>%
  arrange(test, site_label, read_set) %>%
  print(n = Inf)
```

```{r cmh-plot, fig.width = 5.5, fig.height = 3.5}
occ_mean <- occ_df %>%
  group_by(site_label, read_set, condition) %>%
  summarise(occupancy = mean(occupancy, na.rm = TRUE), .groups = "drop")

# p-value label height: one per ROW, because facet_grid(scales = "free_y")
# gives each row a single y axis shared across the panels in that row
label_y <- occ_df %>%
  group_by(read_set) %>%
  summarise(y = max(occupancy, na.rm = TRUE) * 1.12, .groups = "drop")

lbl <- p_cmh %>%
  left_join(label_y, by = "read_set") %>%
  mutate(label = ifelse(is.na(p_value), "p = NA",
                        paste0("p = ", formatC(p_value, format = "g", digits = 3))))

p_occ_cmh <- ggplot(occ_mean, aes(x = condition, y = occupancy, fill = condition)) +
  geom_col(width = 0.65, colour = "black", linewidth = 0.2) +
  geom_line(data = occ_df, aes(x = condition, y = occupancy, group = replicate),
            inherit.aes = FALSE, colour = "grey35", linewidth = 0.3) +
  geom_point(data = occ_df,
             aes(x = condition, y = occupancy, fill = condition, shape = replicate),
             size = 1.8, colour = "black", stroke = 0.3) +
  geom_text(data = lbl, aes(x = 1.5, y = y, label = label),
            inherit.aes = FALSE, size = 2.8) +
  facet_grid(read_set ~ site_label, scales = "free_y") +
  scale_fill_manual(values = c(DMSO = "#99000D", dTag = "#FC9272")) +
  scale_shape_manual(values = c("Rep 1" = 21, "Rep 2" = 24)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = NULL, y = "TF footprint occupancy (%)",
       fill = "SNAI1/2", shape = NULL,
       title = "TF footprint occupancy at motif, SNAI1/2 DMSO vs dTag (1 min DAF)",
       subtitle = paste("bars = mean per-bp occupancy over the motif;",
                        "p = Cochran-Mantel-Haenszel on per-read footprint calls,",
                        "strata = replicates")) +
  theme_classic() +
  theme(legend.position  = "right",
        strip.background = element_blank(),
        strip.text       = element_text(face = "bold", size = 8),
        plot.title       = element_text(size = 9),
        plot.subtitle    = element_text(size = 7),
        panel.spacing    = unit(0.4, "lines"))

p_occ_cmh
ggsave(fig("TFoccupancy_bar_Enh2_Enh5_all_vs_accessible_CMH_v297.pdf"),
       p_occ_cmh, width = 5.5, height = 3.5)
```

# 8. Session info

```{r session-info}
sessionInfo()
```
