---
title: "ATAC-seq fold-change at distal summits as a function of distance to TF-motif-bound summits"
subtitle: "SNAI2 and TWIST1 degron (dTAG) analyses"
author: "Lucia Ichino"
date: "2026-07-30"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: flatly
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# About this document

Minimal version of the merged SNAI2 / TWIST1 analysis: only the code required to produce
the four figures is kept. For each factor:

1. read per-summit Tn5 cut counts (`bedtools coverage -counts`, one BED per sample),
2. DESeq2 on the summit counts, 3 h dTAG vs DMSO contrast,
3. define *bound* motifs = FIMO hits within ±100 bp of a ChIP-seq peak summit,
4. keep distal summits only (`enhancer` / `other_distal` in `data_v4.rds`),
5. bin summits by distance to the nearest motif-positive summit and plot mean log2FC ± SEM.

The two analyses differ only in their inputs, their DESeq2 design and the figure cosmetics,
so the shared steps live in one set of helper functions and the inputs are all declared in
the configuration chunk below.

# Configuration

All paths and all sample orders are here; nothing below this chunk needs editing to run on
another machine.

```{r config}
## ---- inputs shared by both analyses ---------------------------------------
annotation_rds <- ".../data_v4.rds"
distal_classes <- c("enhancer", "other_distal")

## ---- SNAI2 inputs ----------------------------------------------------------
snai2_counts_dir <- ".../ATAC_region_summits"
snai2_peak_file  <- ".../s8_lib1_CNCCp4_SNAI2a3029_DMSO_r2_aSNAI2_reltodTag_peaks_minusINPUT_filtered.narrowPeak"
snai2_fimo_bed   <- ".../SNAI2_DBD_jolma2013_fimo/fimo_sorted.bed"

# Column order matters: it is what makes the sampleTable below correct.
snai2_samples <- c(
  "ATAC_SNAI12_c5946b16_DMSO_rep1", "ATAC_SNAI12_c5946b16_3h_rep1",
  "ATAC_SNAI12_c5946b16_1d_rep1",   "ATAC_SNAI12_c5946b16_3d_rep1",
  "ATAC_SNAI12_c5946b27_DMSO_rep1", "ATAC_SNAI12_c5946b27_3h_rep1",
  "ATAC_SNAI12_c5946b27_1d_rep1",   "ATAC_SNAI12_c5946b27_3d_rep1",
  "ATAC_SNAI12_c5946b16_DMSO_rep2", "ATAC_SNAI12_c5946b16_3h_rep2",
  "ATAC_SNAI12_c5946b16_1d_rep2",   "ATAC_SNAI12_c5946b16_3d_rep2",
  "ATAC_SNAI12_c5946b27_DMSO_rep2", "ATAC_SNAI12_c5946b27_3h_rep2",
  "ATAC_SNAI12_c5946b27_1d_rep2",   "ATAC_SNAI12_c5946b27_3d_rep2"
)

## ---- TWIST1 inputs ---------------------------------------------------------
twist1_counts_dir <- ".../ATAC_region_summitsTW"
twist1_peak_file  <- ".../lib11_TWIST1_peaks_DMSOintersect_minusINPUT.narrowPeak"
twist1_fimo_bed   <- ".../coordinator_fimo/fimo_sorted.bed"

twist1_samples <- c(
  "GSM7213525_TWIST1FV_dTAG0h_r1_ATAC", "GSM7213528_TWIST1FV_dTAG0h_r2_ATAC",
  "GSM7213526_TWIST1FV_dTAG3h_r1_ATAC", "GSM7213529_TWIST1FV_dTAG3h_r2_ATAC"
)

## ---- outputs ---------------------------------------------------------------
fig_dir <- "/Users/luciaichino/Library/CloudStorage/Dropbox/lucia (1)/Lucia 09_05_16/postdoc/Wysocka_lab_personal_folder/SNAI_paper/figures_testing"

fig_snai2         <- file.path(fig_dir, "ATAC_foldchanges_bySNAIdist.pdf")
fig_snai2_nolabel <- file.path(fig_dir, "ATAC_foldchanges_bySNAIdist_notext.pdf")
fig_twist1         <- file.path(fig_dir, "ATAC_foldchanges_byTWIST1dist.pdf")
fig_twist1_nolabel <- file.path(fig_dir, "ATAC_foldchanges_byTWIST1dist_notext.pdf")

fig_width  <- 4
fig_height <- 3.5

dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

## ---- analysis parameters ---------------------------------------------------
motif_flank <- 100   # bp around each ChIP peak summit used to call a motif "bound"
min_count   <- 5     # DESeq2 pre-filter: counts >= min_count ...
dist_breaks <- c(0, 1, 300, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 3000, Inf)
x_max       <- 3050  # coord_cartesian upper limit
```

# Packages


```{r libraries}
library(GenomicRanges)
library(rtracklayer)
library(DESeq2)
library(tidyverse)
library(ggrepel)
```

# Helper functions

```{r helpers}
## Read every *_counts_ATAC_summits.bed in a directory and return one wide table
## (one column per sample, plus the chr_start_end `region` key).
read_summit_counts <- function(counts_dir, sample_order) {
  files <- list.files(counts_dir,
                      pattern = "_counts_ATAC_summits\\.bed$",
                      full.names = TRUE)
  stopifnot(length(files) > 0)

  long_df <- purrr::map_dfr(files, function(f) {
    df <- readr::read_tsv(f, col_names = FALSE, show_col_types = FALSE)
    tibble::tibble(
      region = paste(df$X1, df$X2, df$X3, sep = "_"),
      sample = sub("_counts.*", "", basename(f)),
      # last column is always the count from `bedtools coverage -counts`
      count  = df[[ncol(df)]]
    )
  })

  wide <- tidyr::pivot_wider(long_df, names_from = sample, values_from = count)
  unique(dplyr::select(wide, dplyr::all_of(sample_order), region))
}

## DESeq2 on the summit counts, returning the 3h-vs-DMSO contrast as a data frame
## with the region key in column `X` (the column name the downstream code expects).
deseq_3h_vs_dmso <- function(counts_tbl, sample_order, sample_table, design,
                             smallest_group_size, min_count = 5) {
  mat <- as.data.frame(counts_tbl[, sample_order])
  rownames(mat) <- counts_tbl$region
  rownames(sample_table) <- sample_order

  dds <- DESeqDataSetFromMatrix(countData = mat,
                                colData   = sample_table,
                                design    = design)

  keep <- rowSums(counts(dds) >= min_count) >= smallest_group_size
  dds  <- dds[keep, ]
  dds  <- DESeq(dds)

  res <- results(dds, contrast = c("condition", "3h", "DMSO"))
  res <- res[order(res$padj), ]
  tibble::rownames_to_column(as.data.frame(res), "X")
}

## FIMO motif hits restricted to +/- `flank` bp around ChIP-seq peak summits.
bound_motifs <- function(peak_file, fimo_bed, flank = 100) {
  peaks_gr <- rtracklayer::import(peak_file, format = "narrowPeak")

  fimo_df <- read.table(fimo_bed, sep = "\t", header = FALSE,
                        stringsAsFactors = FALSE, comment.char = "#")

  motifs_gr <- GRanges(
    seqnames = fimo_df$V1,
    ranges   = IRanges(start = fimo_df$V2 + 1, end = fimo_df$V3),  # BED is 0-based
    strand   = if (ncol(fimo_df) >= 6) fimo_df$V6 else "*"
  )

  summit_pos <- start(peaks_gr) + peaks_gr$peak
  summit_windows <- GRanges(
    seqnames = seqnames(peaks_gr),
    ranges   = IRanges(start = summit_pos - flank, end = summit_pos + flank)
  )

  subsetByOverlaps(motifs_gr, summit_windows)
}

## `data_v4.rds` -> GRanges of distal regions only.
load_distal_regions <- function(rds_file, classes) {
  ann <- readRDS(rds_file)
  parts <- do.call(rbind, strsplit(as.character(ann$region), "_"))
  gr <- GRanges(
    seqnames = parts[, 1],
    ranges   = IRanges(start = as.numeric(parts[, 2]) + 1,   # region strings are 0-based here
                       end   = as.numeric(parts[, 3])),
    classification = ann$classification
  )
  gr[gr$classification %in% classes]
}

## Distal summits + distance from each summit centre to the nearest
## motif-positive summit centre. Returns a data frame with NA distances dropped.
distal_summits_with_distance <- function(res_df, distal_gr, motifs_gr) {
  ## res_df$X is chr_start_end and is already 1-based (it comes from a GRanges),
  ## so no +1 on start here.
  coords <- stringr::str_split_fixed(res_df$X, "_", 3)

  res_gr <- GRanges(
    seqnames = coords[, 1],
    ranges   = IRanges(start = as.numeric(coords[, 2]),
                       end   = as.numeric(coords[, 3])),
    log2FoldChange = res_df$log2FoldChange
  )

  res_gr <- res_gr[overlapsAny(res_gr, distal_gr)]
  res_gr$has_motif <- overlapsAny(res_gr, motifs_gr)

  ## regions are 300bp summit-centred windows, so the midpoint IS the summit
  centers_gr <- GRanges(
    seqnames = seqnames(res_gr),
    ranges   = IRanges(start = round((start(res_gr) + end(res_gr)) / 2), width = 1),
    strand   = "*"
  )
  mcols(centers_gr) <- mcols(res_gr)

  motif_centers_gr <- centers_gr[centers_gr$has_motif]
  hits <- distanceToNearest(centers_gr, motif_centers_gr, ignore.strand = TRUE)

  dist_to_motif <- rep(NA_integer_, length(centers_gr))
  dist_to_motif[queryHits(hits)] <- mcols(hits)$distance
  res_gr$dist_to_nearest_motif <- dist_to_motif

  out <- as.data.frame(res_gr)
  out[!is.na(out$dist_to_nearest_motif), ]   # summits on contigs with no motif+ region
}

## Mean log2FC +/- SEM per distance bin, with the bin labels used on the plot.
bin_by_distance <- function(plot_df, breaks) {
  plot_df %>%
    dplyr::mutate(
      bin_idx  = cut(dist_to_nearest_motif, breaks = breaks,
                     include.lowest = TRUE, labels = FALSE),
      dist_bin = cut(dist_to_nearest_motif, breaks = breaks, include.lowest = TRUE)
    ) %>%
    dplyr::group_by(dist_bin, bin_idx) %>%
    dplyr::summarise(
      mean_l2fc = mean(log2FoldChange),
      sem       = sd(log2FoldChange) / sqrt(dplyr::n()),
      mid       = median(dist_to_nearest_motif),
      n         = dplyr::n(),
      .groups   = "drop"
    ) %>%
    dplyr::mutate(
      bin_low   = breaks[bin_idx],
      bin_high  = breaks[bin_idx + 1],
      bin_label = ifelse(is.infinite(bin_high),
                         paste0(">", bin_low, "bp"),
                         paste0(bin_low, "-", bin_high, "bp")),
      label     = paste0(bin_label, "\nn=", n)
    )
}

## The figure. `show_labels = FALSE` gives the clean panel version.
plot_binned <- function(binned_df, colour, x_lab, y_lab,
                        show_labels = TRUE, xmax = 3050) {
  p <- ggplot(binned_df, aes(x = mid, y = mean_l2fc)) +
    geom_ribbon(aes(ymin = mean_l2fc - sem, ymax = mean_l2fc + sem),
                alpha = 0.3, fill = colour) +
    geom_line(color = colour, linewidth = 1) +
    geom_point(size = 2, color = colour)

  if (show_labels) {
    p <- p + geom_text_repel(aes(label = label), size = 3.6, lineheight = 0.85,
                             min.segment.length = 0, segment.size = 0.3,
                             box.padding = 0.3, max.overlaps = Inf)
  }

  p +
    labs(x = x_lab, y = y_lab) +
    ggtitle("ATAC-seq foldchange at distal ATAC summits") +
    theme_bw() +
    coord_cartesian(xlim = c(0, xmax))
}
```

# Distal region annotation

Shared by both analyses, so it is loaded once.

```{r load-annotation}
distal_gr <- load_distal_regions(annotation_rds, distal_classes)
```

# SNAI2 analysis

Degron of SNAI2 in CNCC-like cells, two clones (`c5946b16`, `c5946b27`), two replicates,
four time points → 16 libraries. `condition` cycles through the four time points every
4 columns, `clone` splits the first 8 columns from the last 8, `replicate` alternates in
blocks of 4 — which is exactly the column order set in `snai2_samples`.

```{r snai2-deseq}
snai2_counts <- read_summit_counts(snai2_counts_dir, snai2_samples)

snai2_sample_table <- data.frame(
  condition = factor(rep(c("DMSO", "3h", "1day", "3days"), 4),
                     levels = c("DMSO", "3h", "1day", "3days")),
  clone     = factor(rep(c("c5946b16", "c5946b27"), each = 8)),
  replicate = factor(rep(rep(c("rep1", "rep2"), each = 4), 2))
)

snai2_res_3h <- deseq_3h_vs_dmso(
  counts_tbl          = snai2_counts,
  sample_order        = snai2_samples,
  sample_table        = snai2_sample_table,
  design              = ~ clone + replicate + condition,
  smallest_group_size = 4,
  min_count           = min_count
)
```

```{r snai2-distance}
snai2_motifs <- bound_motifs(snai2_peak_file, snai2_fimo_bed, flank = motif_flank)

snai2_plot_df   <- distal_summits_with_distance(snai2_res_3h, distal_gr, snai2_motifs)
snai2_binned_df <- bin_by_distance(snai2_plot_df, dist_breaks)
```

Motif-positive summits are in the query set as well, so they get distance 0 — that is what
the leading `0–1 bp` bin isolates.

```{r snai2-plot, fig.width=4, fig.height=3.5}
snai2_xlab <- "Distance to nearest SNAI2-motif region (bp, bin midpoint)"
snai2_ylab <- "Mean ATAC log2FC (dTag/DMSO) ± SEM"

p_snai2 <- plot_binned(snai2_binned_df, "firebrick", snai2_xlab, snai2_ylab,
                       show_labels = TRUE, xmax = x_max)
p_snai2
ggsave(fig_snai2, plot = p_snai2, width = fig_width, height = fig_height)

p_snai2_nolabel <- plot_binned(snai2_binned_df, "firebrick", snai2_xlab, snai2_ylab,
                               show_labels = FALSE, xmax = x_max)
p_snai2_nolabel
ggsave(fig_snai2_nolabel, plot = p_snai2_nolabel, width = fig_width, height = fig_height)
```

# TWIST1 analysis

Same pipeline on the public TWIST1-FV degron ATAC data (GEO `GSM7213525/6/8/9`): two
conditions, two replicates, no clone term, so the design is `~ replicate + condition` and
the pre-filter group size drops to 2. Motifs here are coordinator FIMO hits restricted to
TWIST1 ChIP peak summits.

```{r twist1-deseq}
twist1_counts <- read_summit_counts(twist1_counts_dir, twist1_samples)

twist1_sample_table <- data.frame(
  condition = factor(rep(c("DMSO", "3h"), each = 2), levels = c("DMSO", "3h")),
  replicate = factor(rep(c("rep1", "rep2"), 2))
)

twist1_res_3h <- deseq_3h_vs_dmso(
  counts_tbl          = twist1_counts,
  sample_order        = twist1_samples,
  sample_table        = twist1_sample_table,
  design              = ~ replicate + condition,
  smallest_group_size = 2,
  min_count           = min_count
)
```

```{r twist1-distance}
twist1_motifs <- bound_motifs(twist1_peak_file, twist1_fimo_bed, flank = motif_flank)

twist1_plot_df   <- distal_summits_with_distance(twist1_res_3h, distal_gr, twist1_motifs)
twist1_binned_df <- bin_by_distance(twist1_plot_df, dist_breaks)
```

```{r twist1-plot, fig.width=4, fig.height=3.5}
twist1_xlab <- "Distance to nearest TWIST1-motif region (bp, bin midpoint)"
twist1_ylab <- "Mean ATAC log2FC (TWIST1-FV dTag/DMSO) ± SEM"

p_twist1 <- plot_binned(twist1_binned_df, "steelblue", twist1_xlab, twist1_ylab,
                        show_labels = TRUE, xmax = x_max)
p_twist1
ggsave(fig_twist1, plot = p_twist1, width = fig_width, height = fig_height)

p_twist1_nolabel <- plot_binned(twist1_binned_df, "steelblue", twist1_xlab, twist1_ylab,
                                show_labels = FALSE, xmax = x_max)
p_twist1_nolabel
ggsave(fig_twist1_nolabel, plot = p_twist1_nolabel, width = fig_width, height = fig_height)
```

# Session info

```{r session-info}
sessionInfo()
```
