---
title: 'Curate motifs from the human CNCC ChromBPNet model'
author: "Kaelan Brennan"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    depth: 3
    theme: sandstone
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyfoot[CO,CE]{1028-02098-001-001}
- \fancyfoot[LE,RO]{\thepage}
- \fancypagestyle{plain}{\pagestyle{fancy}}
editor_options:
  chunk_output_type: console
---

# 1. Introduction

Here, I will curate the CWMs of interest from a model trained on DMSO-treated CNCC ATAC-seq data. Three single-task models (DMSO, dTAG-3h, dTAG-1d) were trained across the same region set derived from CNCC ATAC-seq peaks. CWMs were clustered using tfmodisco-lite and were mapped using FINEMO.In this analysis, I will collect them, fix any width issues, remove redundant patterns, and calculate .bw-derived coverage/contribution scores.

# 2. Computational Setup

```{r, warning=F, message=F}

#Standard packages

library(rtracklayer) ; library(GenomicRanges); library(magrittr) ; library(Biostrings); library(ggseqlogo);
library(ggplot2) ; library(reshape2); library(plyranges); library(Rsamtools); library(ggpubr); library(BSgenome.Hsapiens.UCSC.hg38); library(parallel); library(tidyr); library(tibble); library(ggrastr); library(dplyr); library(data.table); library(patchwork); library(readr); library(testit); library(cowplot); library(universalmotif)

setwd("/storage/kaelanb/analysis/lucia/analysis/")
options(knitr.figure_dir="/storage/kaelanb/analysis/lucia/analysis/figures/1_curate_motifs/", java.parameters = "- Xmx6g")

#Sourced code

source("/storage/kaelanb/analysis/oct4_public_data/scripts/knitr_common.r")
source("/storage/kaelanb/analysis/oct4_public_data/scripts/caching.r")

#Pre-existing variables

library(TxDb.Hsapiens.UCSC.hg38.knownGene)
bsgenome <- BSgenome.Hsapiens.UCSC.hg38
rmsk <- rtracklayer::import("/work/kaelanb/indexes/hg38_rmsk.bed") %>% keepStandardChromosomes(pruning.mode="coarse")
blacklist <- rtracklayer::import("/work/kaelanb/indexes/hg38-blacklist.v2.bed")
chrombpnet.blacklisted.gr <- import("/work/kaelanb/public_data/chrombpnet_tutorial/data/downloads/blacklist.bed.gz")
promoters_txdb <- promoters(genes(TxDb.Hsapiens.UCSC.hg38.knownGene), upstream = 250, downstream = 250) %>% keepStandardChromosomes(pruning.mode="coarse")

```

# 3. Import relevant data

```{r import}

# ChromBPNet training regions

cncc_all_peaks.gr <- import('/storage/kaelanb/analysis/lucia/data/data_forKaelan/20260607_cncc_training_atac_peaks.narrowpeak')
cncc_all_peaks.path <- "/storage/kaelanb/analysis/lucia/data/data_forKaelan/20260607_cncc_training_atac_peaks.narrowpeak"

# Contribution scores

atac_contrib.list <- list(all_peaks_DMSO = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/bw/all_peaks.counts_scores.bw",
                          all_peaks_dTAG_3h = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/dtag_3h/all_peaks/bw/all_peaks.counts_scores.bw",
                          all_peaks_dTAG_1d = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/dtag_1d/all_peaks/bw/all_peaks.counts_scores.bw")

# Observed ATAC-seq data

atac_obs.list <- list(DMSO = "/storage/kaelanb/analysis/lucia/data/bw/ATAC_DMSO_merged.bw",
                      dTAG_3h = "/storage/kaelanb/analysis/lucia/data/bw/ATAC_3h_merged.bw",
                      dTAG_1d = "/storage/kaelanb/analysis/lucia/data/bw/ATAC_1d_merged.bw")

```

# 4. Select patterns of interest

Note that patterns are selected from: `/storage/kaelanb/analysis/lucia/model/cncc_all_peaks/modisco/trimmed_logos/` based on prior knowledge of motifs and seqlet counts.

## 4.1. Import patterns of interest

### 4.1.1. Define peak-level information 

```{r level}

region_width <- 1000  #parameters.json informed parameter

peaks.qc <- readr::read_tsv("/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/hitcall/all_peaks/peaks_qc.tsv") %>%
  dplyr::mutate(peak_center = peak_region_start + region_width / 2) %>%
  dplyr::select(peak_id, peak_center)

central_width <- 400   # Standard Modisco focus point
half_central <- central_width / 2

```

### 4.1.2. Import

```{r patterns}

generate_pattern_vec <- function(numbers) {
  paste0("pos_patterns.pattern_", numbers)
}

# All peaks

all_pattern.vec <- generate_pattern_vec(c(1, 2, 3, 5, 6, 7, 8, 9, 10, 11, 13, 16, 17, 19, 23, 28, 31, 32))

all_cwms.gr <- readr::read_tsv('/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/hitcall/all_peaks/hits_unique.tsv') %>% 
  dplyr::left_join(peaks.qc, by = "peak_id") %>%
  dplyr::mutate(hit_center = (start + end) / 2,
                dist_from_peak_center = abs(hit_center - peak_center)) %>%
  makeGRangesFromDataFrame(keep.extra.columns = T) %>% 
  dplyr::filter(motif_name %in% all_pattern.vec | motif_name == "neg_patterns.pattern_6") %>%
  dplyr::mutate(motif_name = case_when(
    motif_name == "pos_patterns.pattern_1" ~ "Coordinator",
    motif_name == "pos_patterns.pattern_2" ~ "TFAP2A",
    motif_name == "pos_patterns.pattern_3" ~ "SP",
    motif_name == "pos_patterns.pattern_5" ~ "NR2F",
    motif_name == "pos_patterns.pattern_6" ~ "CTCF",
    motif_name == "pos_patterns.pattern_7" ~ "SIX",
    motif_name == "pos_patterns.pattern_8" ~ "ETV",
    motif_name == "pos_patterns.pattern_9" ~ "TEAD",
    motif_name == "pos_patterns.pattern_10" ~ "ASCL1",
    motif_name == "pos_patterns.pattern_11" ~ "SOX9_4bp",
    motif_name == "pos_patterns.pattern_13" ~ "SOX9_3bp",
    motif_name == "pos_patterns.pattern_16" ~ "RFX",
    motif_name == "pos_patterns.pattern_17" ~ "NRF",
    motif_name == "pos_patterns.pattern_19" ~ "NFY",
    motif_name == "pos_patterns.pattern_23" ~ "ZN787",
    motif_name == "pos_patterns.pattern_28" ~ "ZNF",
    motif_name == "pos_patterns.pattern_31" ~ "CREB",
    motif_name == "pos_patterns.pattern_32" ~ "REST",
    motif_name == "neg_patterns.pattern_6" ~ "SNAI2",
    TRUE ~ motif_name))

all_cwms.gr$dist_from_peak_center %>% summary()

all_cwms.gr$motif_name %>% table()

all_cwms.gr %>%
  dplyr::group_by(motif_name) %>%
  dplyr::summarise(mean_dist = mean(dist_from_peak_center),
                   median_dist = median(dist_from_peak_center)) %>%
  as.data.frame()

```

## 4.2. Curate motif redundancy

Prevent motif double-counting and remove palindromic patterns. 

```{r curate}

# Function: kudos Melanie Weilert, Zeitlinger lab

remove_palindromic_motifs_from_bpnet_instances<-function(dfi, pattern_to_filter,
                                                         chrom_name = 'example_chrom',
                                                         start_name = 'pattern_start_abs',
                                                         end_name = 'pattern_end_abs',
                                                         value_name = 'contrib_weighted',
                                                         motif_name = 'pattern_name',
                                                         region_name = 'example_idx',
                                                         starts.in.df.are.0based = F){

  if(class(dfi)[1]=="GRanges"){
    dfi.gr<-dfi
  } else{
    #Convert to GRanges
    dfi.gr<-makeGRangesFromDataFrame(dfi, keep.extra.columns = T, starts.in.df.are.0based = starts.in.df.are.0based,
                                     seqnames.field = chrom_name, start.field = start_name, end.field = end_name)
  }

  #Subset by motifs
  dfi_filt.gr<-dfi.gr %>% dplyr::filter(!!rlang::sym(motif_name)==pattern_to_filter)

  #Create islands of overlapping motifs
  dfi_islands.gr<-GenomicRanges::reduce(dfi_filt.gr, ignore.strand = T)
  dfi_islands.gr$island_idx<-1:length(dfi_islands.gr)

  #Remap island ids to the motifs
  dfi_vs_islands.ov<-findOverlaps(dfi_filt.gr, dfi_islands.gr, ignore.strand = T)
  testit::assert("Motifs overlap redundantly...", (dfi_vs_islands.ov@from %>% unique() %>% length) == (dfi_vs_islands.ov@from %>% length))
  dfi_filt.gr$island_idx<-dfi_islands.gr$island_idx[dfi_vs_islands.ov@to]

  #For each island id, select motif with the best importance
  dfi_unique.gr <- dfi_filt.gr %>% as.data.frame %>%
    dplyr::group_by(island_idx, !!rlang::sym(region_name)) %>% #Group by both island_id and example_idx to keep the same motifs mapped across different task-annotated windows
    dplyr::slice_max(order_by = !!rlang::sym(value_name), n = 1, with_ties = FALSE) %>% as.data.frame() %>%
    makeGRangesFromDataFrame(keep.extra.columns = T, starts.in.df.are.0based = F) %>%
    plyranges::select(-island_idx)

  message(length(dfi_filt.gr)-length(dfi_unique.gr), " palindromic motifs removed of ", length(dfi_filt.gr))
  return(dfi_unique.gr)
}

# Import and reduce peaks
all_peaks.gr <- rtracklayer::import(cncc_all_peaks.path) %>%
  GenomicRanges::resize(., 2032, fix = "center") %>%
  GenomicRanges::reduce(., ignore.strand = TRUE) %>%
  plyranges::mutate(region_id_tmp = 1:length(.))

# Prepare GRanges
all_cncc_cwms_dedup.gr <- all_cwms.gr %>%
  plyranges::mutate(
    start = start(.),
    end = end(.),
    region_id_tmp = NA
  )

# Map motifs to peak regions
ov <- findOverlaps(all_cncc_cwms_dedup.gr, all_peaks.gr, ignore.strand = TRUE)
all_cncc_cwms_dedup.gr$region_id_tmp[queryHits(ov)] <- all_peaks.gr$region_id_tmp[subjectHits(ov)]

# Remove motifs outside peaks
all_cncc_cwms_dedup.gr <- all_cncc_cwms_dedup.gr %>% 
  plyranges::filter(!is.na(region_id_tmp))

# Deduplicate palindromic motifs per motif name
all_cncc_cwms_dedup.gr <- lapply(unique(all_cncc_cwms_dedup.gr$motif_name), function(motif) {
  motifs.gr <- all_cncc_cwms_dedup.gr %>% plyranges::filter(motif_name == motif)
  motifs_dedup.gr <- remove_palindromic_motifs_from_bpnet_instances(
    dfi = motifs.gr,
    pattern_to_filter = motif,
    chrom_name = "seqnames",
    start_name = "start",
    end_name = "end",
    value_name = "hit_importance",
    motif_name = "motif_name",
    region_name = "region_id_tmp",
    starts.in.df.are.0based = FALSE
  )
  message(length(motifs_dedup.gr), " kept from ", length(motifs.gr), " for motif ", motif)
  return(motifs_dedup.gr)
}) %>% 
  as("GRangesList") %>% 
  unlist() %>%
  plyranges::select(-region_id_tmp)

```

## 4.3. Summarize so far

```{r sum}

all_cncc_cwms_dedup.gr <- all_cncc_cwms_dedup.gr %>%
  dplyr::select(-start_untrimmed, -end_untrimmed)

all_cncc_cwms_dedup.gr$motif_name %>% table()

```

## 4.4. Fix motif offset

This is due to the mapping treating + and - strand motifs differently, thus creating a 1bp offfset with respect to strand

```{r offset}

# Function to fix

shift_minus_5p <- function(gr, bp = 1) {
  idx <- as.character(strand(gr)) == "-"
  
  start(gr)[idx] <- start(gr)[idx] + bp
  end(gr)[idx]   <- end(gr)[idx]   + bp
  
  gr
}

# Fix

all_cncc_cwms_dedup.gr <- all_cncc_cwms_dedup.gr %>%
  shift_minus_5p(., 1)

```

## 4.5 Remove sub-motifs that are truly part of a larger motif

Motifs may contain sub-components (e.g., the Double-Ebox is actually part of a Coordinator)

  + Coordinator > Double_Ebox
  + SOX9_4bp > SOX9_3bp
  + Double_Ebox > SNAI2
  + Coordinator > SNAI2

```{r submotifs}

sub_pairs <- list(
  c("SOX9_3bp", "SOX9_4bp"),
  c("SNAI2", "Coordinator"))

# Write function to prioritize the selected motif

remove_redundant_motif <- function(gr, small, composite) {
  small.gr <- gr %>% plyranges::filter(motif_name == small)
  composite.gr <- gr %>% plyranges::filter(motif_name == composite)

  if (length(small.gr) == 0 || length(composite.gr) == 0) {
    return(gr)
  }

  small.keep <- small.gr %>%
    plyranges::filter(
      !overlapsAny(., composite.gr, ignore.strand = TRUE)
    )

  removed_n <- length(small.gr) - length(small.keep)
  message(
    paste0(
      small, " motifs removed due to overlap with ",
      composite, ": ", removed_n
    )
  )

  # Return GRanges with redundant small motifs removed
  c(
    gr %>% plyranges::filter(motif_name != small),
    small.keep
  )
}

# Remove sub-motifs

for (pair in sub_pairs) {
  all_curated_cncc_cwms.gr <- remove_redundant_motif(
    all_cncc_cwms_dedup.gr,
    small = pair[1],
    composite = pair[2]
  )
}

```

## 4.5. Fix motif lengths

I noticed that REST, Coordinator, and TFAP2A were arbitrarily a bit shorter upon CWM scanning by FINEMO. I will correct them here based on the consensus patterns.

```{r fix}

# REST: extend by 1bp on both the 5' end and the 3' end
rest_idx <- which(all_curated_cncc_cwms.gr$motif_name == "REST")
rest_gr <- all_curated_cncc_cwms.gr[rest_idx]

rest_gr <- resize(rest_gr, width = width(rest_gr) + 1, fix = "start")  
rest_gr <- resize(rest_gr, width = width(rest_gr) + 1, fix = "end")

all_curated_cncc_cwms.gr[rest_idx] <- rest_gr

# Coordinator: extend by 1bp on the 3' end only

coord_idx <- which(all_curated_cncc_cwms.gr$motif_name == "Coordinator")
coord_gr <- all_curated_cncc_cwms.gr[coord_idx]

coord_gr <- resize(coord_gr, width = width(coord_gr) + 1, fix = "start")

all_curated_cncc_cwms.gr[coord_idx] <- coord_gr

# TFAP2A: shift one bp towards 5'

tfap2a_idx <- which(all_curated_cncc_cwms.gr$motif_name == "TFAP2A")
tfap2a_gr <- all_curated_cncc_cwms.gr[tfap2a_idx]

tfap2a_gr <- resize(tfap2a_gr, width = width(tfap2a_gr) + 1, fix = "start") 
tfap2a_gr <- resize(tfap2a_gr, width = width(tfap2a_gr) - 1, fix = "end")

all_curated_cncc_cwms.gr[tfap2a_idx] <- tfap2a_gr

```

# 5. Append useful information

Here, I will include the following annotations:

  + Whether a pattern overlaps an annotated promoter (promoter_ov)
  + Motif sequence (motif_seq)

Additionally, I will calculate the following genomic signals:

  + ATAC-seq Counts Contribution from DMSO treatment
  + ATAC-seq Counts Contribution from SNAI2 dTAG treatment for 3h
  + ATAC-seq Counts Contribution from SNAI2 dTAG treatment for 1d
  + Observed ATAC-seq from all conditions in a 500bp window surrounding the motif

I will also annotate SNAI2 motifs as bound or not based on ChIP-seq data

```{r calc}

# Define functions

check_coverage_argument <- function(cvg, regions=NULL) {
  if(class(cvg) == "character") {
    if(is.null(regions)) {
      cvg <- rtracklayer::import(cvg, as="RleList")
    } else {
      stopifnot(file.exists(cvg))
      cvg <- rtracklayer::import(cvg, which=regions, as="RleList")
    }
  } 
  cvg
}

regionSums <- function(regions, cvg) {
  cvg <- check_coverage_argument(cvg, regions)
  seqlevels(regions) <- names(cvg)
  oo <- order(regions)
  regions <- regions[oo]
  ans <- unlist(
           viewSums(
             #Views(cvg, as(regions, "GRangesList"))
             Views(cvg, split(regions, seqnames(regions)))
           ), use.names=FALSE)
  ans[oo] <- ans  # restore original order
  ans
}

# Calculate

all_curated_cncc_cwms.gr <- all_curated_cncc_cwms.gr %>%
  dplyr::mutate(promoter_ov = overlapsAny(., promoters_txdb)) %>%
  dplyr::mutate(motif_seq = getSeq(bsgenome, ., as.character = T)) %>%
  dplyr::mutate(DMSO_obs_atac = regionSums(resize(., 500, "center"), atac_obs.list$DMSO),
                dTAG_3h_obs_atac = regionSums(resize(., 500, "center"), atac_obs.list$dTAG_3h),
                dTAG_1d_obs_atac = regionSums(resize(., 500, "center"), atac_obs.list$dTAG_1d)) %>%
  dplyr::mutate(DMSO_contrib = regionSums(., atac_contrib.list$all_peaks_DMSO),
                dTAG_3h_contrib = regionSums(., atac_contrib.list$all_peaks_dTAG_3h),
                dTAG_1d_contrib = regionSums(., atac_contrib.list$all_peaks_dTAG_1d)) %>%
  dplyr::mutate(DMSO_contrib_norm = DMSO_contrib / width(all_curated_cncc_cwms.gr),
                dTAG_3h_contrib_norm = dTAG_3h_contrib / width(all_curated_cncc_cwms.gr),
                dTAG_1d_contrib_norm = dTAG_1d_contrib / width(all_curated_cncc_cwms.gr)) %>%
  dplyr::mutate(dTAG_3h_diff_DMSO = dTAG_3h_contrib - DMSO_contrib,
                dTAG_1d_diff_DMSO = dTAG_1d_contrib - DMSO_contrib)

# Add SNAI2 binding information

all_curated_cncc_cwms.gr <- all_curated_cncc_cwms.gr %>%
  plyranges::mutate(
    bound = dplyr::case_when(motif_name == "SNAI2" ~ overlapsAny(., import("/storage/lucia/SNAI_paper/peaks/s8_lib1_CNCCp4_SNAI2a3029_DMSO_r2_aSNAI2_reltodTag_peaks_minusINPUT_filtered.narrowPeak"), maxgap = 2000), TRUE ~ NA))

# Quick summary

all_curated_cncc_cwms.gr %>%
  dplyr::group_by(motif_name) %>%
  dplyr::summarise(DMSO_contrib_mean = mean(DMSO_contrib),
                   dTAG_3h_contrib_mean = mean(dTAG_3h_contrib),
                   dTAG_1d_contrib_mean = mean(dTAG_1d_contrib),
                   DMSO_contrib_median = median(DMSO_contrib),
                   dTAG_3h_contrib_median = median(dTAG_3h_contrib),
                   dTAG_1d_contrib_median = median(dTAG_1d_contrib)
                   ) %>%
  as.data.frame()

# Export data as RData, bed, and CSV

saveRDS(all_curated_cncc_cwms.gr, "/storage/kaelanb/analysis/lucia/motifs/all_peaks_curated_cncc_cwms_w_metadata.rdata")
as.data.frame(all_curated_cncc_cwms.gr) %>% readr::write_csv("/storage/kaelanb/analysis/lucia/motifs/all_peaks_curated_cncc_cwms_w_metadata.csv")

## Export as bed for IGV viewing

all_curated_cncc_cwms_w_metadata.bed <- all_curated_cncc_cwms.gr %>%
  as.data.frame() %>%
  dplyr::rename(name = motif_name)
rtracklayer::export(all_curated_cncc_cwms_w_metadata.bed, "/storage/kaelanb/analysis/lucia/motifs/all_peaks_curated_cncc_cwms.bed")

# Summarize motif widths

as.data.frame(all_curated_cncc_cwms.gr) %>%
  group_by(motif_name) %>%
  summarise(
    widths = paste(unique(width), collapse = ", ")
  )

```

# 6. Plot Summary plots

## 6.1. Write functions

```{r func}

filter_motif_strand_position <- function(df, motif_settings) {
  
  df %>%
    dplyr::rowwise() %>%
    dplyr::filter(
      motif_name %in% names(motif_settings),
      strand == motif_settings[[motif_name]]$strand,
      position %in% motif_settings[[motif_name]]$positions
    ) %>%
    dplyr::ungroup()
}

```

## 6.2. Get the number and name panel

```{r panel}

# Order by contribution

contrib.order <- all_curated_cncc_cwms.gr %>%
  dplyr::group_by(motif_name) %>%
  dplyr::summarise(mean_DMSO = mean(DMSO_contrib)) %>%
  as.data.frame() %>%
  dplyr::arrange(mean_DMSO) %>%
  dplyr::select(motif_name) %>%
  .$motif_name 

# Count patterns per motif_name

pattern_counts <- all_curated_cncc_cwms.gr %>%
  as.data.frame() %>%
  dplyr::count(motif_name, name = "n_patterns")

motif_order <- rev(contrib.order)

label_df <- pattern_counts %>%
  mutate(
    motif_name = factor(motif_name, levels = motif_order),
    label = paste0(motif_name, "\n(n = ", n_patterns, ")")
  ) %>%
  arrange(motif_name)

p_labels <- ggplot(label_df, aes(x = 1, y = 1, label = label)) +
  geom_text(fontface = "plain", size = 3.8) +
  facet_wrap(~motif_name, ncol = 1, strip.position = "left") +
  theme_void() +
  theme(strip.text = element_blank())

```

## 6.3. Plot all CWMS

```{r cwm}

filter_motif_strand_position <- function(df, motif_settings) {
  
  df %>%
    dplyr::rowwise() %>%
    dplyr::filter(
      motif_name %in% names(motif_settings),
      strand == motif_settings[[motif_name]]$strand,
      position %in% motif_settings[[motif_name]]$positions
    ) %>%
    dplyr::ungroup()
}

# Get info and clean up

cwm_long <- read_tsv("/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/hitcall/all_peaks/cwms_long_full.tsv") %>%
  dplyr::filter(motif_name %in% all_pattern.vec | motif_name == "neg_patterns.pattern_6") %>%
  dplyr::mutate(motif_name = case_when(
    motif_name == "pos_patterns.pattern_1" ~ "Coordinator",
    motif_name == "pos_patterns.pattern_2" ~ "TFAP2A",
    motif_name == "pos_patterns.pattern_3" ~ "SP",
    motif_name == "pos_patterns.pattern_5" ~ "NR2F",
    motif_name == "pos_patterns.pattern_6" ~ "CTCF",
    motif_name == "pos_patterns.pattern_7" ~ "SIX",
    motif_name == "pos_patterns.pattern_8" ~ "ETV",
    motif_name == "pos_patterns.pattern_9" ~ "TEAD",
    motif_name == "pos_patterns.pattern_10" ~ "ASCL1",
    motif_name == "pos_patterns.pattern_11" ~ "SOX9_4bp",
    motif_name == "pos_patterns.pattern_13" ~ "SOX9_3bp",
    motif_name == "pos_patterns.pattern_16" ~ "RFX",
    motif_name == "pos_patterns.pattern_17" ~ "NRF",
    motif_name == "pos_patterns.pattern_19" ~ "NFY",
    motif_name == "pos_patterns.pattern_23" ~ "ZN787",
    motif_name == "pos_patterns.pattern_28" ~ "ZNF",
    motif_name == "pos_patterns.pattern_31" ~ "CREB",
    motif_name == "pos_patterns.pattern_32" ~ "REST",
    motif_name == "neg_patterns.pattern_6" ~ "SNAI2",
    TRUE ~ motif_name))

motif_settings <- list(
  Coordinator = list(
    strand = "+",
    positions = 22:38
  ),
  TFAP2A = list(
    strand = "+",
    positions = 20:29
  ),
  SP = list(
    strand = "+",
    positions = 20:30
  ),
  NR2F = list(
    strand = "-",
    positions = 14:20
  ),
  CTCF = list(
    strand = "-",
    positions = 22:36
  ),
  SIX = list(
    strand = "-",
    positions = 25:37
  ),
  ETV = list(
    strand = "+",
    positions = 17:24
  ),
  TEAD = list(
    strand = "-",
    positions = 16:23
  ),
  ASCL1 = list(
    strand = "+",
    positions = 24:31
  ),
  SOX9_4bp = list(
    strand = "-",
    positions = 17:33
  ),
  SOX9_3bp = list(
    strand = "-",
    positions = 19:34
  ),
  RFX = list(
    strand = "+",
    positions = 21:35
  ),
  NRF = list(
    strand = "-",
    positions = 22:32
  ),
  NFY = list(
    strand = "-",
    positions = 16:23
  ),
  ZN787 = list(
    strand = "+",
    positions = 13:25
  ),
  ZNF = list(
    strand = "-",
    positions = 17:36
  ),
  CREB = list(
    strand = "+",
    positions = 22:30
  ),
  REST = list(
    strand = "+",
    positions = 11:29
  ),
  SNAI2 = list(
    strand = "-",
    positions = 17:23
  )
)

cwm_long <- filter_motif_strand_position(cwm_long, motif_settings)

# Format

library(tidyverse)

motif_list <- cwm_long %>%
  arrange(position) %>%
  group_by(motif_name) %>%
  group_split(.keep = TRUE) %>%
  set_names(map_chr(., ~ as.character(unique(.x$motif_name)))) %>%
  map(function(df) {
    mat <- t(as.matrix(df[, c("A_scaled", "C_scaled", "G_scaled", "T_scaled")]))
    rownames(mat) <- c("A", "C", "G", "T")
    mat
  })

motif_list <- motif_list[rev(contrib.order)]

# Plot

seq.plot <- ggseqlogo(motif_list, method = "custom", ncol = 1, scales = "free") +
  ylab("Contribution score (motif_scale-corrected)") +
  theme_cowplot() +
  ylab("ChromBPNet Counts Contribution") + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), strip.background = element_blank(), strip.text = element_blank(), axis.line.x = element_blank())

```

## 6.4. Plot motif contribution summary

```{r summary}

contrib_summary.df <- all_curated_cncc_cwms.gr %>%
  dplyr::group_by(motif_name) %>%
  dplyr::summarise(DMSO_contrib_median = mean(DMSO_contrib),
                   dTAG_3h_contrib_median = mean(dTAG_3h_contrib),
                   dTAG_1d_contrib_median = mean(dTAG_1d_contrib)) %>%
  as.data.table() %>%
  melt.data.table(id.vars = "motif_name", measure.vars = patterns("_contrib_median") ,value.name = 'value', variable.name = 'signal') %>%
  mutate(signal_type = gsub("_contrib_median", "", signal))

contrib_summary.df$motif_name <- factor(contrib_summary.df$motif_name, levels = contrib.order)
contrib_summary.df$signal_type <- factor(contrib_summary.df$signal_type, levels = c("DMSO", "dTAG_3h", "dTAG_1d"))

contrib.plot <- ggplot(contrib_summary.df, aes(x = signal_type, y = motif_name, fill = value)) +
  geom_tile(color = "black", linewidth = 0.5) +
  ggtitle("Average counts contribution to chromatin accessibility\nacross ChromBPNet-mapped motifs") +
  scale_fill_gradientn(colours = c("#4C78A8", "#729FCF", "#A9C7E2", "white", "#F2B6A0", "#D9826B", "#B85C4A"),
                       values = scales::rescale(c(-0.031, -0.02, -0.01, 0, max(contrib_summary.df$value, na.rm = TRUE) / 2, max(contrib_summary.df$value, na.rm = TRUE) * 0.75, max(contrib_summary.df$value, na.rm = TRUE))), name = "ATAC-seq\ncounts contrib.") +
  theme_cowplot() +
  scale_x_discrete(position = "top", labels = c("DMSO" = "DMSO", "dTAG_3h" = "dTAG: 3 hours", "dTAG_1d" = "dTAG: 1 day")) +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.ticks = element_blank(), axis.line.x = element_blank(), axis.line.y = element_blank(),
        axis.text.y = element_text(size = 12, color = "black", face = "italic"), plot.title = element_text(size = 15, face = "bold"))

# Append all

total_summary.plot <- p_labels + seq.plot + contrib.plot + plot_layout(widths = c(1, 4, 6))
total_summary.plot

ggsave("motif_summary.pdf", plot = total_summary.plot, path = "/storage/kaelanb/analysis/lucia/analysis/figures/1_curate_motifs/", width = 16, height = 12)
ggsave("motif_summary.png", plot = total_summary.plot, path = "/storage/kaelanb/analysis/lucia/analysis/figures/1_curate_motifs/", width = 16, height = 12)

```

# 7. Quantify Snail dTAG effects

```{r quantify}

snai2_all.tidy <- all_curated_cncc_cwms.gr %>%
  dplyr::filter(motif_name == "SNAI2") %>%
  as.data.table() %>%
   melt.data.table(id.vars = "motif_name", measure.vars = c("DMSO_contrib", "dTAG_3h_contrib", "dTAG_1d_contrib") ,value.name = 'value', variable.name = 'signal')

# Plot

all_peaks.quant <- ggplot(snai2_all.tidy, aes(x = signal, y = value, fill = signal)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("#8F8F8F", "#9F7ACD", "#442768")) +
  stat_compare_means(method = "wilcox.test", comparisons = list(c("DMSO_contrib", "dTAG_3h_contrib"), c("DMSO_contrib", "dTAG_1d_contrib"), c("dTAG_3h_contrib", "dTAG_1d_contrib")), paired = TRUE, label = "p.signif", label.y = c(0.25, 0.30, 0.37)) +
  ylab("ChromBPNet Counts Contribution") +
  xlab("") + 
  ggtitle("SNAI2 motif contribution increases upon SNAI2 degradation") +
  coord_cartesian(ylim = c(-0.3, 0.4)) +
  theme_cowplot() +
  theme(legend.position = "none")

all_peaks.quant

ggsave("snai_dtag_quantification_all_peaks.pdf", plot = all_peaks.quant, path = "/storage/kaelanb/analysis/lucia/analysis/figures/1_curate_motifs/", width = 8, height = 12)
ggsave("snai_dtag_quantification_all_peaks.png", plot = all_peaks.quant, path = "/storage/kaelanb/analysis/lucia/analysis/figures/1_curate_motifs/", width = 8, height = 12)

```

# 8. Conclusions

Motifs have been successfully curated and summarized.

# 9. Session Information

```{r sessioninfo}
sessionInfo()
```
