---
title: 'Prepare custom sequences for in silico perturbations'
author: "Kaelan Brennan"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    depth: 3
    theme: sandstone
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyfoot[CO,CE]{1028-02098-001-001}
- \fancyfoot[LE,RO]{\thepage}
- \fancypagestyle{plain}{\pagestyle{fancy}}
editor_options:
  chunk_output_type: console
---

# 1. Introduction

Here, we will prepare sequences to test with in silico predictions. 

# 2. Computational Setup

```{r, warning=F, message=F}

#Standard packages

library(rtracklayer) ; library(GenomicRanges); library(magrittr) ; library(Biostrings); library(ggseqlogo);
library(ggplot2) ; library(reshape2); library(plyranges); library(Rsamtools); library(ggpubr); library(BSgenome.Hsapiens.UCSC.hg38); library(parallel); library(tidyr); library(tibble); library(ggrastr);
library(dplyr); library(data.table); library(patchwork); library(readr); library(testit); library(cowplot); library(universalmotif)

setwd("/storage/kaelanb/analysis/lucia/analysis/")
options(knitr.figure_dir="/storage/kaelanb/analysis/lucia/analysis/figures/1_curate_motifs/", java.parameters = "- Xmx6g")

#Sourced code

source("/storage/kaelanb/analysis/oct4_public_data/scripts/granges_common.r")
source("/storage/kaelanb/analysis/oct4_public_data/scripts/metapeak_common.r")
source("/storage/kaelanb/analysis/oct4_public_data/scripts/knitr_common.r")
source("/storage/kaelanb/analysis/oct4_public_data/scripts/caching.r")
source("/storage/kaelanb/analysis/oct4_public_data/scripts/metapeak_functions.r")
source("/storage/kaelanb/analysis/oct4_public_data/scripts/metapeak_extra.r")
source("/storage/kaelanb/analysis/oct4_public_data/scripts/metapeak_functions_kjb.R")
source("/storage/kaelanb/analysis/oct4_public_data/scripts/metapeak_plotting_functions.R")
source("/storage/kaelanb/analysis/oct4_public_data/scripts/motif_functions.r")

#Pre-existing variables

library(TxDb.Hsapiens.UCSC.hg38.knownGene)
bsgenome <- BSgenome.Hsapiens.UCSC.hg38
rmsk <- rtracklayer::import("/work/kaelanb/indexes/hg38_rmsk.bed") %>% keepStandardChromosomes(pruning.mode="coarse")
blacklist <- rtracklayer::import("/work/kaelanb/indexes/hg38-blacklist.v2.bed")
chrombpnet.blacklisted.gr <- import("/work/kaelanb/public_data/chrombpnet_tutorial/data/downloads/blacklist.bed.gz")
promoters_txdb <- promoters(genes(TxDb.Hsapiens.UCSC.hg38.knownGene), upstream = 250, downstream = 250) %>% keepStandardChromosomes(pruning.mode="coarse")

```

# 3. Import relevant data

```{r import}

# Coverage information

contrib.list <- list(DMSO = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/bw/all_peaks.counts_scores.bw",
                     dTAG_3h = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/dtag_3h/all_peaks/bw/all_peaks.counts_scores.bw",
                     dTAG_1d = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/dtag_1d/all_peaks/bw/all_peaks.counts_scores.bw")

pred.list <- list(DMSO = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/bw/all_peaks_chrombpnet.bw",
                     dTAG_3h = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/dtag_3h/all_peaks/bw/all_peaks_chrombpnet.bw",
                     dTAG_1d = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/dtag_1d/all_peaks/bw/all_peaks_chrombpnet.bw")

obs.list <- list(DMSO = "/storage/kaelanb/analysis/lucia/data/bw/ATAC_DMSO_merged.bw",
                     dTAG_3h = "/storage/kaelanb/analysis/lucia/data/bw/ATAC_3h_merged.bw",
                     dTAG_1d = "/storage/kaelanb/analysis/lucia/data/bw/ATAC_1d_merged.bw")

atac_contrib.list <- list(all_peaks_DMSO = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/bw/all_peaks.counts_scores.bw",
                          all_peaks_dTAG_3h = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/dtag_3h/all_peaks/bw/all_peaks.counts_scores.bw",
                          all_peaks_dTAG_1d = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks/dtag_1d/all_peaks/bw/all_peaks.counts_scores.bw",
                          no_prom_DMSO = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks_no_prom/bw/all_peaks_no_prom.counts_scores.bw",
                          no_prom_dTAG_3h = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks_no_prom/dtag_3h/all_peaks_no_prom/bw/all_peaks_no_prom.counts_scores.bw",
                          no_prom_dTAG_1d = "/storage/kaelanb/analysis/lucia/model/snail_modisco_test/all_peaks_no_prom/dtag_1d/all_peaks_no_prom/bw/all_peaks_no_prom.counts_scores.bw")

# Motifs

all_curated_cncc_cwms.gr <- readRDS("/storage/kaelanb/analysis/lucia/motifs/all_peaks_curated_cncc_cwms_w_metadata.rdata")

```

# 4. Select motifs and regions of interest

## 4.1. Parse SNAI2 motifs

Here, we are interested in those SNAI2 motifs that have a negative influence on accessibility because we intend to test the predicted effects upon the in silico perturbation of SNAI2.

```{r select}

snai2_motifs.gr <- all_curated_cncc_cwms.gr %>%
  dplyr::filter(motif_name == "SNAI2") %>%
  dplyr::filter(DMSO_contrib < 0) %>%
  dplyr::filter(dTAG_3h_diff_DMSO > 0) %>%
  dplyr::filter(dTAG_1d_diff_DMSO > 0) %>%
  dplyr::mutate(obs_dmso = regionSums(resize(.,500,"center"), obs.list$DMSO),
                obs_3h = regionSums(resize(.,500,"center"), obs.list$dTAG_3h),
                obs_1d = regionSums(resize(.,500,"center"), obs.list$dTAG_1d),
                pred_dmso = regionSums(resize(.,500,"center"), pred.list$DMSO),
                pred_3h = regionSums(resize(.,500,"center"), pred.list$dTAG_3h),
                pred_1d = regionSums(resize(.,500,"center"), pred.list$dTAG_1d))
  
snai2_motifs_parsed.gr <- snai2_motifs.gr %>%
  dplyr::filter(bound == TRUE) %>%
  dplyr::filter(obs_3h > obs_dmso & obs_1d > obs_dmso) %>%
  dplyr::filter(pred_3h > pred_dmso & pred_1d > pred_dmso)

```

## 4.2. Parse Lucia's enhancer regions for SNAI2 motifs

```{r enhancers}

enhancers.gr <- import("/storage/kaelanb/analysis/lucia/data/narrowpeak/regions_bound_distal_UPK27andATAC.bed") %>%
  dplyr::mutate(unique_id = 1:length(.))

snai2_at_enhancers.gr <- snai2_motifs_parsed.gr %>%
  subsetByOverlaps(., enhancers.gr, ignore.strand = T)

snai2_at_enhancers.gr$motif_seq <- as.character(reverseComplement(DNAStringSet(snai2_at_enhancers.gr$motif_seq)))

snai2_at_enhancers.gr$motif_seq %>% ggseqlogo()

# High priority regions

priority.gr <- c(GRanges('chr20', IRanges(start = 17168168, end = 17171604), strand = '*', name = 'ENH4'),
  GRanges('chr20', IRanges(start = 57256219, end = 57260565), strand = '*', name = 'ENH8'),
  GRanges('chr11', IRanges(start = 12318027, end = 12321872), strand = '*', name = 'ENH10'),
  GRanges('chr6', IRanges(start = 79609697, end = 79612498), strand = '*', name = 'ENH15'),
  GRanges('chr3', IRanges(start = 69115593, end = 69118574), strand = '*', name = 'ENH30'),
  GRanges('chr11', IRanges(start = 8707968, end = 8709964), strand = '*', name = 'ENH33'))

# Bring in previous SNAI2 and correct lengths

previous_snai2.gr <- read.table("/storage/lucia/fimo/SNAI2_DBD_jolma2013_fimo/fimo_sorted.bed", header = FALSE, sep = "\t", col.names = c("chr", "start", "end", "name", "score", "strand", "empty", "log_pvalue", "pvalue", "qvalue", "sequence")) %>%
    dplyr::select(-empty) %>%
    makeGRangesFromDataFrame(keep.extra.columns = TRUE) %>%
    subsetByOverlaps(., priority.gr, ignore.strand = T)


pos_snai2 <- previous_snai2.gr %>% as.data.frame() %>% dplyr::filter(strand == "+") %>% makeGRangesFromDataFrame(keep.extra.columns = T)
pos_snai2 <- resize(pos_snai2, width = width(pos_snai2) - 3, fix = "end") %>% dplyr::mutate(motif_seq = getSeq(bsgenome, ., as.character = T))

neg_snai2 <- previous_snai2.gr %>% as.data.frame() %>% dplyr::filter(strand == "-") %>% makeGRangesFromDataFrame(keep.extra.columns = T)
neg_snai2 <- resize(neg_snai2, width = width(neg_snai2) - 2, fix = "end")
neg_snai2 <- resize(neg_snai2, width = width(neg_snai2) - 1, fix = "start") %>% dplyr::mutate(motif_seq = getSeq(bsgenome, ., as.character = T))

previous_snai2.gr <- c(pos_snai2, neg_snai2) %>% dplyr::select(name, motif_seq)

snai2_at_priority.gr <- c(all_curated_cncc_cwms.gr %>%
                            dplyr::filter(motif_name == "SNAI2") %>%
                            subsetByOverlaps(., priority.gr, ignore.strand = T) %>%
                            dplyr::mutate(motif_seq = as.character(reverseComplement(DNAStringSet(.$motif_seq)))),
                          previous_snai2.gr) %>%
  dplyr::mutate(motif_name = "SNAI2",
                priority = TRUE)

# Total SNAI2 to test

snai2_to_test.gr <- c(snai2_at_priority.gr, snai2_at_enhancers.gr)

snai2_to_test.gr %>% .$motif_seq %>% ggseqlogo()

```

## 4.3. Re-calculate coverage info

```{r calc}

snai2_to_test.gr <- snai2_to_test.gr %>%
  dplyr::mutate(obs_dmso = regionSums(resize(.,500,"center"), obs.list$DMSO),
                obs_3h = regionSums(resize(.,500,"center"), obs.list$dTAG_3h),
                obs_1d = regionSums(resize(.,500,"center"), obs.list$dTAG_1d),
                pred_dmso = regionSums(resize(.,500,"center"), pred.list$DMSO),
                pred_3h = regionSums(resize(.,500,"center"), pred.list$dTAG_3h),
                pred_1d = regionSums(resize(.,500,"center"), pred.list$dTAG_1d)) %>%
  dplyr::mutate(DMSO_contrib = regionSums(., atac_contrib.list$all_peaks_DMSO),
                dTAG_3h_contrib = regionSums(., atac_contrib.list$all_peaks_dTAG_3h),
                dTAG_1d_contrib = regionSums(., atac_contrib.list$all_peaks_dTAG_1d)) %>%
  dplyr::mutate(DMSO_contrib_norm = DMSO_contrib / width(all_curated_cncc_cwms.gr),
                dTAG_3h_contrib_norm = dTAG_3h_contrib / width(all_curated_cncc_cwms.gr),
                dTAG_1d_contrib_norm = dTAG_1d_contrib / width(all_curated_cncc_cwms.gr)) %>%
  dplyr::mutate(dTAG_3h_diff_DMSO = dTAG_3h_contrib - DMSO_contrib,
                dTAG_1d_diff_DMSO = dTAG_1d_contrib - DMSO_contrib) %>%
  dplyr::mutate(unique_id = 1:length(.))

```

# 5. Design sequences

## 5.1. Function to generate sequences

Here, I will shuffle the motif until the core GGT pattern is gone, as these are the strongest bases in the SNAI2 motif. I am doing it this way because, since the motif is 7bp long, there just is not enough sequence content to do the dinucleotide shuffling. 

```{r function}

randomize_motif <- function(seq_str, core_patterns = c("GGT"),
                             bg_freqs = c(A = 0.25, C = 0.25, G = 0.25, T = 0.25),
                             max_iter = 20000, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  n <- nchar(seq_str)
  orig_bases <- strsplit(seq_str, "")[[1]]

  for (i in 1:max_iter) {
    candidate <- paste(sample(names(bg_freqs), n, replace = TRUE, prob = bg_freqs), collapse = "")
    cand_bases <- strsplit(candidate, "")[[1]]

    all_positions_differ <- all(cand_bases != orig_bases)
    contains_core <- any(sapply(core_patterns, function(p) grepl(p, candidate, fixed = TRUE)))

    if (all_positions_differ && !contains_core) {
      return(candidate)
    }
  }
  warning("Could not generate a fully-different, core-free replacement for ", seq_str,
          " after ", max_iter, " iterations; returning best-effort candidate")
  return(candidate)
}

# Estimate A/C/G/T frequencies from the sequence flanking the motif within its window, to use as a locally-matched background for randomize_motif().

flank_bg_freqs <- function(window_seq, motif_start_in_window, motif_end_in_window) {
  flank_seq <- paste0(
    substr(window_seq, 1, motif_start_in_window - 1),
    substr(window_seq, motif_end_in_window + 1, nchar(window_seq))
  )
  counts <- table(factor(strsplit(flank_seq, "")[[1]], levels = c("A", "C", "G", "T")))
  counts <- counts + 1  # pseudocount so no base ever hits exactly 0
  as.list(counts / sum(counts))
}

scramble_snai2_motif <- function(motif_gr, genome, window_size = 2114,
                                  core = "GGT", seed_base = 42) {
  stopifnot(length(motif_gr) == 1)

  chrom  <- as.character(seqnames(motif_gr))
  strnd  <- as.character(strand(motif_gr))
  m_start <- start(motif_gr)
  m_end   <- end(motif_gr)

  # Motif center in genomic (+strand) coordinates -- strand doesn't affect
  # the coordinate system, only which direction the sequence is read.
  motif_center <- m_start + floor((m_end - m_start) / 2)

  # Build the +strand window
  half <- floor(window_size / 2)
  win_start <- motif_center - (half - 1)
  win_end   <- motif_center + half
  win_gr <- GRanges(seqnames = chrom, ranges = IRanges(win_start, win_end), strand = "+")

  wt_window_seq <- as.character(getSeq(genome, win_gr))
  stopifnot("Window sequence is not expected length" = nchar(wt_window_seq) == window_size)

  # Motif position within the +strand window
  motif_start_in_window <- m_start - win_start + 1
  motif_end_in_window   <- m_end   - win_start + 1
  motif_seq_plus <- substr(wt_window_seq, motif_start_in_window, motif_end_in_window)

  # Strand-aware "sense" sequence -- the motif as it's actually read 5'->3'
  if (strnd == "-") {
    sense_seq <- as.character(reverseComplement(DNAString(motif_seq_plus)))
  } else {
    sense_seq <- motif_seq_plus
  }

  # Locally-matched background composition (from this window's flanks)
  bg_freqs <- flank_bg_freqs(wt_window_seq, motif_start_in_window, motif_end_in_window)

  # Replace in sense orientation so "GGT" always refers to the real core
  sense_scrambled <- randomize_motif(sense_seq, core_patterns = core,
                                      bg_freqs = unlist(bg_freqs), seed = seed_base)

  # Convert scrambled motif back to +strand orientation for insertion
  if (strnd == "-") {
    scrambled_plus <- as.character(reverseComplement(DNAString(sense_scrambled)))
  } else {
    scrambled_plus <- sense_scrambled
  }

  mutant_window_seq <- paste0(
    substr(wt_window_seq, 1, motif_start_in_window - 1),
    scrambled_plus,
    substr(wt_window_seq, motif_end_in_window + 1, nchar(wt_window_seq))
  )
  stopifnot(nchar(mutant_window_seq) == window_size)

  # Position of the motif in chrombpnet's cropped 1000bp output space
  crop <- floor((window_size - 1000) / 2)

  list(
    chrom = chrom,
    strand = strnd,
    genomic_start = m_start,
    genomic_end = m_end,
    window_start = win_start,
    window_end = win_end,
    motif_start_in_window = motif_start_in_window,
    motif_end_in_window = motif_end_in_window,
    motif_start_in_output = motif_start_in_window - crop,
    motif_end_in_output   = motif_end_in_window - crop,
    motif_seq_sense_wt = sense_seq,
    motif_seq_sense_scrambled = sense_scrambled,
    wt_seq = wt_window_seq,
    mutant_seq = mutant_window_seq
  )
}

# Loop over all SNAI2 motifs, write one combined FASTA (wt + scramble pairs) and one metadata CSV with positions and motif sequences.

process_all_snai2_motifs <- function(snai2_gr, genome, output_dir,
                                      prefix = "snai2", core = "GGT",
                                      id_col = "unique_id") {
  if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

  stopifnot("id_col not found in mcols(snai2_gr)" = id_col %in% colnames(mcols(snai2_gr)))
  ids <- mcols(snai2_gr)[[id_col]]
  stopifnot("unique_id column has duplicate or missing values" =
              !any(is.na(ids)) && !any(duplicated(ids)))

  fasta_path <- file.path(output_dir, paste0(prefix, "_scramble_2114bp.fa"))
  meta_path  <- file.path(output_dir, paste0(prefix, "_scramble_metadata.csv"))

  fasta_lines <- character(0)
  meta_list <- vector("list", length(snai2_gr))

  for (i in seq_along(snai2_gr)) {
    uid <- as.character(ids[i])
    res <- scramble_snai2_motif(snai2_gr[i], genome, core = core, seed_base = 42 + i)

    # FASTA headers are keyed on unique_id so wt/mutant predictions can be
    # traced back to the same motif -- and to snai2_to_test.gr -- downstream.
    wt_header       <- paste0("wt_", uid)
    scramble_header <- paste0("scramble_", uid)

    fasta_lines <- c(fasta_lines,
                      paste0(">", wt_header), res$wt_seq,
                      paste0(">", scramble_header), res$mutant_seq)

    # One row per motif, both sequences included, joined by unique_id --
    # makes it easy to pull predictions for wt/mutant into the same row
    # for plotting effect sizes.
    meta_list[[i]] <- data.frame(
      unique_id = uid,
      wt_header = wt_header,
      scramble_header = scramble_header,
      chrom = res$chrom,
      strand = res$strand,
      genomic_start = res$genomic_start,
      genomic_end = res$genomic_end,
      window_start = res$window_start,
      window_end = res$window_end,
      motif_start_in_window = res$motif_start_in_window,
      motif_end_in_window = res$motif_end_in_window,
      motif_start_in_output = res$motif_start_in_output,
      motif_end_in_output = res$motif_end_in_output,
      motif_seq_wt = res$motif_seq_sense_wt,
      motif_seq_scrambled = res$motif_seq_sense_scrambled,
      wt_seq = res$wt_seq,
      mutant_seq = res$mutant_seq,
      stringsAsFactors = FALSE
    )

    cat(sprintf("[%d/%d] %s | WT: %s -> Scrambled: %s\n",
                i, length(snai2_gr), uid,
                res$motif_seq_sense_wt, res$motif_seq_sense_scrambled))
  }

  writeLines(fasta_lines, con = fasta_path)
  meta_df <- do.call(rbind, meta_list)
  write.csv(meta_df, meta_path, row.names = FALSE)

  cat("\nFASTA written to:", fasta_path, "\n")
  cat("Metadata written to:", meta_path, "\n")

  return(meta_df)
}

```

## 5.2. Generate sequences and record info

```{r generate}

output_dir <- "/storage/kaelanb/analysis/lucia/analysis/custom_seq/"
genome <- bsgenome

snai2_metadata <- process_all_snai2_motifs(snai2_to_test.gr, genome, output_dir, prefix = "snai2", id_col = "unique_id")

```

## 5.3. Showcase SNAI2

```{r showcase}

head(snai2_to_test.gr)


snai2_to_test.gr %>%
  dplyr::filter(unique_id == 2 | unique_id == 8 | unique_id == 18 | unique_id == 19 | unique_id == 20 | unique_id == 23 | unique_id == 26)

```

# 6 Conclusions

Here, I have generated sequences where the SNAI2 motif has been computationally mutated. The predictions will be performed in `3_predict_motif_effects.ipynb`

# 7. Session Information

For the purposes of reproducibility, the R/Bioconductor session information is printed below:

```{r sessioninfo}
sessionInfo()
```